### 2014-12-03 Composition_Ecology.Rmd
Initial commit.  A problem with a couple of the tifs that won't get ignored for some reason.

Land use patterns and historical contingency change climate optima for key forest genera in the Northeastern United States.
========================================================
What is the climate space of the northeastern United States?

```{r, fig.width=5}
# Load climate data and generate the pre-sett and modern bounds:
#  Pre-settlement climate: 1900 - 1930
#  Modern climate: 1960 - 1990

source('R/Setup_R.R')
source('R/load_climate.R') # gives then.clim & now.clim
source('R/load_fia.R')     # gives agg.dens
source('R/load_ncdf.R')    # gives western and eastern

```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

### 2014-12-03 R/Setup_R.R

@@ -1,0 +1,9 @@
+#  Setup some data and libraries:
+library(raster)
+library(ggplot2)
+library(mgcv)
+library(reshape2)
+library(plyr)
+
+base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
+base.rast <- setValues(base.rast, NA)

### 2014-12-03 R/load_climate.R

@@ -1,0 +1,69 @@
+#  Load the climate data:
+path.root <- getwd()
+
+if(regexpr('goring', path.root)>0){
+  path.gis <- 'C:/Users/goring/Documents/GriddedClimate/CIEE_data'
+}
+if(regexpr('Goring', path.root)>0){
+  path.gis = 'C:/Users/Simon Goring/Documents/GriddedClimate/CIEE_data'
+}
+
+## climate normal layers:
+#naming convention: biox_norm1 = 1900-1975; biox_norm2 = 1985-2010
+#  
+then.files = list.files(path.gis, pattern = "_norm1.asc", 
+                        recursive=TRUE, full.names = TRUE)
+now.files = list.files(path.gis, pattern = "_norm2.asc", 
+                       recursive = TRUE, full.names = TRUE)
+
+load.stack <- function(filenames){
+  #  Loads a raster stack based on the filenames provided, then crops to
+  #  the extent provided in 'crop.ext'
+  loaded.stack <- stack(filenames)
+  proj4string(loaded.stack) <- '+proj=longlat +ellps=WGS84'
+  
+  return(loaded.stack)
+}  
+
+#  For each year, by variable:
+## year-specific climate layers
+bioclim.files <- list.files(path.gis, pattern=".asc$", recursive = TRUE)
+bioclim.files <- bioclim.files[!(regexpr('norm', bioclim.files)>0 | regexpr('elevstd', bioclim.files) > 0)]
+
+#  Get just the bioclim variable names, not the full path.
+bio.vals <- unique(substr(bioclim.files, 
+                          regexpr('/', bioclim.files) + 1,
+                          regexpr('_', bioclim.files)-1))
+
+right.clim <- function(x)regexpr(x, bioclim.files) > 0
+
+#  Load each of the bioclim variables.  Takes some time.  This could
+#  be sped up a bit if we parallelize.
+#  The result is a list of RasterBricks for each bioclim value, 
+#  representing each year (in sequence) from 1901 - 2010.
+
+if(!'then.rast.RData' %in% list.files(path='data/output')){
+  then.rast <- lapply(bio.vals, function(x){
+    load.stack(paste0(path.gis, '/', bioclim.files[right.clim(x)][1:30]))
+  })
+  
+  then.rast <- lapply(then.rast, 
+                      function(x)resample(projectRaster(x, crs=CRS('+init=epsg:3175')), 
+                                          base.rast))
+
[…]

### 2014-12-03 R/load_fia.R

@@ -1,0 +1,76 @@
+#  Load & compile the FIA data:
+
+fia.trans <- read.csv('data/FIA_conversion.csv', stringsAsFactors = FALSE)
+fia.trans[fia.trans == ''] <- NA
+
+#  Load FIA composition data:
+get.fia.stack <- function(x, type){
+  stack.files <- list.files(x, full = TRUE)
+  stack.fia <- stack(stack.files)
+  
+  fia.vals <- getValues(stack.fia)
+  name.end <- regexpr(paste('_', type, sep=''), colnames(fia.vals), fixed=TRUE)
+  colnames(fia.vals) <- gsub("_$","", substr(colnames(fia.vals), 1, name.end-1))
+  colnames(fia.vals) <- gsub('_', '.', colnames(fia.vals))
+  
+  fia.vals <- as.data.frame(fia.vals)
+  
+  fia.vals$x <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,1]
+  fia.vals$y <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,2]
+  
+  fia.vals
+}
+
+fia.dens <- get.fia.stack('data/albers/density/', type = 'density')
+
+re.agg <- function(x){
+  # This takes the fia data (by species) and aggregates it to PalEON taxa.
+  orig <- 1:nrow(x)
+  
+  x.test <- x[,!colnames(x) %in% c('x', 'y')]
+  
+  x.test$orig <- orig
+  
+  x.melt <- melt(x.test, na.rm=TRUE, value.name= 'value', id = c('orig'))
+  
+  x.melt$equiv <- fia.trans$PalEON[match(gsub('.', ' ', x.melt$variable, fixed = TRUE), 
+                                         gsub('.', ' ', fia.trans$scientific, fixed = TRUE))]
+  
+  x.melt <- x.melt[x.melt$value > 0,]
+  
+  x.cast <- dcast(x.melt, orig ~ equiv, 
+                  sum, drop = FALSE, 
+                  value.var = 'value')
+  
+  x.cast$x <- x$x[x.cast$orig]
+  x.cast$y <- x$y[x.cast$orig]
+  
+  x.cast$cell <- extract(setValues(base.rast, 1:ncell(base.rast)), x.cast[,c('x', 'y')])
+  
+  x.cast <- x.cast[!is.na(x.cast$cell),!is.na(colnames(x.cast))]
[…]

### 2014-12-03 R/load_ncdf.R

@@ -1,0 +1,22 @@
+library(ncdf4)
+
+original.points <- read.csv('data//westerncompv0.3.csv')
+presence <- data.frame(x = original.points$x,
+                       y = original.points$y,
+                       point = rowSums(!is.na(original.points[,3:ncol(original.points)]))>0)
+
+coordinates(presence) <- ~ x + y
+proj4string(presence) <- CRS('+init=epsg:3175')
+presence <- rasterize(presence, base.rast, field = 'point')
+
+western <- nc_open('data/input//PLScomposition_western_0.2-release.nc', write = FALSE)
+eastern <- nc_open('data/input//PLScomposition_eastern_0.2-release.nc', write = FALSE)
+
+taxa <- unique(c(names(western$var), names(eastern$var)))
+
+ll.ncdf <- list(long = c(ncvar_get(western, 'x'), ncvar_get(eastern, 'x')),
+                lat  = c(ncvar_get(western, 'y'), ncvar_get(eastern, 'y')))
+
+western.grid <- expand.grid(x = ncvar_get(western, 'x'), y = rev(ncvar_get(western, 'y')))
+eastern.grid <- expand.grid(x = ncvar_get(eastern, 'x'), y = rev(ncvar_get(eastern, 'y')))
+

### 2014-12-03 R/resampling_climate.R

@@ -1,0 +1,316 @@
+all.taxa <- names(western$var)
+
+all.taxa <- all.taxa[!all.taxa%in% c('Other hardwood', 'Atlantic White Cedar',
+  'Buckeye', 'Ironwood', 'Walnut', 'Black gum/sweet gum')]
+
+#For each taxon sample:
+#  1.  The FIA data for one of the now.rast years for each of the climate variables.
+#  2.  One of the then.rast years & one of the western layers
+#  3.  Get vectors of values
+
+hellinger <- function(kernel1, kernel2){
+  1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
+}
+
+ranges <- data.frame(min = c(100, 0, 220, 155, -383),
+                     max = c(879, 403, 558, 405, 27))
+
+get_dens_plss <- function(taxon, clim.var){
+  layer_ncdf <- sample(500, 1)
+  layer_clim <- sample(nlayers(then.rast[[1]]), 1)
+  
+  taxon.vals <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
+  
+  values <- data.frame(x = western.grid$x,
+                       y = western.grid$y, 
+                       taxon = as.numeric(taxon.vals[,,layer_ncdf]))
+  
+  coordinates(values) <- ~ x + y
+  proj4string(values) <- CRS('+init=epsg:3175')
+  
+  good.vals <- extract(presence, values)
+  
+  values <- values[good.vals == 1 & !is.na(good.vals), ]
+  
+  clims <- do.call(cbind.data.frame, lapply(then.rast, function(x)extract(x[[layer_clim]], values)))
+  colnames(clims) <- c('bio16', 'bio17', 'bio4', 'bio5', 'bio6')
+  density(clims[,clim.var], weights = values$taxon/sum(values$taxon),
+          from = ranges[clim.var,1], to = ranges[clim.var,2], n = 100)$y
+}
+
+get_dens_fia <- function(taxon, clim.var){
+
+  layer_clim <- sample(nlayers(now.rast[[1]]), 1)
+  
+  taxon.vals <- agg.dens[,taxon]
+  
+  values <- data.frame(x = xyFromCell(base.rast, agg.dens$cell)[,'x'],
+                       y = xyFromCell(base.rast, agg.dens$cell)[,'y'], 
+                       taxon = agg.dens[,taxon])
+  
+  values <- na.omit(values)
+  
[…]

### 2014-12-04 Composition_Ecology.Rmd
Getting the repo set up. * Added `ecology.csl` to allow formatting of `bib` files. * moved in the goringetal bib file, assuming many refs will cross-over. * Added `margins.sty` to set out the pre-formatting for the rendered document. * started commenting `resampling_climate` *  Added the header blob for the paper.

@@
+---
+title: "Composition-climate paper"
+author: "Simon Goring *et al*."
+date: "`r format(Sys.time(), '%d %B, %Y')`"
+output:
+pdf_document:
+pandoc_args: "-H margins.sty"
+bibliography: composition.bib
+csl: harvard1.csl
+---
+
 Land use patterns and historical contingency change climate optima for key forest genera in the Northeastern United States.
 ========================================================
-What is the climate space of the northeastern United States?
+
+*Historic land use and forest change over the centuries since Euroamerican settlement have significantly changed
 

### 2014-12-04 R/load_ncdf.R

@@ -19,4 +19,3 @@ ll.ncdf <- list(long = c(ncvar_get(western, 'x'), ncvar_get(eastern, 'x')),
 
 western.grid <- expand.grid(x = ncvar_get(western, 'x'), y = rev(ncvar_get(western, 'y')))
 eastern.grid <- expand.grid(x = ncvar_get(eastern, 'x'), y = rev(ncvar_get(eastern, 'y')))
-

### 2014-12-04 R/resampling_climate.R

@@ -9,6 +9,7 @@ all.taxa <- all.taxa[!all.taxa%in% c('Other hardwood', 'Atlantic White Cedar',
 #  3.  Get vectors of values
 
 hellinger <- function(kernel1, kernel2){
+  #  Calculate hellinger distance (between two kernels)
   1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
 }
 
@@ -16,18 +17,25 @@ ranges <- data.frame(min = c(100, 0, 220, 155, -383),
                      max = c(879, 403, 558, 405, 27))
 
 get_dens_plss <- function(taxon, clim.var){
-  layer_ncdf <- sample(500, 1)
+  
+  #  Function resamples proportion data from the composition model for the region
+  #  and then draws from the composition model and the climate data.  This is set up
+  #  within a loop further down the code.
+  
+  layer_ncdf <- sample(500, 1)                      # there are 500 layers in the ncdf
   layer_clim <- sample(nlayers(then.rast[[1]]), 1)
   
-  taxon.vals <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
+  taxon.vals.w <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
+  #taxon.vals.e <- ncvar_get(eastern, taxon, c(1,1,1), c(-1, -1, -1))
   
-  values <- data.frame(x = western.grid$x,
-                       y = western.grid$y, 
-                       taxon = as.numeric(taxon.vals[,,layer_ncdf]))
+  values <- data.frame(x = western.grid$x, #c(western.grid$x, eastern.grid$x),
+                       y = western.grid$y, #c(western.grid$y, eastern.grid$y), 
+                       taxon = as.numeric(taxon.vals.w[,,layer_ncdf]))#as.numeric(c(taxon.vals.w[,,layer_ncdf], taxon.vals.e[,,layer_ncdf])))
   
   coordinates(values) <- ~ x + y
   proj4string(values) <- CRS('+init=epsg:3175')
   
+  #  Create the spatial points object above to extract values from the composition raster.
   good.vals <- extract(presence, values)
   
   values <- values[good.vals == 1 & !is.na(good.vals), ]
@@ -39,7 +47,8 @@ get_dens_plss <- function(taxon, clim.var){
 }
 
 get_dens_fia <- function(taxon, clim.var){
-
+  #  Similar to above, but we only have one forest layer, not 500, so
+  #  the model is a bit weaker.
[…]

### 2014-12-15 Composition_Ecology.Rmd
More text contributions.  Mostly focused on the  abstract. *  Started methods. * Added in prep papers to the bib. file. * Minor modifications to the `resampling_climate` file, fixed a typo, and added the call to the the `vegan` library. *  Added more comments to the `resampling_climate` code.

@@
 
-*Historic land use and forest change over the centuries since Euroamerican settlement have significantly changed
+*Historic land use and forest change over the centuries since Euroamerican settlement have significantly changed the structure and composition of forests in the upper Midwestern United States.
+The extent to which this change has affected the realized climate niche for key taxa is less well known.*
+
+*Predictions for future species range shifts are predicated largely on correlative models that relate modern species distributions to climate parameters in the modern era.
+Given the extent of regional forest change and recent losses to key forest taxa including Hemlock, Elm and Chestnut it is possible that certain taxa have shiften with respect to climate change over the last century.
+The additional pressure of climatically biased land use conversion for agriculture along the western and southern border of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*
 
@@
 
-You can also embed plots, for example:
+Introduction
+==============
+
+Climate change is rough as shit yo.
 
 ```{r fig.width=7, fig.height=6}
-plot(cars)
+
 ```
+
+Methods
+===============
+We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015] with annually resolved climate data for the region.
+PLSS vegetation data contains no uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.
+Additionally, data in the northeastern United States [@thompson2013four] from town Proprietor Surveys is recorded arially, with no variance within township polygons.
+To accomodate these different data sources a conditional autoregressive model was applied to the data, providing the ability to obtain uncertainty estimates across the domain [@paciorek2015comp].
+
+To generate the climate niches we
+
+#  Tree size?  Segregate

### 2014-12-15 R/resampling_climate.R

@@ -28,6 +28,8 @@ get_dens_plss <- function(taxon, clim.var){
   taxon.vals.w <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
   #taxon.vals.e <- ncvar_get(eastern, taxon, c(1,1,1), c(-1, -1, -1))
   
+  #  turn everything into a data.frame & then into a SpatialPointsDataFrame
+  #  code is commented out so that when we get full FIA coverage we can do a bit more.
   values <- data.frame(x = western.grid$x, #c(western.grid$x, eastern.grid$x),
                        y = western.grid$y, #c(western.grid$y, eastern.grid$y), 
                        taxon = as.numeric(taxon.vals.w[,,layer_ncdf]))#as.numeric(c(taxon.vals.w[,,layer_ncdf], taxon.vals.e[,,layer_ncdf])))
@@ -48,7 +50,7 @@ get_dens_plss <- function(taxon, clim.var){
 
 get_dens_fia <- function(taxon, clim.var){
   #  Similar to above, but we only have one forest layer, not 500, so
-  #  the model is a bit weaker.
+  #  we only pull from climate layers:
   layer_clim <- sample(nlayers(now.rast[[1]]), 1)
   
   taxon.vals <- agg.dens[,taxon]
@@ -57,6 +59,7 @@ get_dens_fia <- function(taxon, clim.var){
                        y = xyFromCell(base.rast, agg.dens$cell)[,'y'], 
                        taxon = agg.dens[,taxon])
   
+  #  We get NA values if there is no FIA coverage
   values <- na.omit(values)
   
   coordinates(values) <- ~ x + y
@@ -72,11 +75,12 @@ get_dens_fia <- function(taxon, clim.var){
           from = ranges[clim.var,1], to = ranges[clim.var,2], n = 100)$y
 }
 
+#  This gives us 100 samples from the plss data, 
 plss.tables <-   llply(all.taxa,
                         function(x){
                           lapply(1:5, function(y){
                             do.call(rbind.data.frame, 
-                              lapply(1:100, function(z)get_dens_plss(x, y, 500)))
+                              lapply(1:100, function(z)get_dens_plss(x, y)))
                           })
                         }, .progress = 'text')
 
@@ -182,9 +186,9 @@ ggplot(optim.set, aes(taxon, optima, fill=era)) + geom_boxplot() +
         legend.text = element_text(family = 'serif', 
                                    face = 'bold', size = 18, color = 'black'),
[…]

### 2015-02-02 Composition_Ecology.Rmd
Modified `load_climate` to reflect the new model with samples in both time periods for all climate variables.
`resampling_climate` now includes `land.use` information.

Introductory text has been updated.

@@
 
-Climate change is rough as shit yo.
+Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States.
+There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset].
+This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change.
 
-```{r fig.width=7, fig.height=6}
+If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions.
+In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie and savanna to closed forest, in an approixmately west to east direction, and from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
+Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa.
 
-```
+We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for X major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.
 
@@
 
-To generate the climate niches we
+The analysis follows a blocked sampling method.
+We generate probability density functions for actual PLSS presence, by taxon, at the settlement era and for the same PLSS space in the modern era.
+We also generate climate data for the FIA presence, and we generate climate space for modern land use.
+In this way we can test how much of the change in shape is the result of changing land use vs changing climate since the early part of the century.
 
-#  Tree size?  Segregate
+We also develop estimates of the modern climate space that taxa would occupy if they remained in place, with only climate changing, and vice versa (need better explanation here).
+
+Results
+---------------
+
+Discussion
+----------------
+
+Figures
+--------------
+
+References
+---------------

### 2015-02-02 R/load_climate.R

@@ -42,7 +42,7 @@ right.clim <- function(x)regexpr(x, bioclim.files) > 0
 #  The result is a list of RasterBricks for each bioclim value, 
 #  representing each year (in sequence) from 1901 - 2010.
 
-if(!'then.rast.RData' %in% list.files(path='data/output')){
+if(!'then.rast.rds' %in% list.files(path='data/output')){
   then.rast <- lapply(bio.vals, function(x){
     load.stack(paste0(path.gis, '/', bioclim.files[right.clim(x)][1:30]))
   })
@@ -51,19 +51,19 @@ if(!'then.rast.RData' %in% list.files(path='data/output')){
                       function(x)resample(projectRaster(x, crs=CRS('+init=epsg:3175')), 
                                           base.rast))
 
-  save(then.rast, file = 'data/output/then.rast.RData')
+  saveRDS(then.rast, file = 'data/output/then.rast.rds')
 } else {
-  load('data/output/then.rast.RData')
+  then.rast <- readRDS('data/output/then.rast.rds')
 }
 
-if(!'now.rast.RData' %in% list.files(path='data/output')){
+if(!'now.rast.rds' %in% list.files(path='data/output')){
   now.rast <- lapply(bio.vals, function(x){
         load.stack(paste0(path.gis, '/', bioclim.files[right.clim(x)][76:101]))
         })
 
   now.rast <- lapply(now.rast, function(x)resample(projectRaster(x, crs=CRS('+init=epsg:3175')), 
                                              base.rast))
-  save(now.rast, file = 'data/output/now.rast.RData')
+  saveRDS(now.rast, file = 'data/output/now.rast.rds')
 } else {
-  load('data/output/now.rast.RData')
+  now.rast <- readRDS('data/output/now.rast.RData')
 }

### 2015-02-02 R/resampling_climate.R

@@ -1,7 +1,22 @@
+landuse <- list.files('../../../Dropbox/REU Work/Land Use/',pattern='.tif$', full.names=TRUE)
+
+landuse <- landuse[c(1, 4, 6, 7, 8, 9, 10, 16)]
+landuse.r <- (stack(landuse))
+landuse.r[landuse.r>100] <- NA
+landuse <- sum(landuse.r)>0
+
+agg.dens[is.na(agg.dens)] <- 0
+agg.dens[,-1] <- agg.dens[,-1] / rowSums(agg.dens[,-1])
+
 all.taxa <- names(western$var)
 
 all.taxa <- all.taxa[!all.taxa%in% c('Other hardwood', 'Atlantic White Cedar',
   'Buckeye', 'Ironwood', 'Walnut', 'Black gum/sweet gum')]
+
+unit.raster <- setValues(base.rast, 1:ncell(base.rast))
+proj4string(landuse) <- '+proj=longlat +ellps=WGS84'
+
+landuse <- projectRaster(landuse, unit.raster) > 0.9
 
 #For each taxon sample:
 #  1.  The FIA data for one of the now.rast years for each of the climate variables.
@@ -16,81 +31,169 @@ hellinger <- function(kernel1, kernel2){
 ranges <- data.frame(min = c(100, 0, 220, 155, -383),
                      max = c(879, 403, 558, 405, 27))
 
-get_dens_plss <- function(taxon, clim.var){
+# To do the analysis we want to get the climate layers for the past:
+get_dens_era <- function(taxon, clim.var){
   
   #  Function resamples proportion data from the composition model for the region
-  #  and then draws from the composition model and the climate data.  This is set up
-  #  within a loop further down the code.
-  
+  #  and then draws from the composition model and the climate data.
+  #  The function outputs a table with six columns:
+  #  data      - the proportion composition within the cell
+  #  class     - the era of observation
+  #  cell      - the literal cell # in the raster
+  #  plss.clim - the climate extracted from the 1900 - 1930s era
+  #  fia.clim  - the climate extracted from the 1970 - 2000s era
+  #
+  #  Each table has PLSS/pres (taxon present in cell at PLSS)
+  #                 FIA/pres  (taxon present in cell at PLSS)
+  #                 FIA/np    (taxon present in cell at PLSS but not in FIA)
+
   layer_ncdf <- sample(500, 1)                      # there are 500 layers in the ncdf
[…]

### 2015-04-11 Composition_Ecology.Rmd
* Minor changes to the Methods text. * Moved a bunch of new files in. * I've changed how I'm generating climate, and switched to a PRISM dataset on my local hard-drive.  To do this I've had to write `r/building_small_climate.R`, it's the raw code that generates the subset data. * The above change has meant large changes to `R/load_climate.R` which is now considerably simpler.

@@
 The analysis follows a blocked sampling method.
-We generate probability density functions for actual PLSS presence, by taxon, at the settlement era and for the same PLSS space in the modern era.
+We generate probability density functions for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era.
 We also generate climate data for the FIA presence, and we generate climate space for modern land use.
@@
 ---------------
+Across the region PLS cells account for X
 

### 2015-04-11 R/load_climate.R

@@ -1,69 +1,11 @@
 #  Load the climate data:
-path.root <- getwd()
+#  The large prism datasets live on my hard drive, but are accessible through:
+#
 
-if(regexpr('goring', path.root)>0){
-  path.gis <- 'C:/Users/goring/Documents/GriddedClimate/CIEE_data'
-}
-if(regexpr('Goring', path.root)>0){
-  path.gis = 'C:/Users/Simon Goring/Documents/GriddedClimate/CIEE_data'
-}
+#  To crop and reproject we use the R file, included hereL
+#  'R/building_small_climate.R'
 
-## climate normal layers:
-#naming convention: biox_norm1 = 1900-1975; biox_norm2 = 1985-2010
-#  
-then.files = list.files(path.gis, pattern = "_norm1.asc", 
-                        recursive=TRUE, full.names = TRUE)
-now.files = list.files(path.gis, pattern = "_norm2.asc", 
-                       recursive = TRUE, full.names = TRUE)
+clim.files <- list.files('data/input/climate', full.names = TRUE)
 
-load.stack <- function(filenames){
-  #  Loads a raster stack based on the filenames provided, then crops to
-  #  the extent provided in 'crop.ext'
-  loaded.stack <- stack(filenames)
-  proj4string(loaded.stack) <- '+proj=longlat +ellps=WGS84'
-  
-  return(loaded.stack)
-}  
-
-#  For each year, by variable:
-## year-specific climate layers
-bioclim.files <- list.files(path.gis, pattern=".asc$", recursive = TRUE)
-bioclim.files <- bioclim.files[!(regexpr('norm', bioclim.files)>0 | regexpr('elevstd', bioclim.files) > 0)]
-
-#  Get just the bioclim variable names, not the full path.
-bio.vals <- unique(substr(bioclim.files, 
-                          regexpr('/', bioclim.files) + 1,
-                          regexpr('_', bioclim.files)-1))
-
-right.clim <- function(x)regexpr(x, bioclim.files) > 0
-
-#  Load each of the bioclim variables.  Takes some time.  This could
-#  be sped up a bit if we parallelize.
-#  The result is a list of RasterBricks for each bioclim value, 
-#  representing each year (in sequence) from 1901 - 2010.
-
-if(!'then.rast.rds' %in% list.files(path='data/output')){
-  then.rast <- lapply(bio.vals, function(x){
[…]

### 2015-04-11 R/resampling_climate.R

@@ -1,22 +1,29 @@
-landuse <- list.files('../../../Dropbox/REU Work/Land Use/',pattern='.tif$', full.names=TRUE)
+#landuse <- list.files('../../../Dropbox/REU Work/Land Use/',pattern='.tif$', full.names=TRUE)
 
-landuse <- landuse[c(1, 4, 6, 7, 8, 9, 10, 16)]
-landuse.r <- (stack(landuse))
-landuse.r[landuse.r>100] <- NA
-landuse <- sum(landuse.r)>0
+#landuse <- landuse[c(1, 4, 6, 7, 8, 9, 10, 16)]
+# landuse.r <- (stack(landuse))
+# landuse.r[landuse.r>100] <- NA
+# landuse <- sum(landuse.r)>0
+# 
+source('R/load_fia.R')
 
 agg.dens[is.na(agg.dens)] <- 0
 agg.dens[,-1] <- agg.dens[,-1] / rowSums(agg.dens[,-1])
+
+plss.density[,4:ncol(plss.density)] <- plss.density[,4:ncol(plss.density)]/rowSums(plss.density[,4:ncol(plss.density)])
 
 all.taxa <- names(western$var)
 
 all.taxa <- all.taxa[!all.taxa%in% c('Other hardwood', 'Atlantic White Cedar',
   'Buckeye', 'Ironwood', 'Walnut', 'Black gum/sweet gum')]
 
-unit.raster <- setValues(base.rast, 1:ncell(base.rast))
-proj4string(landuse) <- '+proj=longlat +ellps=WGS84'
+colnames(plss.density) <- gsub('.', '/', colnames(plss.density), fixed=TRUE)
 
-landuse <- projectRaster(landuse, unit.raster) > 0.9
+plss.density$Poplar <- plss.density$Poplar + plss.density$'Tulip/poplar'
+colnames(plss.density)[which(colnames(plss.density) == 'Poplar')] <- "Poplar/tulip poplar"
+
+#proj4string(landuse) <- '+proj=longlat +ellps=WGS84'
+#landuse <- projectRaster(landuse, unit.raster) > 0.9
 
 #For each taxon sample:
 #  1.  The FIA data for one of the now.rast years for each of the climate variables.
@@ -28,8 +35,9 @@ hellinger <- function(kernel1, kernel2){
   1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
 }
 
-ranges <- data.frame(min = c(100, 0, 220, 155, -383),
[…]

### 2015-04-12 R/resampling_climate.R
For `resampling_climate.R`: Fixed re-sampling so that it only selects the four climate variables (rather than the five).  Modified plotting, added a `ggsave`, recoded the hellinger distance lapply section, but it's not quite finished.

@@ -103,7 +103,7 @@ get_dens_era <- function(taxon, clim.var){
 #  This gives us 100 samples from the data for both PLSS and FIA: 
 data.tables <-   llply(all.taxa,
                         function(x){
-                          lapply(1:5, function(y){
+                          lapply(1:4, function(y){
                             do.call(rbind.data.frame, 
                               lapply(1, function(z){
                                 aa <- try(get_dens_era(x, y))
@@ -115,7 +115,7 @@ data.tables <-   llply(all.taxa,
                         }, .progress = 'text')
 
 
-data.out <- lapply(1:15, function(x)lapply(1:5, function(y){
+data.out <- lapply(1:15, function(x)lapply(1:4, function(y){
   aa <- data.tables[[x]][[y]]
   aa$taxon <- all.taxa[x]
   aa$climate <- bio.vals[y]
@@ -127,12 +127,14 @@ bb <- do.call(rbind.data.frame, aa)
 w.means <- ddply(bb, .(base, clim, taxon), summarise, 
       x = weighted.mean(climate, data, na.rm=TRUE))
 
-bb$climate <- factor(bb$climate, labels = c('Pwet', 'Pdry', 'Tdiff', 'Twarm', 'Tcold'))
+bb$climate <- factor(bb$climate, labels = c('Pann', 'Tmax', 'Tmean', 'Tcold'))
 
-diff.output <- ggplot(bb, aes(x = clim, fill = base, color = c.ref)) + 
-                      geom_density(alpha = 0.2) + 
+diff.output <- ggplot(subset(na.omit(bb), data > 0.01 & taxon %in% c('Oak', 'Basswood', 'Tamarack')), 
+                      aes(x = clim, fill = base, color = c.ref)) + 
+                      geom_density(alpha = 0.2, trim=TRUE) + 
                       facet_grid(taxon~climate, scale = 'free')
 
+ggsave(filename = '../../../Dropbox/diffoutput.tiff', plot = diff.output, width = 5, height = 15)
 #  Now we have the presence or absence of taxa at PLSS and FIA eras
 #  what are we looking for?
 #  We want to be able to predict the shape of the PDF for a modern taxon distribution
@@ -143,6 +145,21 @@ diff.output <- ggplot(bb, aes(x = clim, fill = base, color = c.ref)) +
[…]

### 2015-04-12 R/load_climate.R
Should mostly be local changes that don't need to be resolved. . .

@@ -65,5 +65,5 @@ if(!'now.rast.rds' %in% list.files(path='data/output')){
                                              base.rast))
   saveRDS(now.rast, file = 'data/output/now.rast.rds')
 } else {
-  now.rast <- readRDS('data/output/now.rast.RData')
+  now.rast <- readRDS('data/output/now.rast.rds')
 }

### 2015-04-12 R/resampling_climate.R

@@ -50,7 +50,7 @@ get_dens_era <- function(taxon, clim.var){
   layer_ncdf <- sample(500, 1)                      # there are 500 layers in the ncdf
   taxon.vals <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
 
-  #  bind all the 
+  #  bind all the values together, with repeating x & y coordinates.
   
   values <- data.frame(x     = c(rep(western.grid$x, 2),
                                  rep(xyFromCell(base.rast, agg.dens$cell)[,'x'],2),
@@ -102,20 +102,28 @@ get_dens_era <- function(taxon, clim.var){
   values@data
 }
 
-#  This gives us 100 samples from the data for both PLSS and FIA: 
-data.tables <-   llply(all.taxa,
-                        function(x){
-                          lapply(1:5, function(y){
-                            do.call(rbind.data.frame, 
-                              lapply(1, function(z){
-                                aa <- try(get_dens_era(x, y))
-                                if(class(aa) == 'try-error'){
-                                  aa <- data.frame(data=NA, base=NA, c.ref = NA, cell=NA, pres=NA, climate = NA)
-                                }
-                                aa}))
-                          })
-                        }, .progress = 'text')
-
+if('data.tables.rds' %in% list.files('data/output')){
+  #  This gives us 100 samples from the data for both PLSS and FIA, binding the climate and
+  #  presence data together, the data is structured as a list of lists with length X, each
+  #  with length 5 for each of the climate variables.
+  
+  #  Some of these tables are returned as NA, these are largely datasets with very few samples.
+  data.tables <-   llply(all.taxa,
+                          function(x){
+                            lapply(1:5, function(y){
+                              do.call(rbind.data.frame, 
+                                lapply(1, function(z){
+                                  aa <- try(get_dens_era(x, y))
+                                  if(class(aa) == 'try-error'){
+                                    aa <- data.frame(data=NA, base=NA, c.ref = NA, cell=NA, pres=NA, climate = NA)
+                                  }
+                                  aa}))
+                            })
+                          }, .progress = 'text')
[…]

### 2015-04-13 Composition_Ecology.Rmd
Should mostly be local changes that don't need to be resolved. . .
Added `gridExtra` to the list of loaded packages.
Big changes to `resampling_climate`, removed early text blobs, incuding `landuse` section.
* Moved `hellinger` function to new file.
* removed the need for a spatial points object in `get_dens_era`
* turned off text output in the bit `data.table` `ldply`.
* removed a whole huge chunk of code to make things easier to do.

@@
 
-```{r, fig.width=5}
+```{r, echo = FALSE, warning = FALSE, message = FALSE}
 # Load climate data and generate the pre-sett and modern bounds:
@@
 source('R/load_climate.R') # gives then.clim & now.clim
-source('R/load_fia.R')     # gives agg.dens
-source('R/load_ncdf.R')    # gives western and eastern
+source('R/load_veg.R')
 
@@
 ---------------
-Across the region PLS cells account for X
+The Public Lands data is spatially extensive, sampling occured across the region in a regular pattern, while the FIA is limited to regions with forest cover.
+As such, the extent of points is not overlapping.
+Add to this the extensive forest cover change in the region and we see significant changes in cover between FIA and PLS data.
+
+```{r loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE}
+
+source('R/figures/gain_loss_plot.R')
+
+```
+**Figure 1**.  *Gain loss and continuous presence for PLSS and FIA tree taxa.*
+
+The shifts in distribution to lead to changes in the apparent climate envelopes of the taxa, but these are dwarfed by the shifts in regional climate across the area.
+```{r densityplots, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width=7}
+
+source('R/resampling_climate.R')
+source('R/figures//climate_boxplots.R')
+
+```
+
+Whe we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
+
+```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE}
+
+source('R/hellinger_dists.R')
+
+ggplot(all.hellinger, aes(x = fixedv_c, y = fixedc_v, color = climate)) +
+  geom_point() + geom_abline(intercept=0, slope = 1) + facet_wrap(~shift) +
+  xlab('Vegetation Fixed - Climate Shifting') +
+  ylab('Climate Fixed - Vegetation Shifting') +
+  theme_bw()
+
+```
+
+This plot makes it clear that precipitation across the region has changed significantly since the PLS era (or the early 1900s), while overall $t_{mean}$ has changed less.
+Among taxa only beech, ironwood, hemlock, fir and tamarak show distributional changes larger than the effect of shifting climate.
+For beech, a shift is seen in $t_{mean}$, where $t_{mean}$ for the PLS era distribution is 6.8^o^C, but is lower for the FIA distribution using PLS era climate, at $t_{mean}$ = 6.6^o^C.
+This is likely due to a northward shift in beech distribution in the current FIA dataset due to the loss of southern beech forests (earlier Figure).
 

### 2015-04-13 R/Setup_R.R

@@ -5,7 +5,7 @@ library(mgcv)
 library(reshape2)
 library(plyr)
 library(ncdf4)
-
+library(gridExtra)
 
 base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
 base.rast <- setValues(base.rast, 1:ncell(base.rast))

### 2015-04-13 R/resampling_climate.R

@@ -1,39 +1,11 @@
-#landuse <- list.files('../../../Dropbox/REU Work/Land Use/',pattern='.tif$', full.names=TRUE)
 
-#landuse <- landuse[c(1, 4, 6, 7, 8, 9, 10, 16)]
-# landuse.r <- (stack(landuse))
-# landuse.r[landuse.r>100] <- NA
-# landuse <- sum(landuse.r)>0
-# 
-source('R/load_fia.R')
-
-agg.dens[is.na(agg.dens)] <- 0
 agg.dens[,-1] <- agg.dens[,-1] / rowSums(agg.dens[,-1])
-
-plss.density[,4:ncol(plss.density)] <- plss.density[,4:ncol(plss.density)]/rowSums(plss.density[,4:ncol(plss.density)])
-
-all.taxa <- names(western$var)
-
-all.taxa <- all.taxa[!all.taxa%in% c('Other hardwood', 'Atlantic White Cedar',
-  'Buckeye', 'Ironwood', 'Walnut', 'Black gum/sweet gum')]
-
-colnames(plss.density) <- gsub('.', '/', colnames(plss.density), fixed=TRUE)
-
-plss.density$Poplar <- plss.density$Poplar + plss.density$'Tulip/poplar'
-colnames(plss.density)[which(colnames(plss.density) == 'Poplar')] <- "Poplar/tulip poplar"
-
-#proj4string(landuse) <- '+proj=longlat +ellps=WGS84'
-#landuse <- projectRaster(landuse, unit.raster) > 0.9
+pls.data[,-1] <- pls.data[,-1]/rowSums(pls.data[,-1])
 
 #For each taxon sample:
 #  1.  The FIA data for one of the now.rast years for each of the climate variables.
 #  2.  One of the then.rast years & one of the western layers
 #  3.  Get vectors of values
-
-hellinger <- function(kernel1, kernel2){
-  #  Calculate hellinger distance (between two kernels)
-  1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
-}
 
 # These are the climate ranges:
 ranges <- data.frame(min = c(400,  15, 0, -30),
@@ -57,48 +29,38 @@ get_dens_era <- function(taxon, clim.var){
   #taxon.vals <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
 
   #  bind all the variables together:  
-  values <- data.frame(x     = c(rep(plss.density$x, 2),
[…]

### 2015-04-13 R/hellinger_dists.R
The new `hellinger` file moves the old function to its own file for ease of editing.  Returns the `all.hellinger` object.
Added `load_veg` to pull in the FIA data and PLSS data together.

Added the newest version (v0.9-4) of the settlement vegetation data.

@@ -1,0 +1,44 @@
+#  To run the hellinger distances:
+
+hellinger <- function(clim1, clim2){
+  
+  center <- mean(c(clim1, clim2), na.rm = TRUE)
+  ranger <- diff(range(c(clim1, clim2),na.rm=TRUE))
+  
+  clim1 <- (clim1 - center) / ranger
+  clim2 <- (clim2 - center) / ranger
+  
+  clim.range <- range(c(clim1, clim2), na.rm=TRUE)
+  
+  kernel1 <- density(clim1, na.rm=TRUE, 
+                     from = clim.range[1], to = clim.range[2])$y
+  kernel2 <- density(clim2, na.rm=TRUE, 
+                     from = clim.range[1], to = clim.range[2])$y
+  
+  #  Calculate hellinger distance (between two kernels)
+  1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
+  
+}
+
+all.hellinger <- ldply(1:length(best.taxa), function(x){
+      ldply(1:4, function(y){
+        data.subset <- vegclim.table[vegclim.table$taxon %in% best.taxa[x] & 
+                                       vegclim.table$climate %in% bio.vals[y] & 
+                                       vegclim.table$data > 0,]
+        
+        p_shift_c <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
+                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
+        f_shift_c <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'PLSS')$clim,
+                               subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim)
+        p_shift_v <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
+                               subset(data.subset, base == 'FIA' & c.ref == 'PLSS')$clim)
+        f_shift_v <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim,
+                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
+        
+        data.frame(taxon = best.taxa[x],
+                   climate = bio.vals[y],
+                   fixedv_c = c(p_shift_c,f_shift_c),
+                   shift=c('PLS', 'FIA'),
+                   fixedc_v = c(p_shift_v, f_shift_v))
+      })
+    })

### 2015-04-13 R/load_veg.R

@@ -1,0 +1,22 @@
+source('R/load_fia.R')
+
+pls.data <- read.csv('data/plss_density_alb_v0.9-4.csv', row.names=1)
+colnames(pls.data) <- tolower(colnames(pls.data))
+
+colnames(agg.dens) <- tolower(gsub(' ', '.', colnames(agg.dens), fixed=TRUE))
+
+colnames(agg.dens) <- gsub('/', '.', colnames(agg.dens), fixed=TRUE)
+
+agg.dens$poplar <- agg.dens$poplar.tulip.poplar
+
+best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
+              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
+              'tamarack', 'maple', 'poplar', 'birch')
+
+agg.dens <- agg.dens[,c('cell', best.taxa)]
+pls.data <- pls.data[,c('cell', best.taxa)]
+
+agg.dens[is.na(agg.dens)] <- 0
+
+cell.xy <- data.frame(cell = 1:ncell(unit.raster),
+                     xyFromCell(unit.raster, 1:ncell(unit.raster)))

### 2015-05-04 Composition_Ecology.Rmd
Minor changes.

@@
 ggplot(all.hellinger, aes(x = fixedv_c, y = fixedc_v, color = climate)) +
-  geom_point() + geom_abline(intercept=0, slope = 1) + facet_wrap(~shift) +
+  geom_point(size = 2) + geom_abline(intercept=0, slope = 1) + facet_wrap(~shift) +
   xlab('Vegetation Fixed - Climate Shifting') +

### 2015-05-04 R/figures/climate_boxplots.R
Added the `figures` R files.  Updated the `.gitignore`.

@@ -1,0 +1,24 @@
+# plot resampling:
+
+ppt.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'ppt'), 
+                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
+  geom_boxplot(alpha=0.2) +
+  xlab('') + ylab('') +
+  facet_grid(taxon ~ climate, scales='free') + theme_bw()
+tma.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmax'), 
+                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
+  geom_boxplot(alpha=0.2) +
+  xlab('') + ylab('') +            
+  facet_grid(taxon ~ climate, scales='free') + theme_bw()
+tmi.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmin'), 
+                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
+  geom_boxplot(alpha=0.2) +
+  xlab('') + ylab('') +
+  facet_grid(taxon ~ climate, scales='free') + theme_bw()
+tme.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmean'), 
+                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
+  geom_boxplot(alpha=0.2) +
+  xlab('') + ylab('') +
+  facet_grid(taxon ~ climate, scales='free') + theme_bw()
+
+grid.arrange(ppt.plot, tma.plot, tmi.plot, tme.plot, nrow=1, sub='')

### 2015-05-04 R/figures/gain_loss_plot.R

@@ -1,0 +1,27 @@
+loss.gain <- data.frame(cell = 1:ncell(unit.raster))
+
+for(i in 1:length(best.taxa)){
+  loss.gain[,(i+1)] <- NA
+  colnames(loss.gain)[i+1] <- best.taxa[i]
+  
+  sub.pls <- na.omit(pls.data[pls.data[,best.taxa[i]] > 0 & !is.na(pls.data$cell),])
+  sub.fia <- na.omit(agg.dens[agg.dens[,best.taxa[i]] > 0 & !is.na(agg.dens$cell),])
+  
+  #  PLS is there, but FIA is not.
+  loss.gain[sub.pls$cell[!sub.pls$cell %in% sub.fia$cell],best.taxa[i]] <- 'Loss'
+  #  PLS and FIA are there:
+  loss.gain[sub.pls$cell[sub.pls$cell %in% sub.fia$cell],best.taxa[i]] <- 'Presence'
+  # FIA is there, PLS is not:
+  loss.gain[sub.fia$cell[!sub.fia$cell %in% sub.pls$cell],best.taxa[i]] <- 'Gain'
+  
+}
+
+loss.melt <- na.omit(melt(loss.gain, id = 'cell'))
+
+loss.melt <- data.frame(xyFromCell(unit.raster, loss.melt$cell),
+                        loss.melt)
+
+ggplot(loss.melt, aes(x = x, y = y, fill = factor(value))) + geom_tile() +
+  facet_wrap(~variable) + theme_bw() +
+  scale_fill_brewer(type='qual')
+

### 2015-05-04 R/figures/gain_loss_plot.R
Making the plots prettier.

@@ -23,5 +23,10 @@ loss.melt <- data.frame(xyFromCell(unit.raster, loss.melt$cell),
 
 ggplot(loss.melt, aes(x = x, y = y, fill = factor(value))) + geom_tile() +
   facet_wrap(~variable) + theme_bw() +
-  scale_fill_brewer(type='qual')
+  scale_fill_brewer(type='qual', name = 'Change State') +
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        strip.text = element_text(family = 'serif', face='bold', size = 14),
+        legend.title = element_text(family = 'serif', face='bold', size = 14),
+        legend.text = element_text(family = 'serif', size = 12))
 

### 2015-05-12 Composition_Ecology.Rmd
Added the Natural Earth module, some text including explanation of climate trends.
Added overlapping distributions and text.

Setup R addes some code to allow NeturalEarth module to run.

Building small climate gives us the cropped PRISM data.

Modifications to Hellinger distance to allow direct comaprisons of intensity of shifts.

--- 2015-04-13 Composition_Ecology.Rmd
@@
 source('R/load_veg.R')
+source('R/natural_earth.R')
 
@@
 
-The shifts in distribution to lead to changes in the apparent climate envelopes of the taxa, but these are dwarfed by the shifts in regional climate across the area.
-```{r densityplots, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width=7}
+```{r changing_climate, warning=FALSE, message=FALSE, echo=FALSE}
+
+source('R/figures/climate_ts.R')
+
+changes <- modern$value - past$value
+```
+
+During this time climate has also changed.
+Mean temperatures in the region show an average increase of `r round(changes[3], 2)`^o^C since the 1895-1925 normal.
+Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2`^o^C)).
+The most significant change has been in annual minimum temperatures, which have increased by `r round(changes[4], 1)` since the 1895-1925 normal.
+This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.
+
+```{r densities}
 
 source('R/resampling_climate.R')
-source('R/figures//climate_boxplots.R')
+
+source('overlapping_distributions.R')
 
 ```
+**Figure** *Changes in the overall climate space in the upper midwestern United States.
+The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*
 
-Whe we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
+While these changes in climate are visible across the whole range, we would expect that species with significant shifts in distribution as a result of land use change in the post-settlement era should show changes that superceed the changes resulting from climate alone.
+In particular, if the land use pressure is biased spatially or climatically then these changes should be even more dramatic.
+Hellinger distance is used to understand the difference in the shape of kernel densities.
+Using the distributions in climate space for each of the key taxa we can shift climate, or vegetation to understand the relative effects of land use change and climatic change across the region.
 
@@
 ggplot(all.hellinger, aes(x = fixedv_c, y = fixedc_v, color = climate)) +
-  geom_point() + geom_abline(intercept=0, slope = 1) + facet_wrap(~shift) +
+  geom_point(size = 2) + 
+  geom_abline(intercept=0, slope = 1) + 
+  facet_grid(shift ~ climate) +
   xlab('Vegetation Fixed - Climate Shifting') +
   ylab('Climate Fixed - Vegetation Shifting') +
-  theme_bw()
+  theme_bw() +
+  theme(axis.text = element_text(family = 'serif', size = 12),
+        axis.title = element_text(family = 'serif', face = 'bold', size = 14),
+        strip.text = element_text(family = 'serif', face='bold', size = 14),
+        legend.title = element_text(family = 'serif', face='bold', size = 14))
 
 ```
+
+The Hellinger distances reveal that
+
+```{r densityplots, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width=7}
+source('R/figures//climate_boxplots.R')
+
+```
+
+Whe we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
 

### 2015-05-12 R/Setup_R.R

@@ -12,3 +12,8 @@ base.rast <- setValues(base.rast, 1:ncell(base.rast))
 
 pls.rast <- raster('data/input//dens_v0.9-4.tif')
 unit.raster <- setValues(pls.rast, 1:ncell(pls.rast))
+
+model.proj <- '+init=epsg:3175'
+
+usa <- readOGR('Maps/us.shp', 'us')
+canada <- readOGR('Maps/PROVINCE.SHP', 'PROVINCE')

### 2015-05-12 R/building_small_climate.R

@@ -95,3 +95,41 @@ regrid <- function(i, class){
 
 #  Very slow, but only needs to be done once:
 lapply(c('norm'), function(x) lapply(1:4, function(y) regrid(y, x)))
+
+
+#  We need to get a plot of each value through time for the domain:
+variables <- c('ppt', 'tmax', 'tmin', 'tmean')
+
+clim <- list()
+
+for(i in 1:4){
+  
+  setfiles <- list.files(paste0('../../GriddedClimate/prism/PRISM_',variables[i], 
+                                '_stable_4kmM2_189501_198012_bil'),full.names=TRUE)
+  setfiles <- setfiles[regexpr('.bil$', setfiles)>0]
+  
+  # This gives all monthly values, we're going to stack them and then
+  #  average all of them:
+  
+  clim[[i]] <- data.frame(param = variables[i],
+                          val =   rep(NA, length(setfiles)),
+                          year =  rep(c(1:12), 86),
+                          month = rep(1895:1980, each=12))
+  
+  for(j in 1:length(setfiles)){
+    
+    norm <- raster(setfiles[j])
+    norm <- projectRaster(crop(norm, extent(c(-99, -64, 35, 51))), paleon.grid)
+    norm[is.na(paleon.grid)] <- NA
+    clim[[i]]$val[j] <- mean(getValues(norm), na.rm=TRUE)
+  }    
+  
+}
+
+clim.year <- data.frame(ppt = as.numeric(dcast(clim[[1]], param ~ month, fun.aggregate = sum, value.var = 'val', na.rm=TRUE)[,-1]),
+                        tmx = as.numeric(dcast(clim[[2]], param ~ month, fun.aggregate = max, value.var = 'val', na.rm=TRUE)[,-1]),
+                        tmi = as.numeric(dcast(clim[[3]], param ~ month, fun.aggregate = min, value.var = 'val', na.rm=TRUE)[,-1]),
+                        tmn = as.numeric(dcast(clim[[4]], param ~ month, fun.aggregate = mean, value.var = 'val', na.rm=TRUE)[,-1]))
+
+clim.yout <- data.frame(melt(clim.year), year = 1895:1980)
+write.csv(clim.yout, 'data//input/annualclim.csv')

### 2015-05-12 R/figures/climate_ts.R

@@ -1,0 +1,42 @@
+climate <- read.csv('data/input/annualclim.csv')[,-1]
+
+#load the PDSI data:
+#  Load the PDSI data for the region:
+
+NCDF <- stack('data/NADApdsi04.nc')
+ncdf.resamp <- resample(projectRaster(NCDF, crs = '+init=epsg:3175'), y = unit.raster)
+pdsi <- getValues(ncdf.resamp)[pls.data$cell,]
+pdsi_df <- data.frame(variable = 'pdsi', 
+                      value = colMeans(pdsi, na.rm=TRUE)[1:205],
+                      year = 2005:1801)
+
+climate <- rbind(climate, pdsi_df)
+
+now.mask <- now.rast
+now.mask[is.na(pls.rast)] <- NA
+
+then.mask <- then.rast
+then.mask[is.na(pls.rast)] <- NA
+
+modern <- colMeans(getValues(now.mask), na.rm=TRUE)
+past   <- colMeans(getValues(then.mask), na.rm=TRUE)
+
+modern <- data.frame(variable = c('ppt', 'tmx', 'tmn', 'tmi'),
+                     value = colMeans(getValues(now.mask), na.rm=TRUE),
+                     year = 1990)
+
+past <- data.frame(variable = c('ppt', 'tmx', 'tmn', 'tmi'),
+                     value = colMeans(getValues(then.mask), na.rm=TRUE),
+                     year = 1905)
+
+clim_plot <- ggplot(data = climate, aes(x = year, y = value)) + geom_line(size = 2) +
+  geom_point(data = modern, aes(x = year, y = value), size = 4, color = 'red') +
+  geom_point(data = past, aes(x = year, y = value), size = 4, color = 'red') +
+  facet_wrap(~variable, ncol = 1, scales='free_y') +
+  theme(axis.text = element_text(family = 'serif', size = 12),
+        axis.title = element_text(family = 'serif', face = 'bold', size = 14),
+        strip.text = element_text(family = 'serif', face='bold', size = 14),
+        legend.position = "none") +
+  theme_bw()
+  
+ggsave(clim_plot, filename = 'figures/clim_plot.tiff', width = 8, height = 6, dpi = 150)

### 2015-05-12 R/figures/density_plots.R

@@ -1,0 +1,13 @@
+#  Forget the backwards plot.
+density.plots <- ggplot(subset(na.omit(vegclim.table), data > 0 & 
+                                 !(taxon %in% 'ironwood') & 
+                                 (base == 'PLSS' | (base == 'FIA' & c.ref == 'FIA')))) + 
+    geom_density(aes(x = clim, y = ..scaled.., size = base, fill = c.ref), alpha = 0.2) +
+  scale_size_manual(values = c(0.5, 1)) +
+    facet_grid(taxon ~ climate, scales = 'free') +
+  theme_bw()  +
+    theme(axis.text = element_text(family = 'serif', size = 12),
+          axis.title = element_text(family = 'serif', face = 'bold', size = 14),
+          strip.text = element_text(family = 'serif', face='bold', size = 14))
+
+ggsave(density.plots, filename = 'figures/densityplots.tiff', height = 8, width = 8, dpi = 150)

### 2015-05-12 R/figures/gain_loss_plot.R

@@ -21,12 +21,18 @@ loss.melt <- na.omit(melt(loss.gain, id = 'cell'))
 loss.melt <- data.frame(xyFromCell(unit.raster, loss.melt$cell),
                         loss.melt)
 
-ggplot(loss.melt, aes(x = x, y = y, fill = factor(value))) + geom_tile() +
-  facet_wrap(~variable) + theme_bw() +
-  scale_fill_brewer(type='qual', name = 'Change State') +
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        strip.text = element_text(family = 'serif', face='bold', size = 14),
-        legend.title = element_text(family = 'serif', face='bold', size = 14),
-        legend.text = element_text(family = 'serif', size = 12))
+loss.plot <- base.map +
+    geom_tile(data = subset(loss.melt, !variable %in% 'ironwood') , aes(x = x, y = y, fill = factor(value))) +
+    facet_wrap(~variable, ncol = 5) + theme_bw() +
+    scale_fill_brewer(palette = 'RdYlGn', name = 'Change State') +
+    geom_path(data = umw.domain, aes(x = long, y = lat, group = group)) +
+    geom_path(data = can.domain, aes(x = long, y = lat, group = group)) +
+    theme(axis.text = element_blank(),
+          axis.title = element_blank(),
+          strip.text = element_text(family = 'serif', face='bold', size = 14),
+          legend.title = element_text(family = 'serif', face='bold', size = 14),
+          legend.text = element_text(family = 'serif', size = 12)) +
+  coord_equal(xlim = ext[1:2], ylim=ext[3:4])
+
+ggsave(plot = loss.plot, filename = 'figures/loss_plots.tiff', dpi = 150, width = 6, height = 4)
 

### 2015-05-12 R/figures/overlapping_distributions.R

@@ -1,0 +1,17 @@
+
+overlapping_clim <- ggplot(vegclim.table, aes(x = clim)) + 
+  geom_density(aes(fill = c.ref), alpha = 0.5) + 
+  facet_wrap(~climate, scale='free', nrow=1) +
+  xlab('Climate Parameter') + ylab('Kernel Density') +
+  theme(axis.text.x = element_text(family = 'serif', size = 12),
+        axis.text.y = element_blank(),
+        axis.title = element_text(family = 'serif', face = 'bold', size = 14),
+        strip.text = element_text(family = 'serif', face='bold', size = 14),
+        legend.title = element_text(family = 'serif', face='bold', size = 14)) +
+  labs(fill = 'Era') +
+  scale_x_continuous(expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0))
+
+ggsave(overlapping_clim, 
+       filename = 'figures/overlappingdensity.tiff', 
+       dpi = 150, width = 8, height = 4)

### 2015-05-12 R/hellinger_dists.R

@@ -10,13 +10,13 @@ hellinger <- function(clim1, clim2){
   
   clim.range <- range(c(clim1, clim2), na.rm=TRUE)
   
-  kernel1 <- density(clim1, na.rm=TRUE, 
-                     from = clim.range[1], to = clim.range[2])$y
-  kernel2 <- density(clim2, na.rm=TRUE, 
-                     from = clim.range[1], to = clim.range[2])$y
+  kernel1 <- scale(density(clim1, na.rm=TRUE, 
+                     from = clim.range[1], to = clim.range[2])$y)
+  kernel2 <- scale(density(clim2, na.rm=TRUE, 
+                     from = clim.range[1], to = clim.range[2])$y)
   
   #  Calculate hellinger distance (between two kernels)
-  1/sqrt(2) * sqrt(sum(sqrt(kernel1) - sqrt(kernel2))^2)
+  1/sqrt(2) * sqrt(sum(sqrt(kernel1 - min(kernel1)) - sqrt(kernel2-min(kernel2)))^2)
   
 }
 
@@ -26,19 +26,19 @@ all.hellinger <- ldply(1:length(best.taxa), function(x){
                                        vegclim.table$climate %in% bio.vals[y] & 
                                        vegclim.table$data > 0,]
         
-        p_shift_c <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
-                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
-        f_shift_c <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'PLSS')$clim,
+        full_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
                                subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim)
-        p_shift_v <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
-                               subset(data.subset, base == 'FIA' & c.ref == 'PLSS')$clim)
-        f_shift_v <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim,
+        
+        clim_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
+                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
+        
+        lu_shift   <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim,
[…]

### 2015-05-12 R/natural_earth.R

@@ -1,0 +1,78 @@
+#  Loading in the Natural Earth datasets:
+#  This works in general, except that
+
+library(rgeos)
+
+umw.domain <- usa
+
+can.domain <- canada
+
+#  Load & clip the base raster.  This is broader than the clipping extent for the 
+#  domain, and it helps speed up the use of the natural earth layers later in the
+#  analysis.
+#  Natural earth data is stored in the Maps folder, and comes from: www.naturalearthdata.com
+#  Here we are using the 10m data.
+
+nat.earth <- crop(stack('Maps/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
+
+xylimits <-  c(-200000, 1150000, 500000, 1700000)
+
+if(model.proj == '+init=epsg:4326'){
+  ext <- c(-98, -83, 42, 50)
+}
+
+if(model.proj == '+init=epsg:3175'){
+  ext <- c(-100000, 1050000, 600000, 1600000)
+}
+
+quick.subset <- function(x, y, ext){
+  # Subsets a shapefile (x = file, y = layer) to a given extent.
+  shape <- readOGR(x, y)
+  
+  shape@data$id <- rownames(shape@data)
+  
+  x.f = fortify(shape, region="id")
+  x.join = join(x.f, shape@data, by="id")
+  
+  aa <- x.join[,1:2]
+  coordinates(aa) <- ~ long + lat
+  proj4string(aa) <- CRS(proj4string(nat.earth))
+  aa <- coordinates(spTransform(aa, CRS(model.proj)))
+  
+  x.join[,1:2] <- aa
+  
+  x.subset <- subset(x.join, x.join$long > xylimits[1] & x.join$long < xylimits[2] &
+                           x.join$lat > xylimits[3] & x.join$lat < xylimits[4])
+  
+  x.subset
+}
+
+lakes.subset <- quick.subset('Maps/NaturalEarth/Lakes/ne_10m_lakes.shp',
+                             'ne_10m_lakes', ext)
+river.subset <- quick.subset('Maps/NaturalEarth/Rivers/ne_10m_rivers_lake_centerlines.shp',
+                             'ne_10m_rivers_lake_centerlines', ext)
+coast.subset <- quick.subset('Maps/NaturalEarth/Coasts/ne_10m_coastline.shp',
+                             'ne_10m_coastline', ext)
+
[…]

### 2015-05-12 R/resampling_climate.R

@@ -2,6 +2,9 @@
 agg.dens[,-1] <- agg.dens[,-1] / rowSums(agg.dens[,-1])
 pls.data[,-1] <- pls.data[,-1]/rowSums(pls.data[,-1])
 
+then.rast <- resample(then.rast, unit.raster)
+now.rast <- resample(now.rast, unit.raster)
+
 #For each taxon sample:
 #  1.  The FIA data for one of the now.rast years for each of the climate variables.
 #  2.  One of the then.rast years & one of the western layers
@@ -30,10 +33,10 @@ get_dens_era <- function(taxon, clim.var){
 
   #  bind all the variables together:  
   values <- data.frame(cell  = c(rep(pls.data$cell, 2), rep(agg.dens$cell, 2)),
-                       x     = c(rep(xyFromCell(base.rast, pls.data$cell)[,'x'],2),
-                                 rep(xyFromCell(base.rast, agg.dens$cell)[,'x'],2)),
-                       y     = c(rep(xyFromCell(base.rast, pls.data$cell)[,'y'],2),
-                                 rep(xyFromCell(base.rast, agg.dens$cell)[,'y'],2)),
+                       x     = c(rep(xyFromCell(unit.raster, pls.data$cell)[,'x'],2),
+                                 rep(xyFromCell(unit.raster, agg.dens$cell)[,'x'],2)),
+                       y     = c(rep(xyFromCell(unit.raster, pls.data$cell)[,'y'],2),
+                                 rep(xyFromCell(unit.raster, agg.dens$cell)[,'y'],2)),
                        data  = c(rep(as.numeric(pls.data[,taxon]), 2),
                                  rep(agg.dens[,taxon],2)),
                        base  = c(rep('PLSS', nrow(pls.data)), 
@@ -46,11 +49,13 @@ get_dens_era <- function(taxon, clim.var){
                                  rep('PLSS', nrow(agg.dens))),
                        stringsAsFactors = FALSE)
   
+#  The problem here is that both 
+  
   plss.plss.clim <- getValues(then.rast)[subset(values, base %in% 'PLSS' & c.ref %in% 'PLSS')$cell, clim.var]
   plss.fia.clim  <- getValues(now.rast)[subset(values, base %in% 'PLSS' & c.ref %in% 'FIA')$cell, clim.var]
   
[…]

### 2015-05-13 R/Setup_R.R
I'm trying to wrap a bunch of the subroutines into functions, even if they're only called once.  This way I'm not creating a whole mess of variables in my `source`d scripts, particularly when they're used only once.
Fixed the plotting order of the PDSI/precip/climate plots, changed the nadapdsi file and added an http reference for the parent file.

Changed file locations for the usa & canada maps and in `natural_earth`.  This needs to be standardized between work & home.

@@ -4,8 +4,8 @@ library(ggplot2)
 library(mgcv)
 library(reshape2)
 library(plyr)
-library(ncdf4)
 library(gridExtra)
+library(rgdal)
 
 base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
 base.rast <- setValues(base.rast, 1:ncell(base.rast))
@@ -15,5 +15,5 @@ unit.raster <- setValues(pls.rast, 1:ncell(pls.rast))
 
 model.proj <- '+init=epsg:3175'
 
-usa <- readOGR('Maps/us.shp', 'us')
-canada <- readOGR('Maps/PROVINCE.SHP', 'PROVINCE')
+usa <- readOGR('data/input/shapes/usa/us.shp', 'us')
+canada <- readOGR('data/input/shapes/canada/PROVINCE.SHP', 'PROVINCE')

### 2015-05-13 R/natural_earth.R

@@ -13,7 +13,7 @@ can.domain <- canada
 #  Natural earth data is stored in the Maps folder, and comes from: www.naturalearthdata.com
 #  Here we are using the 10m data.
 
-nat.earth <- crop(stack('Maps/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
+nat.earth <- crop(stack('data/input/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
 
 xylimits <-  c(-200000, 1150000, 500000, 1700000)
 
@@ -47,11 +47,11 @@ quick.subset <- function(x, y, ext){
   x.subset
 }
 
-lakes.subset <- quick.subset('Maps/NaturalEarth/Lakes/ne_10m_lakes.shp',
+lakes.subset <- quick.subset('data/input/NaturalEarth/Lakes/ne_10m_lakes.shp',
                              'ne_10m_lakes', ext)
-river.subset <- quick.subset('Maps/NaturalEarth/Rivers/ne_10m_rivers_lake_centerlines.shp',
+river.subset <- quick.subset('data/input/NaturalEarth/Rivers/ne_10m_rivers_lake_centerlines.shp',
                              'ne_10m_rivers_lake_centerlines', ext)
-coast.subset <- quick.subset('Maps/NaturalEarth/Coasts/ne_10m_coastline.shp',
+coast.subset <- quick.subset('data/input/NaturalEarth/Coasts/ne_10m_coastline.shp',
                              'ne_10m_coastline', ext)
 
 nat.crop <- crop(projectRaster(nat.earth, crs = model.proj), y=extent(ext))

### 2015-05-13 Composition_Ecology.Rmd
Heavily modified the `climate_boxplots` to show only the major shift from PLS to FIA by each taxon.
Reclassed some factor levels in `vegclim.table`, which may result in some errors down the line, particularly when calling taxa.

Moved text around for the `composition_ecology`, and edited text in the results.

@@
 ---------------
+```{r changing_climate, warning=FALSE, message=FALSE, echo=FALSE}
+
+source('R/figures/climate_ts.R')
+
+changes <- modern$value - past$value
+```
+**Figure 1**.
+*Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
+Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.
+PDSI shows strong coherence with A_ppt_ in this region.*
+
+The choice of a normal for the domain is potentially problematic.
+The time transgressive nature of the PLS Survey means that some of the PLS data comes from records sampled in the early 1800s, while other come from the early 1900s.
+Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
+As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice [ADD JULY temperatures here].
+
+Mean temperatures in the region show an average increase of `r round(changes[3], 2)`^o^C since the 1895-1925 normal.
+Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2`^o^C)).
+The most significant change has been in annual minimum temperatures, which have increased by `r round(changes[4], 1)` since the 1895-1925 normal.
+This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.
+
 The Public Lands data is spatially extensive, sampling occured across the region in a regular pattern, while the FIA is limited to regions with forest cover.
 As such, the extent of points is not overlapping.
-Add to this the extensive forest cover change in the region and we see significant changes in cover between FIA and PLS data.
+Given the extensive use of FIA data in generating and estimating climate responses of tree taxa and forest types to changing climate, the use of the FIA data as an estimate for tree species distributions is acceptable fo estimating shifts in climate space, and attributing the loss or gain of climate niche space due to land use and climate variability and change during the 20^th^ century.
 
@@
 
-```{r changing_climate, warning=FALSE, message=FALSE, echo=FALSE}
-
-source('R/figures/climate_ts.R')
-
-changes <- modern$value - past$value
-```
-
-During this time climate has also changed.
-Mean temperatures in the region show an average increase of `r round(changes[3], 2)`^o^C since the 1895-1925 normal.
-Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2`^o^C)).
-The most significant change has been in annual minimum temperatures, which have increased by `r round(changes[4], 1)` since the 1895-1925 normal.
-This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.
-
 ```{r densities}
@@
 
-source('overlapping_distributions.R')
+source('R/figures/climate_boxplots.R')
+
+boxplot_clim
 

### 2015-05-13 R/figures/climate_boxplots.R

@@ -1,24 +1,15 @@
-# plot resampling:
+# Plotting the total climate-shift for the domain.
 
-ppt.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'ppt'), 
-                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
-  geom_boxplot(alpha=0.2) +
-  xlab('') + ylab('') +
-  facet_grid(taxon ~ climate, scales='free') + theme_bw()
-tma.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmax'), 
-                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
-  geom_boxplot(alpha=0.2) +
-  xlab('') + ylab('') +            
-  facet_grid(taxon ~ climate, scales='free') + theme_bw()
-tmi.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmin'), 
-                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
-  geom_boxplot(alpha=0.2) +
-  xlab('') + ylab('') +
-  facet_grid(taxon ~ climate, scales='free') + theme_bw()
-tme.plot <- ggplot(subset(vegclim.table, data > 0.05 & climate == 'tmean'), 
-                   aes(factor(paste0(base, ', ', c.ref)), clim)) + 
-  geom_boxplot(alpha=0.2) +
-  xlab('') + ylab('') +
-  facet_grid(taxon ~ climate, scales='free') + theme_bw()
+boxplot_clim <- ggplot(data = subset(na.omit(vegclim.table), 
+                                     comb %in% c('FIA Era', 'PLS Era') & data>0)) +
+  geom_boxplot(aes(y = clim, x = taxon, fill = comb)) +
+  scale_fill_grey() +
+  facet_wrap(~climate, ncol = 1, scales='free_y') +
+  theme_bw() +
+  theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
+        axis.title = element_blank(),
+        strip.text = element_text(family = 'serif', face='bold', size = 14),
+        legend.title = element_text(family = 'serif', face='bold', size = 14),
+        legend.text = element_text(family = 'serif', size = 12))
 
-grid.arrange(ppt.plot, tma.plot, tmi.plot, tme.plot, nrow=1, sub='')
+ggsave(plot = boxplot_clim, filename = 'figure/boxplot_clim.tiff', width = 8, height = 6, dpi = 150)

### 2015-05-13 R/figures/gain_loss_plot.R

@@ -43,4 +43,3 @@ loss.plot <- base.map +
   coord_equal(xlim = ext[1:2], ylim=ext[3:4])
 
 ggsave(plot = loss.plot, filename = 'figures/loss_plots.tiff', dpi = 150, width = 6, height = 4)
-

### 2015-05-13 R/resampling_climate.R

@@ -86,3 +86,13 @@ data.out <- lapply(1:15, function(x)lapply(1:4, function(y){
 
 vegclim.table <- do.call(rbind.data.frame, lapply(data.out, function(x) do.call(rbind.data.frame, x)))
 vegclim.table <- vegclim.table[!is.na(vegclim.table$data),]
+
+vegclim.table$comb <- factor(with(vegclim.table, paste0(base, c.ref)), 
+                             levels = c('PLSSPLSS','FIAFIA', 'PLSSFIA', 'FIAPLSS'),
+                             labels = c('PLS Era', 'FIA Era', 'NA1', 'NA2'))
+
+vegclim.table$taxon <- factor(vegclim.table$taxon, 
+                              levels = c('tamarack', 'pine', 'spruce', 'fir', 'hemlock', 'cedar.juniper',
+                                         'poplar', 'maple', 'birch', 'beech', 'ironwood', 'basswood', 'ash', 'elm', 'oak'),
+                              labels = c('Tamarack', 'Pine', 'Spruce', 'Fir', 'Hemlock', 'Cedar',
+                                         'Poplar', 'Maple', 'Birch', 'Beech', 'Ironwood', 'Basswood', 'Ash', 'Elm', 'Oak'))

### 2015-05-14 Composition_Ecology.Rmd
Changed plots for the climate boxplots, added some plotting changes for the hellinger distance plots, and added an ordination plot.  I'm still not super happy about the ordination plot, but it's better.

--- 2015-05-12 Composition_Ecology.Rmd
@@
 
-source('overlapping_distributions.R')
+source('R/figures//climate_boxplots.R')
 
@@
 
-ggplot(all.hellinger, aes(x = fixedv_c, y = fixedc_v, color = climate)) +
-  geom_point(size = 2) + 
-  geom_abline(intercept=0, slope = 1) + 
-  facet_grid(shift ~ climate) +
-  xlab('Vegetation Fixed - Climate Shifting') +
-  ylab('Climate Fixed - Vegetation Shifting') +
-  theme_bw() +
-  theme(axis.text = element_text(family = 'serif', size = 12),
-        axis.title = element_text(family = 'serif', face = 'bold', size = 14),
-        strip.text = element_text(family = 'serif', face='bold', size = 14),
-        legend.title = element_text(family = 'serif', face='bold', size = 14))
+hell.plot
 
@@
 ```{r densityplots, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width=7}
-source('R/figures//climate_boxplots.R')
+
+source('R/figures//ordination_plot.R')
+ordi_plot
 

### 2015-05-14 Composition_Ecology.Rmd

--- 2015-05-13 Composition_Ecology.Rmd
@@
 
-Whe we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
+When we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
 

### 2015-06-20 Composition_Ecology.Rmd
Minor text changes.  Changes at head of Rmd to reflect changes in the sourced files.
Sourced files are now run as functions, keeping local variables internal so we don't have so many variables loaded globally.

Removed `R/load_fia.R` and put the function into `R/load_veg.R`.

@@
 source('R/load_veg.R')
+
+pls_data <- get_pls('0.9-4')
+agg_dens <- get_fia()[,colnames(pls_data)]
+
 source('R/natural_earth.R')
+
+base_map <- natural_earth()
 
@@
 Mean temperatures in the region show an average increase of `r round(changes[3], 2)`^o^C since the 1895-1925 normal.
-Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2`^o^C)).
+Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2)`^o^C)).
 The most significant change has been in annual minimum temperatures, which have increased by `r round(changes[4], 1)` since the 1895-1925 normal.

### 2015-06-20 R/load_climate.R

@@ -5,9 +5,14 @@
 #  To crop and reproject we use the R file, included hereL
 #  'R/building_small_climate.R'
 
-clim.files <- list.files('data/input/climate', full.names = TRUE)
+load_climate <- function(){
+  clim.files <- list.files('data/input/climate', full.names = TRUE)
 
-now.rast <- stack(clim.files[regexpr('norm', clim.files)>0])
-then.rast <- stack(clim.files[regexpr('early', clim.files)>0])
+  now.rast <- stack(clim.files[regexpr('norm', clim.files)>0])
+  then.rast <- stack(clim.files[regexpr('early', clim.files)>0])
 
-bio.vals <- c('ppt', 'tmax', 'tmean', 'tmin')
+  bio.vals <- c('ppt', 'tmax', 'tmean', 'tmin')
+  return(list(now = now.rast, then = then.rast, names = bio.vals))
+}
+
+climate <- load_climate()

### 2015-06-20 R/load_fia.R

@@ -1,77 +1,0 @@
-#  Load & compile the FIA data:
-
-fia.trans <- read.csv('data/FIA_conversion.csv', stringsAsFactors = FALSE)
-fia.trans[fia.trans == ''] <- NA
-
-#  Load FIA composition data:
-get.fia.stack <- function(x, type){
-  stack.files <- list.files(x, full = TRUE)
-  stack.fia <- resample(stack(stack.files), pls.rast, method = 'ngb')
-  stack.fia[is.na(getValues(pls.rast))] <- NA
-  
-  fia.vals <- getValues(stack.fia)
-  name.end <- regexpr(paste('_', type, sep=''), colnames(fia.vals), fixed=TRUE)
-  colnames(fia.vals) <- gsub("_$","", substr(colnames(fia.vals), 1, name.end-1))
-  colnames(fia.vals) <- gsub('_', '.', colnames(fia.vals))
-  
-  fia.vals <- as.data.frame(fia.vals)
-  
-  fia.vals$x <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,1]
-  fia.vals$y <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,2]
-  
-  fia.vals
-}
-
-fia.dens <- get.fia.stack('data/albers/density/', type = 'density')
-
-re.agg <- function(x){
-  # This takes the fia data (by species) and aggregates it to PalEON taxa.
-  orig <- 1:nrow(x)
-  
-  x.test <- x[,!colnames(x) %in% c('x', 'y')]
-  
-  x.test$orig <- orig
-  
-  x.melt <- melt(x.test, na.rm=TRUE, value.name= 'value', id = c('orig'))
-  
-  x.melt$equiv <- fia.trans$PalEON[match(gsub('.', ' ', x.melt$variable, fixed = TRUE), 
-                                         gsub('.', ' ', fia.trans$scientific, fixed = TRUE))]
-  
-  x.melt <- x.melt[x.melt$value > 0,]
-  
-  x.cast <- dcast(x.melt, orig ~ equiv, 
-                  sum, drop = FALSE, 
-                  value.var = 'value')
-  
-  x.cast$x <- x$x[x.cast$orig]
-  x.cast$y <- x$y[x.cast$orig]
-  
-  x.cast$cell <- extract(setValues(unit.raster, 1:ncell(unit.raster)), x.cast[,c('x', 'y')])
-  
[…]

### 2015-06-20 R/load_veg.R

@@ -1,22 +1,98 @@
-source('R/load_fia.R')
+get_fia <- function(x){
+  fia.trans <- read.csv('data/FIA_conversion.csv', stringsAsFactors = FALSE)
+  fia.trans[fia.trans == ''] <- NA
+  
+  #  Load FIA composition data:
+  get.fia.stack <- function(x, type){
+    stack.files <- list.files(x, full = TRUE)
+    stack.fia <- resample(stack(stack.files), pls.rast, method = 'ngb')
+    stack.fia[is.na(getValues(pls.rast))] <- NA
+    
+    fia.vals <- getValues(stack.fia)
+    name.end <- regexpr(paste('_', type, sep=''), colnames(fia.vals), fixed=TRUE)
+    colnames(fia.vals) <- gsub("_$","", substr(colnames(fia.vals), 1, name.end-1))
+    colnames(fia.vals) <- gsub('_', '.', colnames(fia.vals))
+    
+    fia.vals <- as.data.frame(fia.vals)
+    
+    fia.vals$x <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,1]
+    fia.vals$y <- xyFromCell(stack.fia, 1:ncell(stack.fia))[,2]
+    
+    fia.vals
+  }
+  
+  fia.dens <- get.fia.stack('data/albers/density/', type = 'density')
+  
+  re.agg <- function(x){
+    # This takes the fia data (by species) and aggregates it to PalEON taxa.
+    orig <- 1:nrow(x)
+    
+    x.test <- x[,!colnames(x) %in% c('x', 'y')]
+    
+    x.test$orig <- orig
+    
+    x.melt <- melt(x.test, na.rm=TRUE, value.name= 'value', id = c('orig'))
+    
+    x.melt$equiv <- fia.trans$PalEON[match(gsub('.', ' ', x.melt$variable, fixed = TRUE), 
+                                           gsub('.', ' ', fia.trans$scientific, fixed = TRUE))]
+    
+    x.melt <- x.melt[x.melt$value > 0,]
+    
+    x.cast <- dcast(x.melt, orig ~ equiv, 
+                    sum, drop = FALSE, 
+                    value.var = 'value')
+    
+    x.cast$x <- x$x[x.cast$orig]
+    x.cast$y <- x$y[x.cast$orig]
+    
[…]

### 2015-06-20 R/natural_earth.R

@@ -1,11 +1,6 @@
 #  Loading in the Natural Earth datasets:
-#  This works in general, except that
 
 library(rgeos)
-
-umw.domain <- usa
-
-can.domain <- canada
 
 #  Load & clip the base raster.  This is broader than the clipping extent for the 
 #  domain, and it helps speed up the use of the natural earth layers later in the
@@ -13,66 +8,69 @@ can.domain <- canada
 #  Natural earth data is stored in the Maps folder, and comes from: www.naturalearthdata.com
 #  Here we are using the 10m data.
 
-nat.earth <- crop(stack('data/input/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
-
-xylimits <-  c(-200000, 1150000, 500000, 1700000)
-
-if(model.proj == '+init=epsg:4326'){
-  ext <- c(-98, -83, 42, 50)
+natural_earth <- function(){
+  nat.earth <- crop(stack('data/input/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
+  
+  xylimits <-  c(-200000, 1150000, 500000, 1700000)
+  
+  if(model.proj == '+init=epsg:4326'){
+    ext <- c(-98, -83, 42, 50)
+  }
+  
+  if(model.proj == '+init=epsg:3175'){
+    ext <- c(-100000, 1050000, 600000, 1600000)
+  }
+  
+  quick.subset <- function(x, y, ext){
+    # Subsets a shapefile (x = file, y = layer) to a given extent.
+    shape <- readOGR(x, y)
+    
+    shape@data$id <- rownames(shape@data)
+    
+    x.f = fortify(shape, region="id")
+    x.join = join(x.f, shape@data, by="id")
+    
+    aa <- x.join[,1:2]
+    coordinates(aa) <- ~ long + lat
+    proj4string(aa) <- CRS(proj4string(nat.earth))
+    aa <- coordinates(spTransform(aa, CRS(model.proj)))
+    
+    x.join[,1:2] <- aa
+    
+    x.subset <- subset(x.join, x.join$long > xylimits[1] & x.join$long < xylimits[2] &
+                             x.join$lat > xylimits[3] & x.join$lat < xylimits[4])
+    
+    x.subset
+  }
+  
[…]

### 2015-06-24 Composition_Ecology.Rmd
Pushing changes to the structure of the code.  I've been writing all the `source`d files as functions so that there's not as much information in the global environment to worry about causing bugs.
* moved `RColorBrewer` into the setup file.

* Moved `best.taxa` up into a global variable.

* Naturalearth is a function, returns a list.

* `resample_climate` returns a value, it's all wrapped into a function.

* `climate_boxplots` becomes a function (`clim_boxplots`).

* `climate_ts` is now a function.  Returns three tables.

* `get_loss_gain` makes a function that exports the panel plot.

@@
 
+best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
+              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
+              'tamarack', 'maple', 'poplar', 'birch')
+
 source('R/natural_earth.R')
 
-base_map <- natural_earth()
+natural <- natural_earth()
 
@@
 
-changes <- modern$value - past$value
+climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
+
+changes <- climate_values$modern$value - climate_values$past$value
+
 ```
@@
 
+loss_plots <- loss_plot()
+
 ```
@@
 source('R/resampling_climate.R')
+vegclim_table <- resample_climate(climate, agg_dens, pls_data)
 
 source('R/figures/climate_boxplots.R')
-
+boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
 boxplot_clim

### 2015-06-24 R/Setup_R.R

@@ -6,6 +6,7 @@ library(reshape2)
 library(plyr)
 library(gridExtra)
 library(rgdal)
+library(RColorBrewer)
 
 base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
 base.rast <- setValues(base.rast, 1:ncell(base.rast))

### 2015-06-24 R/figures/climate_boxplots.R

@@ -1,15 +1,24 @@
 # Plotting the total climate-shift for the domain.
-
-boxplot_clim <- ggplot(data = subset(na.omit(vegclim.table), 
-                                     comb %in% c('FIA Era', 'PLS Era') & data>0)) +
-  geom_boxplot(aes(y = clim, x = taxon, fill = comb)) +
-  scale_fill_grey() +
-  facet_wrap(~climate, ncol = 1, scales='free_y') +
-  theme_bw() +
-  theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
-        axis.title = element_blank(),
-        strip.text = element_text(family = 'serif', face='bold', size = 14),
-        legend.title = element_text(family = 'serif', face='bold', size = 14),
-        legend.text = element_text(family = 'serif', size = 12))
-
-ggsave(plot = boxplot_clim, filename = 'figure/boxplot_clim.tiff', width = 8, height = 6, dpi = 150)
+clim_boxplots <- function(clim_table = vegclim_table){
+  
+  clim_table$climate <- factor(clim_table$climate, 
+                             levels = c('tmax', 'tmean', 'tmin', 'ppt'),
+                             labels = c('T[max]', 'T[mean]', 'T[min]', 'P[ann]'))
+  
+  boxplot_clim <- ggplot(data = subset(na.omit(clim_table), 
+                                       comb %in% c('FIA Era', 'PLS Era') & data>0 &
+                                         !taxon %in% 'Ironwood')) +
+    geom_boxplot(aes(y = clim, x = taxon, fill = comb)) +
+    scale_fill_grey() +
+    facet_grid(climate~., scales='free_y', labeller = label_parsed) +
+    theme_bw() +
+    theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
+          axis.title = element_blank(),
+          strip.text = element_text(family = 'serif', face='bold', size = 14),
+          legend.title = element_text(family = 'serif', face='bold', size = 14),
+          legend.text = element_text(family = 'serif', size = 12),
+          axis.text.x=element_text(angle=45, hjust = 1))
+  
+  ggsave(plot = boxplot_clim, filename = 'figures/boxplot_clim.tiff', width = 9, height = 6, dpi = 150)
+  boxplot_clim
+}

### 2015-06-24 R/figures/climate_ts.R

@@ -1,50 +1,60 @@
-climate <- read.csv('data/input/annualclim.csv')[,-1]
-
-#load the PDSI data as a function so the memory use is contained in the function.
-#  Load the PDSI data for the region:
-#  download from - http://www.ncdc.noaa.gov/paleo/pdsi.html
-pdsi_out <- function(unit.raster, pls.data){
-  NCDF <- stack('data/input/NADAv2-2008.nc')
-  ncdf.resamp <- resample(projectRaster(NCDF, crs = '+init=epsg:3175'), y = unit.raster)
-  pdsi <- getValues(ncdf.resamp)[pls.data$cell,]
-  pdsi_df <- data.frame(variable = 'pdsi', 
-                        value = colMeans(pdsi, na.rm=TRUE)[1:206],
-                        year = 2006:1801)
-  pdsi_df
-}
-
-pdsi_df <- pdsi_out(unit.raster, pls.data)
-
-climate <- rbind(climate, pdsi_df)
-
-# This removes any raster cells without PLS data to get a regional climate
-#  picture for the plotting:
-clim_pattern <- function(rast.in, pls.rast, year){
-
-  rast.mask <- rast.in
-  rast.mask[is.na(pls.rast)] <- NA
-
-  data.frame(variable = c('ppt', 'tmx', 'tmn', 'tmi'),
-             value = colMeans(getValues(rast.mask), na.rm=TRUE),
-             year = year)
-}
-
-modern <- clim_pattern(now.rast, pls.rast, 1990)
-past   <- clim_pattern(then.rast, pls.rast, 1910)
-
-climate$variable <- factor(climate$variable, 
-                           levels = c('tmx', 'tmn', 'tmi', 'ppt', 'pdsi'))
-
-clim_plot <- ggplot(data = climate, aes(x = year, y = value)) + geom_line(size = 2) +
-  geom_point(data = modern, aes(x = year, y = value), size = 4, color = 'red') +
-  geom_point(data = past, aes(x = year, y = value), size = 4, color = 'red') +
-  facet_wrap(~variable, ncol = 1, scales='free_y') +
-  ylab('') +
-  xlab('Year (CE)') +
-  theme_bw() +
-  theme(axis.text = element_text(family = 'serif', size = 12),
-        axis.title = element_text(family = 'serif', face = 'bold', size = 14),
-        strip.text = element_text(family = 'serif', face='bold', size = 14),
-        legend.position = "none")
[…]

### 2015-06-24 R/figures/gain_loss_plot.R

@@ -1,14 +1,16 @@
 #  Plotting for taxa that shows the change between historical and modern distributions.
 
-get_loss_gain <- function(unit.raster, best.taxa, pls.data, agg.dens){
+get_loss_gain <- function(unit.raster, best.taxa, pls_data, agg_dens){
   loss.gain <- data.frame(cell = 1:ncell(unit.raster))
+  
+  agg_dens[is.na(agg_dens)] <- 0
   
   for(i in 1:length(best.taxa)){
     loss.gain[,(i+1)] <- NA
     colnames(loss.gain)[i+1] <- best.taxa[i]
     
-    sub.pls <- na.omit(pls.data[pls.data[,best.taxa[i]] > 0 & !is.na(pls.data$cell),])
-    sub.fia <- na.omit(agg.dens[agg.dens[,best.taxa[i]] > 0 & !is.na(agg.dens$cell),])
+    sub.pls <- na.omit(pls_data[pls_data[,best.taxa[i]] > 0 & !is.na(pls_data$cell),])
+    sub.fia <- na.omit(agg_dens[agg_dens[,best.taxa[i]] > 0 & !is.na(agg_dens$cell),])
     
     #  PLS is there, but FIA is not.
     loss.gain[sub.pls$cell[!sub.pls$cell %in% sub.fia$cell],best.taxa[i]] <- 'Loss'
@@ -26,20 +28,61 @@ get_loss_gain <- function(unit.raster, best.taxa, pls.data, agg.dens){
   loss.melt
 }
 
-loss_melt <- get_loss_gain(unit.raster, best.taxa, pls.data, agg.dens)
-
-loss.plot <- base.map +
-    geom_tile(data = subset(loss_melt, !variable %in% 'ironwood') , aes(x = x, y = y, fill = factor(value))) +
+loss_plot <- function(){
+  
+  if(model.proj == '+init=epsg:4326'){
+    ext <- c(-98, -83, 42, 50)
+  }
+  
+  if(model.proj == '+init=epsg:3175'){
+    ext <- c(-100000, 1050000, 600000, 1600000)
+  }
+  
+  loss_melt <- get_loss_gain(unit.raster, best.taxa, pls_data, agg_dens)
+  loss_melt$rgb <- brewer.pal(n = 3, 'Dark2')[as.numeric(factor(loss_melt$value, 
+                                                                levels = c('Presence', 'Loss', 'Gain')))]
+  
[…]

### 2015-06-24 R/load_veg.R

@@ -90,9 +90,5 @@ get_pls <- function(version){
   
   agg.dens$poplar <- agg.dens$poplar.tulip.poplar
   
-  best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
-                'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
-                'tamarack', 'maple', 'poplar', 'birch')
-  
   pls.data[,c('cell', best.taxa)]
 }

### 2015-06-24 R/natural_earth.R

@@ -63,7 +63,8 @@ natural_earth <- function(){
                                          NE1_HR_LC_SR_W_DR.3,
                                         1))
   
-  ggplot() + #data = rast.table, aes(x = x, y = y)) +
+  list(rast_table = rast.table, 
+       base_map = ggplot() + #data = rast.table, aes(x = x, y = y)) +
     #scale_alpha_discrete(range=c(1,0)) +
     geom_path(data=river.subset, aes(x = long, y = lat, group = group), color = 'blue', alpha = 0.1) +
     geom_polygon(data=lakes.subset, aes(x = long, y = lat, group = group), fill = '#ADD8E6') +
@@ -72,5 +73,5 @@ natural_earth <- function(){
     coord_cartesian(xlim = ext[1:2], ylim=ext[3:4]) +
     theme(axis.text.x = element_blank(),
           axis.text.y = element_blank()) +
-    xlab('') + ylab('')
+    xlab('') + ylab(''))
 }

### 2015-06-24 R/resampling_climate.R

@@ -1,98 +1,100 @@
-
-agg.dens[,-1] <- agg.dens[,-1] / rowSums(agg.dens[,-1])
-pls.data[,-1] <- pls.data[,-1]/rowSums(pls.data[,-1])
-
-then.rast <- resample(then.rast, unit.raster)
-now.rast <- resample(now.rast, unit.raster)
-
-#For each taxon sample:
-#  1.  The FIA data for one of the now.rast years for each of the climate variables.
-#  2.  One of the then.rast years & one of the western layers
-#  3.  Get vectors of values
-
-# These are the climate ranges:
-ranges <- data.frame(min = c(400,  15, 0, -30),
-                     max = c(2200, 35, 16, 2))
-
-# To do the analysis we want to get the climate layers for the past:
-get_dens_era <- function(taxon, clim.var){
+resample_climate <- function(climate, agg_dens, pls_data){
+  agg_dens[,-1] <- agg_dens[,-1] / rowSums(agg_dens[,-1])
+  pls_data[,-1] <- pls_data[,-1] / rowSums(pls_data[,-1])
   
-  #  Function resamples proportion data from the composition model for the region
-  #  and then draws from the composition model and the climate data.
-  #  The function outputs a table with six columns:
-  #  data      - the proportion composition within the cell
-  #  class     - the era of observation
-  #  cell      - the literal cell # in the raster
-  #  plss.clim - the climate extracted from the 1900 - 1930s era
-  #  fia.clim  - the climate extracted from the 1970 - 2000s era
-  #
-  #  Each table has PLSS/pres (taxon present in cell at PLSS)
-  #                 FIA/pres  (taxon present in cell at PLSS)
-  #                 FIA/np    (taxon present in cell at PLSS but not in FIA)
-  #taxon.vals <- ncvar_get(western, taxon, c(1,1,1), c(-1, -1, -1))
-
-  #  bind all the variables together:  
-  values <- data.frame(cell  = c(rep(pls.data$cell, 2), rep(agg.dens$cell, 2)),
-                       x     = c(rep(xyFromCell(unit.raster, pls.data$cell)[,'x'],2),
-                                 rep(xyFromCell(unit.raster, agg.dens$cell)[,'x'],2)),
-                       y     = c(rep(xyFromCell(unit.raster, pls.data$cell)[,'y'],2),
-                                 rep(xyFromCell(unit.raster, agg.dens$cell)[,'y'],2)),
-                       data  = c(rep(as.numeric(pls.data[,taxon]), 2),
[…]

### 2015-07-06 R/figures/GSA_Presentation_diagrams.R
Changing the code so that `source`d files live as functions, reducing the number of global variables.
Added Jin reference for the NLCD reference.

Working on the Hellinger code, there are some problems right now witht he PLS data not getting added properly.

Clearning out some stuff from `get_pls`.

Temporarily added the `GSA_Presentation_diagrams` so I can pull diagrams from there into the paper.

Modifying `ordination_plot` so that it works a bit more effectively.

@@ -1,0 +1,59 @@
+#  GSA Presentation Examples:
+
+plot.densities <- function(subset.boolean){
+
+  full.shift <- vegclim.table[subset.boolean,]
+  
+  full.shift$climate <- factor(full.shift$climate, levels = c('tmax', 'ppt'),
+                             labels = c('T[max]', 'P[ann]'))
+  
+  full.shift$base <- factor(full.shift$base, levels = c('PLSS', 'FIA'))
+  full.shift$c.ref <- factor(full.shift$c.ref, levels = c('PLSS', 'FIA'))
+  
+  subset.dplot <- ggplot(data = full.shift) +
+    geom_density(aes(x = clim, y = ..scaled.., fill = base, 
+                     linetype = c.ref, size = c.ref),
+                 alpha = 0.2) +
+    scale_size_manual(values = c(1, 1.5)) +
+    scale_fill_manual(values = c('#F8766D', '#00BFC4')) +
+    scale_linetype_manual(values = c(1, 2), drop = FALSE) +
+    facet_grid(taxon~climate, scale = 'free_x', labeller = label_parsed) +
+    theme_bw() +
+    xlab('Climate Variable') +
+    ylab('Kernel Density') +
+    theme(axis.text = element_text(family = 'serif', size = 12),
+          axis.title = element_text(family = 'serif', face = 'bold', size = 14),
+          strip.text = element_text(family = 'serif', face='bold', size = 14),
+          legend.position = 'none')
+  
+  subset.dplot
+}
+
+full.shift <- with(vegclim.table, taxon %in% c('Spruce', 'Hemlock', 'Elm') & 
+                    data > 0 & climate  %in% c('ppt', 'tmax') &
+                    ((base == 'PLSS' & c.ref == 'PLSS') | (base == 'FIA' & c.ref == 'FIA')))
+
+plss.shift <- with(vegclim.table, taxon %in% c('Spruce', 'Hemlock', 'Elm') & 
+                     data > 0 & climate %in% c('ppt', 'tmax') &
+                     (base == 'PLSS'))
+
+land.use   <- with(vegclim.table, taxon %in% c('Spruce', 'Hemlock', 'Elm') & 
+                     data > 0 & climate %in% c('ppt', 'tmax') &
+                     ((base == 'PLSS' & c.ref == 'FIA') | (base == 'FIA' & c.ref == 'FIA')))
+
[…]

### 2015-07-06 R/hellinger_dists.R

@@ -20,57 +20,66 @@ hellinger <- function(clim1, clim2){
   
 }
 
-all.hellinger <- ldply(1:length(best.taxa), function(x){
-      ldply(1:4, function(y){
-        data.subset <- vegclim.table[vegclim.table$taxon %in% best.taxa[x] & 
-                                       vegclim.table$climate %in% bio.vals[y] & 
-                                       vegclim.table$data > 0,]
-        
-        full_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
-                               subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim)
-        
-        clim_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
-                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
-        
-        lu_shift   <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim,
-                               subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
-        
-        data.frame(taxon = best.taxa[x],
-                   climate = bio.vals[y],
-                   full.change = full_shift,
-                   fixedv_c = c(clim_shift - lu_shift))
+hellinger_plot <- function(){
+  
+  best.taxa <- as.character(unique(vegclim_table$taxon))
+  
+  all.hellinger <- ldply(1:length(best.taxa), function(x){
+        ldply(1:4, function(y){
+          data.subset <- vegclim_table[vegclim_table$taxon %in% best.taxa[x] & 
+                                         vegclim_table$climate %in% climate$name[y] & 
+                                         vegclim_table$data > 0,]
+          
+          full_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
+                                 subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim)
+          
+          clim_shift <- hellinger(subset(data.subset, base == 'PLSS' & c.ref == 'PLSS')$clim,
+                                 subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
+          
+          lu_shift   <- hellinger(subset(data.subset, base == 'FIA' & c.ref == 'FIA')$clim,
+                                 subset(data.subset, base == 'PLSS' & c.ref == 'FIA')$clim)
[…]

### 2015-07-06 R/load_veg.R

@@ -83,12 +83,6 @@ get_fia <- function(x){
 get_pls <- function(version){
   pls.data <- read.csv('data/plss_density_alb_v0.9-4.csv', row.names=1)
   colnames(pls.data) <- tolower(colnames(pls.data))
-  
-  colnames(agg.dens) <- tolower(gsub(' ', '.', colnames(agg.dens), fixed=TRUE))
-  
-  colnames(agg.dens) <- gsub('/', '.', colnames(agg.dens), fixed=TRUE)
-  
-  agg.dens$poplar <- agg.dens$poplar.tulip.poplar
-  
+    
   pls.data[,c('cell', best.taxa)]
 }

### 2015-07-06 Composition_Ecology.Rmd
Changing the code so that `source`d files live as functions, reducing the number of global variables.
Added Jin reference for the NLCD reference.

Working on the Hellinger code, there are some problems right now witht he PLS data not getting added properly.

Clearning out some stuff from `get_pls`.

Temporarily added the `GSA_Presentation_diagrams` so I can pull diagrams from there into the paper.

Modifying `ordination_plot` so that it works a bit more effectively.
Working on the final plots for the paper.  Adding in some text results.
Added `analogue` to the setup file, allowing us to use th

@@
 source('R/Setup_R.R')
-source('R/load_climate.R') # gives then.clim & now.clim
+source('R/load_climate.R') # gives `climate` with elements `then`, `now` and `names`
 source('R/load_veg.R')
-
-pls_data <- get_pls('0.9-4')
-agg_dens <- get_fia()[,colnames(pls_data)]
 
@@
               'tamarack', 'maple', 'poplar', 'birch')
+
+pls_data <- get_pls('0.9-4')
+agg_dens <- get_fia()[,colnames(pls_data)]
 
@@
 
+#  This takes a bit of time.
 loss_plots <- loss_plot()
@@
 
-hell.plot
+#  Returns a list with the ggplot (object 1) and all.hellinger.
+hell_plot <- hellinger_plot()
 
 ```
+**Figure X**. *For each of the four variables examined we see*
 
-The Hellinger distances reveal that
+The Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables.
+Maximum temperature shows the greatest change attributed to land use change.
+Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change.
 
-```{r densityplots, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width=7}
+The precipitation panel (Figure X) indicates that precipitation across the region has driven change in the species climate niche for this variable.
+Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.
 
-source('R/figures//ordination_plot.R')
-ordi_plot
-
-```
-
-When we examine the taxa seperately we can see that by far the majority of comparisons show greater differences due to climate change than due to shifts in distribution:
-
-This plot makes it clear that precipitation across the region has changed significantly since the PLS era (or the early 1900s), while overall $t_{mean}$ has changed less.
-Among taxa only beech, ironwood, hemlock, fir and tamarak show distributional changes larger than the effect of shifting climate.
-For beech, a shift is seen in $t_{mean}$, where $t_{mean}$ for the PLS era distribution is 6.8^o^C, but is lower for the FIA distribution using PLS era climate, at $t_{mean}$ = 6.6^o^C.
-This is likely due to a northward shift in beech distribution in the current FIA dataset due to the loss of southern beech forests (earlier Figure).
+Interstingly, since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.
 

### 2015-07-06 R/Setup_R.R

@@ -7,6 +7,7 @@ library(plyr)
 library(gridExtra)
 library(rgdal)
 library(RColorBrewer)
+library(analogue)
 
 base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
 base.rast <- setValues(base.rast, 1:ncell(base.rast))

### 2015-07-06 R/hellinger_dists.R

@@ -78,8 +78,8 @@ hellinger_plot <- function(){
              label='Land Use Dominates', family='serif',
              fontface = 'bold', size = 5)
   
-  ggsave(plot = hell.plot, filename = 'figures/hellingerplot.tiff', 
+  ggsave(plot = hell_plot, filename = 'figures/hellingerplot.tiff', 
          width = 10, height = 8, dpi = 150)
  
-  hell_plot
+  list(hell_plot, all.hellinger)
 }

### 2015-07-06 R/load_veg.R

@@ -77,6 +77,9 @@ get_fia <- function(x){
   
   colnames(agg_dens) <- tolower(gsub(' |[[:punct:]]', '\\.', colnames(agg_dens)))
   colnames(agg_dens)[regexpr('poplar', colnames(agg_dens))>0] <- 'poplar'
+  
+  agg_dens[is.na(agg_dens)] <- 0
+  
   agg_dens
 }
 

### 2015-07-06 R/resampling_climate.R

@@ -1,6 +1,11 @@
+#  This code takes a climate variable and the FIA & PLS data
+#  and returns a table with the climate variable for each taxon
+#  within each cell in each of the four eras (PLS/PLS, FIA/FIA 
+#  and the cross values).
+
 resample_climate <- function(climate, agg_dens, pls_data){
-  agg_dens[,-1] <- agg_dens[,-1] / rowSums(agg_dens[,-1])
-  pls_data[,-1] <- pls_data[,-1] / rowSums(pls_data[,-1])
+  agg_dens[,-1] <- tran(agg_dens[,-1], method = 'proportion')
+  pls_data[,-1] <- tran(pls_data[,-1], method = 'proportion')
   
   then_rast <- resample(climate$then, unit.raster)
   now_rast <- resample(climate$now, unit.raster)
@@ -76,7 +81,7 @@ resample_climate <- function(climate, agg_dens, pls_data){
                                   }
                                   aa}))
                             })
-                          })
+})
   
   data.out <- lapply(1:15, function(x)lapply(1:4, function(y){
     aa <- data.tables[[x]][[y]]

### 2015-07-07 Composition_Ecology.Rmd
Added new figures to help the `md` file render online.
Added code block to get `pandoc.table` to run properly.

Some text edits in the Introduction, adding references &cetera.

@@
 output:
+  md_document:
+    variant: markdown
   pdf_document:
+    pandoc_args: "-V geometry:vmargin=1in -V geometry:hmargin=1in"
 pandoc_args: "-H margins.sty"
-bibliography: composition.bib
-csl: harvard1.csl
+bibliography: goringetal_references.bib
+csl: ecology.csl
 ---
@@
 ========================================================
+
+<!--
+
+```{r setup-knitr, echo = FALSE, message = FALSE}
+knitr::opts_chunk$set(
+  comment = " ",
+  error = TRUE,
+  eval = TRUE,
+  cache = TRUE,
+  tidy = TRUE
+
+)
+
+library("pander")
+library("plyr")
+panderOptions('table.style', 'rmarkdown')
+```
+-->
 
@@
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE}
+```{r, echo = FALSE, warning = FALSE, message = FALSE, message=FALSE}
 # Load climate data and generate the pre-sett and modern bounds:
@@
 Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States.
-There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset].
-This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change.
+There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset;@schulte2007homogenization].
+This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
+One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly the .
 
-If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions.
-In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie and savanna to closed forest, in an approixmately west to east direction, and from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
+```{r}
+
+natural[[2]] + 
+  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
+  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c', '#377eb8'))
+
+```
+**Figure 1.** *Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*
+
+If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions since, by and large, me might expect the climate envelope encompased by the species to remain largely unchanged.
+In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
 Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa.
+Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests, we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
+Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.
 
-We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for X major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.
+We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.
 
@@
 ===============
-We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015] with annually resolved climate data for the region.
+We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.
 PLSS vegetation data contains no uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.
@@
 ```
+![](figures/clim_plot.tiff)
 **Figure 1**.
@@
 #  This takes a bit of time.
-loss_plots <- loss_plot()
+loss_melt <- loss_plot()
+
+loss_table <- as.data.frame.matrix(table(loss_melt$variable, loss_melt$value))
+loss_table$Gain <- loss_table$Gain / (loss_table$Loss + loss_table$Presence)
+loss_table[,2:3] <- tran(loss_table[,2:3], method = 'proportion')
 
 ```
-**Figure 1**.  *Gain loss and continuous presence for PLSS and FIA tree taxa.*
+![](figures/loss_plots.tiff)
 
-```{r densities}
+**Figure 2**.
+*Gain loss and continuous presence for PLSS and FIA tree taxa.
+Continuous presence is indicated by dark green.
+For most taxa it is the central parts of their distribution that show continuous presence.
+Few taxa show novel presence (purple), although taxa such as poplar, ash and basswood do show gains of over 10%.
+Most taxa show significant losses in distribution (orange).*
+
+**Table 1.** *Gain and loss of the various forest tree types of interest.*
+```{r, results=as.is}
+pandoc.table(loss_table, justify = "left", style = 'simple')
+```
+
+```{r densities, warning=FALSE, message=FALSE, echo=FALSE}
 
@@
 ```
-**Figure** *Changes in the overall climate space in the upper midwestern United States.
+
+**Figure 3** *Changes in the overall climate space in the upper midwestern United States.
 The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*
@@
 
+hell_plot[[1]]
+
 ```
-**Figure X**. *For each of the four variables examined we see*
+
+**Figure X**. *For each of the four variables examined we see...*
 

### 2015-07-07 R/Setup_R.R

@@ -15,6 +15,14 @@ base.rast <- setValues(base.rast, 1:ncell(base.rast))
 pls.rast <- raster('data/input//dens_v0.9-4.tif')
 unit.raster <- setValues(pls.rast, 1:ncell(pls.rast))
 
+land_use <- raster('../../../Maps/nlcd_2011_landcover_2011_edition_2014_10_10//nlcd_cropped_repro.tif')
+
+#  We need to change the raster to a data.frame:
+land_use <- data.frame(xyFromCell(land_use, 1:ncell(land_use)),
+                       value = getValues(land_use))
+reassign <- read.csv('data/input/landuse_reassignment.csv', stringsAsFactors=FALSE)
+land_use$class <- factor(reassign[match(land_use$value, reassign[,1]),2])
+
 model.proj <- '+init=epsg:3175'
 
 usa <- readOGR('data/input/shapes/usa/us.shp', 'us')

### 2015-07-07 R/figures/gain_loss_plot.R

@@ -83,6 +83,6 @@ loss_plot <- function(){
   ggsave(plot = short_plot, filename = 'figures/short_plots.tiff', dpi = 150, width = 6, height = 5)
   
   
-  list(loss_plot, short_plot)
+  loss_melt
 }
 

### 2015-07-07 R/natural_earth.R

@@ -28,7 +28,7 @@ natural_earth <- function(){
     shape@data$id <- rownames(shape@data)
     
     x.f = fortify(shape, region="id")
-    x.join = join(x.f, shape@data, by="id")
+    x.join = plyr::join(x.f, shape@data, by="id")
     
     aa <- x.join[,1:2]
     coordinates(aa) <- ~ long + lat

### 2015-07-07 Composition_Ecology.Rmd
Added the `md` file.

@@
 **Table 1.** *Gain and loss of the various forest tree types of interest.*
-```{r, results=as.is}
+```{r, results='as.is'}
 pandoc.table(loss_table, justify = "left", style = 'simple')

### 2015-07-07 Composition_Ecology.Rmd
Minor changes to the figure functions so that the figures are generated from the code itself, not linked in.
Re-numbered figures in text.

@@
 output:
+  word_document:
+    fig_width: 6
+    fig_height: 6
   md_document:
-    variant: markdown
+    variant: markdown_github
   pdf_document:
@@
 
-```{r, echo = FALSE, warning = FALSE, message = FALSE, message=FALSE}
+```{r, echo = FALSE, warning = FALSE, message=FALSE, results='hide'}
 # Load climate data and generate the pre-sett and modern bounds:
@@
 
-```{r}
+```{r figure1, fig.width=6, fig.height=5, echo = FALSE, warning = FALSE, message=FALSE}
 
@@
 ```
+
 **Figure 1.** *Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*
@@
 
+climate_values$plot
+
 ```
-![](figures/clim_plot.tiff)
-**Figure 1**.
+
+**Figure 2.**.
 *Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
@@
 
-loss_table <- as.data.frame.matrix(table(loss_melt$variable, loss_melt$value))
+loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable, loss_melt[[1]]$value))
 loss_table$Gain <- loss_table$Gain / (loss_table$Loss + loss_table$Presence)
@@
 
+loss_melt[[2]]
 ```
-![](figures/loss_plots.tiff)
 
-**Figure 2**.
+**Figure 3.**.
 *Gain loss and continuous presence for PLSS and FIA tree taxa.
@@
 
-**Figure 3** *Changes in the overall climate space in the upper midwestern United States.
+**Figure 4.** *Changes in the overall climate space in the upper midwestern United States.
 The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*
@@
 
-**Figure X**. *For each of the four variables examined we see...*
+**Figure 5**. *For each of the four variables examined we see...*
 

### 2015-07-07 R/figures/climate_ts.R

@@ -56,5 +56,5 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
           legend.position = "none")
     
   ggsave(clim_plot, filename = 'figures/clim_plot.tiff', width = 8, height = 6, dpi = 150)
-  list(modern = modern, past = past, climate = climate)
+  list(modern = modern, past = past, climate = climate, plot = clim_plot)
 }

### 2015-07-07 R/figures/gain_loss_plot.R

@@ -83,6 +83,6 @@ loss_plot <- function(){
   ggsave(plot = short_plot, filename = 'figures/short_plots.tiff', dpi = 150, width = 6, height = 5)
   
   
-  loss_melt
+  list(loss_melt, loss_plot)
 }
 

### 2015-07-09 Composition_Ecology.Rmd
Added some references, and more extensive text.

@@
 This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
-One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly the .
+One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1), where extensive land use conversion has left little natural land.
 
@@
 **Figure 1.** *Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*
+
+The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.
+By deliniating three tpyes of interaction, compunding, confounding and contradicting, it is possible to understand both the nature of the interaction and the potential effect on species resiliance and response to broader regional change.
+
+Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.
+In the case of warming a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.
+Confounding effects may cause little overall effect on species distributions since in this case the land use pressure falls within the central distribution for the environmental variable, they may increase the variability of predictive models, but do little to the overall envelope.
+Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in the warmer tail of the distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the cooler end of the tail excludes the taxa from those regions.
 

### 2015-07-09 Composition_Ecology.Rmd
Added the corn belt reference, added some 'to read' papers to the bibtex as well.
More edits to flesh out the hypothesis.

@@
 
-If modern species distributions have been affected by land use change in a random fashion then we should expect little effect on the resulting predictions since, by and large, me might expect the climate envelope encompased by the species to remain largely unchanged.
 In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
@@
 Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.
+
+Interaction between environmental gradients and land use patterns vary by region and environmental variable.
+We might expect that species in the north experience a confounding land use effect with respect to temperature, but, with respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect this effect to be compounding.
+For southern tree species land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients.
+This should result in contradicting land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the twin effects of land use change and regional climate change.
+
+Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting land use effects may result in the impression of narrower climatic tolerances.
+For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envolope may result in underprediction of species resiliance to climate change.
 
@@
 We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.
-PLSS vegetation data contains no uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.
+PLSS vegetation data contains no reported uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.
 Additionally, data in the northeastern United States [@thompson2013four] from town Proprietor Surveys is recorded arially, with no variance within township polygons.

### 2015-07-14 Composition_Ecology.Rmd
Added new run with PDF and DOCX versions.  Don't have a good change history for the Rmd, something's wonky with RStudio.

@@
 author: "Simon Goring *et al*."
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: '`r format(Sys.time(), ''%d %B, %Y'')`'
 output:
-  word_document:
-    fig_width: 6
-    fig_height: 6
+  pdf_document:
+    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
   md_document:
     variant: markdown_github
-  pdf_document:
-    pandoc_args: "-V geometry:vmargin=1in -V geometry:hmargin=1in"
-pandoc_args: "-H margins.sty"
+  word_document:
+    fig_height: 6
+    fig_width: 6
+csl: ecology.csl
+pandoc_args: -H margins.sty
 bibliography: goringetal_references.bib
-csl: ecology.csl
 ---

### 2015-07-24 Composition_Ecology.Rmd
Changed title to reflect @iceageecologist's edits to the draft.  Added outlined abstract (but note other abstract below).
Merged figures 1 & 2 to resolve issue #1.  Unresolved are references to figures 1 & 2.

@@
 
-Land use patterns and historical contingency change climate optima for key forest genera in the Northeastern United States.
+Effect of historic land use and climate change on tree-climate relationships for key forest genera in the northern United States
 ========================================================
@@
 -->
+
+Abstract
+====
+* Much of our understanding of species-cliamte relationships and functional-community relationships in forests is based off of contemporary observations.
+* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use.  Climates also changing.  
+* Hence the question of whether species-climate relationships today are representative compared to those in past.
+* Here we assess stability of species-climate relationships. Using historic PLS data and FIA data.
+* We show that species-climate relationships have significantly shifted for many taxa.
+* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
+* Relative importance of land use and climate change varies by taxon and climate variable.  
+* Shifts in warmer temperature mainly due to land use change, etc.
+* Implications sentence or two at the end.
 
@@
 This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
-One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1), where extensive land use conversion has left little natural land.
+One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.
 
@@
 
-natural[[2]] + 
+land_use$class[land_use$class == 'Water'] <- NA
+
+land_use_plot <- natural[[2]] + 
   geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
-  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c', '#377eb8'))
+  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c'))
+
+source('R/figures/climate_ts.R')
+
+climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
+
+changes <- climate_values$modern$value - climate_values$past$value
+
+grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
 
@@
 
-**Figure 1.** *Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*
+**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
+Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.
+PDSI shows strong coherence with A_ppt_ in this region.
+(f) Patterns of land use change in the upper Midwestern United States.
+Data from the NLCD [@jin2013comprehensive].*
 
@@
 ---------------
-```{r changing_climate, warning=FALSE, message=FALSE, echo=FALSE}
-
-source('R/figures/climate_ts.R')
-
-climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
-
-changes <- climate_values$modern$value - climate_values$past$value
-
-climate_values$plot
-
-```
-
-**Figure 2.**.
-*Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.
-PDSI shows strong coherence with A_ppt_ in this region.*
 
@@
 The time transgressive nature of the PLS Survey means that some of the PLS data comes from records sampled in the early 1800s, while other come from the early 1900s.
-Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
+Regardless, throughout this period climate variability exists (Figure 1e), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
 As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice [ADD JULY temperatures here].

### 2015-07-24 Composition_Ecology.Rmd
Added some comments based on discussion with @andydawson 24/7/2015.

@@
 
+**Paragraph about data scale between PLS and FIA**
+
+**Paragraph about FIA/PLS absence not really Absence**
+
 The analysis follows a blocked sampling method.
@@
 The time transgressive nature of the PLS Survey means that some of the PLS data comes from records sampled in the early 1800s, while other come from the early 1900s.
-Regardless, throughout this period climate variability exists (Figure 1e), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
+Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
 As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice [ADD JULY temperatures here].
+**Talk about modern climate normal & historical climate normal make it clear that this is the best we can do, and that it's good.**
 

### 2015-07-29 Composition_Ecology.Rmd
Removed "Land use" values for cells without either PLS or FIA data to make the plotting look better.
The changes are largely around an issue in the main `Rmd` file.  It's not recognizing the `clim_changes` variable, for some reason, so I'm getting an eval error when I knit.
Not sure why.

Annual clim is modified so that tmean values are replaced by tdiff.

@@
 author: "Simon Goring *et al*."
-date: '`r format(Sys.time(), ''%d %B, %Y'')`'
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
   eval = TRUE,
-  cache = TRUE,
+  cache = FALSE,
   tidy = TRUE
@@
 
+land_use <- land_use[land_use$cell %in% c(pls_data$cell, agg_dens$cell),]
+
 source('R/natural_earth.R')
@@
 climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
-
-changes <- climate_values$modern$value - climate_values$past$value
 
@@
 We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.
-PLSS vegetation data contains no reported uncertainty, however it is spatially discontinuous, particularly in the south where digitization of PLSS records has occured at a slower rate.
-Additionally, data in the northeastern United States [@thompson2013four] from town Proprietor Surveys is recorded arially, with no variance within township polygons.
-To accomodate these different data sources a conditional autoregressive model was applied to the data, providing the ability to obtain uncertainty estimates across the domain [@paciorek2015comp].
 
@@
 
-Mean temperatures in the region show an average increase of `r round(changes[3], 2)`^o^C since the 1895-1925 normal.
-Annual precipitation has increased slightly (`r round(changes[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(changes[2], 2)`^o^C)).
-The most significant change has been in annual minimum temperatures, which have increased by `r round(changes[4], 1)` since the 1895-1925 normal.
+`r clim_change <- climate_values$modern$value - climate_values$past$value `
+
+Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.
+Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C)).
+The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)` since the 1895-1925 normal.
 This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.

### 2015-07-29 R/figures/gain_loss_plot.R

@@ -39,6 +39,7 @@ loss_plot <- function(){
   }
   
   loss_melt <- get_loss_gain(unit.raster, best.taxa, pls_data, agg_dens)
+
   loss_melt$rgb <- brewer.pal(n = 3, 'Dark2')[as.numeric(factor(loss_melt$value, 
                                                                 levels = c('Presence', 'Loss', 'Gain')))]
   

### 2015-07-29 Composition_Ecology.Rmd
Fixes to get the `clim_change` call to work.  There was an error in `climate_ts` because I forgot to change a period to an underscore.  `knitr` wasn't picking up that error, but instead showing me the inline code error further down.
+ `climate_ts` - changed `now.rast` to `rast_in`.

Output all `tiff` and `rds` files.

@@
 
-```{r figure1, fig.width=6, fig.height=5, echo = FALSE, warning = FALSE, message=FALSE}
+```{r figure1, echo = FALSE, warning = FALSE, message=FALSE, fig.width=6, fig.height=5}
 
@@
 
+clim_change <- climate_values$modern$value - climate_values$past$value
+
 grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
-
 ```
@@
 **Talk about modern climate normal & historical climate normal make it clear that this is the best we can do, and that it's good.**
-
-`r clim_change <- climate_values$modern$value - climate_values$past$value `
 

### 2015-07-29 R/figures/climate_ts.R

@@ -36,7 +36,7 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
   clim_pattern <- function(rast_in, pls_rast, year){
   
     rast_mask <- rast_in
-    rast_mask[!(1:ncell(now.rast))%in%pls_data$cell]<- NA
+    rast_mask[!(1:ncell(rast_in))%in%pls_data$cell]<- NA
   
     data.frame(variable = c('P[ann]', 'T[max]', 'T[diff]', 'T[min]'),
                value = colMeans(getValues(rast_mask), na.rm=TRUE),

### 2015-10-20 Composition_Ecology.Rmd
_Main `Rmd`_ Minor edits to the abstract bullet points.  Added an extent value and built some of the plots a bit better.
Added more references to the introduction, cleaned up and extend the Methods section.  Added blocked sample figure with color legend.

Some editing of results.

_bib file_
Added references here.  Fairly extensive.

*`comp_ecol.R`* is the purled version of the paper.

--- 2015-07-29 Composition_Ecology.Rmd
@@
 output:
+  word_document:
+    fig_height: 6
+    fig_width: 6
   pdf_document:
@@
     variant: markdown_github
-  word_document:
-    fig_height: 6
-    fig_width: 6
 csl: ecology.csl
@@
 ====
-* Much of our understanding of species-cliamte relationships and functional-community relationships in forests is based off of contemporary observations.
-* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use.  Climates also changing.  
-* Hence the question of whether species-climate relationships today are representative compared to those in past.
-* Here we assess stability of species-climate relationships. Using historic PLS data and FIA data.
+* Much of our understanding of species-climate relationships and functional-community relationships in forests is based on contemporary observations of tree distributions.
+* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
+* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
+* We assess the stability of species-climate relationships using historic PLS data and FIA data.
 * We show that species-climate relationships have significantly shifted for many taxa.
@@
 
+ext <- c(-100000, 1050000, 600000, 1600000)
+
 land_use_plot <- natural[[2]] + 
   geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
-  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c'))
+  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
+  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
+  theme(legend.position = 'none')
 
@@
 grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
+
 ```
@@
 In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
-Goring et al [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, but did not examine how these shifts might affect the relationships between climate and distribution for individual taxa.
-Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests, we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
+Goring et al.
+[@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies,@schulte2001original,@schulte2007homogenization,@radeloff2000historical].
+To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa.
+Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
 Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.
@@
 ===============
-We pair pre-settlement vegetation data from the Public Land Survey System aggregated to a 64km^2^ grid [@goring2015preset] with annually resolved climate data for the region.
+We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.
+The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.
+At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.
+Goring et al.
+[-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.
+When examining relationships between FIA and PLS data, Goring et al.
+[-@goring2015] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.
 
-**Paragraph about data scale between PLS and FIA**
+Another complication is the structure of the FIA dataset.
+While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].
+As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA.
+This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree,@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations.
+[needs refs].
 
-**Paragraph about FIA/PLS absence not really Absence**
+The analysis presented here follows a blocked sampling method (Figure 2).
+We generate probability density functions along climate gradients for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era.
+We do the same for FIA presence, and for anthropogenic land use classes.
+Climate data is derived from monthly PRISM data from 1895 - 2014.
+A 30-year climate was generated for modern July temperature, January temperature, the continentality (difference between T_jul_ and T_jan_), and annual precipitation.
 
-The analysis follows a blocked sampling method.
-We generate probability density functions for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era.
-We also generate climate data for the FIA presence, and we generate climate space for modern land use.
-In this way we can test how much of the change in shape is the result of changing land use vs changing climate since the early part of the century.
+![model grid](figures/model_grid.png)
 
-We also develop estimates of the modern climate space that taxa would occupy if they remained in place, with only climate changing, and vice versa (need better explanation here).
+**Figure 2**.
+*Blocked method for analysis.
+Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
+A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*
+
+The choice of a thirty year normal for the PLS data is potentially problematic.
+The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset, @white1983history], and for much of this time period no climatological records exist.
+Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
+As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.
+Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions [@trouet, @mann] tell us _THIS_.
+
+By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation (Figure 2) we can use Hellinger distance to estimate the effects of changing climate (d_H_ between kernel densities in the same column of Figure 2), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).
+In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).
+The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.
 
@@
 ---------------
-
-The choice of a normal for the domain is potentially problematic.
-The time transgressive nature of the PLS Survey means that some of the PLS data comes from records sampled in the early 1800s, while other come from the early 1900s.
-Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
-As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice [ADD JULY temperatures here].
-**Talk about modern climate normal & historical climate normal make it clear that this is the best we can do, and that it's good.**
 
@@
 
-```{r loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE}
+```{r loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
 
@@
 loss_table[,2:3] <- tran(loss_table[,2:3], method = 'proportion')
+
+loss_table[,2:3] <- paste0(round(loss_table[,2:3]*100,1),'%')
 
@@
 
-```{r densities, warning=FALSE, message=FALSE, echo=FALSE}
+```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
 
@@
 
-```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE}
+```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
 

### 2015-10-20 comp_ecol.R

@@ -1,0 +1,94 @@
+## ----setup-knitr, echo = FALSE, message = FALSE--------------------------
+knitr::opts_chunk$set(
+  comment = " ",
+  error = TRUE,
+  eval = TRUE,
+  cache = FALSE,
+  tidy = TRUE
+
+)
+
+library("pander")
+library("plyr")
+panderOptions('table.style', 'rmarkdown')
+
+## ---- echo = FALSE, warning = FALSE, message=FALSE, results='hide'-------
+# Load climate data and generate the pre-sett and modern bounds:
+#  Pre-settlement climate: 1900 - 1930
+#  Modern climate: 1960 - 1990
+
+source('R/Setup_R.R')
+source('R/load_climate.R') # gives `climate` with elements `then`, `now` and `names`
+source('R/load_veg.R')
+
+best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
+              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
+              'tamarack', 'maple', 'poplar', 'birch')
+
+pls_data <- get_pls('0.9-4')
+agg_dens <- get_fia()[,colnames(pls_data)]
+
+land_use <- land_use[land_use$cell %in% c(pls_data$cell, agg_dens$cell),]
+
+source('R/natural_earth.R')
+
+natural <- natural_earth()
+  
+
+## ----figure1, echo = TRUE, warning = FALSE, message=FALSE----------------
+
+land_use$class[land_use$class == 'Water'] <- NA
+
+land_use_plot <- natural[[2]] + 
+  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
+  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c'))
+
+
+source('R/figures/climate_ts.R')
+
+climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
+
+clim_change <- climate_values$modern$value - climate_values$past$value
+
+
+## ----figureTime, fig.width=6, fig.height=5-------------------------------
+grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
+
+clim_change[3]
+
+## ----loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE-------------
+
+source('R/figures/gain_loss_plot.R')
+
+#  This takes a bit of time.
+loss_melt <- loss_plot()
+
[…]

### 2015-10-27 Composition_Ecology.Rmd
Some editing in the abstract and introduction, but snall changes, mostly changing comma separated references to semicolon separated.
Added more text to the 'regional climate change' section and a few references from Viau, and Ahmed.  Could add more, maybe plumb the PalEON climate paper.

Beefed up the caption for Figure 5.

Knit with the changes.

@@
 
-*Predictions for future species range shifts are predicated largely on correlative models that relate modern species distributions to climate parameters in the modern era.
-Given the extent of regional forest change and recent losses to key forest taxa including Hemlock, Elm and Chestnut it is possible that certain taxa have shiften with respect to climate change over the last century.
-The additional pressure of climatically biased land use conversion for agriculture along the western and southern border of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*
+*Predictions for future species range shifts are largely predicated on correlative models that relate modern species distributions to climate parameters in the modern era.
+Given the extent of regional forest change, and recent losses to key forest taxa including Hemlock, Elm and Chestnut, it is possible that certain taxa have shifted their climatic distributions over the last century.
+The additional pressure of climatically biased land use conversion for agriculture along the western and southern margins of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*
 
@@
 Goring et al.
-[@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies,@schulte2001original,@schulte2007homogenization,@radeloff2000historical].
+[@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical].
 To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa.
@@
 When examining relationships between FIA and PLS data, Goring et al.
-[-@goring2015] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.
+[-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.
 
@@
 As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA.
-This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree,@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations.
+This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations.
 [needs refs].
@@
 The choice of a thirty year normal for the PLS data is potentially problematic.
-The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset, @white1983history], and for much of this time period no climatological records exist.
+The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist.
 Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
 As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.
-Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions [@trouet, @mann] tell us _THIS_.
+Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].
 
@@
 
-**Figure 5**. *For each of the four variables examined we see...*
+**Figure 5**.
+*For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.
+Taxa with negative y values have changes in distributions along the climate gradient that are driven largely by land use, while positive y axis values can be largely attributed to regional climate change.*
 
@@
 
-The precipitation panel (Figure X) indicates that precipitation across the region has driven change in the species climate niche for this variable.
+The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.
 Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.

### 2015-10-27 R/Setup_R.R
Added some more specific indications for package use.  Not sure why these things were throwing errors, but the addition of `colorspace` has meant I need to use `:` notation in `gain_loss_plot`.

@@ -1,4 +1,5 @@
 #  Setup some data and libraries:
+library(colorspace)
 library(raster)
 library(ggplot2)
 library(mgcv)

### 2015-10-27 R/figures/gain_loss_plot.R

@@ -43,6 +43,10 @@ loss_plot <- function(){
   loss_melt$rgb <- brewer.pal(n = 3, 'Dark2')[as.numeric(factor(loss_melt$value, 
                                                                 levels = c('Presence', 'Loss', 'Gain')))]
   
+  loss_melt$rgb[loss_melt$value == 'Gain'] <- "#FF0000"
+  loss_melt$rgb[loss_melt$value == 'Presence'] <- colorspace::hex(colorspace::RGB(0.1, 0.1, 0.1))
+  loss_melt$rgb[loss_melt$value == 'Loss'] <- colorspace::hex(colorspace::RGB(2/255,119/255,160/255))
+  
   loss_melt$variable <- factor(loss_melt$variable,
                                levels = c('tamarack', 'pine', 'spruce', 'fir', 'hemlock', 'cedar.juniper',
                                           'poplar', 'maple', 'birch', 'beech', 'ironwood', 'basswood', 'ash', 'elm', 'oak'),
@@ -51,8 +55,8 @@ loss_plot <- function(){
   
   loss_plot <- natural$base_map +
       geom_tile(data = natural$rast_table, aes(x = x, y = y, fill = rgb)) +
-      geom_tile(data = subset(loss_melt, !variable %in% 'Ironwood') , aes(x = x, y = y, fill = rgb)) +
-      facet_wrap(~variable, ncol = 5) + theme_bw() +
+      geom_tile(data = subset(loss_melt, variable %in% 'Pine') , aes(x = x, y = y, fill = rgb)) +
+      #facet_wrap(~variable, ncol = 5) + theme_bw() +
       scale_fill_identity() +
   #    geom_path(data = umw.domain, aes(x = long, y = lat, group = group)) +
   #    geom_path(data = can.domain, aes(x = long, y = lat, group = group)) +

### 2015-10-27 Composition_Ecology.Rmd
More references to build up the introduction.  Mostly trying to clarify the compounding/confounding and contradicting concepts.  Added some climate references too.
Clarified the hypotheses examined in the paper.

@@
 Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.
-By deliniating three tpyes of interaction, compunding, confounding and contradicting, it is possible to understand both the nature of the interaction and the potential effect on species resiliance and response to broader regional change.
+By deliniating three tpyes of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.
 
 Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.
-In the case of warming a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.
-Confounding effects may cause little overall effect on species distributions since in this case the land use pressure falls within the central distribution for the environmental variable, they may increase the variability of predictive models, but do little to the overall envelope.
-Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in the warmer tail of the distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the cooler end of the tail excludes the taxa from those regions.
+In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.
+This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].
+Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental variable.
+Confounding effects may increase the variability of predictive models, but do little to the overall envelope.
+Garry oak distributions in the Pacific Northwest may suffer from confounding effects.
+The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining].
+Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.
+Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.
 
@@
 Goring et al.
-[@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical].
+[-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical].
 To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa.
@@
 
-Interaction between environmental gradients and land use patterns vary by region and environmental variable.
-We might expect that species in the north experience a confounding land use effect with respect to temperature, but, with respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect this effect to be compounding.
-For southern tree species land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients.
-This should result in contradicting land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the twin effects of land use change and regional climate change.
+Interaction between environmental gradients, land use patterns and taxa are likely to vary by region and environmental variable of interest.
+Species in the north may experience a confounding land use effect with respect to temperature, northward shifts due to warming temperatures may not have significantly manifested themselves, and land use conversion is likely to be heterogenous with respect to temperature since there is little agriculture in northern Wisconsin and northeastern Minnesota.
+Thus, with respect to temperature we expect to see no shift in distribution along temperature gradients.
 
-Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting land use effects may result in the impression of narrower climatic tolerances.
-For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envolope may result in underprediction of species resiliance to climate change.
+With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change.
+Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.
 
-We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.
+For southern tree species, land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients.
+This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.
+
+Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty.
+Tree species are clearly responding to climate change within forested landscapes, showing, for example, differences in the distributions of mature trees and seedlings [@monleon2015evidence], but modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts.
+For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.
+
+We use gridded climate data products (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States.
+By examining the interactions between climate, species distributions and land use change since EuroAmerican settlement we examine the possible implications of range shifts for future estimates of species resillience to climate change across the region.
 
@@
 As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA.
-This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations.
-[needs refs].
+This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].
 

### 2015-10-28 Composition_Ecology.Rmd
Added some bib references, worked on introduction too, cleaning up the text around confounding, conflicting and contradicting effects.

@@
 Garry oak distributions in the Pacific Northwest may suffer from confounding effects.
-The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining].
+The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental].
 Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.
@@
 To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa.
-Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
+Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
 Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

### 2015-11-12 Composition_Ecology.Rmd
Tried a new tack to get the subscripting to work.  If it doesn't it's not the end of the world.
Minor text edits in the introduction.

Worked to fix the figures in the paper (see edits in `gain_loss_plot`).

Added text to discussion.

NOTE:  The hypothesis is that we have three different types of land use/climate interactions.  This needs to be followed up in the analysis.

@@
 **Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Low temperatures in the 1970s result in lower T_max_ during the modern era, but climate change has resulted in higher T_min_ values than the early-century normals, along with increasing A_ppt_.
-PDSI shows strong coherence with A_ppt_ in this region.
+Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.
+PDSI shows strong coherence with P~ann~ in this region.
 (f) Patterns of land use change in the upper Midwestern United States.
@@
 Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.
-By deliniating three tpyes of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.
+By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.
 
@@
 This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].
-Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental variable.
+Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient.
 Confounding effects may increase the variability of predictive models, but do little to the overall envelope.
-Garry oak distributions in the Pacific Northwest may suffer from confounding effects.
-The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental].
+Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.
+The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used.
 Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.
@@
 [-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical].
-To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa.
 Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
@@
 
-For southern tree species, land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients.
+For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients.
 This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.
@@
 Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty.
-Tree species are clearly responding to climate change within forested landscapes, showing, for example, differences in the distributions of mature trees and seedlings [@monleon2015evidence], but modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts.
+Tree species are responding to climate change within forested landscapes, but the response is inconsistent.
+Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure].
+Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns.
 For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.
@@
 Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.
-Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C)).
-The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)` since the 1895-1925 normal.
+Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C).
+The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal.
 This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.
@@
 
-loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable, loss_melt[[1]]$value))
-loss_table$Gain <- loss_table$Gain / (loss_table$Loss + loss_table$Presence)
-loss_table[,2:3] <- tran(loss_table[,2:3], method = 'proportion')
+loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable,
+                                         loss_melt[[1]]$value))
+loss_table$Gain <- round(loss_table$Gain / (loss_table$Loss + loss_table$Presence)*100,1)
 
-loss_table[,2:3] <- paste0(round(loss_table[,2:3]*100,1),'%')
+loss_table[,2:3] <- round(tran(loss_table[,2:3], method = 'proportion') * 100,1)
 
@@
 *Gain loss and continuous presence for PLSS and FIA tree taxa.
-Continuous presence is indicated by dark green.
+Loss is assessed where previously PLSS data indicated presence of a taxon but FIA data do not record presence (light blue).
+Continuous presence is indicated by dark gray, for where both FIA and PLSS cells indicate presence.
 For most taxa it is the central parts of their distribution that show continuous presence.
-Few taxa show novel presence (purple), although taxa such as poplar, ash and basswood do show gains of over 10%.
-Most taxa show significant losses in distribution (orange).*
+Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*
 
-**Table 1.** *Gain and loss of the various forest tree types of interest.*
-```{r, results='as.is'}
+**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era.
+Gain since the pre-settlement era is the number of new cells as a proportion of historical cover for the tree taxon, thus positive values are not indicative of higher overall spatial coverage.
+Loss and Presence columns sum to 100% and represent the proprtion of the historical range that has been lost, or shown continuous presence.*
+
+```{r, results='as.is', echo=FALSE}
 pandoc.table(loss_table, justify = "left", style = 'simple')
@@
 
-**Figure 4.** *Changes in the overall climate space in the upper midwestern United States.
-The climate densities largely overlap, although it is possible to see the shifts in precipitation, mean and minium temperatures, where FIA data (pink) has higher values), while the decline in maximum temperatures is also visible, largely at the very highest values.*
+**Figure 4.** *Changes in the realized climate space for genera in the upper Midwestern United States.
+For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.
+At the same time minimum temperatures and annual precipitation have both increased.*
 
@@
 
-Interstingly, since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.
+Since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.
 
@@
 ----------------
+This work adds to our understanding of the relationships between shifts in realized climate niches and the interacting effects of climate change and land use.
 
-Figures
---------------
+Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000].
+Land use change reduces the correlational structure between species and climate [@devineau2011arid].
+While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015], and in the correlational structure between species and climate.
+Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013niche].
+
+From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States.
+However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].
+In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013niche].
 

### 2015-11-12 R/figures/gain_loss_plot.R

@@ -55,8 +55,8 @@ loss_plot <- function(){
   
   loss_plot <- natural$base_map +
       geom_tile(data = natural$rast_table, aes(x = x, y = y, fill = rgb)) +
-      geom_tile(data = subset(loss_melt, variable %in% 'Pine') , aes(x = x, y = y, fill = rgb)) +
-      #facet_wrap(~variable, ncol = 5) + theme_bw() +
+      geom_tile(data = loss_melt , aes(x = x, y = y, fill = rgb)) +
+      facet_wrap(~variable, ncol = 5) + theme_bw() +
       scale_fill_identity() +
   #    geom_path(data = umw.domain, aes(x = long, y = lat, group = group)) +
   #    geom_path(data = can.domain, aes(x = long, y = lat, group = group)) +
@@ -72,7 +72,7 @@ loss_plot <- function(){
   
   short_plot <- natural$base_map +
     geom_tile(data = natural$rast_table, aes(x = x, y = y, fill = rgb)) +
-    geom_tile(data = subset(loss_melt, variable %in% 'Oak') , aes(x = x, y = y, fill = rgb)) +
+    geom_tile(data = loss_melt, aes(x = x, y = y, fill = rgb)) +
     scale_fill_identity() +
     facet_wrap(~variable, ncol = 5) + theme_bw() +
     #geom_path(data = umw.domain, aes(x = long, y = lat, group = group), color = 'black') +

### 2015-11-18 R/Setup_R.R
A number of changes for the next revision:
`climate_boxplots.R` - Changed the legend title on the boxplot figure, as per discussion last week.

`Setup_R.R` - added `maptools` to the set of packages.

`goringetal_references.bib` - added sala, Devineau, a placeholder for Ivory and some others.

`confounding_table.R` this is the new test for whether changes are confounding, compounding or contradicting.

`FigureFour_Panels.R`, the remake of the four panel figure.  Stil lnot sure it's neccessary.  Need to bolster the text around it though.

@@ -9,6 +9,7 @@ library(gridExtra)
 library(rgdal)
 library(RColorBrewer)
 library(analogue)
+library(maptools)
 
 base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
 base.rast <- setValues(base.rast, 1:ncell(base.rast))

### 2015-11-18 R/confounding_table.R

@@ -1,0 +1,69 @@
+#  Okay, I have three types of change:
+#  (1) compounding
+#  (2) confounding
+#  (3) contradicting
+#
+#  * In compounding delta_C & LU should have the same net sign relative to the 
+#  underlying veg data.
+#  * In confounding the LU is in the middle of the distribution and plays little 
+#  to no role
+#  * In contradicting we should see that the LU has a different sign relative
+#  to the clim change.
+
+taxa <- unique(vegclim_table$taxon)
+clim_var <- unique(vegclim_table$climate)
+
+conf_test <- function(taxa, clim_var){
+  clim = t.test(subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
+         subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'FIA' & data > 0)$clim)
+
+  lu   = t.test(subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
+         subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'FIA' & c.ref == 'PLSS' & data > 0)$clim)
+
+  # We want to return the t-test p value & estimate
+  data.frame(taxa = taxa, clim = clim_var,
+             clim_est = diff(clim$estimate), clim_p = clim$p.value,
+             lu_est   = diff(lu$estimate),   lu_p   = lu$p.value)
+  
+}
+
+tests <- do.call(rbind.data.frame,
+        lapply(taxa, function(x){
+  do.call(rbind.data.frame,lapply(clim_var, function(y){conf_test(x, y)}))}))
+
+checker <- function(x){
+  # x will be the row:
+  # compounding:
+  # both ps are sig & both ests are in the same direction:
+  if(as.numeric(x[4]) < (0.05 / 120) & as.numeric(x[6]) < (0.05 / 120)){
+    if((as.numeric(x[3]) * as.numeric(x[5])) > 0){
+      return('+')
+    } else {
+      return('-')
+    }
+  } else {
+    if(as.numeric(x[4]) > (0.05 / 120)){
+      return('n/a')
+    } else {
+      return('.')
+    }
+  }
+}
+
+tests$result <- apply(tests, 1, checker)
[…]

### 2015-11-18 R/figures/FigureFour_Panels.R

@@ -1,0 +1,32 @@
+#  This is a hellinger plot of the four curves for Oak in vegclim_table:
+
+library(png)
+library(grid)
+
+ggplotColours <- function(n=6, h=c(0, 360) +15){
+  if ((diff(h)%%360) < 1) h[2] <- h[2] - 360/n
+  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
+}
+
+model_grid <- readPNG('figures/model_grid.png')
+
+oak <- subset(vegclim_table, taxon == 'Tamarack' & data > 0)
+
+oak_curves <- ggplot(oak) +
+  geom_density(adjust=1, size = 1.2,
+                         aes(x = clim, color = base, linetype = c.ref)) +
+            facet_wrap(~climate, scale = 'free') +
+            theme_bw() +
+            scale_color_manual(name = 'Vegetation', values = rev(ggplotColours(2))) +
+            scale_linetype_manual(name = 'Climate', values = c(2,1)) +
+            theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
+                  axis.title = element_blank(),
+                  strip.text = element_text(family = 'serif', face='bold', size = 14),
+                  legend.title = element_text(family = 'serif', face='bold', size = 14),
+                  legend.text = element_text(family = 'serif', size = 12),
+                  legend.position = 'none')
+
+ggsave(plot=grid.arrange(rasterGrob(model_grid), oak_curves, ncol=2,
+                         widths = c(3,4)), 
+       file = 'figures/figureFour_panels.tiff',
+       width=6, height = 3, dpi = 300)

### 2015-11-18 R/figures/climate_boxplots.R

@@ -9,7 +9,7 @@ clim_boxplots <- function(clim_table = vegclim_table){
                                        comb %in% c('FIA Era', 'PLS Era') & data>0 &
                                          !taxon %in% 'Ironwood')) +
     geom_boxplot(aes(y = clim, x = taxon, fill = comb)) +
-    scale_fill_grey() +
+    scale_fill_grey(name = 'Climate Era') +
     facet_grid(climate~., scales='free_y', labeller = label_parsed) +
     theme_bw() +
     theme(axis.text = element_text(family = 'serif', face='bold', size = 12),

### 2015-11-18 Composition_Ecology.Rmd
Changes to main Rmd: * trying to clear up abstract conflict between bullets & text. * Changed the version of the PLS we're pulling from. * Changed the way the figures are pulled in the text.  Stopped making them in text and now pulling them in from file. * Some renaming of references for proper rendering. * Moved Figure 2 further down in the text. * Trying to clarify the language around the comparisons. * Fixed up Table 1.

@@
 ====
-* Much of our understanding of species-climate relationships and functional-community relationships in forests is based on contemporary observations of tree distributions.
 * Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
@@
 
+*Using contemporary and historical species distributions and climate data we show that modern forest tree genera occupy realized niches with cooler maximum temperatures, higher minimum temperatures and higher rainfall than they have in the past.*
+
 ```{r, echo = FALSE, warning = FALSE, message=FALSE, results='hide'}
@@
 
-pls_data <- get_pls('0.9-4')
+pls_data <- get_pls('0.9-6')
 agg_dens <- get_fia()[,colnames(pls_data)]
@@
 Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States.
-There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than pre-settlement forests [@goring2015preset;@schulte2007homogenization].
+There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].
+
 This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
@@
 
-grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
+ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')
 
 ```
+![figone](figures/figure_one.tiff)
 
@@
 ===============
-We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.
+We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.
 The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.
@@
 
-The analysis presented here follows a blocked sampling method (Figure 2).
-We generate probability density functions along climate gradients for actual PLSS presence, by taxon, at the settlement era and for the same PLS taxa (and point locations) in the modern era.
-We do the same for FIA presence, and for anthropogenic land use classes.
+The analysis presented here follows a blocked sampling method.
+We generate probability density functions along historical climate gradients for taxa within the PLSS and for the same PLSS taxa (and point locations) using climate data for the modern era.
+We do the same for FIA presence (climate at the PLSS and modern eras), and for anthropogenic land use classes.
 Climate data is derived from monthly PRISM data from 1895 - 2014.
-A 30-year climate was generated for modern July temperature, January temperature, the continentality (difference between T_jul_ and T_jan_), and annual precipitation.
-
-![model grid](figures/model_grid.png)
-
-**Figure 2**.
-*Blocked method for analysis.
-Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*
+A 30-year climate normal was generated for modern July temperature ($T_{jul}$), January temperature ($T_{jan}$), the continentality (difference between $T_{jul}$ and $T_{jan}$, $T_{diff}$), and annual precipitation ($P_{ann}$).
 
@@
 
-By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation (Figure 2) we can use Hellinger distance to estimate the effects of changing climate (d_H_ between kernel densities in the same column of Figure 2), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).
+By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation ($PLSS_{veg}$-$PLSS_{clim}$, $PLSS_{veg}$-$FIA_{clim}$, $FIA_{veg}$-$PLSS_{clim}$, $FIA_{veg}$-$FIA_{clim}$) we can use Hellinger distance to estimate the effects of changing climate ($d_{H}$), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).
 In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).
@@
                                          loss_melt[[1]]$value))
-loss_table$Gain <- round(loss_table$Gain / (loss_table$Loss + loss_table$Presence)*100,1)
+loss_table <- data.frame(loss_table, 
+                         Modern = round((loss_table$Gain + loss_table$Presence) * .064, 0),
+                         Historical = round((loss_table$Loss + loss_table$Presence ) * .064, 0))
 
-loss_table[,2:3] <- round(tran(loss_table[,2:3], method = 'proportion') * 100,1)
+loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)
 
-loss_melt[[2]]
+ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
+       dpi=300, width=6, height = 3)
 ```
 
-**Figure 3.**.
+![figuretwo](figures/Figure_panelloss.tiff)
+**Figure 2.**.
 *Gain loss and continuous presence for PLSS and FIA tree taxa.
-Loss is assessed where previously PLSS data indicated presence of a taxon but FIA data do not record presence (light blue).
-Continuous presence is indicated by dark gray, for where both FIA and PLSS cells indicate presence.
-For most taxa it is the central parts of their distribution that show continuous presence.
+Loss is assessed as cells that contained presence during the PLSS era, and no presence in the FIA data (light blue).
+Continuous presence is indicated by dark gray, where both FIA and PLSS cells indicate presence.
 Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*
@@
 **Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era.
-Gain since the pre-settlement era is the number of new cells as a proportion of historical cover for the tree taxon, thus positive values are not indicative of higher overall spatial coverage.
-Loss and Presence columns sum to 100% and represent the proprtion of the historical range that has been lost, or shown continuous presence.*
+The first three columns represent percentages of the total area occupied by the taxon since settlement.
+Modern and Historical values represent 100s of km^2^ occupied by the tree taxon within either the FIA or PLSS data.*
 
@@
 ```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
-
 source('R/resampling_climate.R')
@@
 boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
-boxplot_clim
 
@@
 
-**Figure 4.** *Changes in the realized climate space for genera in the upper Midwestern United States.
+![FigureThree](figures/boxplot_clim.tiff)
+**Figure 3.** *Changes in the realized climate space for genera in the upper Midwestern United States.
 For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.
@@
 
-```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
+```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
+source('R/figures/FigureFour_Panels.R')
+```
+
+![modelgrid](figures/figureFour_panels.tiff)
+**Figure 4**.
+*Blocked method for analysis, with representative normalized densities for climate distributions.
+Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
+A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*
+
+Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables.
+Maximum temperature shows the greatest change attributed to land use change.
+Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change.
+The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.
+Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.
+
+```{r climate_shifts, warning=FALSE, message=FALSE, echo=FALSE}
 
@@
 
-hell_plot[[1]]
-
 ```
+![hellplot](figures/hellingerplot.tiff)
 
@@
 *For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.
-Taxa with negative y values have changes in distributions along the climate gradient that are driven largely by land use, while positive y axis values can be largely attributed to regional climate change.*
-
-The Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables.
-Maximum temperature shows the greatest change attributed to land use change.
-Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change.
-
-The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.
-Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.
+The x-axis represents the total shift (diagonal in Figure 4) from PLSS - FIA for both vegetation and climate, while the y-axis represents the difference between $d_{h}(V_{PLSS}C_{PLSS},V_{PLSS}C_{FIA})$ and $d_{h}(V_{FIA}C_{FIA},V_{PLSS}C_{FIA})$.
+Taxa with negative y-axis values have changes in distributions along the climate gradient that are driven largely by land use, while positive y-axis values can be largely attributed to regional climate change.*
 
 Since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.
+
+**Table 2**.
+*Compounding (+), confounding (.) and contradicting (-) relationships between land use and climate change.
+There is no relationship between spatial correlates and the land use/climate interaction.*
+
+```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is'}
+source('R/confounding_table.R')
+
+pandoc.table(conf_table, justify = "left", style = 'simple')
+
+```
 
@@
 Land use change reduces the correlational structure between species and climate [@devineau2011arid].
-While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015], and in the correlational structure between species and climate.
-Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013niche].
+While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015preset], and in the correlational structure between species and climate.
+Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013building].
 
@@
 However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].
-In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013niche].
+In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013building;@early2014climatic].
 

### 2016-02-01 Composition_Ecology.Rmd
Edits to the Intro & figure captions.  Representing changes throughout.
Some minor fixes to references.

Changed plot output for hellinger plot to PDF to try to get better resolution of text.

`climate_ts` increased the font size of the strip axis.

Pushing figures.

@@
   word_document:
-    fig_height: 6
-    fig_width: 6
   pdf_document:
@@
 library("plyr")
-panderOptions('table.style', 'rmarkdown')
 ```
@@
 
-This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
+This shift in forest composition and co-occurrence has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
 One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.
@@
 Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].
+A second issue is that climate normals may be insufficient for modeling seedling recruitment [@serra2015averaged].
+Given the widespread deforestation following settlement, one or few years of cold or hot weather could affect establishment regionally, resuling in a significantly changed landscape.
 
@@
 ![figuretwo](figures/Figure_panelloss.tiff)
+
 **Figure 2.**.
-*Gain loss and continuous presence for PLSS and FIA tree taxa.
-Loss is assessed as cells that contained presence during the PLSS era, and no presence in the FIA data (light blue).
-Continuous presence is indicated by dark gray, where both FIA and PLSS cells indicate presence.
-Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*
+*Maps showing areas of gain, loss, and continuous presence of tree genera.
+'Loss' indicates areas where PLSS data indicated presence of a genus but FIA data do not record presence (light blue).
+'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red).
+'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray).
+Losses are more common than gains, particularly in the southwestern portion of tree ranges.
+Few taxa show gains, although poplar, ash and basswood show gains of over 10%.*
 
-**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era.
-The first three columns represent percentages of the total area occupied by the taxon since settlement.
-Modern and Historical values represent 100s of km^2^ occupied by the tree taxon within either the FIA or PLSS data.*
+**Table 1.** *Gain and loss of the analyzed tree genera since the pre-settlement era.
+The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras.
+Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.
+Taxa are sorted by the percent area lost (Column 3).*
 
-```{r, results='as.is', echo=FALSE}
-pandoc.table(loss_table, justify = "left", style = 'simple')
+```{r table1, results='as.is', echo=FALSE, eval=TRUE}
+
+pandoc.table(loss_table, justify = "left")
 ```
@@
 ![FigureThree](figures/boxplot_clim.tiff)
-**Figure 3.** *Changes in the realized climate space for genera in the upper Midwestern United States.
-For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.
-At the same time minimum temperatures and annual precipitation have both increased.*
+
+**Figure 3.** *Changes in the realized climate distributions for tree genera in the upper Midwestern United States.
+In the box-and-whisker plots, the central line indicates the median, box margins indicate 25th and 75th percentiles, whisker tips indicate 5th and 95th percentiles, and dots indicate outliers beyong the 95th percentile.
+These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land use change.*
 
@@
 ![modelgrid](figures/figureFour_panels.tiff)
+
 **Figure 4**.
-*Blocked method for analysis, with representative normalized densities for climate distributions.
-Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*
+*(a) Diagram and legend for 2x2 factorial design for diagnosing the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships.
+(b-e).
+Representative normalized densities for Larix relative to each of the four climate variables.
+We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).*
 
@@
   **Figure 5**.
-*For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.
-The x-axis represents the total shift (diagonal in Figure 4) from PLSS - FIA for both vegetation and climate, while the y-axis represents the difference between $d_{h}(V_{PLSS}C_{PLSS},V_{PLSS}C_{FIA})$ and $d_{h}(V_{FIA}C_{FIA},V_{PLSS}C_{FIA})$.
-Taxa with negative y-axis values have changes in distributions along the climate gradient that are driven largely by land use, while positive y-axis values can be largely attributed to regional climate change.*
+*The total shift in climatic distribution over the last 100-150 years for each tree genus plotted against the amount of shift caused by historic climate change vs.
+changes in tree distributions, the latter mainly due to historic land use and pathogen outbreaks.
+Results are plotted separately for each of the four climate variables considered here.
+Total shifts in distributions ($D_{tot}$) are calculated using the Hellinger distance between climatic distributions for tree genera for the presettlement and modern time periods.
+The importance of historic climate shifts on shifting climate distributions of tree genera is determined by calculating the Hellinger distance between PLSS tree distributions overlaid with late 19th century climates versus modern climates ($D_{C}$).
+The importance of historic vegetation change on shifting climate distributions of tree genera is determined by calculating the Hellinger distance between PLSS and FIA tree distributions when both are overlaid on late 19th-century climates ($D_{V}$).
+$D_{C}$ and $D_{V}$ are then differenced from each other, such that negative values indicate a stronger influence of historic land use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.*
 
@@
 
-```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is'}
+```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is', eval=TRUE}
 source('R/confounding_table.R')
 
-pandoc.table(conf_table, justify = "left", style = 'simple')
+rownames(conf_table) <- NULL
+
+pander::pandoc.table(conf_table, justify = "left")
 
 ```
+
+Our results show that the largest class of land use/climate interaction (Table 2) since the onset of EuroAmerican settlement in the Upper Midwest is Compounding, accounting for `r table(unlist(conf_table[,2:5]))['+']/(nrow(conf_table)*4) * 100`% of the possible interactions.
+Confounding interactions account for `r table(unlist(conf_table[,2:5]))['.']/(nrow(conf_table)*4)*100`% of interactions, while only `r table(unlist(conf_table[,2:5]))['-']/(nrow(conf_table)*4)*100`% of interactions are contradicting.
+
+There is no clear relationship between taxa with confounding, contradicting or compounding interactions and the spatial distribution of those taxa.
+For climate variables, the proportion of compounding interactions is greatest for $T_{max}$, and least for $T_{min}$.
+For other climate variables we see a general balance of Only Elm (for $T_{max}$) and Tamarack (for $T_{min}$) show contradicting changes for climate/land use interactions.
 
@@
 
-Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000].
+Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
 Land use change reduces the correlational structure between species and climate [@devineau2011arid].

### 2016-02-01 R/confounding_table.R

@@ -67,3 +67,4 @@ conf_table$mean_y <- sapply(taxa, mean_y)
 conf_table$mean_x <- sapply(taxa, mean_x)
 
 conf_table <- conf_table[order(conf_table$mean_x),]
+

### 2016-02-01 R/figures/climate_ts.R

@@ -60,7 +60,7 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
     theme_bw() +
     theme(axis.text = element_text(family = 'serif', size = 12),
           axis.title = element_text(family = 'serif', face = 'bold', size = 14),
-          strip.text = element_text(family = 'serif', face='bold', size = 14),
+          strip.text = element_text(family = 'serif', face='bold', size = 16),
           legend.position = "none")
     
   ggsave(clim_plot, filename = 'figures/clim_plot.tiff', width = 8, height = 6, dpi = 150)

### 2016-02-01 R/hellinger_dists.R

@@ -78,7 +78,7 @@ hellinger_plot <- function(){
              label='Land Use Dominates', family='serif',
              fontface = 'bold', size = 5)
   
-  ggsave(plot = hell_plot, filename = 'figures/hellingerplot.tiff', 
+  ggsave(plot = hell_plot, filename = 'figures/hellingerplot.pdf', 
          width = 10, height = 8, dpi = 150)
  
   list(hell_plot, all.hellinger)

### 2016-02-26 Composition_Ecology.Rmd
Working on the manuscript.  Added `panderOptions`, revised the introdction.
Figure captions are all revised.  Edits throughout the methods.  Results & Discussion mostly unchanged.

Number of article references added.

@@
 author: "Simon Goring *et al*."
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: '`r format(Sys.time(), "%d %B, %Y")`'
 output:
-  word_document:
   pdf_document:
@@
     variant: markdown_github
+  word_document: null
 csl: ecology.csl
@@
 library("plyr")
+
+panderOptions('missing', '--')
+panderOptions('digits', 4)
+panderOptions('keep.trailing.zeros', TRUE)
+panderOptions('table.continues', NULL)
+panderOptions('table.split.table', Inf)
+pander::panderOptions('table.style', 'rmarkdown')
+grDevices::ps.options(fonts = 'serif')
+
 ```
@@
 ====
-* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
-* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
-* We assess the stability of species-climate relationships using historic PLS data and FIA data.
+* Observational datasets of contemporary species-climate relationships are widely used understand environmental controls on species distriutions and project the responses of species to 21^st^-century climate change.
+* Given recent climate change and continued land use conversion, are species-climate relationships today representative of those in past.
+* We assess the stability of species-climate relationships using historic PLSS data and FIA data.
 * We show that species-climate relationships have significantly shifted for many taxa.
@@
 
-source('R/Setup_R.R')
-source('R/load_climate.R') # gives `climate` with elements `then`, `now` and `names`
-source('R/load_veg.R')
+source("R/Setup_R.R")
+source("R/load_climate.R") # gives `climate` with elements `then`, `now` and `names`
+source("R/load_veg.R")
 
-best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
-              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
-              'tamarack', 'maple', 'poplar', 'birch')
+best.taxa <- c("beech", "ironwood", "hemlock", "fir", "cedar.juniper",
+              "basswood", "spruce", "elm", "ash", "pine", "oak",
+              "tamarack", "maple", "poplar", "birch")
 
-pls_data <- get_pls('0.9-6')
+pls_data <- get_pls("0.9-6")
 agg_dens <- get_fia()[,colnames(pls_data)]
@@
 
-source('R/natural_earth.R')
+source("R/natural_earth.R")
 
@@
 ==============
+4)	Some researchers have focused on historical resurveys to assess climate-driven range shifts (Moritz, Waller, Tingley, etc.)
+5)	Others have examined stability or plasticity of species-climate distributions during the last deglaciation (Pearman, Veloz, AT Petersen, etc.)
+
+Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, long-term datasets to map distributions of species and species traits within environmental space.
+Distributions of species within climate space can be used to assess species risk and resilience to climate change [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels] and these models are then used to inform management and adaptation decisions in the future, although their role in developing planning scenarios is often unclear [@guisan2013predicting].
+Understanding climate-related risk and response for ecological communities within landscapes is critically important for future adaptation planning, but we are often limited by a narrow window of observation that may obscure greater plasticity in species-climate responses [@maiorano2013building], and the constraints that result from historical contingency [@svenning2015influence].
+Process-based species distribution models using physiological data as a primary basis [e.g., @leiblein2015modelling] may address issues of climate disequilibrium within models, focusing on representing the true niche space using experimental data at the individual and community level, however, parameterization of these models is critical, and has been shown to provide poor match at high levels [@matthes].
+Increasing recognition that historical distributions show evidence of lags in species climate responses [@webb1986vegetation;@ordonez2013realized;svenning2015influence] and that habitat loss through human agency can impact apparent species distribtutions at both long [@ivory2015] and shorter time scales [@sarmento2013impacts] point to the need to understand species-climate distributions across time-scales, wherever the model lies along the gradient from correlational to process based [@dormann]and to quantify land use interactions with distribution modelling.
+
+Focus on past distributions has taken multiple approaches, including historical resurveys [@li2014drivers;@tingley2012push] which provide evidence of sustained shifts in distribution and co-occurence [@goring2016presettlement].
+Integration of paleoecological data can provide greater insight into shifting ecological patterns within communities [@jackson2015community], but one of the largest shifts in community composition within the upper Midwestern United States occured with EuroAmerican Settlement in the 1800s [@goringetal2016;@otherrefs] as a result of changing land use, and simultaneously, regional shifts in climate at the end of the Little Ice Age [@paquette2013climatic;others].
+Both land use change and climate change are ongoing, and the challenge of disentangling the influence of each on modern species distributions is complex.
+The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+Studies into the interactions between land use and climate change have been largely focused on future impacts [@oliver2014interactions;@jetz2007projected], but widespread land use conversion has already occured and has had significant impacts on modern vegetation structure and composition at regional [@gehrig2007tree] and global scales [@].
+A useful framework for assessing the interaction of climate and land use change considers three types of interaction: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat].
+Using this framework it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to regional climate change.
+
+Compounding effects, when climatic change and land use change are biased in the same direction, are seen in butterfly species in the northern Sierra Nevada mountains of California, where warming, compounded by land use conversion at the lowest elevations drives species upslope more rapidly than might otherwise be expected [@forister2010compounded].
+Confounding effects may cause little overall change on species distributions, land use pressure falls within the central distribution for the environmental gradient.
+The spatial extent of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little, but land use conversion has excluded it from the deep soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], thus predictive models for Garry oak [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used.
+Contradicting effects result in a narrower climate niche for taxa.
+Some bird species in the Nierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land use conversion in the lowlands.
 
@@
 There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].
-
-This shift in forest composition and co-occurrence has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
+Shifts in forest composition and co-occurrence has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
 One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.
@@
 
-land_use$class[land_use$class == 'Water'] <- NA
+land_use$class[land_use$class == "Water"] <- NA
 
@@
   coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
-  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
-  theme(legend.position = 'none')
+  scale_fill_manual(values = c("#ff7f00", "#e41a1c", "#33a02c")) +
+  theme(legend.position = "none")
 
-source('R/figures/climate_ts.R')
+source("R/figures/climate_ts.R")
 
@@
 
-ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')
+ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename="figures/figure_one.tiff")
 
 ```
-![figone](figures/figure_one.tiff)
+![figone](figures/Fig_1_Composite.png)
 
@@
 (f) Patterns of land use change in the upper Midwestern United States.
-Data from the NLCD [@jin2013comprehensive].*
-
-The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].
-Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.
-By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.
-
-Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.
-In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.
-This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].
-Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient.
-Confounding effects may increase the variability of predictive models, but do little to the overall envelope.
-Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.
-The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used.
-Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.
-Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.
+Land use data from the NLCD [@jin2013comprehensive].*
 
@@
 Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.
-
 For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients.
@@
 Tree species are responding to climate change within forested landscapes, but the response is inconsistent.
-Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure].
+Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but appear to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure].
 Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns.
-For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.
+For species that reach their northern or western range limits in the upper Midwestern United States (*e.g.*, hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, due to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.
 
@@
 ===============
-We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.
-The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.
+Pre-settlement era vegetation data from the Public Land Survey System (PLSS), aggregated to a 64 $km^{2}$ grid [@goring2016preset] is paired with annually resolved climate data for the region.
+The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLSS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.
 At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.
@@
 [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.
-When examining relationships between FIA and PLS data, Goring et al.
+When examining relationships between FIA and PLSS data, Goring et al.
 [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.
@@
 Another complication is the structure of the FIA dataset.
-While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].
+While the PLSS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].
 As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA.
@@
 
-The choice of a thirty year normal for the PLS data is potentially problematic.
-The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist.
+The choice of a thirty year normal for the PLSS data is potentially problematic.
+The time transgressive nature of the PLSS Survey means that some of the PLSS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist.
 Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
@@
 
-source('R/figures/gain_loss_plot.R')
+source("R/figures/gain_loss_plot.R")
 
@@
 
-loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)
+loss_table[,1:3] <- round(tran(loss_table[,1:3], method = "proportion") * 100,1)
 
-ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
+ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
        dpi=300, width=6, height = 3)
@@
 
-```{r table1, results='as.is', echo=FALSE, eval=TRUE}
+```{r table1, results="as.is", echo=FALSE, eval=TRUE}
 
-pandoc.table(loss_table, justify = "left")
+pandoc.table(loss_table, justify = "left", style = "rmarkdown")
 ```
@@
 ```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
-source('R/resampling_climate.R')
+source("R/resampling_climate.R")
 vegclim_table <- resample_climate(climate, agg_dens, pls_data)
 
-source('R/figures/climate_boxplots.R')
+source("R/figures/climate_boxplots.R")
 boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
@@
 ```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
-source('R/figures/FigureFour_Panels.R')
+source("R/figures/FigureFour_Panels.R")
 ```
@@
 
-source('R/hellinger_dists.R')
+source("R/hellinger_dists.R")
 
@@
 
-```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is', eval=TRUE}
-source('R/confounding_table.R')
+```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results="as-is", eval=TRUE}
+source("R/confounding_table.R")
 
@@
 
-pander::pandoc.table(conf_table, justify = "left")
+pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
 
@@
 
-Our results show that the largest class of land use/climate interaction (Table 2) since the onset of EuroAmerican settlement in the Upper Midwest is Compounding, accounting for `r table(unlist(conf_table[,2:5]))['+']/(nrow(conf_table)*4) * 100`% of the possible interactions.
-Confounding interactions account for `r table(unlist(conf_table[,2:5]))['.']/(nrow(conf_table)*4)*100`% of interactions, while only `r table(unlist(conf_table[,2:5]))['-']/(nrow(conf_table)*4)*100`% of interactions are contradicting.
+Our results show that the largest class of land use/climate interaction (Table 2) since the onset of EuroAmerican settlement in the Upper Midwest is Compounding, accounting for `r table(unlist(conf_table[,2:5]))["+"]/(nrow(conf_table)*4) * 100`% of the possible interactions.
+Confounding interactions account for `r table(unlist(conf_table[,2:5]))["."]/(nrow(conf_table)*4)*100`% of interactions, while only `r table(unlist(conf_table[,2:5]))["-"]/(nrow(conf_table)*4)*100`% of interactions are contradicting.
 

### 2016-02-26 R/Setup_R.R
Some changes to the R files.  Mostly stylistic (with the addition of `lintr` integration).

@@ -11,22 +11,27 @@ library(RColorBrewer)
 library(analogue)
 library(maptools)
 
-base.rast <- raster('data/albers/density//Abies_balsamea_density_alb.tif')
-base.rast <- setValues(base.rast, 1:ncell(base.rast))
+base_rast <- raster("data/albers/density//Abies_balsamea_density_alb.tif")
+base_rast <- setValues(base_rast, 1:ncell(base_rast))
 
-pls.rast <- raster('data/input//dens_v0.9-4.tif')
-unit.raster <- setValues(pls.rast, 1:ncell(pls.rast))
+pls_rast <- raster("data/input//dens_v0.9-4.tif")
+unit_raster <- setValues(pls_rast, 1:ncell(pls_rast))
 
-land_use <- raster('../../../Maps/nlcd_2011_landcover_2011_edition_2014_10_10//nlcd_cropped_repro.tif')
+land_use <- raster(paste0("../../../Maps/",
+                          "nlcd_2011_landcover_2011_edition",
+                          "_2014_10_10//nlcd_cropped_repro.tif"))
 
 #  We need to change the raster to a data.frame:
-land_use <- data.frame(cell = extract(unit.raster, xyFromCell(land_use, 1:ncell(land_use))),
+land_use <- data.frame(cell = extract(unit_raster,
+                                      xyFromCell(land_use, 1:ncell(land_use))),
                        xyFromCell(land_use, 1:ncell(land_use)),
                        value = getValues(land_use))
-reassign <- read.csv('data/input/landuse_reassignment.csv', stringsAsFactors=FALSE)
+
+reassign <- read.csv("data/input/landuse_reassignment.csv",
+                     stringsAsFactors = FALSE)
 land_use$class <- factor(reassign[match(land_use$value, reassign[,1]),2])
 
-model.proj <- '+init=epsg:3175'
+model_proj <- "+init=epsg:3175"
 
-usa <- readOGR('data/input/shapes/usa/us.shp', 'us')
-canada <- readOGR('data/input/shapes/canada/PROVINCE.SHP', 'PROVINCE')
+usa <- readOGR("data/input/shapes/usa/us.shp", "us")
+canada <- readOGR("data/input/shapes/canada/PROVINCE.SHP", "PROVINCE")

### 2016-02-26 R/figures/gain_loss_plot.R

@@ -1,13 +1,13 @@
 #  Plotting for taxa that shows the change between historical and modern distributions.
 
-get_loss_gain <- function(unit.raster, best.taxa, pls_data, agg_dens){
-  loss.gain <- data.frame(cell = 1:ncell(unit.raster))
+get_loss_gain <- function(unit_raster, best.taxa, pls_data, agg_dens){
+  loss.gain <- data.frame(cell = 1:ncell(unit_raster))
   
   agg_dens[is.na(agg_dens)] <- 0
   
-  for(i in 1:length(best.taxa)){
-    loss.gain[,(i+1)] <- NA
-    colnames(loss.gain)[i+1] <- best.taxa[i]
+  for(i in 1:length(best.taxa)) {
+    loss.gain[,(i + 1)] <- NA
+    colnames(loss.gain)[i + 1] <- best.taxa[i]
     
     sub.pls <- na.omit(pls_data[pls_data[,best.taxa[i]] > 0 & !is.na(pls_data$cell),])
     sub.fia <- na.omit(agg_dens[agg_dens[,best.taxa[i]] > 0 & !is.na(agg_dens$cell),])
@@ -23,22 +23,22 @@ get_loss_gain <- function(unit.raster, best.taxa, pls_data, agg_dens){
   
   loss.melt <- na.omit(melt(loss.gain, id = 'cell'))
   
-  loss.melt <- data.frame(xyFromCell(unit.raster, loss.melt$cell),
+  loss.melt <- data.frame(xyFromCell(unit_raster, loss.melt$cell),
                           loss.melt)
   loss.melt
 }
 
 loss_plot <- function(){
   
-  if(model.proj == '+init=epsg:4326'){
+  if(model_proj == '+init=epsg:4326'){
     ext <- c(-98, -83, 42, 50)
   }
   
-  if(model.proj == '+init=epsg:3175'){
+  if(model_proj == '+init=epsg:3175'){
     ext <- c(-100000, 1050000, 600000, 1600000)
   }
   
-  loss_melt <- get_loss_gain(unit.raster, best.taxa, pls_data, agg_dens)
+  loss_melt <- get_loss_gain(unit_raster, best.taxa, pls_data, agg_dens)
 
   loss_melt$rgb <- brewer.pal(n = 3, 'Dark2')[as.numeric(factor(loss_melt$value, 
                                                                 levels = c('Presence', 'Loss', 'Gain')))]
[…]

### 2016-02-26 R/load_veg.R

@@ -4,13 +4,13 @@ get_fia <- function(x){
   
   #  Load FIA composition data:
   get.fia.stack <- function(x, type){
-    stack.files <- list.files(x, full = TRUE)
-    stack.fia <- resample(stack(stack.files), pls.rast, method = 'ngb')
-    stack.fia[is.na(getValues(pls.rast))] <- NA
+    stack.files <- list.files(x, full.names = TRUE)
+    stack.fia <- resample(stack(stack.files), pls_rast, method = 'ngb')
+    stack.fia[is.na(getValues(pls_rast))] <- NA
     
     fia.vals <- getValues(stack.fia)
-    name.end <- regexpr(paste('_', type, sep=''), colnames(fia.vals), fixed=TRUE)
-    colnames(fia.vals) <- gsub("_$","", substr(colnames(fia.vals), 1, name.end-1))
+    name.end <- regexpr(paste('_', type, sep = ''), colnames(fia.vals), fixed = TRUE)
+    colnames(fia.vals) <- gsub("_$","", substr(colnames(fia.vals), 1, name.end - 1))
     colnames(fia.vals) <- gsub('_', '.', colnames(fia.vals))
     
     fia.vals <- as.data.frame(fia.vals)
@@ -45,7 +45,7 @@ get_fia <- function(x){
     x.cast$x <- x$x[x.cast$orig]
     x.cast$y <- x$y[x.cast$orig]
     
-    x.cast$cell <- extract(setValues(unit.raster, 1:ncell(unit.raster)), x.cast[,c('x', 'y')])
+    x.cast$cell <- extract(setValues(unit_raster, 1:ncell(unit_raster)), x.cast[,c('x', 'y')])
     
     x.cast <- x.cast[!is.na(x.cast$cell),!is.na(colnames(x.cast))]
     x.melt_2 <- melt(x.cast[ ,!colnames(x.cast)%in% c('orig')], 

### 2016-02-26 R/natural_earth.R

@@ -13,11 +13,11 @@ natural_earth <- function(){
   
   xylimits <-  c(-200000, 1150000, 500000, 1700000)
   
-  if(model.proj == '+init=epsg:4326'){
+  if(model_proj == '+init=epsg:4326'){
     ext <- c(-98, -83, 42, 50)
   }
   
-  if(model.proj == '+init=epsg:3175'){
+  if(model_proj == '+init=epsg:3175'){
     ext <- c(-100000, 1050000, 600000, 1600000)
   }
   
@@ -33,7 +33,7 @@ natural_earth <- function(){
     aa <- x.join[,1:2]
     coordinates(aa) <- ~ long + lat
     proj4string(aa) <- CRS(proj4string(nat.earth))
-    aa <- coordinates(spTransform(aa, CRS(model.proj)))
+    aa <- coordinates(spTransform(aa, CRS(model_proj)))
     
     x.join[,1:2] <- aa
     
@@ -50,10 +50,10 @@ natural_earth <- function(){
   coast.subset <- quick.subset('data/input/NaturalEarth/Coasts/ne_10m_coastline.shp',
                                'ne_10m_coastline', ext)
   
-  umw.domain <- spTransform(usa, CRS(model.proj))
-  can.domain <- spTransform(canada, CRS(model.proj))
+  umw.domain <- spTransform(usa, CRS(model_proj))
+  can.domain <- spTransform(canada, CRS(model_proj))
   
-  nat.crop <- crop(projectRaster(nat.earth, crs = model.proj), y=extent(ext))
+  nat.crop <- crop(projectRaster(nat.earth, crs = model_proj), y=extent(ext))
   
   rast.table <- data.frame(xyFromCell(nat.crop, 1:ncell(nat.crop)),
                            getValues(nat.crop/255))

### 2016-02-26 R/resampling_climate.R

@@ -7,8 +7,8 @@ resample_climate <- function(climate, agg_dens, pls_data){
   agg_dens[,-1] <- tran(agg_dens[,-1], method = 'proportion')
   pls_data[,-1] <- tran(pls_data[,-1], method = 'proportion')
   
-  then_rast <- resample(climate$then, unit.raster)
-  now_rast <- resample(climate$now, unit.raster)
+  then_rast <- resample(climate$then, unit_raster)
+  now_rast <- resample(climate$now, unit_raster)
   
   #For each taxon sample:
   #  1.  The FIA data for one of the now_rast years for each of the climate variables.
@@ -38,10 +38,10 @@ resample_climate <- function(climate, agg_dens, pls_data){
   
     #  bind all the variables together:  
     values <- data.frame(cell  = c(rep(pls_data$cell, 2), rep(agg_dens$cell, 2)),
-                         x     = c(rep(xyFromCell(unit.raster, pls_data$cell)[,'x'],2),
-                                   rep(xyFromCell(unit.raster, agg_dens$cell)[,'x'],2)),
-                         y     = c(rep(xyFromCell(unit.raster, pls_data$cell)[,'y'],2),
-                                   rep(xyFromCell(unit.raster, agg_dens$cell)[,'y'],2)),
+                         x     = c(rep(xyFromCell(unit_raster, pls_data$cell)[,'x'],2),
+                                   rep(xyFromCell(unit_raster, agg_dens$cell)[,'x'],2)),
+                         y     = c(rep(xyFromCell(unit_raster, pls_data$cell)[,'y'],2),
+                                   rep(xyFromCell(unit_raster, agg_dens$cell)[,'y'],2)),
                          data  = c(rep(as.numeric(pls_data[,taxon]), 2),
                                    rep(agg_dens[,taxon],2)),
                          base  = c(rep('PLSS', nrow(pls_data)), 

### 2016-06-17 R/Setup_R.R
Added a header block to uncomment ifpeople run this from a clone.  Added the `captioner` package to help captioning the Figures & Tables.

@@ -1,5 +1,10 @@
 #  Setup some data and libraries:
+#install.packages(c('colorspace', 'raster', 'ggplot2', 'mgcv', 'reshape2', 'plyr', 
+#                   'gridExtra', 'rgdal', 'RColorBrewer', 'analogue', 'maptools',
+#                   'captioner'))
+
 library(colorspace)
+library(captioner)
 library(raster)
 library(ggplot2)
 library(mgcv)
@@ -10,6 +15,9 @@ library(rgdal)
 library(RColorBrewer)
 library(analogue)
 library(maptools)
+
+fig_nums <- captioner(prefix = "Figure")
+tab_nums <- captioner(prefix = "Table")
 
 base_rast <- raster("data/albers/density//Abies_balsamea_density_alb.tif")
 base_rast <- setValues(base_rast, 1:ncell(base_rast))

### 2016-06-17 R/confounding_table.R
Mostly `lint`ing, adding some p adjustment for the estimates and cleaning up the table output so that it shows up as symbols rather than values.

@@ -13,7 +13,7 @@
 taxa <- unique(vegclim_table$taxon)
 clim_var <- unique(vegclim_table$climate)
 
-conf_test <- function(taxa, clim_var){
+conf_test <- function(taxa, clim_var) {
   clim = t.test(subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
          subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'FIA' & data > 0)$clim)
 
@@ -28,21 +28,22 @@ conf_test <- function(taxa, clim_var){
 }
 
 tests <- do.call(rbind.data.frame,
-        lapply(taxa, function(x){
-  do.call(rbind.data.frame,lapply(clim_var, function(y){conf_test(x, y)}))}))
+        lapply(taxa, function(x) {
+  do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y)}))}))
 
-checker <- function(x){
+checker <- function(x) {
   # x will be the row:
   # compounding:
   # both ps are sig & both ests are in the same direction:
-  if(as.numeric(x[4]) < (0.05 / 120) & as.numeric(x[6]) < (0.05 / 120)){
-    if((as.numeric(x[3]) * as.numeric(x[5])) > 0){
+  if (p.adjust(as.numeric(x[4]), method = "bonferroni", n = 120) < 0.05 & 
+      p.adjust(as.numeric(x[6]), method = "bonferroni", n = 120) < 0.05) {
+    if ((as.numeric(x[3]) * as.numeric(x[5])) > 0) {
       return('+')
     } else {
       return('-')
     }
   } else {
-    if(as.numeric(x[4]) > (0.05 / 120)){
+    if (as.numeric(x[4]) > (0.05 / 120)) {
       return('n/a')
     } else {
       return('.')
@@ -55,16 +56,27 @@ conf_table <- dcast(tests, taxa ~ clim, value.var = 'result')
 
 conf_table <- conf_table[match(taxa, conf_table$taxa),]
 
-mean_y <- function(taxa){
+mean_y <- function(taxa) {
   mean(subset(vegclim_table, taxon == taxa & base == 'PLSS' & data > 0)$y)
 }
 
-mean_x <- function(taxa){
+mean_x <- function(taxa) {
[…]

### 2016-06-17 R/figures/climate_ts.R
Mostly `lint`ing, added some explicit package calls to some of the functions.  This should really become practice.

@@ -12,16 +12,21 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
   #  Load the PDSI data for the region:
   #  download from - http://www.ncdc.noaa.gov/paleo/pdsi.html
   #  We shortcut this if the pdsi_df file exists:
-  if('pdsi_df.rds' %in% list.files('data/output/')){
+  if ('pdsi_df.rds' %in% list.files('data/output/')) {
     pdsi_df <- readRDS('data/output/pdsi_df.rds')  
-  } else{
-    pdsi_out <- function(unit.raster, pls_data){
+  } else {
+    pdsi_out <- function(unit.raster, pls_data) {
       NCDF <- stack('data/input/NADAv2-2008.nc')
-      ncdf.resamp <- resample(projectRaster(NCDF, crs = '+init=epsg:3175'), y = unit.raster)
-      pdsi <- getValues(ncdf.resamp)[pls_data$cell,]
+      
+      ncdf.resamp <- raster::resample(raster::projectRaster(NCDF, 
+                                                            crs = '+init=epsg:3175'), 
+                                      y = unit.raster)
+      
+      pdsi <- raster::getValues(ncdf.resamp)[pls_data$cell,]
+      
       pdsi_df <- data.frame(variable = 'pdsi', 
-                            value = colMeans(pdsi, na.rm=TRUE)[1:206],
-                            year = 2006:1801)
+                            value    = colMeans(pdsi, na.rm = TRUE)[1:206],
+                            year     = 2006:1801)
       pdsi_df
     }
     
@@ -36,11 +41,11 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
   clim_pattern <- function(rast_in, pls_rast, year){
   
     rast_mask <- rast_in
-    rast_mask[!(1:ncell(rast_in))%in%pls_data$cell]<- NA
+    rast_mask[!(1:ncell(rast_in)) %in% pls_data$cell] <- NA
   
     data.frame(variable = c('P[ann]', 'T[max]', 'T[diff]', 'T[min]'),
-               value = colMeans(getValues(rast_mask), na.rm=TRUE),
-               year = year)
+               value    = colMeans(getValues(rast_mask), na.rm = TRUE),
+               year     = year)
   }
   
   modern <- clim_pattern(now_rast, pls_rast, 1995)
@@ -51,16 +56,16 @@ climate_ts <- function(unit.raster, pls_data, now_rast, then_rast){
[…]

### 2016-06-17 Composition_Ecology.Rmd
Broad text edits to reflect changes by @IceAgeEcologist and mine as well.  Also reformatting for Ecol. Letters.

--- 2015-11-18 Composition_Ecology.Rmd
@@
 ---
-title: "Composition-climate paper"
-author: "Simon Goring *et al*."
-date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
-  word_document:
-    fig_height: 6
-    fig_width: 6
+  word_document: null
   pdf_document:
@@
 ---
-
-Effect of historic land use and climate change on tree-climate relationships for key forest genera in the northern United States
-========================================================
 
@@
 library("plyr")
-panderOptions('table.style', 'rmarkdown')
+
+panderOptions('missing', '--')
+panderOptions('digits', 4)
+panderOptions('keep.trailing.zeros', TRUE)
+panderOptions('table.continues', NULL)
+pander::panderOptions('table.style', 'rmarkdown')
+pander::panderOptions('table.split.table', 300)
+grDevices::ps.options(fonts = 'serif')
+
 ```
@@
 
-Abstract
-====
-* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
-* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
-* We assess the stability of species-climate relationships using historic PLS data and FIA data.
-* We show that species-climate relationships have significantly shifted for many taxa.
-* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
-* Relative importance of land use and climate change varies by taxon and climate variable.  
-* Shifts in warmer temperature mainly due to land use change, etc.
-* Implications sentence or two at the end.
+# Effect of historic land-use and climate change on tree-climate relationships in the northern United States
 
-*Historic land use and forest change over the centuries since Euroamerican settlement have significantly changed the structure and composition of forests in the upper Midwestern United States.
-The extent to which this change has affected the realized climate niche for key taxa is less well known.*
+**Statement of Authorship**
 
-*Predictions for future species range shifts are largely predicated on correlative models that relate modern species distributions to climate parameters in the modern era.
-Given the extent of regional forest change, and recent losses to key forest taxa including Hemlock, Elm and Chestnut, it is possible that certain taxa have shifted their climatic distributions over the last century.
-The additional pressure of climatically biased land use conversion for agriculture along the western and southern margins of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*
+* Simon J Goring: Conceived paper, performed analysis, developed figures and wrote the paper
+* John "Jack" W Williams: Wrote the paper
 
-*Using contemporary and historical species distributions and climate data we show that modern forest tree genera occupy realized niches with cooler maximum temperatures, higher minimum temperatures and higher rainfall than they have in the past.*
+**Data Accessibility**
+
+* All data for this manuscript, along with code and version-control is available at http://github.com/SimonGoring/CompositionClimate.
+In addition, an archived copy of the published version is available with the following DOI.
+
+**Authors**
+
+* Simon J Goring
+  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
+* Jack W Williams
+  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
+  * Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706
+
+**Keywords**
+
+* climate change, land use, historical ecology, realized niche, anthropocene, public land survey, forest inventory and analysis
+
+**Running Title**
+
+* Land use change forces climate shifts in trees
+
+**Article Type**
+
+* Letters
+
+**Number of Words**
+
+* Main Text:
+* Text boxes:
+
+**Number of References**
+
+**Number of Figures**
+
+**Corresponding Author**
+
+* Simon J Goring
+  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
+
+# Abstract
+
+Forest inventory data are essential for understanding tree species distributions along environmental gradients, and for projecting forest responses to climate change.
+However, the stability and representativeness of contemporary tree-climate distributions is poorly understood.
+Here we show that historic land use and climate change over the last two centuries have significantly altered tree-climate relationships, using contemporary and historical data for climates and 15 common tree genera in the northern US.
+Realized niches have shifted towards cooler maximum temperatures, higher minimum temperatures, and higher rainfall.
+A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.
+The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions.
+Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.
+
+/pagebreak
 
@@
 
-Introduction
-==============
+# Introduction
 
-Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States.
-There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].
+*The introduction should summarize briefly the background and aims, and end with a very brief statement of what has been achieved by the work.*
 
-This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].
-One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.
+Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space REFS.
+These distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
+Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+
+Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
+Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence].
+Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence].
+Habitat loss through human agency can impact apparent species distributions at timescales of thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche].
+In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.
+In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].
+
+Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [@gehrig2007tree] REFS.
+Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.
+Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].
+
+The interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat].
+Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.
+For example, for butterfly species in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving species upslope [@forister2010compounded].
+Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.
+The overall spatial and climatic distribution of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little over the past several centuries, but land-use conversion has excluded it from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental].
+Hence, predictive models for Garry oak [@pellatt2012using] using only contemporary data may be confounded by this range-wide interaction between human land use and soil characteristics.
+Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
+For example, bird species in the Sierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
 
@@
   geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
-  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
-  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
-  theme(legend.position = 'none')
+  coord_equal(xlim = ext[1:2], 
+              ylim = ext[3:4]) + 
+  scale_fill_manual(values = c("#ff7f00", "#e41a1c", "#33a02c")) +
+  theme(legend.position = "none")
 
-source('R/figures/climate_ts.R')
+source("R/figures/climate_ts.R")
 
@@
 
-ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')
+ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow = 1, widths = c(3, 4)), 
+       filename = "figures/figure_one.tiff")
 
 ```
-![figone](figures/figure_one.tiff)
+![figone](figures/Fig_1_Composite.png)
 
-**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.
+`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
+Low temperatures in the 19XXs result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.
 PDSI shows strong coherence with P~ann~ in this region.
-(f) Patterns of land use change in the upper Midwestern United States.
-Data from the NLCD [@jin2013comprehensive].*
+(f) Patterns of land-use change in the upper Midwestern United States.
+Land use data from the NLCD [@jin2013comprehensive].*")`
 
-The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].
-Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.
-By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.
+In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2015preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
+Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].
+There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).
+Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.
+In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.
+Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS.
+These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.
 
-Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.
-In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.
-This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].
-Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient.
-Confounding effects may increase the variability of predictive models, but do little to the overall envelope.
-Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.
-The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used.
-Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.
-Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.
+Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and assess the relative impacts of regional climate shifts and vegetation change in driving these changes.
+To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change.
+We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts.
+We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.
 
-In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.
+# Materials and Methods
+
+*This section should contain sufficient detail so that all procedures can be repeated (in conjunction with cited references).
+A checklist is provided so that authors can check that their methods report details which our editors regard as essential.
+This checklist is available here.
+Where specific equipment and materials are named, the manufacturer's name, city and country should be given (generally in parentheses after first mention).*
+
+## Historical and Modern Vegetation Data:
+
+`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km2)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.
+Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.
+The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.")`
+
+```{r, echo = FALSE, message = FALSE, warning=FALSE, results='asis'}
+
+area_table <- data.frame(Genera = colnames(pls_data)[-1],
+                         `PLS Coverage\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
+                                                    big.mark = ',', digits = 3)[-1],
+                         `FIA Coverage\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
+                                                    big.mark = ',', digits = 3)[-1])
+
+# Capitalize the taxa and sort them by PFT:
+area_table[,1] <- stringi::stri_trans_totitle(area_table[,1])
+area_table <- area_table[c(9,6,1,15,8,2,13,11,14,
+                           5,4,3,10,7,12),]
+
+rownames(area_table) <- NULL
+
+pander::pandoc.table(area_table, style = 'rmarkdown')
+
+```
+
+Settlement-era forest data are based on survey data from the Public Land Survey System [PLSS: @white1983history;@schulte2001original;@goring2015preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2015preset].
+We briefly review the methods here.
+PLSS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history].
+At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree.
 Goring et al.
-[-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical].
-Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.
-Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.
+[-@goring2015preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
+This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2015preset].
+Using the count data produced by Goring et al.
+[-@goring2015preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (Table `r tab_nums("genera_table", display = "cite")`).
+Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar).
+Here presence is defined as at least one recorded tree in any 8x8 km grid cell.
 
-Interaction between environmental gradients, land use patterns and taxa are likely to vary by region and environmental variable of interest.
-Species in the north may experience a confounding land use effect with respect to temperature, northward shifts due to warming temperatures may not have significantly manifested themselves, and land use conversion is likely to be heterogenous with respect to temperature since there is little agriculture in northern Wisconsin and northeastern Minnesota.
-Thus, with respect to temperature we expect to see no shift in distribution along temperature gradients.
+Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure].
+FIA datasets are publicly available surveys of forest composition (for all regions with > 10% forest cover) [@woudenberg2010forest].
+FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.
+Each Phase 2/3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
 
-With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change.
-Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.
+Comparisons between historical and modern vegetation must address differences in sampling design between the PLSS and FIA datasets [@goring2015preset;@kronenfeld2014validating;@kronenfeld2010influence].
+Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
+Nonetheless, differences persist.
+For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLSS data.
+This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence/absence patterns.
+Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLSS datasets (`r tab_nums("genera_table", display="cite")`).
+Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA.
+However the practice of using FIA data as a surrogate for tree presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.
+In prior comparative analyses of the PLSS and FIA data, Goring et al.
+[-@goring2015preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
 
-For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients.
-This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.
+## Historical and Modern Climate Data
 
-Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty.
-Tree species are responding to climate change within forested landscapes, but the response is inconsistent.
-Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure].
-Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns.
-For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.
+Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism].
+The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.
+Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLSS and FIA data using bilinear interpolation.
+Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
+There is an unavoidable temporal mismatch between the historical climate data and the PLSS survey period, on the order of 50 to 70 years.
+Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
 
-We use gridded climate data products (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States.
-By examining the interactions between climate, species distributions and land use change since EuroAmerican settlement we examine the possible implications of range shifts for future estimates of species resillience to climate change across the region.
+## Shifts in Species Distributions in Climate Space
 
-Methods
-===============
-We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.
-The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.
-At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.
-Goring et al.
-[-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.
-When examining relationships between FIA and PLS data, Goring et al.
-[-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.
+![modelgrid](figures/figureFour_panels.tiff)
 
-Another complication is the structure of the FIA dataset.
-While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].
-As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA.
-This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].
+`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships.
+We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
+(b-e) Representative normalized probability densities for Larix for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`
 
-The analysis presented here follows a blocked sampling method.
-We generate probability density functions along historical climate gradients for taxa within the PLSS and for the same PLSS taxa (and point locations) using climate data for the modern era.
-We do the same for FIA presence (climate at the PLSS and modern eras), and for anthropogenic land use classes.
-Climate data is derived from monthly PRISM data from 1895 - 2014.
-A 30-year climate normal was generated for modern July temperature ($T_{jul}$), January temperature ($T_{jan}$), the continentality (difference between $T_{jul}$ and $T_{jan}$, $T_{diff}$), and annual precipitation ($P_{ann}$).
+Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.
+Differences in distributions are quantified using Hellinger distances [@matthes2015benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively).
+The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) is determined by calculating the Hellinger distance between PLSS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of$\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.
 
-The choice of a thirty year normal for the PLS data is potentially problematic.
-The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist.
-Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period.
-As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.
-Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].
+To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.
+For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (land-use change) distributions are significantly different.
+If both changes are significant and of the same sign, then the shift is defined as a compounding change.
+If both are significant and are in opposite directions then the change is defined as counteracting.
+In the case that either or both changes are not significant then the change is defined as confounding.
 
-By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation ($PLSS_{veg}$-$PLSS_{clim}$, $PLSS_{veg}$-$FIA_{clim}$, $FIA_{veg}$-$PLSS_{clim}$, $FIA_{veg}$-$FIA_{clim}$) we can use Hellinger distance to estimate the effects of changing climate ($d_{H}$), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).
-In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).
-The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.
+Analysis in this paper is undertaken using R and R Markdown.
+All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).
+Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.
+Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.
 
-Results
----------------
+# Results
 
-Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.
-Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C).
-The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal.
-This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.
+*The Results section should present the experiments that support the conclusions to be drawn later in the Discussion.
+The Results section should conform to a high standard of rigour.
+Extended lines of inference, arguments or speculations should not be placed in the Results.*
 
-The Public Lands data is spatially extensive, sampling occured across the region in a regular pattern, while the FIA is limited to regions with forest cover.
-As such, the extent of points is not overlapping.
-Given the extensive use of FIA data in generating and estimating climate responses of tree taxa and forest types to changing climate, the use of the FIA data as an estimate for tree species distributions is acceptable fo estimating shifts in climate space, and attributing the loss or gain of climate niche space due to land use and climate variability and change during the 20^th^ century.
+Regionally, January (minimum) temperatures have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal `r fig_nums("land_use_clim", display = "cite")`, maximum annual temperature has declined (`r round(clim_change[2], 1)`^o^C), while mean annual temperatures increased by `r round(clim_change[3], 1)`^o^C (not shown).
+Because of the large shift in minimum temperatures, Tdiff has changed by .
+Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).
+The Palmer Drought Severity Index is closely coherent with Pann trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
 
@@
 
-loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)
+loss_table[,1:3] <- round(tran(loss_table[,1:3], method = "proportion") * 100,1)
 
-ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
-       dpi=300, width=6, height = 3)
+ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
+       dpi = 300, 
+       width = 6, 
+       height = 3)
 ```
@@
 ![figuretwo](figures/Figure_panelloss.tiff)
-**Figure 2.**.
-*Gain loss and continuous presence for PLSS and FIA tree taxa.
-Loss is assessed as cells that contained presence during the PLSS era, and no presence in the FIA data (light blue).
-Continuous presence is indicated by dark gray, where both FIA and PLSS cells indicate presence.
-Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*
 
-**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era.
-The first three columns represent percentages of the total area occupied by the taxon since settlement.
-Modern and Historical values represent 100s of km^2^ occupied by the tree taxon within either the FIA or PLSS data.*
+`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera.
+'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue).
+'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red).
+'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray).
+Losses are more common than gains, particularly in the southwestern portion of tree ranges.
+Few taxa show gains, although Populus, Fraxinus and Tilia show gains of over 10% (Table 2).")`
 
-```{r, results='as.is', echo=FALSE}
-pandoc.table(loss_table, justify = "left", style = 'simple')
+```{r echo = FALSE}
+tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era.
+The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras.
+Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.
+Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
 ```
 
+For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3).
+Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
+For Ulmus and Larix, however, losses occur throughout the range.
+Losses are highest for Ostrya/Carpinus and Larix when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
+Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for Populus (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
+Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
+Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.
+Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
+*Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.
+
+`r tab_nums("tab_gain_loss", display = "full")`
+
+```{r table1, results="asis", echo=FALSE, eval=TRUE}
+
+pandoc.table(loss_table, justify = "left", style = "rmarkdown")
+```
+
+Changes in spatial coverage and changes in regional climate result in overall changes in the climate envelopes occupied by genera in the upper Midwest (`r fig_nums("boxplots", display = "cite")`).
+
 ```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
-source('R/resampling_climate.R')
+source("R/resampling_climate.R")
 vegclim_table <- resample_climate(climate, agg_dens, pls_data)
 
-source('R/figures/climate_boxplots.R')
+source("R/figures/climate_boxplots.R")
 boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
@@
 ![FigureThree](figures/boxplot_clim.tiff)
-**Figure 3.** *Changes in the realized climate space for genera in the upper Midwestern United States.
-For forest taxa maximum temperatures decline across the region from the PLS era to the modern era.
-At the same time minimum temperatures and annual precipitation have both increased.*
 
-While these changes in climate are visible across the whole range, we would expect that species with significant shifts in distribution as a result of land use change in the post-settlement era should show changes that superceed the changes resulting from climate alone.
-In particular, if the land use pressure is biased spatially or climatically then these changes should be even more dramatic.
-Hellinger distance is used to understand the difference in the shape of kernel densities.
-Using the distributions in climate space for each of the key taxa we can shift climate, or vegetation to understand the relative effects of land use change and climatic change across the region.
+`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States.
+In the box-and-whisker plots, the central line indicates the median, box margins indicate 25th and 75th percentiles, whisker tips indicate 5th and 95th percentiles, and dots indicate outliers beyong the 95th percentile.
+These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`
+
+The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.
+For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
+Historical land use also plays a role in shifting distributions, particularly for annual maximum temperature where losses at the highest Tmax values are visible from pre-settlement to modern, however, land use appears toplay a smaller role for the other climate variables.
+
+Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).
+The directionality of shifts between historical and modern climate distributions are consistent across genera: modern maximum annual temperatures are lower for all genera, while minimum temperatures are higher (`r fig_nums("boxplot", display = "cite")`).
+Across all genera modern annual precipitation is also higher than in the past.
+*Larix* and *Ulmus* show the most extreme changes in climate space overall.
+*Ulmus* shows the largest increase in minimum temperature ($\delta_{tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*.
+Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{tmax}$ = -0.4 ^o^C).
+*Ulmus* shows the largest increase in mean precipitation ($\delta_{ppt}$ = 86mm), the smallest change is for *Larix* ($\delta_{ppt}$ = 42mm).
 
 ```{r panel_densities, warning=FALSE, message=FALSE, echo=FALSE}
-source('R/figures/FigureFour_Panels.R')
+source("R/figures/FigureFour_Panels.R")
 ```
-
-![modelgrid](figures/figureFour_panels.tiff)
-**Figure 4**.
-*Blocked method for analysis, with representative normalized densities for climate distributions.
-Data presented represent one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-A diagonal shift from the top left to bottom right represents the true change in forest-climate relationships.*
-
-Hellinger distances reveal that while taxa may show the imprint of land use change on shifting climatic niches, this shift is not uniform across taxa not climatic  variables.
-Maximum temperature shows the greatest change attributed to land use change.
-Tamarack, elm, poplar, spruce beech, cedar, and fir all show a greater influence for land use change than regional climate change.
-The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.
-Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.
 
@@
 
-**Figure 5**.
-*For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.
-The x-axis represents the total shift (diagonal in Figure 4) from PLSS - FIA for both vegetation and climate, while the y-axis represents the difference between $d_{h}(V_{PLSS}C_{PLSS},V_{PLSS}C_{FIA})$ and $d_{h}(V_{FIA}C_{FIA},V_{PLSS}C_{FIA})$.
-Taxa with negative y-axis values have changes in distributions along the climate gradient that are driven largely by land use, while positive y-axis values can be largely attributed to regional climate change.*
+`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historic climate change vs.
+changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historic land-use and pathogen outbreaks.
+Results are plotted separately for each of the four climate variables considered here.
+Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig.
+2).
+For $d_{C}-d_{V} (Y-axis), negative values indicate a stronger influence of historic land-use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.")`
 
-Since the pre-settlement era maximum temperatures have declined, while minimum temperatures have increased strongly.
+Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
+Shifts in realized climate distributions over the last 100 to 150 years are largest for Ulmus (TJan, TJul), Larix (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).
+Attribution analyses based on the d index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for Larix, Ulmus, and Fagus.
+The large shift in the distribution of Tsuga relative to July temperatures, conversely, appears to be mainly attributable to historic climate change.
+For minimum temperature (Figure 5X), shifting distributions of Ulmus and Cedar can be mainly attributed to land use change, while Larix and Betula have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
+The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only Populus and Picea have seen climate niche shifts due to land-use that are larger than the climate shifts.
 
-**Table 2**.
-*Compounding (+), confounding (.) and contradicting (-) relationships between land use and climate change.
-There is no relationship between spatial correlates and the land use/climate interaction.*
+`r tab_nums("compTable", "Compounding (+), confounding (.) and counteracting (-) relationships between land-use and climate change.
+When the land-use/climate interaction was regressed against spatial vaariables (latitude, longitude) no significant relationship was found.")`
 
-```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results='as-is'}
-source('R/confounding_table.R')
+```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results="asis", eval=TRUE}
+source("R/confounding_table.R")
 
-pandoc.table(conf_table, justify = "left", style = 'simple')
+rownames(conf_table) <- NULL
+
+pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
 
@@
 
-Discussion
-----------------
-This work adds to our understanding of the relationships between shifts in realized climate niches and the interacting effects of climate change and land use.
+The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show evidence of counteracting effects (`r tab_nums("compTable", display = "cite")`).
+Compounding interactions are particularly strong for Tjul (Table 3), and are consistent with the observed trend of lowered Tmin (Figure 1) andlosses concentrated in the souther portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
+These analyses are insenstitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of confounding relationships and increases counts in both counteracting and compounding interactions.
+Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\Chi^{2}}_4$ = 15.22), but are not significantly correlated for other climate variables.
 
-Land use is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000].
-Land use change reduces the correlational structure between species and climate [@devineau2011arid].
-While rates of land use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land use are likely to persist, both in reorganizing forested landscapes [@goring2015preset], and in the correlational structure between species and climate.
+For climate variables, the proportion of compounding interactions is greatest for $T_{max}$, and least for $T_{min}$, for which confounding interactions dominate.
+For $P_{ann}$ and $T_{diff}$ we see a general balance of compounding and confounding interactions.
+Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).
+
+Several genera show common patterns in their relationships to climate/land-use interactions.
+Poplar, spruce, oak, maple and birch all show compounding relationships for both $P_{ann}$ and $T_{max}$ and either compounding or confounding interactions for $T_{diff}$.
+All five taxa show confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).
+These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with little new distribution in those regions.
+
+Fraxinus, Tsuga and Tilia all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.
+Each of these taxa show losses throughout historical distribution, but without clear spatial structure (`r fig_nums("gain_loss", display="cite")`).
+No taxa show patterns that are identical to another genera.
+
+# Discussion
+
+*The Discussion section should be separate from the Results section.
+It allows authors to propose their interpretation of the results, and to suggest what they might mean in a wider context.
+It should end with a clear statement of the main conclusions of the research, and a clear explanation of their importance and relevance.*
+
+This work demonstrates that historic land use can substantially alter the realized climate niches of species and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
+While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2015preset], and the correlational structure between species and climate.
 Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013building].
+
+Shifts in realized climate space since Settlement may have important implications for the management of forest communities, and the fact that historical realized niches are broader than modern niches might lead us to be more optimistic than otherwise, however, projected climate change far exceeds the shifts due to historical land use and climate change.
+In addition, coupled migration and adaptation in response to future climate change is likely to be impeeded by the exitrpation of communities due to land use change.
+Species migration and genetic adaptation are coupled across time-scales [@davis2001range].
+The loss of forest communities along the forest-prairie boundary in the Upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.
+The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.
+Long distance gene flow of adaptive traits to central and leading populations from the lagging edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
+Thus net gene flow in populations along the remaining "lagging" edge will be from the central and leading populations, individuals maladapted to warmer conditions.
+This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].
 
@@
 
+The use of correlational species-distribution models may be heavily impacted by historical land-use change and the sampling design of modern survey data.
+Dorman et al.
+[@dormann2012correlation] propose a gradient of models, from correlative to process-based models for predicting species distributions.
+While process based models may provide more accurate predictions of species distributions given the loss of spatial coverage as a result of land-use conversion, spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population.
+The inability to acurately reflect phenotypic variability along gradients of adaptation could then lead to weaker process-based models.
+In this case, we might assume that both correlative and process based models would be affected by spatially biased land-use conversion, particularly at "lagging" margins of a species distribution.
+
+At a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale].
+Human agency in these marginal environments ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological], but may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
+Detailled sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species may provide broader coverage of a species' fundamental climate niche, but care must be taken to account for the intensive management that may be required to maintain species beyond their realized niche.
+
+This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future.
+We show that modern inventory data samples only a subset of historical distribution, both spatially and within multidimentsional climate space.
+Critically, we show that the combination of regional land-use change and climate change leads to the under-prediction of species resilliance to climate change regionally.
+Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate].
+Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates.
+Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the CMIP3 intercomparison [@wicci2011climate].
+Higher temperatures and lower precipitation means for taxa in the historic period indicate the potential for more robust responses to climate change than might be assessed using modern data only.
+
+Increasing forest fragmentation as a result of land use conversion overlies long term climate patterns.
+In regions where land use conversion is climatically biased, conversion can affect the realized niche of tree species, particularly when sampling of those species is biased by a survey design that is itself biased by land use converison.
+In the Upper Midwestern United States maximum temperatures have declined over the last century, but increases in minimum annual temperature, and land use conversion along the warm and dry forest-prairie margin have resulted in shifts that compound the apparent climatic shifts (`r tab_nums("compTable", display = "cite")`), resulting in a realized niche for most species that is less robust to future shifts in climate, toward warmer conditions with, potentially higher evapotranspiration.
+
+While historical datasets may add to our ability to understand species realized climate niches and their adaptive potential in the face of climate change they also serve to underscore the important interactions between climate change and land use change at a population level.
+The homogenization of midwestern forests evidenced through forest composition data [@goring2015preset;@rhemtulla2009legacies;@schulte2007homogenization;@hanberry2012comparison] may well be reflected in genetic data, although this is less well understood, and may be lost to us due to the broad-scale disturbance resulting from EuroAmerican Settlement in the region.
+Ultimately, land-use history is a critical component of modern ecology that is under-represented in the literature [although see: @hamburg1986disturbance;@compton2000long;@foster1998land].
+
 References

### 2016-06-17 Composition_Ecology.Rmd
Finished key text edits.  Will go through a printed version to check over tomorrow morning.  Added references, still missing full citation to @IceAgeEcologist & Burke paper.  Removed duplicate reference.
Added Early & Sax referecne & mention early in Introduction.

@@
 Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence].
-Habitat loss through human agency can impact apparent species distributions at timescales of thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche].
+Habitat loss through human agency can impact apparent species distributions at timescales of decades [@early2014climatic], thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche].
 In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.
@@
 We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-(b-e) Representative normalized probability densities for Larix for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`
+(b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`
 
@@
 Losses are more common than gains, particularly in the southwestern portion of tree ranges.
-Few taxa show gains, although Populus, Fraxinus and Tilia show gains of over 10% (Table 2).")`
+Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`
 
@@
 Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
-For Ulmus and Larix, however, losses occur throughout the range.
-Losses are highest for Ostrya/Carpinus and Larix when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
-Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for Populus (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
+For *Ulmus* and *Larix*, however, losses occur throughout the range.
+Losses are highest for Ostrya/Carpinus and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
+Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed fo *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
 Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
@@
 Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
-Shifts in realized climate distributions over the last 100 to 150 years are largest for Ulmus (TJan, TJul), Larix (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).
+Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* (TJan, TJul), *Larix* (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).
 Attribution analyses based on the d index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
-Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for Larix, Ulmus, and Fagus.
+Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for *Larix*, *Ulmus*, and Fagus.
 The large shift in the distribution of Tsuga relative to July temperatures, conversely, appears to be mainly attributable to historic climate change.
-For minimum temperature (Figure 5X), shifting distributions of Ulmus and Cedar can be mainly attributed to land use change, while Larix and Betula have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
-The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only Populus and Picea have seen climate niche shifts due to land-use that are larger than the climate shifts.
+For minimum temperature (Figure 5X), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
+The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.
 
@@
 While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2015preset], and the correlational structure between species and climate.
-Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013building].
+This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future [@moritz2013future;@maguire2015modeling].
+Modern surveys sample only a subset of historical distribution, both spatially and within multidimensional climate space.
 
-Shifts in realized climate space since Settlement may have important implications for the management of forest communities, and the fact that historical realized niches are broader than modern niches might lead us to be more optimistic than otherwise, however, projected climate change far exceeds the shifts due to historical land use and climate change.
-In addition, coupled migration and adaptation in response to future climate change is likely to be impeeded by the exitrpation of communities due to land use change.
+Previous studies have shown that the realized climate niches of tree species also shifted during the climate changes accompanying the last deglaciation.
+To a first order, many species accommodated past climate changes by shifting their ranges; i.e.
+they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
+These past range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
+However, species tracking of past climatic change was imperfect and other studies indicates shifts in species realized over time [@maiorano2013building;@pearman2008niche].
+Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], species occupying non-analogue climates [@veloz2012no], or shifts in competition resulting from changing patterns of co-occurrence through time [@maiorano2013building].
+This study adds to this prior work by showing that the large land use changes of the last several centuries can also significantly modify species-climate relationships.
+
+In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.
+First, climates have changed over the 20th and early 21st centuries, with rising winter minimum temperatures, declining summer maximum temperatures, and increasing precipitation.
+Second, tree species distributions have not yet fully adjusted to these climate changes, causing the environmental niches to shift.
+Third, anthropogenic land use has selectively occupied some portions of geographic and environmental space, making these regions less available to tree genera.
+Because much of the land-use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009historical], the effects of climate change and land use are compounding.
+Northern species show shifts away from drier habitat, while land-use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.
+For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to temperature and precipitation gradients.
+
+Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.
+In particular, the PLS survey design resulted in more survey points across the landscape than in the FIA dataset, but fewer trees surveyed per point.
+The FIA is more confined to forested and semi-forested areas because plot surveys are limited to plots with > 10% tree cover [@woudenberg2010forest].
+Hence, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
+Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
+Hence, any incompleteness in the representation of realized climate niches based on FIA data may result in inaccurate model projections.
+
+On the other hand, anthropogenic agency may be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale].
+For example, human agency causes urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].
+Human agency may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
+Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].
+However, usage of species distributional data from highly managed ecosystems must account for the intensive management that may be required to maintain species beyond their realized niche.
+
+These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
+On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change.
+Critically, the combination of regional land-use change and climate change based on contemporary data alone leads to the under-prediction of species resilience to climate change regionally.
+Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate].
+Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates.
+Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the Climate Model Intercomparison Project [@wicci2011climate].
+Higher temperatures and lower precipitation means for taxa for the pre-settlement period indicate the potential for more resilience to climate change than might be assessed using modern data only.
+This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., Tsuga, Fagus), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.
+
+However, projected climate change far exceeds the shifts due to historical land use and climate change.
+In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity.
 Species migration and genetic adaptation are coupled across time-scales [@davis2001range].
-The loss of forest communities along the forest-prairie boundary in the Upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.
+The loss of forest communities along the forest-prairie boundary in the upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.
 The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.
@@
 
-From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States.
-However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].
-In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013building;@early2014climatic].
+Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by historical land-use change and the sampling design of modern survey data.
+However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant].
+Spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population.
+In this case, both correlative and process based models might be affected by spatially biased land-use conversion.
 
-The use of correlational species-distribution models may be heavily impacted by historical land-use change and the sampling design of modern survey data.
-Dorman et al.
-[@dormann2012correlation] propose a gradient of models, from correlative to process-based models for predicting species distributions.
-While process based models may provide more accurate predictions of species distributions given the loss of spatial coverage as a result of land-use conversion, spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population.
-The inability to acurately reflect phenotypic variability along gradients of adaptation could then lead to weaker process-based models.
-In this case, we might assume that both correlative and process based models would be affected by spatially biased land-use conversion, particularly at "lagging" margins of a species distribution.
-
-At a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale].
-Human agency in these marginal environments ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological], but may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
-Detailled sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species may provide broader coverage of a species' fundamental climate niche, but care must be taken to account for the intensive management that may be required to maintain species beyond their realized niche.
-
-This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future.
-We show that modern inventory data samples only a subset of historical distribution, both spatially and within multidimentsional climate space.
-Critically, we show that the combination of regional land-use change and climate change leads to the under-prediction of species resilliance to climate change regionally.
-Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate].
-Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates.
-Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the CMIP3 intercomparison [@wicci2011climate].
-Higher temperatures and lower precipitation means for taxa in the historic period indicate the potential for more robust responses to climate change than might be assessed using modern data only.
-
-Increasing forest fragmentation as a result of land use conversion overlies long term climate patterns.
-In regions where land use conversion is climatically biased, conversion can affect the realized niche of tree species, particularly when sampling of those species is biased by a survey design that is itself biased by land use converison.
-In the Upper Midwestern United States maximum temperatures have declined over the last century, but increases in minimum annual temperature, and land use conversion along the warm and dry forest-prairie margin have resulted in shifts that compound the apparent climatic shifts (`r tab_nums("compTable", display = "cite")`), resulting in a realized niche for most species that is less robust to future shifts in climate, toward warmer conditions with, potentially higher evapotranspiration.
-
-While historical datasets may add to our ability to understand species realized climate niches and their adaptive potential in the face of climate change they also serve to underscore the important interactions between climate change and land use change at a population level.
-The homogenization of midwestern forests evidenced through forest composition data [@goring2015preset;@rhemtulla2009legacies;@schulte2007homogenization;@hanberry2012comparison] may well be reflected in genetic data, although this is less well understood, and may be lost to us due to the broad-scale disturbance resulting from EuroAmerican Settlement in the region.
-Ultimately, land-use history is a critical component of modern ecology that is under-represented in the literature [although see: @hamburg1986disturbance;@compton2000long;@foster1998land].
+In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand species realized climate niches and their adaptive potential in the face of climate change.
+They also serve to underscore the important interactions between climate change and land use change at a population level.
+The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
+Given the importance of modern distributional datasets for predicting species range shifts, compounding may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.
+At the same time, confounding interactions between historic climate change and land-use effects may increase predictive uncertainty.
 

### 2016-06-17 Composition_Ecology.Rmd
Bib and csl in yaml header moved.

@@
 ---
+bibliography: goringetal_references.bib
+csl: ecology.csl
 output:
@@
     variant: markdown_github
-csl: ecology.csl
 pandoc_args: -H margins.sty
-bibliography: goringetal_references.bib
 ---

### 2016-06-19 Composition_Ecology.Rmd
Most text changes complete, minor typos & reformatting.

@@
 A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.
-The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions.
+The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer and wetter distributions across the historical period.
 Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.
@@
 
-Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space REFS.
-These distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
-Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space [@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
+Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
+Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
@@
 
-Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [@gehrig2007tree] REFS.
+Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree].
 Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.
@@
 `r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Low temperatures in the 19XXs result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.
-PDSI shows strong coherence with P~ann~ in this region.
+Regionally low temperatures in the 1990s result in lower $T_{max}$ during the modern era, but climate change has resulted in higher $T_{min}$ values than the early-century normals, along with increasing $P_{ann}$.
+PDSI shows strong coherence with $P_{ann}$ in this region.
 (f) Patterns of land-use change in the upper Midwestern United States.
@@
 
-In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2015preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].
+In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@gorng2016preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
+Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization].
 There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).
@@
 
-Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and assess the relative impacts of regional climate shifts and vegetation change in driving these changes.
-To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we assess the relative impacts of regional climate shifts and vegetation change in driving these changes.
+To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change.
@@
 
-`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km2)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.
+`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km^2^)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.
 Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.
@@
 area_table <- data.frame(Genera = colnames(pls_data)[-1],
-                         `PLS Coverage\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
+                         `PLS\\sCoverage\\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
                                                     big.mark = ',', digits = 3)[-1],
-                         `FIA Coverage\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
+                         `FIA\\sCoverage\\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
                                                     big.mark = ',', digits = 3)[-1])
+
+colnames(area_table)[2] <- "PLS\\sCoverage\\nkm^2^"
+colnames(area_table)[3] <- "FIA\\sCoverage\\nkm^2^"
 
@@
 
-Settlement-era forest data are based on survey data from the Public Land Survey System [PLSS: @white1983history;@schulte2001original;@goring2015preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2015preset].
+Settlement-era forest data are based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset].
 We briefly review the methods here.
-PLSS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history].
+PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history].
 At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree.
 Goring et al.
-[-@goring2015preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
-This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2015preset].
+[-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
+This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset].
 Using the count data produced by Goring et al.
-[-@goring2015preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (Table `r tab_nums("genera_table", display = "cite")`).
+[-@gorng2016preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`).
 Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar).
@@
 Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure].
-FIA datasets are publicly available surveys of forest composition (for all regions with > 10% forest cover) [@woudenberg2010forest].
+FIA datasets are publicly available surveys of forest composition across the United States [@woudenberg2010forest].
 FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.
-Each Phase 2/3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
+Each Phase 2 and 3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
 
-Comparisons between historical and modern vegetation must address differences in sampling design between the PLSS and FIA datasets [@goring2015preset;@kronenfeld2014validating;@kronenfeld2010influence].
+Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
 Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
 Nonetheless, differences persist.
-For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLSS data.
-This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence/absence patterns.
-Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLSS datasets (`r tab_nums("genera_table", display="cite")`).
+For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data.
+This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns.
+Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`).
 Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA.
-However the practice of using FIA data as a surrogate for tree presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.
-In prior comparative analyses of the PLSS and FIA data, Goring et al.
-[-@goring2015preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
+However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.
+In prior comparative analyses of the PLS and FIA data, Goring et al.
+[-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
 
@@
 
-Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism].
+Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004].
 The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.
-Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLSS and FIA data using bilinear interpolation.
+Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation.
 Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
-There is an unavoidable temporal mismatch between the historical climate data and the PLSS survey period, on the order of 50 to 70 years.
-Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
+There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.
+Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period, although recent methodological advances [@tiptoninreview] may provide estimates from oberver datasets as far back as 1825.
 To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
@@
 
-![modelgrid](figures/figureFour_panels.tiff)
+![](figures/figureFour_panels.tiff)
 
@@
 We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-(b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`
+(b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`
 
 Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.
-Differences in distributions are quantified using Hellinger distances [@matthes2015benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively).
+Differences in distributions are quantified using Hellinger distances [@matthes2016benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively).
 The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) is determined by calculating the Hellinger distance between PLSS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of$\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.
+The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.
 
@@
 
-Analysis in this paper is undertaken using R and R Markdown.
+Analysis in this paper is undertaken using R [@Team2014] and R Markdown [@rmarkdown2016].
 All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).
@@
 
-Regionally, January (minimum) temperatures have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal `r fig_nums("land_use_clim", display = "cite")`, maximum annual temperature has declined (`r round(clim_change[2], 1)`^o^C), while mean annual temperatures increased by `r round(clim_change[3], 1)`^o^C (not shown).
-Because of the large shift in minimum temperatures, Tdiff has changed by .
+Regionally, $T_{min}$ has increased by `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal `r fig_nums("land_use_clim", display = "cite")`, $T_{max}$ has declined (`r round(clim_change[2], 1)`^o^C).
+Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C.
 Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).
-The Palmer Drought Severity Index is closely coherent with Pann trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
+The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
 
@@
 
-![figuretwo](figures/Figure_panelloss.tiff)
+![](figures/Figure_panelloss.tiff)
 
 `r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera.
-'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue).
-'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red).
-'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray).
+'Loss' indicates 8x8km grid cells where PLS data indicated presence of a genus but FIA data do not record presence (light blue).
+'Gain' indicates areas where a genus is absent in the PLS data but present in the FIA data (red).
+'Continuous presence' represents locations where both FIA and PLS data indicate presence (dark gray).
 Losses are more common than gains, particularly in the southwestern portion of tree ranges.
@@
 The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras.
-Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.
+Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.
 Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
@@
 For *Ulmus* and *Larix*, however, losses occur throughout the range.
-Losses are highest for Ostrya/Carpinus and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
-Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed fo *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
+Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
+Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
 Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
@@
 
-![FigureThree](figures/boxplot_clim.tiff)
+![](figures/boxplot_clim.tiff)
 
 `r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States.
-In the box-and-whisker plots, the central line indicates the median, box margins indicate 25th and 75th percentiles, whisker tips indicate 5th and 95th percentiles, and dots indicate outliers beyong the 95th percentile.
+In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyong the 95^th^ percentile.
 These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`
@@
 For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
-Historical land use also plays a role in shifting distributions, particularly for annual maximum temperature where losses at the highest Tmax values are visible from pre-settlement to modern, however, land use appears toplay a smaller role for the other climate variables.
+Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern, however, land use appears to play a smaller role for the other climate variables.
 
 Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).
-The directionality of shifts between historical and modern climate distributions are consistent across genera: modern maximum annual temperatures are lower for all genera, while minimum temperatures are higher (`r fig_nums("boxplot", display = "cite")`).
+The directionality of shifts between historical and modern climate distributions are consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
 Across all genera modern annual precipitation is also higher than in the past.
 *Larix* and *Ulmus* show the most extreme changes in climate space overall.
-*Ulmus* shows the largest increase in minimum temperature ($\delta_{tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*.
-Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{tmax}$ = -0.4 ^o^C).
-*Ulmus* shows the largest increase in mean precipitation ($\delta_{ppt}$ = 86mm), the smallest change is for *Larix* ($\delta_{ppt}$ = 42mm).
+*Ulmus* shows the largest increase in minimum temperature ($\delta_{Tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*.
+Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
+*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).
 
@@
 ```
-![hellplot](figures/hellingerplot.tiff)
+![](figures/hellingerplot.tiff)
 
@@
 2).
-For $d_{C}-d_{V} (Y-axis), negative values indicate a stronger influence of historic land-use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.")`
+For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historic land-use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.")`
 
 Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
-Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* (TJan, TJul), *Larix* (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).
-Attribution analyses based on the d index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
-Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for *Larix*, *Ulmus*, and Fagus.
-The large shift in the distribution of Tsuga relative to July temperatures, conversely, appears to be mainly attributable to historic climate change.
-For minimum temperature (Figure 5X), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
-The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.
+Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
+Attribution analyses based on the $d$ index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus*.
+The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change.
+For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
+The climatic signal is strongest for $P_{ann}$, with regional changes being largely responsible for shifting climate niches (`r fig_nums("hellPlot", display = "cite")`), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.
 
@@
 The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show evidence of counteracting effects (`r tab_nums("compTable", display = "cite")`).
-Compounding interactions are particularly strong for Tjul (Table 3), and are consistent with the observed trend of lowered Tmin (Figure 1) andlosses concentrated in the souther portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
+Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the souther portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
 These analyses are insenstitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of confounding relationships and increases counts in both counteracting and compounding interactions.
-Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\Chi^{2}}_4$ = 15.22), but are not significantly correlated for other climate variables.
+Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\Chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
 
@@
 Several genera show common patterns in their relationships to climate/land-use interactions.
-Poplar, spruce, oak, maple and birch all show compounding relationships for both $P_{ann}$ and $T_{max}$ and either compounding or confounding interactions for $T_{diff}$.
+*Populus*, *Picea*, *Quercus*, *Acer* and *Betula* all show compounding relationships for both $P_{ann}$ and $T_{max}$ and either compounding or confounding interactions for $T_{diff}$.
 All five taxa show confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).
@@
 
-Fraxinus, Tsuga and Tilia all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.
+*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.
 Each of these taxa show losses throughout historical distribution, but without clear spatial structure (`r fig_nums("gain_loss", display="cite")`).
@@
 Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
-While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2015preset], and the correlational structure between species and climate.
+While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@gorng2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
 This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future [@moritz2013future;@maguire2015modeling].
@@
 Previous studies have shown that the realized climate niches of tree species also shifted during the climate changes accompanying the last deglaciation.
-To a first order, many species accommodated past climate changes by shifting their ranges; i.e.
-they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
+To a first order, many species accommodated past climate changes by shifting their ranges; *i.e.* they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
 These past range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
@@
 In particular, the PLS survey design resulted in more survey points across the landscape than in the FIA dataset, but fewer trees surveyed per point.
-The FIA is more confined to forested and semi-forested areas because plot surveys are limited to plots with > 10% tree cover [@woudenberg2010forest].
+The FIA is confined to forested and semi-forested areas because surveys are limited to plots with > 10% tree cover [@woudenberg2010forest].
 Hence, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
@@
 Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the Climate Model Intercomparison Project [@wicci2011climate].
-Higher temperatures and lower precipitation means for taxa for the pre-settlement period indicate the potential for more resilience to climate change than might be assessed using modern data only.
-This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., Tsuga, Fagus), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.
+Higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.
+This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., *Tsuga*, *Fagus*), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.
 

### 2016-08-19 Composition_Ecology.Rmd
Working through text changes.  Added a new template file for Word so that the style changes are less problematic.

@@
 output:
-  word_document: null
+  word_document: 
+    reference_docx: word_template.docx  
   pdf_document:
@@
 
-# Effect of historic land-use and climate change on tree-climate relationships in the northern United States
+# Effect of historic land-use, pathogens and climate change on tree-climate relationships in the northern United States
 
@@
 
-* Simon J Goring: Conceived paper, performed analysis, developed figures and wrote the paper
-* John "Jack" W Williams: Wrote the paper
+* Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
+* John "Jack" W Williams: Co-conceived paper, and co-wrote the paper
 
@@
 
-* All data for this manuscript, along with code and version-control is available at http://github.com/SimonGoring/CompositionClimate.
-In addition, an archived copy of the published version is available with the following DOI.
+* All data for this manuscript, along with code and version-control are available at http://github.com/SimonGoring/CompositionClimate.
+In addition, an archived copy of the submitted version is available with the following DOI: XXXX
 
@@
   * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
-* Jack W Williams
+* John W Williams
   * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
@@
 
-* climate change, land use, historical ecology, realized niche, anthropocene, public land survey, forest inventory and analysis
+* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land use, niche shift, Public Land Survey System, realized niche
 
@@
 
-* Land use change forces climate shifts in trees
+* Causes of recent climate shifts in trees
 
-**Article Type**
-
-* Letters
+**Article Type**: Letters
 
@@
 
-**Number of Figures**
+**Number of Figures, Tables & Text Boxes**: 6 (4 figures, 2 tables)
 
@@
 * Simon J Goring
-  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
+  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706, goring@wisc.edu
 
@@
 
-Forest inventory data are essential for understanding tree species distributions along environmental gradients, and for projecting forest responses to climate change.
-However, the stability and representativeness of contemporary tree-climate distributions is poorly understood.
-Here we show that historic land use and climate change over the last two centuries have significantly altered tree-climate relationships, using contemporary and historical data for climates and 15 common tree genera in the northern US.
-Realized niches have shifted towards cooler maximum temperatures, higher minimum temperatures, and higher rainfall.
-A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.
-The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer and wetter distributions across the historical period.
-Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.
+(143 words)
+
+Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
+Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historic land use, pathogen outbreaks, and climate change.
+Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
+A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables.
+Most climate/land-use interactions are compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions.
+The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.
 

### 2016-08-19 Composition_Ecology.Rmd
Added the figures folder to the `gitignore`.
Text changes up through the Introduction to the paper.

Three new references for the `bib` file.

@@
 
-*The introduction should summarize briefly the background and aims, and end with a very brief statement of what has been achieved by the work.*
-
-Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space [@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
+Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change  [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
 Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
-Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, been used, for example, to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
 Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
-Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence].
-Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence].
-Habitat loss through human agency can impact apparent species distributions at timescales of decades [@early2014climatic], thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche].
+Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of anthropogenic land use, and other historical legacies on species distributions [@svenning2015influence].
+Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
 In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.
 In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].
+In contrast t the several studies examining the stability of the climatic niche at timescales of millenia and longer, the effect of 19^th^ and 20^th^ century climate change and land use on species-climate relationships remains largely unknown and understudied.
 
-Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree].
-Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.
+Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic; @sarmento2013impacts; @ivory2016niche].
+Anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.
 Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
-This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].
+This biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].
 
@@
 Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.
-For example, for butterfly species in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving species upslope [@forister2010compounded].
+For example, in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving butterfly species upslope [@forister2010compounded].
 Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.
@@
 Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
-For example, bird species in the Sierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
+For example, bird species in the Sierra Nevadas are expected to be pulled downslope by increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
 
@@
 ```
-![figone](figures/Fig_1_Composite.png)
+![](figures/Fig_1_Composite.png)
 
@@
 
-In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@gorng2016preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization].
-There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).
-Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.
-In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.
-Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS.
-These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.
+In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
+Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization].
+Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
+In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].
+Additionally, formerly dominant tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut].
+These vegetation changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.
 
-Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we assess the relative impacts of regional climate shifts and vegetation change in driving these changes.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.
 To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
-We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change.
+We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
+We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change.
 We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts.
-We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.
+We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.
 

### 2016-08-22 Composition_Ecology.Rmd
Started workon the methods but got distracted.
Adding a `tif` and a figure output.

@@
 
-*This section should contain sufficient detail so that all procedures can be repeated (in conjunction with cited references).
-A checklist is provided so that authors can check that their methods report details which our editors regard as essential.
-This checklist is available here.
-Where specific equipment and materials are named, the manufacturer's name, city and country should be given (generally in parentheses after first mention).*
-
 ## Historical and Modern Vegetation Data:
@@
 
-Settlement-era forest data are based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset].
+Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset].
 We briefly review the methods here.

### 2016-08-22 Composition_Ecology.Rmd
Changes up to the end of the Methods.  Added reference to Mann *et al*.

@@
 PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history].
-At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree.
+At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree.
 Goring et al.
@@
 This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset].
-Using the count data produced by Goring et al.
-[-@gorng2016preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`).
-Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar).
-Here presence is defined as at least one recorded tree in any 8x8 km grid cell.
+Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`).
+A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar.
+'Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell.
 
-Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure].
-FIA datasets are publicly available surveys of forest composition across the United States [@woudenberg2010forest].
-FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.
-Each Phase 2 and 3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
+Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provides detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure], for all regions with >10% forest cover [@woudenberg2010forest].
+FIA uses a nationally standardized sampling procedure, that has evolved over time.
+In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
 
 Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
-Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
+Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
 Nonetheless, differences persist.
@@
 There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.
-Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period, although recent methodological advances [@tiptoninreview] may provide estimates from oberver datasets as far back as 1825.
-To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
+This temporal mismatch is unavoidable; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
+The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against tree-ring-based paleoclimate reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
 
@@
 
-Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.
-Differences in distributions are quantified using Hellinger distances [@matthes2016benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively).
-The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+Distributions of tree genera presences in climate space are described as univariate density functions (PDFs) for both the pre-settlement and modern eras.
+Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] to compare PDFs among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions).
+The total shift ($d_{tot}$) between pre-settlement and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
 The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) betweem the PLSS tree distributions when overlaid on historical vs.
+modern climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; Figure 2a; top row).
 We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.
@@
 To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.
-For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (land-use change) distributions are significantly different.
-If both changes are significant and of the same sign, then the shift is defined as a compounding change.
-If both are significant and are in opposite directions then the change is defined as counteracting.
-In the case that either or both changes are not significant then the change is defined as confounding.
+For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.
+If both changes are significant and of the same sign, then the shift is defined as compounding.
+If both are significant and are in opposite directions then the shift is counteracting.
+In the case that either or both changes are insignificant, then the change is defined as confounding.
 
-Analysis in this paper is undertaken using R [@Team2014] and R Markdown [@rmarkdown2016].
-All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).
-Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.
+All analysis makes use of R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`.
+All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate).
 Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.
@@
 # Results
-
-*The Results section should present the experiments that support the conclusions to be drawn later in the Discussion.
-The Results section should conform to a high standard of rigour.
-Extended lines of inference, arguments or speculations should not be placed in the Results.*
 

### 2016-08-23 Composition_Ecology.Rmd
Revisions up through the Results now.  Only the Discussion & references to work through.  Some figure work too, likely.

@@
 
-Regionally, $T_{min}$ has increased by `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal `r fig_nums("land_use_clim", display = "cite")`, $T_{max}$ has declined (`r round(clim_change[2], 1)`^o^C).
+Regionally, $T_{Jan}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{Jul}$ declined `r round(clim_change[2], 1)`^o^C.
 Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C.
-Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).
-The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
+Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.
+The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
 
@@
 tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era.
-The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras.
+The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
 Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.
@@
 
-For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3).
-Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
+For most tree genera, historic range losses dominate over range gains and are spatially structured (Table 1, Figure 3).
+Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
 For *Ulmus* and *Larix*, however, losses occur throughout the range.
-Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
+Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
 Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
@@
 Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
-*Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.
+*Pinus*, *Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.
 
@@
 ```
-
-Changes in spatial coverage and changes in regional climate result in overall changes in the climate envelopes occupied by genera in the upper Midwest (`r fig_nums("boxplots", display = "cite")`).
 
@@
 The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.
-For *Larix* we see clear shifts in climate space (from solid to dashed lines) whether modern or presettlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
-Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern, however, land use appears to play a smaller role for the other climate variables.
+For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
+Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig.
+2d).
+However, for *Larix*, land use overall appears to play a smaller role than climate change.
 
 Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).
-The directionality of shifts between historical and modern climate distributions are consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
-Across all genera modern annual precipitation is also higher than in the past.
-*Larix* and *Ulmus* show the most extreme changes in climate space overall.
-*Ulmus* shows the largest increase in minimum temperature ($\delta_{Tmin}$ = 4.2 ^o^C) and the smallest increase occurs for *Larix*.
-Maximum annual temperature declines most for *Ostrya/Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
-*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).
+The directionality of shifts between historical and modern climate distributions is consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
+Across all genera, modern $P_{ann}$ is higher than in the past.
+*Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while the smallest increase occurs for *Larix*.
+Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
+*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), while the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).
 
@@
 
-Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
+Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on the $d$ index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+Attribution analyses based on $d$ (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches (), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
 The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus*.
 The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change.
-For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.
-The climatic signal is strongest for $P_{ann}$, with regional changes being largely responsible for shifting climate niches (`r fig_nums("hellPlot", display = "cite")`), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.
+For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.
 
@@
 
-The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show evidence of counteracting effects (`r tab_nums("compTable", display = "cite")`).
-Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the souther portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
-These analyses are insenstitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of confounding relationships and increases counts in both counteracting and compounding interactions.
-Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\Chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
-
-For climate variables, the proportion of compounding interactions is greatest for $T_{max}$, and least for $T_{min}$, for which confounding interactions dominate.
-For $P_{ann}$ and $T_{diff}$ we see a general balance of compounding and confounding interactions.
-Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).
+The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).
+Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the southern portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
+Compounding interactions are also strong for $P_{ann}$ and are caused by both land use and climate change shifting species to a wetter portion of climate space than they occupied historically.
+These analyses are insensitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of insignificant (confounding) relationships and increases the number of  counteracting and compounding interactions.
+Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
 
 Several genera show common patterns in their relationships to climate/land-use interactions.
-*Populus*, *Picea*, *Quercus*, *Acer* and *Betula* all show compounding relationships for both $P_{ann}$ and $T_{max}$ and either compounding or confounding interactions for $T_{diff}$.
-All five taxa show confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).
-These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with little new distribution in those regions.
+*Populus*, *Picea*, *Quercus*, *Acer* and *Betula* all show compounding relationships for both $P_{ann}$ and $T_{max}$, either compounding or confounding interactions for $T_{diff}$.
+and confounding interactions for $T_{min}$ (`r tab_nums("compTable", display = "cite")`).
+These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with few gains in those regions.
 
@@
 No taxa show patterns that are identical to another genera.
+Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).
 

### 2016-08-23 Composition_Ecology.Rmd
Edits for the Results, all the way through into the Discussion.

@@
 
-*The Discussion section should be separate from the Results section.
-It allows authors to propose their interpretation of the results, and to suggest what they might mean in a wider context.
-It should end with a clear statement of the main conclusions of the research, and a clear explanation of their importance and relevance.*
-
-This work demonstrates that historic land use can substantially alter the realized climate niches of species and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
+Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.
+These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate chnage over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
 Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
 While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@gorng2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
-This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future [@moritz2013future;@maguire2015modeling].
-Modern surveys sample only a subset of historical distribution, both spatially and within multidimensional climate space.
 

### 2016-08-24 Composition_Ecology.Rmd
Changes through the Results and into the Discussion.  Fixed `R/Setup_R.R` to move the nldc raster into `data/input`. This will allow complete reproduction of the paper from a git clone.
Removed the `gitignore` from the `data/albers/density` folder.

@@
 In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization].
+Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
 Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
@@
 Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.
-To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
@@
 
-Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@gorng2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@gorng2016preset].
+Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2016preset].
 We briefly review the methods here.
@@
 Goring et al.
-[-@gorng2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
-This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@gorng2016preset].
+[-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
+This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset].
 Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`).
@@
 
-Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@gorng2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
+Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
 Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
@@
 In prior comparative analyses of the PLS and FIA data, Goring et al.
-[-@gorng2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
+[-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
 
@@
 Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
-While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@gorng2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
+While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
 

### 2016-08-24 R/Setup_R.R

@@ -25,9 +25,7 @@ base_rast <- setValues(base_rast, 1:ncell(base_rast))
 pls_rast <- raster("data/input//dens_v0.9-4.tif")
 unit_raster <- setValues(pls_rast, 1:ncell(pls_rast))
 
-land_use <- raster(paste0("../../../Maps/",
-                          "nlcd_2011_landcover_2011_edition",
-                          "_2014_10_10//nlcd_cropped_repro.tif"))
+land_use <- raster("data/input/nlcd_cropped_repro.tif")
 
 #  We need to change the raster to a data.frame:
 land_use <- data.frame(cell = extract(unit_raster,

### 2016-08-24 Composition_Ecology.Rmd
Pushing latest `docx` build of the paper, with edits through the results.  Commiting Rmd, it's slightly ahead of the `docx`.

@@
 First, climates have changed over the 20th and early 21st centuries, with rising winter minimum temperatures, declining summer maximum temperatures, and increasing precipitation.
-Second, tree species distributions have not yet fully adjusted to these climate changes, causing the environmental niches to shift.
-Third, anthropogenic land use has selectively occupied some portions of geographic and environmental space, making these regions less available to tree genera.
+These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@stocker2013ipcc;@swanson2009long;@estrada2013statistically].
+Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress] and environmental niches to shift.
+Third, anthropogenic land use has selectively occupied some portions of geographic and environmental space, making these regions less available to tree species.
 Because much of the land-use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009historical], the effects of climate change and land use are compounding.
-Northern species show shifts away from drier habitat, while land-use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.
-For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to temperature and precipitation gradients.
+Northern species show shifts away from drier and warmer habitat.
+Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient, to apparent wetter climates.
+For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to and in the same direction as temperature and precipitation gradients.
 
@@
 Hence, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
+If so, estimates of forest loss (Table 1) and niche shifts (Figures 4 & 5) may be overestimated.
 Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
-Hence, any incompleteness in the representation of realized climate niches based on FIA data may result in inaccurate model projections.
+Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect model projections based on FIA data.
 

### 2016-08-24 Composition_Ecology.Rmd
Cleaned up the references.  Finished substantial text edits.

@@
 These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
-On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change.
-Critically, the combination of regional land-use change and climate change based on contemporary data alone leads to the under-prediction of species resilience to climate change regionally.
-Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate].
-Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates.
-Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the Climate Model Intercomparison Project [@wicci2011climate].
-Higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.
-This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., *Tsuga*, *Fagus*), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.
+On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
+Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate] and $P_{ann}$ projections range from -17 to 110mm by 2055 relative to 1980s baselines, based on fourteen global climate models from the Climate Model Intercomparison Project [@wicci2011climate].
+Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates.
+These higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.
+This evidence for historical resilience to warmer and dryer climates may be particularly critical for species that reach their northern or western range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).
 
-However, projected climate change far exceeds the shifts due to historical land use and climate change.
+However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.
 In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity.
-Species migration and genetic adaptation are coupled across time-scales [@davis2001range].
-The loss of forest communities along the forest-prairie boundary in the upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.
-The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.
-Long distance gene flow of adaptive traits to central and leading populations from the lagging edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
-Thus net gene flow in populations along the remaining "lagging" edge will be from the central and leading populations, individuals maladapted to warmer conditions.
+Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation].
+Hence, the loss of genotypes adapted to warmer and drier conditions (the trailing edge of distributions for many taxa in the Upper Midwest) caused by land use may mean more rapid losses at the trailing edges of the species range.
+Long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
+Thus net gene flow in populations along the remaining trailing edge will be from the central and leading populations, individuals maladapted to warmer conditions.
 This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].
 
-Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by historical land-use change and the sampling design of modern survey data.
+Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by the use of contemporary observational datasets that are biased by historical climate and land-use change and climatic disequilibrium.
 However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant].
-Spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population.
-In this case, both correlative and process based models might be affected by spatially biased land-use conversion.
+In this case, both correlative and process based models might be affected by the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.
 
-In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand species realized climate niches and their adaptive potential in the face of climate change.
-They also serve to underscore the important interactions between climate change and land use change at a population level.
+In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
 The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
-Given the importance of modern distributional datasets for predicting species range shifts, compounding may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.
-At the same time, confounding interactions between historic climate change and land-use effects may increase predictive uncertainty.
+Given the importance of modern distributional datasets for predicting species range shifts, compounding interactions between historical climate and land use change may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.
+Both empirical and process-based vegetation models need to combine, whenever possible, both contemporary and historical information on tree distributions, tree-climate relationships, and their respective shifts through time.
+
+Acknowledgements
+---------------------
+The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
+This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.
+Maps were made using data from Natural Earth (http://naturalearth.com), which provides raster & vector map data in the public domain.
 

### 2016-08-31 Composition_Ecology.Rmd
Minor edits based on comments from @andydawson (thanks!).  Now working on shaving off 1000 words.  THough these monstly look like major changes, they generally represent minor changes at the line-level.

@@
 
-* Main Text:
-* Text boxes:
+* Main Text: 6070
+* Text boxes: 0
 
-**Number of References**
+**Number of References**: 77
 
@@
 
-Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change  [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
+Contemporary distributions of species, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
 Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
-Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, been used, for example, to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
 Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
-Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of anthropogenic land use, and other historical legacies on species distributions [@svenning2015influence].
+Projections that rely only on contemporary observational data use a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], may incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
 Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
 In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.
-In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].
-In contrast t the several studies examining the stability of the climatic niche at timescales of millenia and longer, the effect of 19^th^ and 20^th^ century climate change and land use on species-climate relationships remains largely unknown and understudied.
+In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close;@matthes2016benchmarking].
 
-Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic; @sarmento2013impacts; @ivory2016niche].
-Anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.
-Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
+Anthropogenic land use is highly structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
 This biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].
 
-The interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat].
-Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.
-For example, in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving butterfly species upslope [@forister2010compounded].
-Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.
-The overall spatial and climatic distribution of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little over the past several centuries, but land-use conversion has excluded it from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental].
-Hence, predictive models for Garry oak [@pellatt2012using] using only contemporary data may be confounded by this range-wide interaction between human land use and soil characteristics.
+Interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat].
+Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction alog an environmental gradient.
+For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, compounding the upslope distributional shifts for butterfly species [@forister2010compounded].
+Confounding effects may cause little overall change on species distributions, with land-use pressure occuring randomly across the species' environmental gradient.
+predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but does not broadly shift distributions in climate space.
 Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
@@
 `r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Regionally low temperatures in the 1990s result in lower $T_{max}$ during the modern era, but climate change has resulted in higher $T_{min}$ values than the early-century normals, along with increasing $P_{ann}$.
+Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.
 PDSI shows strong coherence with $P_{ann}$ in this region.
-(f) Patterns of land-use change in the upper Midwestern United States.
+(f) Patterns of land-use change in the upper Midwestern United States, with the "Tension Zone", differentiating sub-boreal from southern broadleafed forests superimposed.
 Land use data from the NLCD [@jin2013comprehensive].*")`
 
-In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
-Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
+Tree distributions and community composition in the upper Midwestern US have been heavily influenced by the land use conversion associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
+Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
+Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
 In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].
-Additionally, formerly dominant tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut].
-These vegetation changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.
+Formerly dominant tree species, *Ulmus americana* and *Castanea dentata* (at the very south-eastern edge of the domain) have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut].
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.
-To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and land use changes on these shifts.
+We use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
@@
 
-`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km^2^)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.
-Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.
-The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.")`
-
-```{r, echo = FALSE, message = FALSE, warning=FALSE, results='asis'}
-
-area_table <- data.frame(Genera = colnames(pls_data)[-1],
-                         `PLS\\sCoverage\\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
-                                                    big.mark = ',', digits = 3)[-1],
-                         `FIA\\sCoverage\\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
-                                                    big.mark = ',', digits = 3)[-1])
-
-colnames(area_table)[2] <- "PLS\\sCoverage\\nkm^2^"
-colnames(area_table)[3] <- "FIA\\sCoverage\\nkm^2^"
-
-# Capitalize the taxa and sort them by PFT:
-area_table[,1] <- stringi::stri_trans_totitle(area_table[,1])
-area_table <- area_table[c(9,6,1,15,8,2,13,11,14,
-                           5,4,3,10,7,12),]
-
-rownames(area_table) <- NULL
-
-pander::pandoc.table(area_table, style = 'rmarkdown')
-
-```
-
-Estimates of settlement-era forest composition and tree distribution based on survey data from the Public Land Survey System [PLS: @white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2016preset].
+Estimates of settlement-era forest composition and tree distribution based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset].
 We briefly review the methods here.
@@
 A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar.
-'Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell.
+Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.
 
@@
 The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) betweem the PLSS tree distributions when overlaid on historical vs.
+Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) between the PLSS tree distributions when overlaid on historical vs.
 modern climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; Figure 2a; top row).
@@
 
-loss_table[,1:3] <- round(tran(loss_table[,1:3], method = "proportion") * 100,1)
+loss_table[,1:3] <- round(analogue::tran(loss_table[,1:3], method = "proportion") * 100,1)
 
-ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
+ggplot2::ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
        dpi = 300, 
@@
 ```{r echo = FALSE}
-tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era.
-The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
+tab_nums("tab_gain_loss", "Gain and loss (%) of the analyzed tree genera since the pre-settlement era, based on presence or absence within each 8x8km grid cell.
+Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.
+The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.
+Modern and historical occurence are reported, followed by percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
 Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.
 Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
+
 ```
@@
 
-Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5).
+Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on $d$ (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
-The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches (), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
-The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus*.
+Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
+The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
 The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change.
@@
 
-`r tab_nums("compTable", "Compounding (+), confounding (.) and counteracting (-) relationships between land-use and climate change.
+`r tab_nums("compTable", "Compounding (■), confounding (-) and counteracting (•) relationships between land-use and climate change.
 When the land-use/climate interaction was regressed against spatial vaariables (latitude, longitude) no significant relationship was found.")`
@@
 The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).
-Compounding interactions are particularly strong for $T_{max}$ (Table 3), and are consistent with the observed trend of lowered $T_{min}$ (Figure 1) and losses concentrated in the southern portions of species ranges (Figure 2); both tend to shift tree-climate distributions to cooler portions of climate space.
+Compounding interactions, where both climate and land use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space.
 Compounding interactions are also strong for $P_{ann}$ and are caused by both land use and climate change shifting species to a wetter portion of climate space than they occupied historically.
@@
 Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
+
+Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and
 
@@
 Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.
-These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate chnage over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate change over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
 Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
-While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
+While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse;@rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
 
@@
 For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to and in the same direction as temperature and precipitation gradients.
+
+In addition to land use changes, the near total regional collapse of chestnut and elm distributions provides another case in which human agency (in this case, introduced pathogens) has impacted distributions through near extirpation [@anagnostakis1987chestnut;@barnes1979stem], although some reintroduction efforts are underway [e.g., @jacobs2007toward].
+These cases do not align neatly with the land use/climate attribution index, since the decline in regional presence due to mortality, is assigned to land use change.
+Regardless, the extirpation was paired with regional logging, both regionally and elsewhere [@elliott2008long].
+While we might question whether these changes can be attributed to "land use", they do fall under a broader set of anthropogenic changes.
 
@@
 Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by the use of contemporary observational datasets that are biased by historical climate and land-use change and climatic disequilibrium.
-However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant].
+However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.* @wullschleger2014plant].
 In this case, both correlative and process based models might be affected by the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.

### 2016-08-31 R/confounding_table.R
Added axis labels to the Hellinger plot for clarity.  Trying to change symbols in the confounding table so we don't get confused between - and + and actual warmer & cooler directional changes.

@@ -38,15 +38,15 @@ checker <- function(x) {
   if (p.adjust(as.numeric(x[4]), method = "bonferroni", n = 120) < 0.05 & 
       p.adjust(as.numeric(x[6]), method = "bonferroni", n = 120) < 0.05) {
     if ((as.numeric(x[3]) * as.numeric(x[5])) > 0) {
-      return('+')
+      return('■')
     } else {
-      return('-')
+      return('•')
     }
   } else {
     if (as.numeric(x[4]) > (0.05 / 120)) {
       return('n/a')
     } else {
-      return('.')
+      return('-')
     }
   }
 }
@@ -70,9 +70,9 @@ conf_table$mean_x <- sapply(taxa, mean_x)
 conf_table <- conf_table[,1:5]
 
 conf_cor <- conf_table[,2:5]
-conf_cor[conf_cor == '+'] <- 1
-conf_cor[conf_cor == '.'] <- 0
-conf_cor[conf_cor == '-'] <- -1
+conf_cor[conf_cor == '■'] <- 1
+conf_cor[conf_cor == '-'] <- 0
+conf_cor[conf_cor == '•'] <- -1
 
 rownames(conf_cor) <- conf_table[,1]
 cluster <- hclust(dist(conf_cor))

### 2016-08-31 R/hellinger_dists.R

@@ -63,8 +63,8 @@ hellinger_plot <- function(){
     theme_bw() +
     scale_x_sqrt() +
     #scale_y_sqrt() +
-    xlab('Total Change - PLS - FIA') +
-    ylab('Climate Change - Land Use Change') +
+    xlab('Total Change - PLS - FIA - d_{tot}') +
+    ylab('Climate Change - Land Use Change - d_{c} - d_{v}') +
     theme(axis.text = element_text(family = 'serif', size = 12),
           axis.title = element_text(family = 'serif', face = 'bold', size = 14),
           strip.text = element_text(family = 'serif', face='bold', size = 14),

### 2016-09-27 Composition_Ecology.Rmd
Changed caching to `TRUE`, minor text edits to make the paper formatting fit *Ecology Letters* style guide.

@@
   eval = TRUE,
-  cache = FALSE,
+  cache = TRUE,
   tidy = TRUE
@@
 
-# Effect of historic land-use, pathogens and climate change on tree-climate relationships in the northern United States
+# Effect of historic land-use and climate change on tree-climate relationships in the northern United States
 
@@
 * Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
-* John "Jack" W Williams: Co-conceived paper, and co-wrote the paper
+* John W Williams: Co-conceived paper, and co-wrote the paper
 
@@
 
-* Simon J Goring
-  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
-* John W Williams
-  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
-  * Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706
+* Simon J Goring, Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
+* John W Williams, Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706; Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706
 
@@
 
-* Main Text: 6070
+* Main Text: 4900
 * Text boxes: 0
@@
 
-**Number of Figures, Tables & Text Boxes**: 6 (4 figures, 2 tables)
+**Number of Figures, Tables & Text Boxes**: 6 (5 figures, 1 table)
 
@@
 A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables.
-Most climate/land-use interactions are compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions.
+Most climate/land-use interactions are compounding, in which historic land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions.
 The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.
@@
 
-Contemporary distributions of species, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
-Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
+Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
+These distributions of species in geographic and environmental space, combined with future climate scenarios, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
 Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
-Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
-Projections that rely only on contemporary observational data use a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], may incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
+Increasingly, the predictive capacity of ecological models is being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
+Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
 Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
-In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.
-In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close;@matthes2016benchmarking].
+In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.
+In practice, however, the parameterization of these models is poorly constrained for processes operating at large temporal and spatial scales [@moorcroft2006close;@matthes2016benchmarking].
 
-Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
-Anthropogenic land use is highly structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
-This biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].
+Historic land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
+Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].
 
-Interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat].
-Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction alog an environmental gradient.
-For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, compounding the upslope distributional shifts for butterfly species [@forister2010compounded].
-Confounding effects may cause little overall change on species distributions, with land-use pressure occuring randomly across the species' environmental gradient.
-predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but does not broadly shift distributions in climate space.
+Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat].
+Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.
+For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded].
+Confounding effects may cause little overall change on species distributions, with land-use pressure occurring across the species' environmental gradient.
+Predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.
 Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
-For example, bird species in the Sierra Nevadas are expected to be pulled downslope by increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
+For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
 
@@
 PDSI shows strong coherence with $P_{ann}$ in this region.
-(f) Patterns of land-use change in the upper Midwestern United States, with the "Tension Zone", differentiating sub-boreal from southern broadleafed forests superimposed.
+(f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed.
 Land use data from the NLCD [@jin2013comprehensive].*")`
 
-Tree distributions and community composition in the upper Midwestern US have been heavily influenced by the land use conversion associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
+Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
+Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
 Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
 In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].
-Formerly dominant tree species, *Ulmus americana* and *Castanea dentata* (at the very south-eastern edge of the domain) have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut].
+Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.)
 
@@
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
-We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change.
+We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historic vegetation and climate change.
 We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts.
@@
 
-Estimates of settlement-era forest composition and tree distribution based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset].
+Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset].
 We briefly review the methods here.
-PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history].
+PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history].
 At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree.
@@
 This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset].
-Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (`r tab_nums("genera_table", display = "cite")`).
-A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus* to Ironwood, and *Thuja* and *Juniperus* to Cedar.
-Here 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.
+Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera.
+A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*.
+'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.
 
-Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provides detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure], for all regions with >10% forest cover [@woudenberg2010forest].
-FIA uses a nationally standardized sampling procedure, that has evolved over time.
+Modern forest data are from the US FIA program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest].
+The FIA uses a nationally standardized sampling procedure that has evolved over time.
 In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
@@
 Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
-Here the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
+Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
 Nonetheless, differences persist.
-For example, the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data.
-This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence-absence patterns.
-Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLS datasets (`r tab_nums("genera_table", display="cite")`).
-Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA.
-However the practice of using FIA data as a surrogate for tree presence-absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.
+The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets.
+Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA.
+However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general.
 In prior comparative analyses of the PLS and FIA data, Goring et al.
@@
 Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004].
-The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.
-Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLS and FIA data using bilinear interpolation.
-Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
-There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years.
-This temporal mismatch is unavoidable; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
+The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.
+Estimates for July temperature ($T_{max}$), January temperature ($T_{min}$), annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data.
+Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
+There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
 The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
-To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against tree-ring-based paleoclimate reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
 
@@
 
-`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships.
-We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
+`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on observed shifts in tree-climate relationships.
+Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
 (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`
 
-Distributions of tree genera presences in climate space are described as univariate density functions (PDFs) for both the pre-settlement and modern eras.
-Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] to compare PDFs among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions).
-The total shift ($d_{tot}$) between pre-settlement and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) between PLS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-Conversely, the effect of historical vegetation change on tree-climate distributions is determined by calculating the Hellinger distance ($d_{l}$) between the PLSS tree distributions when overlaid on historical vs.
-modern climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; Figure 2a; top row).
-We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.
+Distributions of tree genera within climate space are described as univariate density functions (PDFs) for the pre-settlement and modern eras.
+Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions).
+The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+The effect of historical climate change on shifting climate distributions of tree genera is $d_{C}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The effect of historical land use change on tree-climate distributions is $d_{l}$, the difference between the PLSS tree distributions when overlaid on historical vs.
+modern climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; Figure 2a; top row).
 
-To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.
+We develop an attribution index, ($\delta_{d} = d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.
+
+To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.
 For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.
-If both changes are significant and of the same sign, then the shift is defined as compounding.
-If both are significant and are in opposite directions then the shift is counteracting.
-In the case that either or both changes are insignificant, then the change is defined as confounding.
+If both changes are significant and of the same sign, then the shift is compounding.
+If both are significant and in opposite directions then the shift is counteracting.
+If either or both changes are not significant, then the change is defined as confounding.
 
-All analysis makes use of R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`.
+All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`.
 All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate).
-Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.
+Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.
 
@@
 
-Regionally, $T_{Jan}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{Jul}$ declined `r round(clim_change[2], 1)`^o^C.
-Because of the large shift in minimum temperatures, $T_{diff}$ decreased by `r round(clim_change[3], 1)`^o^C.
+Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C.
 Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.
-The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.
+The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid-19^th^ century climates.
 
@@
 Losses are more common than gains, particularly in the southwestern portion of tree ranges.
-Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`
+Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Supplemental Material).")`
 
@@
 For *Ulmus* and *Larix*, however, losses occur throughout the range.
-Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
-Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
+Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
+Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
 Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
@@
 Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
-*Pinus*, *Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.
+*Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.
 
@@
 `r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States.
-In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyong the 95^th^ percentile.
+In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile.
 These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`
 
-The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* in Figure 2.
+The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).
 For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
-Historical land use also plays a role in shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig.
+Historical land use contributes to shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig.
 2d).
-However, for *Larix*, land use overall appears to play a smaller role than climate change.
+However, for *Larix*, land use overall appears to exert a weaker influence than climate change.
 
 Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).
-The directionality of shifts between historical and modern climate distributions is consistent across genera: modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
+The directionality of shifts between historical and modern climate distributions is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
 Across all genera, modern $P_{ann}$ is higher than in the past.
@@
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa.
+Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historic land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa.
 The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
@@
 
-`r tab_nums("compTable", "Compounding (■), confounding (-) and counteracting (•) relationships between land-use and climate change.
+`r tab_nums("compTable", "Compounding ($\\Box$), confounding (-) and counteracting ($\\bigcirc$) relationships between land-use and climate change.
 When the land-use/climate interaction was regressed against spatial vaariables (latitude, longitude) no significant relationship was found.")`
@@
 
-Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and
+Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and *Larix* ($T_{min}$)
 
@@
 
-*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$.
-Each of these taxa show losses throughout historical distribution, but without clear spatial structure (`r fig_nums("gain_loss", display="cite")`).
-No taxa show patterns that are identical to another genera.
-Only *Ulmus* (for $T_{max}$) and *Larix* (for $T_{min}$) show counteracting changes for climate/land-use interactions (`r tab_nums("compTable", display = "cite")`).
+*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$ (`r tab_nums("compTable", display = "cite")`).
+Each of these taxa experiences range losses, but patterns of gain and loss differ among taxa.
 
@@
 
-This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
+This work reinforces the importance of historical datasets for examining relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
 Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.
-These analyses demonstrate that species-climate relationships have already begun to change, as a result of land use and climate change over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+Our analyses demonstrate that species-climate relationships are changing as a result of land use and climate change over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
 Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
@@
 
-Previous studies have shown that the realized climate niches of tree species also shifted during the climate changes accompanying the last deglaciation.
+Previous studies have shown that the realized climate niches of tree species shifted during the climate changes accompanying the last deglaciation.
 To a first order, many species accommodated past climate changes by shifting their ranges; *i.e.* they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
-These past range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
-However, species tracking of past climatic change was imperfect and other studies indicates shifts in species realized over time [@maiorano2013building;@pearman2008niche].
-Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], species occupying non-analogue climates [@veloz2012no], or shifts in competition resulting from changing patterns of co-occurrence through time [@maiorano2013building].
-This study adds to this prior work by showing that the large land use changes of the last several centuries can also significantly modify species-climate relationships.
+Range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
+However, species tracking of past climatic change was imperfect and other studies indicates shifts in species realized niches over time [@maiorano2013building;@pearman2008niche].
+Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], occupying non-analogue climates [@veloz2012no], or shifts in competition resulting from changing patterns of co-occurrence through time [@maiorano2013building].
+This study adds to prior work by showing that the land use changes of the last several centuries can also significantly modify species-climate relationships.
 
 In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.
-First, climates have changed over the 20th and early 21st centuries, with rising winter minimum temperatures, declining summer maximum temperatures, and increasing precipitation.
-These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@stocker2013ipcc;@swanson2009long;@estrada2013statistically].
-Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress] and environmental niches to shift.
-Third, anthropogenic land use has selectively occupied some portions of geographic and environmental space, making these regions less available to tree species.
-Because much of the land-use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009historical], the effects of climate change and land use are compounding.
-Northern species show shifts away from drier and warmer habitat.
-Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient, to apparent wetter climates.
-For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to and in the same direction as temperature and precipitation gradients.
+First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, declining summer temperatures, and higher precipitation.
+These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@estrada2013statistically].
+Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress].
+Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges.
+Agricultural conversion has largely eliminated open forests in warm and dry regions, while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times.
+Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient.
 
-In addition to land use changes, the near total regional collapse of chestnut and elm distributions provides another case in which human agency (in this case, introduced pathogens) has impacted distributions through near extirpation [@anagnostakis1987chestnut;@barnes1979stem], although some reintroduction efforts are underway [e.g., @jacobs2007toward].
-These cases do not align neatly with the land use/climate attribution index, since the decline in regional presence due to mortality, is assigned to land use change.
-Regardless, the extirpation was paired with regional logging, both regionally and elsewhere [@elliott2008long].
-While we might question whether these changes can be attributed to "land use", they do fall under a broader set of anthropogenic changes.
+In addition to land use changes, the near total regional collapse of *Ulmus* distributions provides another case in which human agency (in this case, accidentally introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem], although reintroduction efforts are underway.
+*Castanea dentata* was also effectively wiped out by chestnut blight, with reintroduction efforts underway, and *Tsuga canadensis* and *Fraxinus* spp.
+are now experiencing rapid mortality due to exotic pests.
+These pathogen-induced changes in tree distributions do not align neatly with the land use/climate attribution index described here, but they do fall under a broader set of vegetation changes caused by anthropogenic changesfactors, both inadvertent and intentional.
 
 Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.
-In particular, the PLS survey design resulted in more survey points across the landscape than in the FIA dataset, but fewer trees surveyed per point.
-The FIA is confined to forested and semi-forested areas because surveys are limited to plots with > 10% tree cover [@woudenberg2010forest].
-Hence, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
-If so, estimates of forest loss (Table 1) and niche shifts (Figures 4 & 5) may be overestimated.
+In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
+If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated.
 Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
-Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect model projections based on FIA data.
+Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
 
-On the other hand, anthropogenic agency may be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale].
-For example, human agency causes urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].
+Anthropogenic agency may also be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale;@groffman2014ecological].
 Human agency may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
@@
 
-These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
+Compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
 On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
@@
 However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.
-In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity.
+In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of adapted populations due to land use change and loss of genetic diversity.
 Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation].
@@
 The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
+In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group.
+The authors than Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript.
 This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.
-Maps were made using data from Natural Earth (http://naturalearth.com), which provides raster & vector map data in the public domain.
+Maps were made using data from Natural Earth (http://naturalearth.com).
 

### 2016-09-27 comp_ecol.R
Removed old files (both `RDS` data files & the clear code version).
Removed unused `yml` file.

@@ -1,94 +1,0 @@
-## ----setup-knitr, echo = FALSE, message = FALSE--------------------------
-knitr::opts_chunk$set(
-  comment = " ",
-  error = TRUE,
-  eval = TRUE,
-  cache = FALSE,
-  tidy = TRUE
-
-)
-
-library("pander")
-library("plyr")
-panderOptions('table.style', 'rmarkdown')
-
-## ---- echo = FALSE, warning = FALSE, message=FALSE, results='hide'-------
-# Load climate data and generate the pre-sett and modern bounds:
-#  Pre-settlement climate: 1900 - 1930
-#  Modern climate: 1960 - 1990
-
-source('R/Setup_R.R')
-source('R/load_climate.R') # gives `climate` with elements `then`, `now` and `names`
-source('R/load_veg.R')
-
-best.taxa <- c('beech', 'ironwood', 'hemlock', 'fir', 'cedar.juniper', 
-              'basswood', 'spruce', 'elm', 'ash', 'pine', 'oak', 
-              'tamarack', 'maple', 'poplar', 'birch')
-
-pls_data <- get_pls('0.9-4')
-agg_dens <- get_fia()[,colnames(pls_data)]
-
-land_use <- land_use[land_use$cell %in% c(pls_data$cell, agg_dens$cell),]
-
-source('R/natural_earth.R')
-
-natural <- natural_earth()
-  
-
-## ----figure1, echo = TRUE, warning = FALSE, message=FALSE----------------
-
-land_use$class[land_use$class == 'Water'] <- NA
-
-land_use_plot <- natural[[2]] + 
-  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
-  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c'))
-
-
-source('R/figures/climate_ts.R')
-
-climate_values <- climate_ts(unit.raster, pls_data, climate$now, climate$then)
-
-clim_change <- climate_values$modern$value - climate_values$past$value
-
-
-## ----figureTime, fig.width=6, fig.height=5-------------------------------
-grid.arrange(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4))
-
-clim_change[3]
-
-## ----loss_gain_fig, warning=FALSE, message=FALSE, echo=FALSE-------------
-
-source('R/figures/gain_loss_plot.R')
-
-#  This takes a bit of time.
-loss_melt <- loss_plot()
-
[…]

### 2016-09-27 R/confounding_table.R
* Changed symbols in `confounding_table.R` * `lintr` informed changes in `hellinger_dist`, changed taxon outputs to scientific names.  Changed plotting options. * Added `if` statement to `natural_earth` to stop the run if people try it, with informative stop methods. * Re-mapped taxon names to scientific in `climate_boxplots.R` and changed plotting options, also some `lintr` options. * Added figures.

@@ -38,9 +38,9 @@ checker <- function(x) {
   if (p.adjust(as.numeric(x[4]), method = "bonferroni", n = 120) < 0.05 & 
       p.adjust(as.numeric(x[6]), method = "bonferroni", n = 120) < 0.05) {
     if ((as.numeric(x[3]) * as.numeric(x[5])) > 0) {
-      return('■')
+      return('$\\Box$')
     } else {
-      return('•')
+      return('$\\bigcirc$')
     }
   } else {
     if (as.numeric(x[4]) > (0.05 / 120)) {
@@ -70,9 +70,9 @@ conf_table$mean_x <- sapply(taxa, mean_x)
 conf_table <- conf_table[,1:5]
 
 conf_cor <- conf_table[,2:5]
-conf_cor[conf_cor == '■'] <- 1
+conf_cor[conf_cor == '$\\Box$'] <- 1
 conf_cor[conf_cor == '-'] <- 0
-conf_cor[conf_cor == '•'] <- -1
+conf_cor[conf_cor == '$\\bigcirc$'] <- -1
 
 rownames(conf_cor) <- conf_table[,1]
 cluster <- hclust(dist(conf_cor))

### 2016-09-27 R/figures/climate_boxplots.R

@@ -1,24 +1,31 @@
 # Plotting the total climate-shift for the domain.
 clim_boxplots <- function(clim_table = vegclim_table){
   
+  clim_table$taxon <- mapvalues(clim_table$taxon, from = levels(clim_table$taxon),
+                                to = c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
+                                       "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
+                                       "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus"))
+  
   clim_table$climate <- factor(clim_table$climate, 
                              levels = c('tmax', 'tmean', 'tmin', 'ppt'),
                              labels = c('T[max]', 'T[mean]', 'T[min]', 'P[ann]'))
   
   boxplot_clim <- ggplot(data = subset(na.omit(clim_table), 
-                                       comb %in% c('FIA Era', 'PLS Era') & data>0 &
+                                       comb %in% c('FIA Era', 'PLS Era') & data > 0 &
                                          !taxon %in% 'Ironwood')) +
-    geom_boxplot(aes(y = clim, x = taxon, fill = comb)) +
+    geom_boxplot(aes(y = clim, x = taxon, fill = comb), lwd = 0.2, width = 1, outlier.size = 0.5) +
     scale_fill_grey(name = 'Climate Era') +
-    facet_grid(climate~., scales='free_y', labeller = label_parsed) +
+    facet_grid(climate~., scales = 'free_y', labeller = label_parsed) +
     theme_bw() +
-    theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
+    theme(axis.text = element_text(family = 'serif', face = 'bold', size = 12),
           axis.title = element_blank(),
-          strip.text = element_text(family = 'serif', face='bold', size = 14),
-          legend.title = element_text(family = 'serif', face='bold', size = 14),
+          strip.text = element_text(family = 'serif', face = 'bold', size = 14),
+          legend.title = element_text(family = 'serif', face = 'bold', size = 14),
           legend.text = element_text(family = 'serif', size = 12),
-          axis.text.x=element_text(angle=45, hjust = 1))
+          axis.text.x = element_text(angle = 45, hjust = 1, face = 'bold.italic'),
[…]

### 2016-09-27 R/hellinger_dists.R

@@ -3,24 +3,29 @@
 hellinger <- function(clim1, clim2){
   
   center <- mean(c(clim1, clim2), na.rm = TRUE)
-  ranger <- diff(range(c(clim1, clim2),na.rm=TRUE))
+  ranger <- diff(range(c(clim1, clim2),na.rm = TRUE))
   
   clim1 <- (clim1 - center) / ranger
   clim2 <- (clim2 - center) / ranger
   
-  clim.range <- range(c(clim1, clim2), na.rm=TRUE)
+  clim.range <- range(c(clim1, clim2), na.rm = TRUE)
   
-  kernel1 <- scale(density(clim1, na.rm=TRUE, 
+  kernel1 <- scale(density(clim1, na.rm = TRUE, 
                      from = clim.range[1], to = clim.range[2])$y)
-  kernel2 <- scale(density(clim2, na.rm=TRUE, 
+  kernel2 <- scale(density(clim2, na.rm = TRUE, 
                      from = clim.range[1], to = clim.range[2])$y)
   
   #  Calculate hellinger distance (between two kernels)
-  1/sqrt(2) * sqrt(sum(sqrt(kernel1 - min(kernel1)) - sqrt(kernel2-min(kernel2)))^2)
+  1/sqrt(2) * sqrt(sum(sqrt(kernel1 - min(kernel1)) - sqrt(kernel2 - min(kernel2))) ^ 2)
   
 }
 
 hellinger_plot <- function(){
+  
+  vegclim_table$taxon <- mapvalues(vegclim_table$taxon, from = levels(vegclim_table$taxon),
+                                to = c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
+                                       "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
+                                       "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus"))
   
   best.taxa <- as.character(unique(vegclim_table$taxon))
   
@@ -50,36 +55,32 @@ hellinger_plot <- function(){
                                      'Maximum Temperature',
                                      'Mean Temperature',
                                      'Minimum Temperature')
-  
-  levels(all.hellinger$taxon) <- c('Beech', 'Ironwood', 'Hemlock', 'Fir', 'Cedar',
-                                   'Basswood', 'Spruce', 'Elm', 'Ash', 'Pine',
-                                   'Oak', 'Tamarack', 'Maple', 'Poplar', 'Birch')
-  
[…]

### 2016-09-27 R/natural_earth.R

@@ -9,6 +9,11 @@ library(rgeos)
 #  Here we are using the 10m data.
 
 natural_earth <- function(){
+  
+  if(!"NaturalEarth" %in% list.files('data/input', include.dirs = TRUE)){
+    stop("NaturalEarth is needed to fully render this Rmd.  Please download - `ne_10m_coastline.zip`, `ne_10m_lakes`, `NE1_HR_LC_SR_W_DR` and `ne_10m_rivers_lake_centerlines`.")
+  }
+  
   nat.earth <- crop(stack('data/input/NaturalEarth/BaseRaster/NE1_HR_LC_SR_W_DR.tif'), y=extent(c(-98-1, -83+1, 42-1, 50+1)))
   
   xylimits <-  c(-200000, 1150000, 500000, 1700000)

### 2016-11-11 Composition_Ecology.Rmd
All changes form *historic* to *historical*.

@@
 
-# Effect of historic land-use and climate change on tree-climate relationships in the northern United States
+# Effect of historical land-use and climate change on tree-climate relationships in the northern United States
 
@@
 Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
-Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historic land use, pathogen outbreaks, and climate change.
+Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
-A new attribution method implicates both historic climate change and land use in these shifts, with the relative importance varying among genera and climate variables.
-Most climate/land-use interactions are compounding, in which historic land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions.
+A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables.
+Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions.
 The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.
@@
 
-Historic land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
+Historical land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
 Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
@@
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and land use changes on these shifts.
-We use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historical climate and land use changes on these shifts.
+We use gridded historical and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
-We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historic vegetation and climate change.
+We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change.
 We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts.
@@
 
-`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on observed shifts in tree-climate relationships.
+`r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships.
 Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
@@
 
-For most tree genera, historic range losses dominate over range gains and are spatially structured (Table 1, Figure 3).
+For most tree genera, historical range losses dominate over range gains and are spatially structured (Table 1, Figure 3).
 Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
@@
 Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
-*Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historic range, or beyond it.
+*Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historical range, or beyond it.
 
@@
 
-The joint effects of historic climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).
+The joint effects of historical climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).
 For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
@@
 
-`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historic climate change vs.
-changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historic land-use and pathogen outbreaks.
+`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs.
+changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historical land-use and pathogen outbreaks.
 Results are plotted separately for each of the four climate variables considered here.
@@
 2).
-For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historic land-use on shifting climate distributions, while positive values indicate a stronger influence of historic climate change on shifting distributions.")`
+For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
 
-Both the total shift in realized tree-climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
+Both the total shift in realized tree-climate distributions and the attribution of these changes to historical land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historic land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa.
+Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa.
 The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
 The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
-The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historic climate change.
-For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.
+The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change.
+For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.
 
@@
 
-Compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
+Compounding and confounding interactions between historical climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
 On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.

### 2016-11-11 Response_to_reviewers.Rmd
Added the **Response to Reviewers** document, `Response_to_reviewers.md`.

---
title: "Response to Reviewers"
author: "Simon Goring"
date: "November 11, 2016"
output: word_document
---

# Response to Reviewers

## Reviewer 1:

Reviewer 1 was generally positive.  I have pulled out the specific comment from the broad overview that addresses a broader limitation of the methods.  I also include line by line changes.

> My primary critique (which I admit is relatively minor) is that I found the logic revolving around the “confounding effects” and how the results relate to that a bit difficult. In particular, there likely needs to be a bit of clarification around the concept and a justification for why treating two non-significant effects and a set of mixed effects (one significant and one not so) as equivalent within this framework.

> Title: Here and throughout, replace “historic” with “historical” as the former refers to an exceptional event in the past while the latter refers to something in the past.

> Line 143-144: Shouldn’t miles be replaced with SI? I realize that the miles are in reference to the historical context of the PLS, but it might be good to keep things consistent throughout.

> Lines 174-176: This is a key point, but expanding a bit would be helpful. For example, did the paper in review examine issues of detectability? I realize this is addressed in the discussion, but a bit more detail would be good here.

> Line 179: Which PRISM data are the authors using, AN or LT. The LT data, for which one has to pay, uses only longer weather station records, which tends toward more reliable temporal trends. This seems like it would be important for the current study. Additionally, it would be good for the authors to at least acknowledge that increasingly sparse data early in the PRSIM record (i.e., late 19th and early 20th centuries) could contribute to greater uncertainties in historical vs. contemporary records. This might make historical climate relationships more difficult to identify.

> Line 195: Since the author’s use the abbreviation “PDF”, it would improve clarity to refer to “univariate probability density functions”.

> Line 207: Here and in a few other places, d_l is used, but later d_v is used. Please make symbology consistent throughout the paper.

> Lines 217-218: I found the labeling of lack of significance in “either or both changes” as confounding to be hard to connect with the definition provided at Line 93. For example, Fagus shows a strong overall change in distribution relative to precipitation, but is classes as confounding. Presumably, this is a situation where climate was important, but history was not. I suppose a bit more clarification regarding what is meant by confounding and why grouping taxa where neither climate nor history has an effect with taxa where only one has an effect is necessary.

> Lines 231-232: Looking at the maps, it seems that this statement is limited by the fact that the current study does not incorporate the northern range limit for many species. Can the authors address this limitation at some point?

> Lines 337-351: The authors discuss agriculture, but not timber harvesting and changes in fire regime in terms of the land-use drivers. It seems that the latter two are potentially important for some of the regional patterns. Additionally, I wonder if some of this could be discussed by looking at gains and losses relative to NLCD land-use layers. Perhaps that was in the supplement and I missed it.

> Figure 5: The x – axis needs more tick marks and numbers. At first, my mind assumed a linear scale, which would imply negative distances, but then I thought maybe this was some sort of log scale, which would make more sense with a distance metric. Some additional clarity is needed.

> Table 1: Some of the taxa are listed with their scientific names and others by common names. Replace common names with scientific names

## Referee: 2

Referee 2 was more critical of the paper than Reviewer 1.

> I think the paper does not adequately point out the limitations of the data sets used for the analyses and that the authors overstate the likely differences in realized niches between the historic and current time periods.

> The paper would be strengthened by stating more strongly the limitations of the methods and that the results may be over estimates and that data are not currently available to more accurately estimate the effect.

> Moreover, other data sets should be called upon to try to corroborate the range shifts and changes in climate tree relationships that emerge from the analyses.  Also the ecological mechanisms that might underlay the effects of land use and climate on tree species distributions should be more fully discussed.

More specific comments are below.

Line reference
2  “north-central” US would be more accurate? Also change this throughout the ms.

175-176 While that study found little effect of the differences between the PLS and FIA data sets, be sure to cover fully the potential effect of these differences on results of this study.  One has to presume that the differences in realized niches between historic and present would be substantially smaller if FIA was sampling low density and widespread occurrence of species as did the PLS.

181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.  It is important to do your analysis on climate variables that are known to be ecologically relevant to tree niche dimensions.

183-185   First, it is a serious stretch to use 1895-1919 climate to represent tree/climate relationships during 1830-1910.  The verbal statement (without statistical analysis) that the dendro-based PDSI did not change over that period is not very convincing.  Moreover, the analysis specifically should be done on the climate predictors used in the niche modeling.  Also, evaluate to what extent does this PDSI index represent trends in Tmin and the other predictors used?

Another issue is lack of discussion about the time period when the forests sampled by FIA got established. It is typical in SDM studies to test the strength of models from different time periods in tree SDM work and select the time period that is most strongly associated with species presence.  In the western US, we found 1950-1980 to produce stronger models than 1980-2010.

A third issue is the role of climate variability over the 1895 to 2014 time period.  Decadal oscillations are common in this period in many parts of the US.  Trees are likely integrating over 2 or more of these climate cycles.  SDMs should be done to best represent the climate and variability that selected for tree presence when the trees were measured.  Your analysis would be strengthened with statistical analysis of climate patters from 1895 to present for your predictor variables and an analysis of which time period produces the best SDMs.

195             It is important to discuss why univariate climate tree presence functions are relevant to the climate factors that strongly influence tree species presence or absence.  SDM modeling typically shows that a particular combination of climate and habitat factors is much more strongly associate with tree presence and any random climate variable.  Your analysis would be stronger to do model selection for best multivariate SDM functions and then compare these functions between historic and present.

211-218 The analyses should control for spatial autocorrelation, simple t tests would seem to have an inflated sample size due not including spatial autocorrelation in the model.

226-230 and Figure 1.  Statistical analysis should be done of the climate trends, showing a least a trend line and a 95% confidence interval.  A moving window regression might better show breakpoints in the trend line.  This is needed because the data show high interannual variability and it is not obvious by eyeing these graphics for which variables trends exist or not. Also, say in the legend what the red dots are.

231-239 I find the maps of range loss in Fig 3 to be stunningly large.  I am not sure anything in the literature would document such large losses in tree species presence from land use or 100 yrs of climate change.  It would seem important to use other data sets to validate these estimates of range loss.  Are the patterns in these maps real or mostly due to the differences among the PLS and FIA data sets?  If real, what are the mechanisms of local extinction?

347-350 Is it accurate to infer that ag has “largely eliminated open forests…” which does not sample low density forests? It seems likely that most of these tree species are present in at least low densities in the areas where your maps show they are not present.

Figure 2 b-e    The axes should be labeled and units indicated.  Consider reformatting these figures.  The current versions make visual separation of the lines hard to detect.

Lines 361-364.  The topic of this sentence should be elaborated upon to more fully point out the limitations of the study and appropriate inference. Say how much the estimates of forest loss and niche shifts may be overestimated.

365-370 These sentences are a bit of bait and switch.  Yes, readers will agree that SDM based on FIA may not fully represent the realized niche of tree species.  However, some of your argument hinges on the magnitude of the error that is made by using FIA.  I think your paper way over states that error.  The paper would have more impact if you were careful to emphasize the limitations of the data sets for this analysis and say that the degree to which FIA leads to SDM error is not knowable with available data sets.

## Referee 3

Referee 3 was most critical of the paper.

> Firstly, its rigor and clarity do not meet the standards for Ecology Letters. For example,

> Title: The authors used the upper Midwestern United States throughout the manuscript. But why the authors used the northern United States in the title?
This study equally emphasized the distribution changes for 15 common tree genera over the last two centuries and the effects of climate change and land use change on these changes. Was it better to say “the effects of historical land use and climate change on species distribution changes” in the tittle?

> Abstract: P2-L34-37, forest inventory data cannot be used to project forest responses to climate change. Rather, forest inventory data were used in models to make projections. It was distribution-climate relationships. In my opinion, this study was not about the stability and representativeness of distribution-climate. This study was to build up or extend the distribution-climate relationships.

> 3) One of the objectives in this manuscript was to investigate the effects of climate change, land use change, and pathogen outbreaks on distribution changes (L125-126). However, the pathogen outbreaks were nearly ignored in the title and methods, results, and discussion sections.

> Most importantly, the way that the authors included the effects of land use change in data analyses (let’s forget pathogen outbreaks for now) was not justified. Thus, I was not convinced the data analyses were robust to answer the question about the effects of climate change, land use change, and pathogen outbreaks on distribution changes. Although the authors argued the land use effects were represented as the difference between VHCH, VMCH. Why?

> Firstly, it is not possible to get the distribution data under VMCH. Secondly, the distribution changes over the last two centuries were caused by many factors including land use change, disturbances (e.g., pathogen outbreaks, fire, harvest), and endogenous succession processes. Why the difference VH and VM could represent the land use change effects? This was too arbitrary. A stronger rationale is needed.

> The authors should use present tense not past tense in the methods and results sections.

> The results section was not well organized. Subtitles should be used to describe the overall distribution changes, shifts along climate change, the relative importance, and interactions. The four paragraphs of interaction subsection in the results section were not concise and like bullets.

> Secondly, the materials and method section was vague and not logically articulated.
1)      July temperature (Tmax) was the minimum, maximum, or mean temperature in July?
January temperature (Tmin) was the minimum, maximum, or mean temperature in January?
Annual temperature range (Tdiff) was the difference between the minimum and maximum temperature over the year?
Annual precipitation (Pann) was the minimum, maximum, or mean precipitation across over the year?
2)      More detailed information in the materials and method section was needed to show how the authors analyzed the overall distribution changes, climate change effects, and the relative importance of land use and climate change. It should not be only in the figure titles. For example:
The authors used gain, loss, and continuous persistence to demonstrate the distribution changes over the last two centuries in Figure 3. However, this approach was not described in the materials and method section but was only in the title of Figure 3.
The authors used the box-and-whisker plots to demonstrate the distribution changes along Tmax, Tmin, and Pann. Likewise, this approach was not described in the materials and method section but was only in the title of Figure 4.
The approach used in Figure 5 was not described in the materials and method section but was only in the title of Figure 5. More detailed information should be given to better analyze the relative importance of land use change and climate change effects.
More detailed information about the univariate density functions (PDFs) and Hellinger distances was needed to better describe the approach.
3)      Again, how the effects land use and exotic pathogens were included in the analyses?
4)      The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a). The historical distribution data for VHCH scenario were derived from PLS and the contemporary distribution data for VMCM were derived from US FIA data. Then, what were the data for VHCM and VMCH scenarios. It was impossible to get the data for VHCM and VMCH scenarios, right? If my understanding was wrong, then the more detailed information was needed to better describe the factorial design and data used for each scenario.
5)      About the t-tests to assess the interaction effects between land use and climate change (e.g., compounding, confounding, counteracting), what the responsible variables were used for the comparisons between the VHCH and VHCM, VHCH and VMCH. More detailed information was needed to better describe these tests.

> Thirdly, the discussion section was not well written. Many contents in the discussion section ventured beyond the domain of the results. I suggest that the authors should set the results in a broader context and logically discuss the implications for distribution-climate relationships, distribution modeling, historical database, and management.
> 1) L316-L322 were not the findings of this study. How were these points related to the findings? They read like introduction and should not be in discussion section.
> 2) Likewise, L323-l336 how were these points related to your findings? It stood alone now.
> 3) L337-l351 should be better related to your findings. The first and second points stood alone now.
> 4) L352-L360 here, the authors finally mentioned the pathogen outbreaks effects. However, the pathogen outbreaks effect were not included in the data analyses. This seems just speculation.

Specific comments:

> 1) P2-L36, change “is poorly understood” to “are poorly understood”.
> 2) P2-L41, change “implicates” to “implicated”.
> 3) P2-L43, change “are compounding” and “reinforces” to “were compounding” and “reinforced”.
> 4) The figures were not well made. The axis titles in Figure 1 were two small. The four lines in the Figure 2 b-e were difficult to read. The genera names in Figure 5 were too crowd and hard to tell. It may be better to use shorter names instead.
> 5) All figure titles were too long. Many of these contents were approach and should be put in methods section.

Overall, this is an interesting topic. Especially, the interaction between land use change and climate change on distribution changes is the most interesting part, because the overall distribution changes have been extensively studied. However, its rigor and clarity (especially methods and discussion sections) do not meet the standard for Ecology Letters. I do not think it warrants publication in its present form.

******************************************
Editor
Editors Comments for the Author(s):
The reviews highlight the importance and novelty of the main idea - that past land-use complicates our estimates of forest community responses to climate change. However, they express concern about the way that the results are presented. There are many suggestions for clarification that should be addressed in revision. In particular, it's important to address the limitations of comparing the two data types, as this was a concern expressed by each reviewer.

### 2016-11-11 Composition_Ecology.Rmd
Changing imperial units to SI.

@@
 We briefly review the methods here.
-PLS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history].
+PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history].
 At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree.

### 2016-11-11 R/hellinger_dists.R
Hellinger plot output goes to `Final_Figures`, not somewhere else.

@@ -81,6 +81,8 @@ hellinger_plot <- function(){
   
   ggsave(plot = hell_plot, filename = 'figures/hellingerplot.pdf', 
          width = 6.7, height = 6, dpi = 300)
- 
+  ggsave(plot = hell_plot, filename = 'Final_Figures/hellingerplot.pdf', 
+         width = 6.7, height = 6, dpi = 300)
+  
   list(hell_plot, all.hellinger)
 }

### 2016-11-24 Response_to_reviewers.Rmd
Changes to title and spatial references (from northern to north-central).
Added statement about species ranges to qualify the region.

Added statement about the choice of climate data.

Standardized the use of $d_{v}$ vs $d_{l}$ in discussing the attribution index.

Explicit call to `pandoc.table`.

Added a period on line 278.

Added the point of *Tsuga* and the difference between land clearance and logging/fire.

Associated changes in `Response_to_reviewers.Rmd`.

@@
 
+We have changed all occurrences of *historic* to *historical*.  (commit: 4cf75f8b)
+
 > Line 143-144: Shouldn’t miles be replaced with SI? I realize that the miles are in reference to the historical context of the PLS, but it might be good to keep things consistent throughout.
+
+We have changed to *km*, retaining the reported miles in parentheses.
 
@@
 I realize this is addressed in the discussion, but a bit more detail would be good here.
+
+This paper is now accepted.
 
@@
 
+We use the PRISM LT data.  This has been clarified in the paper.
+
 > Line 195: Since the author’s use the abbreviation “PDF”, it would improve clarity to refer to “univariate probability density functions”.
 
+Changed.
+
 > Line 207: Here and in a few other places, d_l is used, but later d_v is used. Please make symbology consistent throughout the paper.
+
+We have changed this sentence in particular to read: *The difference between the PLSS and modern tree distributions ($d_{V}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest* and changed other references to $d_{l}$ to $d_{v}$.
 
@@
 
+We have added the paragraph:  *For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.* to indicate this limitation.
+
 > Lines 337-351: The authors discuss agriculture, but not timber harvesting and changes in fire regime in terms of the land-use drivers.
@@
 Perhaps that was in the supplement and I missed it.
+
+We have added the sentence: *While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of* Tsuga.
 
@@
 
+<!-- need to add a paragraph & add the curve from John's figure. -->
+
 > The paper would be strengthened by stating more strongly the limitations of the methods and that the results may be over estimates and that data are not currently available to more accurately estimate the effect.
 
-> Moreover, other data sets should be called upon to try to corroborate the range shifts and changes in climate tree relationships that emerge from the analyses.
+> Moreover, other data sets should be called upon to try to corroborate the range shifts and changes in climate-tree relationships that emerge from the analyses.
 Also the ecological mechanisms that might underlay the effects of land use and climate on tree species distributions should be more fully discussed.
+
+<!-- Pollen data, Jeannine & others. add in John's paper we can point to long term ecological datasets -->
 
@@
 Line reference
-2  “north-central” US would be more accurate? Also change this throughout the ms.
+> 2  "north-central" US would be more accurate? Also change this throughout the ms.
 
-175-176 While that study found little effect of the differences between the PLS and FIA data sets, be sure to cover fully the potential effect of these differences on results of this study.
+We have replaced instances of "northern" with north-central.
+
+> 175-176 While that study found little effect of the differences between the PLS and FIA data sets, be sure to cover fully the potential effect of these differences on results of this study.
 One has to presume that the differences in realized niches between historic and present would be substantially smaller if FIA was sampling low density and widespread occurrence of species as did the PLS.
 
-181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.
+<!-- The challenge is double edged.
+Area lost (definite) vs.
+frequency of occurrence.
+Also most of the tree species are common, frequency in the area lost, vs frequency in the core.
+Add to supplemental information.
+-->
+
+> 181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.
 It is important to do your analysis on climate variables that are known to be ecologically relevant to tree niche dimensions.
 
-183-185   First, it is a serious stretch to use 1895-1919 climate to represent tree/climate relationships during 1830-1910.
+We have added the sentences: *These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
+Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.*
+
+> 183-185   First, it is a serious stretch to use 1895-1919 climate to represent tree/climate relationships during 1830-1910.
 The verbal statement (without statistical analysis) that the dendro-based PDSI did not change over that period is not very convincing.
@@
 
-Another issue is lack of discussion about the time period when the forests sampled by FIA got established.
+> Another issue is lack of discussion about the time period when the forests sampled by FIA got established.
 It is typical in SDM studies to test the strength of models from different time periods in tree SDM work and select the time period that is most strongly associated with species presence.
@@
 
-A third issue is the role of climate variability over the 1895 to 2014 time period.
+<!-- We could play with SDMs and run them back and forth with windows. -->
+
+> A third issue is the role of climate variability over the 1895 to 2014 time period.
 Decadal oscillations are common in this period in many parts of the US.
@@
 
-195             It is important to discuss why univariate climate tree presence functions are relevant to the climate factors that strongly influence tree species presence or absence.
+<!-- We intentionally spaced the records. Species distribution models are not part of this discussion, we recognize that different time periods might affect this. -->
+
+> 195 It is important to discuss why univariate climate tree presence functions are relevant to the climate factors that strongly influence tree species presence or absence.
 SDM modeling typically shows that a particular combination of climate and habitat factors is much more strongly associate with tree presence and any random climate variable.
@@
 
-211-218 The analyses should control for spatial autocorrelation, simple t tests would seem to have an inflated sample size due not including spatial autocorrelation in the model.
+<!-- We are not building predictive models, with SDMs you get the blend, but it may not be clear which elements are directly impacting the model response. -->
 
-226-230 and Figure 1.
+> 211-218 The analyses should control for spatial autocorrelation, simple t-tests would seem to have an inflated sample size due not including spatial autocorrelation in the model.
+
+<!-- I will look into this. -->
+
+> 226-230 and Figure 1.
 Statistical analysis should be done of the climate trends, showing a least a trend line and a 95% confidence interval.
@@
 
-231-239 I find the maps of range loss in Fig 3 to be stunningly large.
+<!-- This is what Jack & I talked about, adding the moving window -->
+
+> 231-239 I find the maps of range loss in Fig 3 to be stunningly large.
 I am not sure anything in the literature would document such large losses in tree species presence from land use or 100 yrs of climate change.
@@
 
-347-350 Is it accurate to infer that ag has “largely eliminated open forests…” which does not sample low density forests?
+<!-- This partly a function of undersampling by the FIA structure (?) We need to clarify this. We add a supplemental version with "only FIA & PLSS overlap" -->
+
+> 347-350 Is it accurate to infer that ag has “largely eliminated open forests” which does not sample low density forests?
 It seems likely that most of these tree species are present in at least low densities in the areas where your maps show they are not present.
 
-Figure 2 b-e    The axes should be labeled and units indicated.  Consider reformatting these figures.  The current versions make visual separation of the lines hard to detect.
+<!-- Point to the LC survey, loss accoring to Rhemtulla, and the maintenance in anthropic biomes. -->
 
-Lines 361-364.
+> Figure 2 b-e, The axes should be labeled and units indicated.  Consider reformatting these figures. The current versions make visual separation of the lines hard to detect.
+
+> Lines 361-364.
 The topic of this sentence should be elaborated upon to more fully point out the limitations of the study and appropriate inference.
@@
 
-365-370 These sentences are a bit of bait and switch.
+> 365-370 These sentences are a bit of bait and switch.
 Yes, readers will agree that SDM based on FIA may not fully represent the realized niche of tree species.
@@
 The paper would have more impact if you were careful to emphasize the limitations of the data sets for this analysis and say that the degree to which FIA leads to SDM error is not knowable with available data sets.
+
+<!-- this repeats ome of the other comments -->
 
@@
 > Title: The authors used the upper Midwestern United States throughout the manuscript. But why the authors used the northern United States in the title?
-This study equally emphasized the distribution changes for 15 common tree genera over the last two centuries and the effects of climate change and land use change on these changes.
+
+We have changed this throughout the paper.
+
+> This study equally emphasized the distribution changes for 15 common tree genera over the last two centuries and the effects of climate change and land use change on these changes.
 Was it better to say “the effects of historical land use and climate change on species distribution changes” in the tittle?
+
+We believe the title should stand as is.
+We are not interested in the changes in spatial distributions, except inasmuch as they affect the climate-space occupied by the tree species, thus the "tree-climate relationship".
 
@@
 
+We appreciate the opinion of the reviewer in this context.  We have tried to use these datasets to show this -  <!-- point to Jack & David's paper. -->
+
 > 3) One of the objectives in this manuscript was to investigate the effects of climate change, land use change, and pathogen outbreaks on distribution changes (L125-126).
 However, the pathogen outbreaks were nearly ignored in the title and methods, results, and discussion sections.
+
+We will reframe "land use" to vegetation change, and then explain in the introduction how these all fold in.
 
@@
 
+We believe that our argument is valid, but may have been unclear in the framing.
+First, we believe that attributing the distributional changes from VH to VM to land use is justified, although, as noted above, we will change this to "vegetation change".
+Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement.
+Initial clearance for agricultural and urban development was contemporaneous with massive logging operations, and, in the late 1800s, widespread fire originiating from human sources.
+Given this, and the fact that very little "old growth" forest remains, modern forests are indellibly imprinted with the signal of human land use, whether agricultural conversion or timbering.
+The IPCC considers forestry to be a component of land use.
+The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging.
+
 > The authors should use present tense not past tense in the methods and results sections.
@@
 Annual precipitation (Pann) was the minimum, maximum, or mean precipitation across over the year?
-2)      More detailed information in the materials and method section was needed to show how the authors analyzed the overall distribution changes, climate change effects, and the relative importance of land use and climate change.
+
+> 2)  More detailed information in the materials and method section was needed to show how the authors analyzed the overall distribution changes, climate change effects, and the relative importance of land use and climate change.
 It should not be only in the figure titles.
 For example:
-The authors used gain, loss, and continuous persistence to demonstrate the distribution changes over the last two centuries in Figure 3.
+
+> The authors used gain, loss, and continuous persistence to demonstrate the distribution changes over the last two centuries in Figure 3.
 However, this approach was not described in the materials and method section but was only in the title of Figure 3.
@@
 Likewise, this approach was not described in the materials and method section but was only in the title of Figure 4.
-The approach used in Figure 5 was not described in the materials and method section but was only in the title of Figure 5.
+
+> The approach used in Figure 5 was not described in the materials and method section but was only in the title of Figure 5.
 More detailed information should be given to better analyze the relative importance of land use change and climate change effects.
 More detailed information about the univariate density functions (PDFs) and Hellinger distances was needed to better describe the approach.
-3)      Again, how the effects land use and exotic pathogens were included in the analyses?
-4)      The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a).
+
+> 3) Again, how the effects land use and exotic pathogens were included in the analyses?
+
+> 4) The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a).
 The historical distribution data for VHCH scenario were derived from PLS and the contemporary distribution data for VMCM were derived from US FIA data.
@@
 If my understanding was wrong, then the more detailed information was needed to better describe the factorial design and data used for each scenario.
-5)      About the t-tests to assess the interaction effects between land use and climate change (e.g., compounding, confounding, counteracting), what the responsible variables were used for the comparisons between the VHCH and VHCM, VHCH and VMCH.
+
+<!-- Clarify this in the review -->
+
+> 5) About the t-tests to assess the interaction effects between land use and climate change (e.g., compounding, confounding, counteracting), what the responsible variables were used for the comparisons between the VHCH and VHCM, VHCH and VMCH.
 More detailed information was needed to better describe these tests.
+
+<!-- catch them up? but also make sure it's clarified better. -->
 
@@
 > 1) L316-L322 were not the findings of this study. How were these points related to the findings? They read like introduction and should not be in discussion section.
+
+<!-- This is just reframing & revisiting the argument. -->
+
 > 2) Likewise, L323-l336 how were these points related to your findings? It stood alone now.

### 2016-11-24 Composition_Ecology.Rmd
Changes to title and spatial references (from northern to north-central).
Added statement about species ranges to qualify the region.

Added statement about the choice of climate data.

Standardized the use of $d_{v}$ vs $d_{l}$ in discussing the attribution index.

Explicit call to `pandoc.table`.

Added a period on line 278.

Added the point of *Tsuga* and the difference between land clearance and logging/fire.

Associated changes in `Response_to_reviewers.Rmd`.

@@
 
-# Effect of historical land-use and climate change on tree-climate relationships in the northern United States
+# Effect of historical land-use and climate change on tree-climate relationships in the north-central United States
 
@@
 Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
-Here we show that tree-climate relationships for 15 common tree genera in the northern US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change.
+Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
@@
 Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historical climate and land use changes on these shifts.
-We use gridded historical and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
@@
 
+For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.
+
 ## Historical and Modern Climate Data
 
-Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism2004].
+Historical and modern climate data are derived from monthly PRISM LT data for 1895 to 2014 CE [@prism2004].
 The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.
 Estimates for July temperature ($T_{max}$), January temperature ($T_{min}$), annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data.
+These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
+Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.
+
 Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
@@
 
-Distributions of tree genera within climate space are described as univariate density functions (PDFs) for the pre-settlement and modern eras.
+Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) for the pre-settlement and modern eras.
 Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions).
 The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-The effect of historical climate change on shifting climate distributions of tree genera is $d_{C}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-The effect of historical land use change on tree-climate distributions is $d_{l}$, the difference between the PLSS tree distributions when overlaid on historical vs.
-modern climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; Figure 2a; top row).
+The effect of historical climate change on shifting climate distributions of tree genera is $d_{c}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The difference between the PLSS and modern tree distributions ($d_{v}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest; Figure 2a; top row).
 
-We develop an attribution index, ($\delta_{d} = d_{c} - d_{l}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.
+We develop an attribution index, ($\delta_{d} = d_{c} - d_{v}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.
 
@@
 
-pandoc.table(loss_table, justify = "left", style = "rmarkdown")
+pander::pandoc.table(loss_table, justify = "left", style = "rmarkdown")
 ```
@@
 `r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs.
-changes in tree distributions ($d_{C} - d_{V}$), the latter mainly due to historical land-use and pathogen outbreaks.
+changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use and pathogen outbreaks.
 Results are plotted separately for each of the four climate variables considered here.
@@
 2).
-For $d_{C}-d_{V}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
+For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
 
@@
 
-Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and *Larix* ($T_{min}$)
+Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and *Larix* ($T_{min}$).
 
@@
 Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient.
+While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of *Tsuga*.
 

### 2016-11-24 Response_to_reviewers.Rmd
Fixed the `hell_plot`, with associated comment in the response to reviewers.  Axis is now $log_{10}$, with some extra ticks.
Also added clarification as to which climate variables were used, and what they actually measure.  With response to Reviewer 3 (line 168).

Fixed the `loss_table` table, changed rownames to scientific names.

@@
 
+We have changed the x axis to a log10 scale and added more tick marks to the x-axis to make it clear that it is non-linear.
+
 > Table 1: Some of the taxa are listed with their scientific names and others by common names. Replace common names with scientific names
+
+I have corrected the names so that only scientific names are displayed.
 
@@
 
+We have clarified these variables: "*Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) .
+.
+.*"
+
 > 2)  More detailed information in the materials and method section was needed to show how the authors analyzed the overall distribution changes, climate change effects, and the relative importance of land use and climate change.
@@
 However, this approach was not described in the materials and method section but was only in the title of Figure 3.
-The authors used the box-and-whisker plots to demonstrate the distribution changes along Tmax, Tmin, and Pann.
+
+> The authors used the box-and-whisker plots to demonstrate the distribution changes along Tmax, Tmin, and Pann.
 Likewise, this approach was not described in the materials and method section but was only in the title of Figure 4.
@@
 More detailed information should be given to better analyze the relative importance of land use change and climate change effects.
-More detailed information about the univariate density functions (PDFs) and Hellinger distances was needed to better describe the approach.
+
+> More detailed information about the univariate density functions (PDFs) and Hellinger distances was needed to better describe the approach.
 

### 2016-11-24 Composition_Ecology.Rmd
Fixed the `hell_plot`, with associated comment in the response to reviewers.  Axis is now $log_{10}$, with some extra ticks.
Also added clarification as to which climate variables were used, and what they actually measure.  With response to Reviewer 3 (line 168).

Fixed the `loss_table` table, changed rownames to scientific names.

@@
 The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.
-Estimates for July temperature ($T_{max}$), January temperature ($T_{min}$), annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data.
+Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data.
 These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
@@
 
-pander::pandoc.table(loss_table, justify = "left", style = "rmarkdown")
+sci_names <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
+               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
+               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")
+
+rownames(loss_table) <- paste0("*", sci_names, "*")
+
+pander::pandoc.table(loss_table, justify = "left", style = "rmarkdown",
+                     emphasize.rownames = FALSE)
 ```

### 2016-11-24 R/hellinger_dists.R

@@ -62,7 +62,8 @@ hellinger_plot <- function(){
               family = 'serif', size = 4, fontface = 'italic') +
     facet_wrap(~climate) +
     theme_bw() +
-    scale_x_sqrt() +
+    scale_x_log10(breaks = c(2, 10, 25, 50, 125), minor_breaks = NULL) +
+    scale_y_continuous(minor_breaks = NULL) +
     xlab(expression("Total Change"*" - "*PLS*" - "*FIA*" - "*d[tot])) +
     ylab(expression('Climate Change - Land Use Change - '*d[c]*' - '*d[v])) +
     theme(axis.text = element_text(family = 'serif', size = 12),
@@ -70,7 +71,7 @@ hellinger_plot <- function(){
           strip.text = element_text(family = 'serif', face = 'bold', size = 12),
           legend.position = "none",
           plot.margin = grid::unit(c(10,10,10,10), "mm")) +
-    coord_cartesian(xlim = c(0,125), ylim = c(-65, 85)) +
+    coord_cartesian(xlim = c(1,125), ylim = c(-65, 85)) +
     geom_abline(intercept = 0, slope = 0, alpha = 0.5) +
     annotate(geom = 'text', x = 7, y = 70, 
              label = 'Climate Dominates', family = 'serif',

### 2016-11-25 Response_to_reviewers.Rmd
More text updates and reviewr comments.
Information about detectability & heterogeneity differences between FIA & PLS, some simple analysis of climate normals, added a whole chunk of supplementary code to look at how choices of normals would affect the envelopes.

Added some references focused on historical temperature estimates.

Changed the structure of the `clim_boxplot()` function so it accepts filenames as an argument.  I really should change `confounding_table` so it's a function, then I could pass in the difference `vegclim_table` alternates.

@@
 
-This paper is now accepted.
+This Goring et al. paper is now accepted.  We have added the following paragraph:
+
+*FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset].
+PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
+Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.*
 
@@
 
+We have added context to this choice and the implications.
+Using long term records reported in Baker et al (1985) and Burnett et al.
+(2010) we indicate that the long term trend is overall warming in the region.
+We had previously used PDSI since this was the only available long term record we had access to.
+This change resulted in the inclusion of the folowing senteces in the Methods:
+
+*Baker et al.
+(1985) show long term increases in mean annual temperature at Ft.
+Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas Burnette et al.
+(2010) show increasing temperatures in all seasons, with faster increases during winter (0.10oC decade-1) than during summer (0.07oC decade-1).
+This implies that historical estimates may be biased by +0.35oC for $t_{max}$, by +0.5oC for $t_{min}$ and +0.15oC for $t_{diff}$.*
+
+and in the Results:
+
+*When long term records (Burnette et al., 2010; Baker et al., 1985) are considered, $\Delta t_{min}$ increases to approximately `r round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `r round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `r round(clim_change[2], 1) - 0.15`^o^C.*
+
+Interestingly, these shifts do not result in significant changes to the overall climate envelope for the tree species (compare the paper figure with the figure below), indicating the influence of land use is much larger than for climate for most taxa.
+
+![](figures/boxplot_corrected.png)
+
+In addition, the change, while large, remains small enough that the shifts related to land use remain larger than changes due to climate, and the directionality of all changes remains constant.
+We have added the code into the supplemental material.
+
+In the results we note:
+
+*Analysis included in the supplemental material indicates that a shift to an earlier baseline (using the estimated change rate from [@burnette2010daily] in Manhattan, Kansas) results in no change in this pattern (see "R/confounding_table.R" in S1).*
+
 > Another issue is lack of discussion about the time period when the forests sampled by FIA got established.
@@
 
-<!-- We could play with SDMs and run them back and forth with windows. -->
+We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks.
+This is a well established phenomena, and we appreciate the reviewer for pointing this out.
+In examining the normals for the region, beginning with a normal ending 1980, gives a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.
+These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively.
+Given the variability within windows, and the relatively young age of most stands in WI, MN and MI (FIA data shows mean ages of 52, 56 and 49 years respectively), we believe that the choice of normal is acceptable.
 

### 2016-11-25 Composition_Ecology.Rmd
More text updates and reviewr comments.
Information about detectability & heterogeneity differences between FIA & PLS, some simple analysis of climate normals, added a whole chunk of supplementary code to look at how choices of normals would affect the envelopes.

Added some references focused on historical temperature estimates.

Changed the structure of the `clim_boxplot()` function so it accepts filenames as an argument.  I really should change `confounding_table` so it's a function, then I could pass in the difference `vegclim_table` alternates.

@@
 [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
+FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset].
+PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
+Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.
 
@@
 Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
-There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.
+There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.
 The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
-To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`), and test July temperature reconstructions from observer data against our $T_{max}$ (Supplement 1).
+Baker et al.
+[-@baker1985minnesota] show long term increases in mean annual temperature at Ft.
+Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas [@burnette2010daily] show increasing temperatures in all seasons, with faster increases during winter (0.10^o^C decade^-1^) than during summer (0.07^o^C decade^-1^).
+This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.
 
@@
 The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid-19^th^ century climates.
+When long term records [@burnette2010daily; @baker1985minnesota] are considered, $\Delta t_{min}$ increases to approximately `r round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `r round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `r round(clim_change[2], 1) - 0.15`^o^C.
 
@@
 source("R/figures/climate_boxplots.R")
-boxplot_clim <- clim_boxplots(clim_table = vegclim_table)
+boxplot_clim <- clim_boxplots(clim_table = vegclim_table, 
+                              filename = 'boxplot_clim')
+
+# Do it with the correction for early historical:
+newveg <- vegclim_table
+topveg <- vegclim_table
+allveg <- vegclim_table
+
+newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tmax"] <- newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tmax"] - 0.35
+newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tmin"] <- newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tmin"] - 0.5
+newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tdiff"] <- newveg$clim[newveg$c.ref == "PLSS" & newveg$climate == "tdiff"] + 0.15
+
+topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tmax"] <- topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tmax"] + 0.84
+topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tmin"] <- topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tmin"] - 0.84
+topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tdiff"] <- topveg$clim[topveg$c.ref == "FIA" & topveg$climate == "tdiff"] + 1.15
+
+allveg$clim[allveg$c.ref == "FIA"] <- topveg$clim[topveg$c.ref == "FIA"]
+allveg$clim[allveg$c.ref == "PLSS"] <- newveg$clim[newveg$c.ref == "PLSS"]
+
+boxplot_clim <- clim_boxplots(clim_table = allveg,
+                              filename = "boxplot_corrected_all")
+
+boxplot_clim <- clim_boxplots(clim_table = newveg,
+                              filename = "boxplot_corrected_pls")
+
+boxplot_clim <- clim_boxplots(clim_table = topveg,
+                              filename = "boxplot_corrected_fia")
+
+boxplot_clim <- clim_boxplots(clim_table = vegclim_table,
+                              filename = "boxplot_original")
 
@@
 pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
-
+knitr::kable(conf_table_corr)
 ```
@@
 Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
+Analysis included in the supplemental material indicates that a shift to an earlier baseline (using the estimated change rate from [@burnette2010daily] in Manhattan, Kansas) results in no change in this pattern (see "R/confounding_table.R" in S1).
 

### 2016-11-25 R/confounding_table.R

@@ -13,12 +13,12 @@
 taxa <- unique(vegclim_table$taxon)
 clim_var <- unique(vegclim_table$climate)
 
-conf_test <- function(taxa, clim_var) {
-  clim = t.test(subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
-         subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'FIA' & data > 0)$clim)
+conf_test <- function(taxa, clim_var, climtable) {
+  clim = t.test(subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
+         subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'FIA' & data > 0)$clim)
 
-  lu   = t.test(subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
-         subset(vegclim_table, taxon == taxa & climate == clim_var & base == 'FIA' & c.ref == 'PLSS' & data > 0)$clim)
+  lu   = t.test(subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
+         subset(climtable, taxon == taxa & climate == clim_var & base == 'FIA' & c.ref == 'PLSS' & data > 0)$clim)
 
   # We want to return the t-test p value & estimate
   data.frame(taxa = taxa, clim = clim_var,
@@ -29,7 +29,11 @@ conf_test <- function(taxa, clim_var) {
 
 tests <- do.call(rbind.data.frame,
         lapply(taxa, function(x) {
-  do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y)}))}))
+          do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = vegclim_table)}))}))
+
+tests_corrected <- do.call(rbind.data.frame,
+                           lapply(taxa, function(x) {
+                             do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = newveg)}))}))
 
 checker <- function(x) {
   # x will be the row:
[…]

### 2016-11-25 R/figures/climate_boxplots.R

@@ -1,5 +1,5 @@
 # Plotting the total climate-shift for the domain.
-clim_boxplots <- function(clim_table = vegclim_table){
+clim_boxplots <- function(clim_table = vegclim_table, filename){
   
   clim_table$taxon <- mapvalues(clim_table$taxon, from = levels(clim_table$taxon),
                                 to = c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
@@ -25,7 +25,7 @@ clim_boxplots <- function(clim_table = vegclim_table){
           axis.text.x = element_text(angle = 45, hjust = 1, face = 'bold.italic'),
           plot.margin = grid::unit(c(10,10,10,10), "mm"))
   
-  ggsave(plot = boxplot_clim, filename = 'figures/boxplot_clim.pdf', width = 6.7, height = 4.5)
-  ggsave(plot = boxplot_clim, filename = 'Final_Figures/boxplot_clim.pdf', width = 6.7, height = 4.5)
+  ggsave(plot = boxplot_clim, filename = paste0('figures/',filename,'.png'), width = 6.7, height = 4.5)
+  ggsave(plot = boxplot_clim, filename = paste0('Final_Figures/',filename,'.pdf'), width = 6.7, height = 4.5)
   boxplot_clim
 }

### 2016-11-25 Response_to_reviewers.Rmd
Adding the changes related to normals & SDMs from reviewer 2.

@@
 
-We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks.
-This is a well established phenomena, and we appreciate the reviewer for pointing this out.
-In examining the normals for the region, beginning with a normal ending 1980, gives a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.
-These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively.
+We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks, we believe that our contribution further underscores this point.
+The normals for the region, beginning with a normal ending 1980 and continuing to a normal ending 2014, give a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.
+These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively, but, as pointed out in Baker *et al*.
+(2016; doi:10.1111/gcb.13273) uncertainty in the underlying estimates are often overlooked.
 Given the variability within windows, and the relatively young age of most stands in WI, MN and MI (FIA data shows mean ages of 52, 56 and 49 years respectively), we believe that the choice of normal is acceptable.
+Having said that, we examined the importance using both shifts in PLS era, and shifts in the modern climate normal.
+The largest effect is in $t_{max}$, since the most recent normals all show a significant increase in $t_{max}$.
+
+![](figures/corrected_all.png)
 
@@
 
-<!-- We intentionally spaced the records. Species distribution models are not part of this discussion, we recognize that different time periods might affect this. -->
+I believe that this is addressed in the response to the prior comment.
 
@@
 
-<!-- We are not building predictive models, with SDMs you get the blend, but it may not be clear which elements are directly impacting the model response. -->
+We apreciate this point and thank the reviewer for bringing it up.
+The challenge in this case is that a multivariate SDM blends the climate variables.
+Further, a multi-model SDM approach might weight climate variables differently across models, through time, making sensitivity analysis for the impact of each climate variable challenging.
+We point out that this analysis is not intended to focus on the construction of SDMs, but rather to show the relationship between land use and the individual climate variables, through the distribution of tree species.
+Thus, disentagling land use and climate shifts on a set of multivariate combination of variables from each time slice becomes extremely challenging, and, is effectively a separate paper.
+
+What this paper intends to point to is that land use is impacting climate variables differentially across species as a result of land use, and that this may have implications for SDM construction.
+Shifts in the strength of SDMs between time scales and potential utility of climate variables in SDM construction is a separate analysis and paper.
 
@@
 
-<!-- I will look into this. -->
+We have included spatial autocorrelation in the t-tests using the `SpatialPack` function`modified.ttest`.
+We use confidence intervals based on the raw t.test scores, but report significance based on the modified.ttest calculation.
 

### 2016-11-27 R/confounding_table.R
Revisited the `confounding_table` based on request from Reviewer #2.  To calculate the degrees of freedom when there is spatial autocorrelation we use the estimate from the pairwise t-test implemented by in `modified.ttest`, but apply it to the t estimate from `t.test`.

@@ -14,16 +14,72 @@ taxa <- unique(vegclim_table$taxon)
 clim_var <- unique(vegclim_table$climate)
 
 conf_test <- function(taxa, clim_var, climtable) {
-  clim = t.test(subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
-         subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'FIA' & data > 0)$clim)
 
-  lu   = t.test(subset(climtable, taxon == taxa & climate == clim_var & base == 'PLSS' & c.ref == 'PLSS' & data > 0)$clim,
-         subset(climtable, taxon == taxa & climate == clim_var & base == 'FIA' & c.ref == 'PLSS' & data > 0)$clim)
-
+  require(dplyr)
+  require(SpatialPack)
+  
+  coords <- climtable %>% select(cell, x, y) %>% distinct
+  
+  # Get all values, join to coords so we have the same number of cells across all datasets
+  # then make sure we have only distinct sets.
+  
+  VHCH <- climtable %>% filter(c.ref == "PLSS" & 
+                               climate == clim_var & 
+                               taxon == taxa &
+                               base == "PLSS") %>% 
+    select(cell, x, y, clim, data) %>% 
+    left_join(coords, .)
+  
+  VHCH$clim[VHCH$data == 0] <- NA
+  
+  VHCM <- climtable %>% filter(c.ref == "FIA" & 
+                                 climate == clim_var & 
+                                 taxon == taxa &
+                                 base == "PLSS") %>% 
+    select(cell, x, y, clim, data) %>% 
+    left_join(coords, .) %>% distinct(cell, .keep_all = TRUE)
+  
+  VHCM$clim[VHCM$data == 0] <- NA
+  
+  VMCH <- climtable %>% filter(c.ref == "PLSS" & 
+                                 climate == clim_var & 
+                                 taxon == taxa &
+                                 base == "FIA") %>% 
+    select(cell, x, y, clim) %>% 
+    left_join(coords, .) %>% distinct(cell, .keep_all = TRUE)
+  
+  VMCH$clim[VMCH$data == 0] <- NA
+  
+  all_cell <- coords$cell[coords$cell %in% VHCH$cell &
[…]

### 2016-11-27 Response_to_reviewers.Rmd
Added header body to reviewer responses to allow it to knit propoerly.
Continued working on responses to reviewers.

Made lots of changes but haven't been committing as regularly as I should.

Mostly focused on responding to Reviewer 3.

@@
 date: "November 11, 2016"
-output: word_document
+bibliography: goringetal_references.bib
+csl: ecology.csl
+output:
+  word_document: 
+    reference_docx: word_template.docx  
+  pdf_document:
+    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
+  md_document:
+    variant: markdown_github
+pandoc_args: -H margins.sty
 ---
@@
 PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
-Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.*
+Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.
+We have selected the 15 most common tree species in the region .
+.
+.
+*
 
@@
 
-We use the PRISM LT data.  This has been clarified in the paper.
+We use the PRISM LT data.  This has been clarified in the paper.  NEED TO ADD DATA SPARSENESS.
 
@@
 
-We have changed this sentence in particular to read: *The difference between the PLSS and modern tree distributions ($d_{V}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest* and changed other references to $d_{l}$ to $d_{v}$.
+We have changed this sentence in particular to read: "*The difference between the PLSS and modern tree distributions ($d_{V}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest*"" and changed other references to $d_{l}$ to $d_{v}$.
 
@@
 I suppose a bit more clarification regarding what is meant by confounding and why grouping taxa where neither climate nor history has an effect with taxa where only one has an effect is necessary.
+
+We've changed the table to make things more clear.
 
@@
 
+![](figures/hellingerplot.png)
+
 > Table 1: Some of the taxa are listed with their scientific names and others by common names. Replace common names with scientific names
@@
 
-<!-- need to add a paragraph & add the curve from John's figure. -->
-
 > The paper would be strengthened by stating more strongly the limitations of the methods and that the results may be over estimates and that data are not currently available to more accurately estimate the effect.
@@
 
-<!-- Pollen data, Jeannine & others. add in John's paper we can point to long term ecological datasets -->
+We thank the reviewer for these comments.
+We believe that we now more fully address the limitations -- We have added references to long term weather records and used their early 19th century estimates as the basis for some sensitivity analysis (see below), we have used a moving window to examine the effect of 21st century climate change on the choice of the modern climate normal.
+We have also strengthed the section that indicates prior support for widespread land use change, and its impacts on tree species distributions in the region.
+Lastly, in the **Discussion** we have reitterated what we believe are some of the limitations of the current study.
 
-More specific comments are below.
+### More specific comments
 
-Line reference
 > 2  "north-central" US would be more accurate? Also change this throughout the ms.
@@
 
-<!-- The challenge is double edged.
-Area lost (definite) vs.
-frequency of occurrence.
-Also most of the tree species are common, frequency in the area lost, vs frequency in the core.
-Add to supplemental information.
--->
-
 > 181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.
@@
 Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.*
+
+Add Woodward - moisture availability -
 
@@
 
-Interestingly, these shifts do not result in significant changes to the overall climate envelope for the tree species (compare the paper figure with the figure below), indicating the influence of land use is much larger than for climate for most taxa.
+Interestingly, these shifts in PLS baseline climate do not result in significant changes to the overall climate envelope for the tree species (compare the paper figure with the figure below), indicating the influence of land use is much larger than for climate for most taxa, and supercedes the shifts from the early 19th to the early 20th century.
 
-![](figures/boxplot_corrected.png)
+![](figures/boxplot_corrected_pls.png)
 
@@
 We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks, we believe that our contribution further underscores this point.
+
 The normals for the region, beginning with a normal ending 1980 and continuing to a normal ending 2014, give a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.
@@
 
-![](figures/corrected_all.png)
+We have added the sentence: *"Recent climate change"*
+
+![](figures/boxplot_corrected_all.png)
 
@@
 Further, a multi-model SDM approach might weight climate variables differently across models, through time, making sensitivity analysis for the impact of each climate variable challenging.
+
 We point out that this analysis is not intended to focus on the construction of SDMs, but rather to show the relationship between land use and the individual climate variables, through the distribution of tree species.
@@
 We have included spatial autocorrelation in the t-tests using the `SpatialPack` function`modified.ttest`.
-We use confidence intervals based on the raw t.test scores, but report significance based on the modified.ttest calculation.
+Unfortunately, the `modified.ttest` only performs a pairwise t-test, which is inappropriate in this situation since we would not be able to assess the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison since only the overlapping points would be included and we would assess zero change.
+
+The `modified.ttest` calculates a t value, but also calculates a modified $df$ that takes into account spatial autocorrelation based on Moran's I for each of the two spatial processes.
+To then produce a difference of the means that is not pairwise, we obtain the corrected $df$ from `modified.ttest`, and recalculate the $p$-value for the t-test using the t estimate from the t.test.
+This is all coded in the `confounding_table.R` file.
 
@@
 
-<!-- This is what Jack & I talked about, adding the moving window -->
+We have added a LOESS curve onto Figure 1, representing a span of 20 years, along with the SE fit.
+We have added light shading over the normal boxes to highlight the normal period.
+As indicated above, we have provided a more detailled analysis of the choice of normal periods and described the effect of the choice on outcomes.
 
@@
 
-<!-- This partly a function of undersampling by the FIA structure (?) We need to clarify this. We add a supplemental version with "only FIA & PLSS overlap" -->
+I have added an explicit statement to quantify losses in the paragraph beginning at **l**104.
+There is strong support for extensive loss of coverage since the PLSS for certain species, and expansion for others, all referenced in this paragraph.
+These patterns are well supported in the literature.
+Rhemtulla et al (2007) cite estimates of land use conversion up to 79% for southern deciduous savanna, and Rhemtulla et al (2009) show that losses in oak-hickory forest complexes surpass 5,000,000 ha.
+As noted in the new sentences: "*Rhemtulla et al.
+[-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin.
+Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (Figure Xf), where 79% of deciduous savanna was converted to cropland [@rhemtulla2007regional].*"
+
+Given the extensive support for broad scale land use change, and the associated impacts of fire & logging, we believe that these patterns are real.
 
@@
 
-<!-- Point to the LC survey, loss accoring to Rhemtulla, and the maintenance in anthropic biomes. -->
+As above, we believe that these patterns are real.  We extend this point however beginning at **l**371, where we discuss the potential use of isolated trees or trees unsampled in the FIA data.
 
 > Figure 2 b-e, The axes should be labeled and units indicated.  Consider reformatting these figures. The current versions make visual separation of the lines hard to detect.
+
+Working on this. . .
 
@@
 Say how much the estimates of forest loss and niche shifts may be overestimated.
+
+See the following note.
 
@@
 
-<!-- this repeats ome of the other comments -->
+I believe the reviewer possibly overstates the case for limitations.
+There is strong support for changes since EuroAmerican settlement, both in vegetation-climate relationships, in the spatial extent of vegetation cover, and in pollen-vegetation relationships in the region, indicating a long-term shift in vegetation representation in climate space.
+However, we have qualified the statements made in this section, and added a short paragraph pointing to the potential effects of the choice of climate normal as well.
+
+*Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.
+In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
+If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
+Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
+Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
+As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.*
+
+*Artifacts of the choice of climate normal (1995 - 2014) for the modern data mean an exaggerated effect for $t_{max}$, however, supplemental analysis shows that the patterns of change for the other variables, $t_{min}$, $t_{diff}$ and $p_{ann}$ remain robust.
+Analysis using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships [@jacques2008pre] following Euro-American settlement in the region providing further support that modern vegetaton-climate relationships have changed since settlement.*
 
@@
 
-We appreciate the opinion of the reviewer in this context.  We have tried to use these datasets to show this -  <!-- point to Jack & David's paper. -->
+We appreciate the opinion of the reviewer in this context. We address the first point by changing the introductory sentence to:
+
+*Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood*
+
+We believe that the issue of stability or extending the dist-clim relationship is a matter of opinion.
+The paper shows that, for many taxa, land use following EuroAmerican settlement in the north-central US limits the representitiveness of modern inventory-based distribution models.
+We do believe that these historical datases can be used to extend distribution-climate relationships.
+The exlicit tests in this paper focus on shifts in climate-distribution relationships and the attribution for these shifts, thus, the stabilty of the relationships.
 
@@
 
-We will reframe "land use" to vegetation change, and then explain in the introduction how these all fold in.
+We agree.
+The impact of pathogen outbreaks are significant for *Ulmus*, and (outside the current range) for *Castanea*, however, given the current data sources, they are difficult to properly assess.
+We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions.
 
@@
 Why?
+
+Previous authors have clearly indicated that EuroAmerican settlement in the north-central United States can be almost entirely attributed to human agency.
+In responding to Reviewer 2 we make the same point, and make this point below.
+EuroAmerican Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture.
+Thus, VH, from the PLSS forest survey (ignoring climate data for now), represents forest structure and distribution in the absence of settlement impacts.
+VM represents a modern landscape, sampled over the last decade as part of the FIA program.
+Distribution loss is almost entirely attributable to anthropogenic agency, either loss of forested land (oak-hickory forest lost 5x10^6^ ha of coverage since settlement due to agricultural conversion, as pointed out in the response to Reviewer 2), thus we feel confident in ascribing change from VC -> VM to land use.
+We believe that changes in the introduction and discussion, strengthening the argument for widespread vegetation change help resolve this issue.
+
+To further clarify, we have modified the explanation of Figure 2 to read, in part:
+
+*The difference between the PLSS and modern tree distributions ($d_{v}$) is attributed largely to land use change, and we will refer to it as such throughout.
+As described above, widespread post-settlement disturbance in the upper Midwest included conversion to agriculture, logging and widespread fire in the late 19th century, largely attributed to slash burning and logging.
+When the vegetation change is overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) we show the change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.*
 
@@
 We believe that our argument is valid, but may have been unclear in the framing.
-First, we believe that attributing the distributional changes from VH to VM to land use is justified, although, as noted above, we will change this to "vegetation change".
+First, we believe that attributing the distributional changes from VH to VM to land use is justified.
 Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement.
@@
 The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging.
+While sucession processes occurred since clearance and settlement, these were directly affected by the initial land clearance.
+For example, Rhemtulla et al.
+(2009) indicate that the loss of hemlock was caused both by logging, and recruitment failure as the result of the loss of seed source because of the extensive logging, thus both cause & effect are rooted in EuroAmerican settlement.
 
@@
 
+We have added the following sentence to the Methods:
+
+*In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.*
+
 > The authors used the box-and-whisker plots to demonstrate the distribution changes along Tmax, Tmin, and Pann.
@@
 > More detailed information about the univariate density functions (PDFs) and Hellinger distances was needed to better describe the approach.
+
+We have modified the sentence discussing the density functions to read:
+
+*Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) for the pre-settlement and modern eras, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.*
+
+We have added the equation for the Hellinger distance:
+
+*Hellinger distance for two discrete probability distributions is defined as:*
+
+H(P, Q) = \frac{1}{\sqrt{2}} \; \sqrt{\sum_{i=1}^k (\sqrt{p_i} - \sqrt{q_i})^2}
+
+*where $P$ and $Q$ are the distributions, with a common discrete index $i$.*
 
@@
 
-<!-- Clarify this in the review -->
+We have attempted to clarify this description by re-writing this paragraph:
+
+*The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+These conditions are observed or recorded in data.
+The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.
+The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.
+We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.*
 
@@
 > 2) Likewise, L323-l336 how were these points related to your findings? It stood alone now.
+
+This paragraph is intended to provide stronger links to other mechanisms by which species may not be in equilibrium with occupied climate space.
+We believe that the final sentence of the paragraph is sufficient as is.
+
 > 3) L337-l351 should be better related to your findings. The first and second points stood alone now.
+
+We have added a reference back to figure 1 which shows current land use in the region, clearly indicating the
+
 > 4) L352-L360 here, the authors finally mentioned the pathogen outbreaks effects. However, the pathogen outbreaks effect were not included in the data analyses. This seems just speculation.
 
-Specific comments:
+## Specific comments:
 
 > 1) P2-L36, change “is poorly understood” to “are poorly understood”.
+
+The sentence refers to the "stability and representativeness", which is singular.
+
 > 2) P2-L41, change “implicates” to “implicated”.
+
+The new attribution index is within the current paper, we believe that the present tense is valid.
+
 > 3) P2-L43, change “are compounding” and “reinforces” to “were compounding” and “reinforced”.
+
+Again, this is presnt tense as it refers to the current study.
+
 > 4) The figures were not well made.
@@
 It may be better to use shorter names instead.
+
 > 5) All figure titles were too long. Many of these contents were approach and should be put in methods section.
 
-Overall, this is an interesting topic.
+> Overall, this is an interesting topic.
 Especially, the interaction between land use change and climate change on distribution changes is the most interesting part, because the overall distribution changes have been extensively studied.

### 2016-11-27 Composition_Ecology.Rmd
Adding the changes related to normals & SDMs from reviewer 2.
Began removing some of the *pathogen* content, or at least it's emphasis in the purpose of the paper.
Began changing tense in the methods to present where appropriate based on the advice of Reviewer 3.

Better explanation of the terminology later used in figure captions, based on the advice of Reviewer 3.

Explained why we used the normal we did.

Added in the Hellinger equation and rephrased the explanation of how it was used.

Added some packages into the **Analysis** section.

Added references to Rhemtulla et al 2007 in the Discu

@@
 
-Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
-Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change.
+Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
+Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
@@
 Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
-Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).
+Rhemtulla et al.
+[-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin.
+Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].
 In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].
-Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.)
+Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.).
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historical climate and land use changes on these shifts.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change and land use, and we diagnose the relative influence of historical climate and land use changes on these shifts.
 We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
@@
 A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*.
-'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.
+We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.
 
@@
 PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
+
 Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.
-
+In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.
 For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.
@@
 
-Historical and modern climate data are derived from monthly PRISM LT data for 1895 to 2014 CE [@prism2004].
+We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] to generate historical and modern climate data.
 The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.
-Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data.
+We resample estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) to the 8 x 8km grid used for PLS and FIA data.
 These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
@@
 The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
-To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`), and test July temperature reconstructions from observer data against our $T_{max}$ (Supplement 1).
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
 Baker et al.
@@
 This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.
+The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study.
 
@@
 
-Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) for the pre-settlement and modern eras.
-Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions).
+Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) of tree species presence ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.
+Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions).
+Hellinger distance for two discrete probability distributions is defined as:
+
+$H(P, Q) = \frac{1}{\sqrt{2}} \; \sqrt{\sum_{i=1}^k (\sqrt{p_i} - \sqrt{q_i})^2}$
+
+where $P$ and $Q$ are the distributions, with a common discrete index $i$.
+
 The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-The effect of historical climate change on shifting climate distributions of tree genera is $d_{c}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-The difference between the PLSS and modern tree distributions ($d_{v}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest; Figure 2a; top row).
+These conditions are observed or recorded in data.
+The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.
+The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.
+We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.
 
@@
 
-All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`.
+## Analysis
+
+All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`.
 All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate).
@@
 In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile.
-These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`
+These shifts in realized climate distributions are produced by the joint effects of historical climate change and land-use change.")`
 
@@
 `r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs.
-changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use and pathogen outbreaks.
+changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use.
 Results are plotted separately for each of the four climate variables considered here.
@@
 `r tab_nums("compTable", "Compounding ($\\Box$), confounding (-) and counteracting ($\\bigcirc$) relationships between land-use and climate change.
-When the land-use/climate interaction was regressed against spatial vaariables (latitude, longitude) no significant relationship was found.")`
+When the land-use/climate interaction was regressed against spatial variables (latitude, longitude) no significant relationship was found.")`
 
@@
 pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
-knitr::kable(conf_table_corr)
+
 ```
@@
 Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress].
-Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges.
-Agricultural conversion has largely eliminated open forests in warm and dry regions, while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times.
+Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges (Figure 1f).
+Agricultural conversion has largely eliminated open forests in warm and dry regions [@rhemtulla2007], while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times.
 Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient.
@@
 In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
-If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated.
+If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
 Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
 Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
+As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.
+
+Artifacts of the choice of climate normal (1995 - 2014) for the modern data mean an exaggerated effect for $t_{max}$, however, supplemental analysis shows that the patterns of change for the other variables, $t_{min}$, $t_{diff}$ and $p_{ann}$ remain robust.
+Analysis using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships [@jacques2008pre] following Euro-American settlement in the region providing further support that modern vegetaton-climate relationships have changed since settlement.
 

### 2016-11-27 Response_to_reviewers.Rmd
Found a bug in `confounding_table`, the `VMCH` wasn't eliminating 0 data rows properly, so it was giving the whole domain, and not the species range.
Reviewer responses, finished the response about the modern climate normal.  Clarified pathogen stuff for Reviewer 3.

@@
 
-We use the PRISM LT data.  This has been clarified in the paper.  NEED TO ADD DATA SPARSENESS.
+We use the PRISM LT data.  This has been clarified in the paper.
 
@@
 
-Add Woodward - moisture availability -
+**JWW to add Woodward - moisture availability -**
 
@@
 
-We have added the sentence: *"Recent climate change"*
+We have added the sentence: *"The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study."*
 
@@
 The impact of pathogen outbreaks are significant for *Ulmus*, and (outside the current range) for *Castanea*, however, given the current data sources, they are difficult to properly assess.
-We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions.
+We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions in the Discussion and in the Introduction.
 
@@
 Previous authors have clearly indicated that EuroAmerican settlement in the north-central United States can be almost entirely attributed to human agency.
-In responding to Reviewer 2 we make the same point, and make this point below.
+In responding to Reviewer 2, we make the same point, and make this point again below.
 EuroAmerican Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture.

### 2016-11-27 R/confounding_table.R

@@ -45,7 +45,7 @@ conf_test <- function(taxa, clim_var, climtable) {
                                  climate == clim_var & 
                                  taxon == taxa &
                                  base == "FIA") %>% 
-    select(cell, x, y, clim) %>% 
+    select(cell, x, y, clim, data) %>% 
     left_join(coords, .) %>% distinct(cell, .keep_all = TRUE)
   
   VMCH$clim[VMCH$data == 0] <- NA
@@ -84,20 +84,24 @@ conf_test <- function(taxa, clim_var, climtable) {
 }
 
 tests <- do.call(rbind.data.frame,
-        lapply(taxa, function(x) {
-          do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = vegclim_table)}))}))
+        lapply(unique(vegclim_table$taxon), function(x) {
+          do.call(rbind.data.frame,lapply(unique(vegclim_table$climate), 
+                                          function(y) {conf_test(x, y, climtable = vegclim_table)}))}))
 
 tests_pls <- do.call(rbind.data.frame,
                            lapply(taxa, function(x) {
-                             do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = newveg)}))}))
+                             do.call(rbind.data.frame,
+                                     lapply(clim_var, function(y) {conf_test(x, y, climtable = newveg)}))}))
 
 tests_fia <- do.call(rbind.data.frame,
                            lapply(taxa, function(x) {
-                             do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = topveg)}))}))
+                             do.call(rbind.data.frame,
+                                     lapply(clim_var, function(y) {conf_test(x, y, climtable = topveg)}))}))
 
 tests_all <- do.call(rbind.data.frame,
                      lapply(taxa, function(x) {
-                       do.call(rbind.data.frame,lapply(clim_var, function(y) {conf_test(x, y, climtable = allveg)}))}))
+                       do.call(rbind.data.frame,
+                               lapply(clim_var, function(y) {conf_test(x, y, climtable = allveg)}))}))
 
 checker <- function(x) {
   # x will be the row:

### 2016-11-27 Response_to_reviewers.Rmd
Pushing everyhtng up for trade-off with @iceageecologist.

@@
 
-*When long term records (Burnette et al., 2010; Baker et al., 1985) are considered, $\Delta t_{min}$ increases to approximately `r round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `r round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `r round(clim_change[2], 1) - 0.15`^o^C.*
+*When long term records (Burnette et al., 2010; Baker et al., 1985) are considered, $\Delta t_{min}$ increases to approximately `round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `round(clim_change[2], 1) - 0.15`^o^C.*
 
@@
 I believe that this is addressed in the response to the prior comment.
+We have added a LOESS curve to Figure 1, we have selected a normal that reflects the period of sampling, and have provided additional support to the choice of the modern normal.
 
@@
 In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
-If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
+If so, estimates of forest loss (`fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
 Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
@@
 
-*The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+*The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
 These conditions are observed or recorded in data.
 The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.
-The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `fig_nums("fourPanel", display = "cite")`a; left column).
 We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.

### 2016-11-27 R/figures/FigureFour_Panels.R

@@ -3,30 +3,38 @@
 library(png)
 library(grid)
 
-ggplotColours <- function(n=6, h=c(0, 360) +15){
-  if ((diff(h)%%360) < 1) h[2] <- h[2] - 360/n
+ggplotColours <- function(n=6, h=c(0, 360) + 15){
+  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
   hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
 }
 
 model_grid <- readPNG('figures/model_grid.png')
 
-oak <- subset(vegclim_table, taxon == 'Tamarack' & data > 0)
+fourpanel <- function(clim_table, taxon) {
+  veg <- clim_table %>% filter(taxon == taxon & data > 0 & climate == "tmax")
+  
+  levels(veg$)
+  
+  curves <- ggplot(veg) +
+    geom_density(adjust = 1, size = 1.2,
+                           aes(x = clim, color = base, linetype = c.ref)) +
+              facet_grid(base ~ c.ref, scale = 'free') +
+              theme_bw() +
+              scale_color_manual(name = 'Vegetation', values = rev(ggplotColours(2))) +
+              scale_linetype_manual(name = 'Climate', values = c(2,1)) +
+              theme(axis.text = element_text(family = 'serif', face = 'bold', size = 12),
+                    axis.title = element_blank(),
+                    strip.text = element_text(family = 'serif', face = 'bold', size = 14),
+                    legend.title = element_text(family = 'serif', face = 'bold', size = 14),
+                    legend.text = element_text(family = 'serif', size = 12),
+                    legend.position = 'none')
+  
+  #ggsave(plot=grid.arrange(rasterGrob(model_grid), oak_curves, ncol=2,
+  #                         widths = c(3,4)), 
+  #       file = 'figures/figureFour_panels.tiff',
+  #       width=6, height = 3, dpi = 300)
+  
+  curves
+}
 
-oak_curves <- ggplot(oak) +
-  geom_density(adjust=1, size = 1.2,
-                         aes(x = clim, color = base, linetype = c.ref)) +
-            facet_wrap(~climate, scale = 'free') +
-            theme_bw() +
-            scale_color_manual(name = 'Vegetation', values = rev(ggplotColours(2))) +
-            scale_linetype_manual(name = 'Climate', values = c(2,1)) +
-            theme(axis.text = element_text(family = 'serif', face='bold', size = 12),
-                  axis.title = element_blank(),
[…]

### 2016-11-27 R/figures/FigureFour_Panels.R
Changed Figure 2 - added it into the response to reviewers.  This involved a change to the `R/figures/FigureFour_Panels.R` file.
Not sure why the png isn't showing up in the commits. . .

@@ -10,31 +10,34 @@ ggplotColours <- function(n=6, h=c(0, 360) + 15){
 
 model_grid <- readPNG('figures/model_grid.png')
 
-fourpanel <- function(clim_table, taxon) {
-  veg <- clim_table %>% filter(taxon == taxon & data > 0 & climate == "tmax")
+  fourpanel <- function(clim_table, taxon) {
+    veg <- clim_table %>% filter(taxon == taxon & data > 0 & climate == "ppt")
+    
+    veg$base <- factor(veg$base, levels = c("PLSS", "FIA"))
+    
+    veg$c.ref = factor(veg$c.ref, levels = c("PLSS", "FIA"))
+    
+    curves <- ggplot(veg) +
+      geom_density(adjust = 1, size = 1.1,
+                             aes(x = clim, color = base, linetype = c.ref)) +
+      scale_color_manual(name = 'Vegetation', values = (ggplotColours(2))) +
+      scale_linetype_manual(name = 'Climate', values = c(1,3)) +
+      facet_grid(c.ref ~ base) +
+      theme_bw() +
+      scale_x_continuous(breaks = c(500, 700, 900)) +
+      theme(axis.text = element_text(family = 'serif', face = 'bold', size = 12),
+            axis.title = element_blank(),
+            strip.text = element_text(family = 'serif', face = 'bold', size = 14),
+            legend.title = element_text(family = 'serif', face = 'bold', size = 14),
+            legend.text = element_text(family = 'serif', size = 12),
+            legend.position = 'none')
+    
+    ggsave(plot = grid.arrange(rasterGrob(model_grid), curves, ncol = 2,
+                             widths = c(3,4)), 
+           file = 'figures/figureFour_panels.tiff',
+           width = 6, height = 3, dpi = 300)
+    
+    curves
+  }
   
-  levels(veg$)
-  
-  curves <- ggplot(veg) +
-    geom_density(adjust = 1, size = 1.2,
-                           aes(x = clim, color = base, linetype = c.ref)) +
-              facet_grid(base ~ c.ref, scale = 'free') +
-              theme_bw() +
-              scale_color_manual(name = 'Vegetation', values = rev(ggplotColours(2))) +
-              scale_linetype_manual(name = 'Climate', values = c(2,1)) +
-              theme(axis.text = element_text(family = 'serif', face = 'bold', size = 12),
-                    axis.title = element_blank(),
[…]

### 2016-12-02 Response_to_reviewers.Rmd
Changed Figure 2 - added it into the response to reviewers.  This involved a change to the `R/figures/FigureFour_Panels.R` file.
Not sure why the png isn't showing up in the commits. . .
I had to soft-reset a bunch of changes to unde the accidental inclusion of two large RDS files.
Changes include all revisions up to the last pass-off to @iceageecologist & include code changes to improve the modularity of the code.

@@
 ---
-title: "Response to Reviewers"
-author: "Simon Goring"
-date: "November 11, 2016"
 bibliography: goringetal_references.bib
@@
 
+Dear Dr. Thrall:
+
+We submit for your consideration the revised manuscript "Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States."
+
+We have carefully revised the manuscript in response to reviewer comments and have paid particular attention to the potential limitations associated with 1) comparing the PLS and FIA data and 2) choice of climate normal.
+We completed several new analyses to test the sensitivity of our results to choice of climate normal and created a separate subsection within the Discussion assessing the robustness of our findings.
+
+We have edited the manuscript for clarity and to respond to the other points raised by the reviewers.
+We include below a point-by-point response to all reviews.
+The revised version has XX  words and 6 figures and tables.
+We have expanded the Supplementary Information to show several additional analyses.
+
+We thank you and especially the reviewers for their thoughtful and constructive criticism – we believe that the manuscript has greatly improved as a result.
+Thank you too for extending the deadline as requested.
+
+Sincerely,
+
+Simon Goring
+Jack Williams
+
+## Editor
+
+*Editors Comments for the Author(s): The reviews highlight the importance and novelty of the main idea - that past land-use complicates our estimates of forest community responses to climate change.
+However, they express concern about the way that the results are presented.
+There are many suggestions for clarification that should be addressed in revision.
+In particular, it's important to address the limitations of comparing the two data types, as this was a concern expressed by each reviewer.*
+
+Thank you.  Please see responses below.
+
 ## Reviewer 1:
 
-Reviewer 1 was generally positive.  I have pulled out the specific comment from the broad overview that addresses a broader limitation of the methods.  I also include line by line changes.
+> The manuscript written by Goring and Williams (“Effect of historic land-use and climate change on tree-climate relationships in the northern United States”; ELE-00978-2016) addresses an important topic in global change research: what is the relative influence of climate vs.
+land-use change on contemporary species distributions.
+The question is relevant as the degree to which the two drivers are both important indicates potentially major errors in future projections previously published using species distribution models (SDMs).
+The authors present quite a clever method for addressing the issue: i.e., building SDMs with both historical and contemporary vegetation and climate data.
+This paper provides novel perspective concerning how our understanding of the role of climate change in vegetation change is still limited.
+The paper is exceedingly well written and the conclusions seem well-supported by the methods and results.
+
+We thank the reviewer for the positive feedback and constructive criticisms.  We address specific points below.
 
@@
 In particular, there likely needs to be a bit of clarification around the concept and a justification for why treating two non-significant effects and a set of mixed effects (one significant and one not so) as equivalent within this framework.
+
+To address this point, we’ve expanded Table 1 to indicate the individual terms (dV and dL) and their significance.
+This lets the reader see for themselves what goes into 'confounding'.
+We also added to the Introduction (4th paragraph) the sentence “These confounding effects may make it more difficult to detect climatic forcing of historical species distributions.”  As it turns out, in the primary analyses presented here, all instances of confounding involve a significant change in climate and no significant change in vegetation (Table 1)
 
@@
 
-Working on this. . .
+We have updated this figure to make the comparison clearer.
+
+![](figures/figure_two.png)
 
@@
 > The authors should use present tense not past tense in the methods and results sections.
+
+We have changed this throughout.
+We except descriptions of methods applied in earlier papers, or in the preparation of data prior to the current study (for example, when summarizing the methods of Goring et al.
+2016).
 

### 2016-12-02 Composition_Ecology.Rmd
I had to soft-reset a bunch of changes to unde the accidental inclusion of two large RDS files.
Changes include all revisions up to the last pass-off to @iceageecologist & include code changes to improve the modularity of the code.

@@
 
-# Effect of historical land-use and climate change on tree-climate relationships in the north-central United States
+# Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States
 
@@
 
-Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood.
-Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use and climate change.
+Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood.
+Here we show that tree-climate relationships for 15 common tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
@@
 Contemporary species occurrence databases, and, increasingly, paleohistorical datasets are used to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche].
-These distributions of species in geographic and environmental space, combined with future climate scenarios, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
-Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+These distributions of species in geographic and environmental space, combined with future climate scenarios, are used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
+Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
-Increasingly, the predictive capacity of ecological models is being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].
+Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling].
 Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
@@
 In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.
-In practice, however, the parameterization of these models is poorly constrained for processes operating at large temporal and spatial scales [@moorcroft2006close;@matthes2016benchmarking].
+In practice, however, the parameterization of these models is poorly constrained for processes operating at long time scales [@moorcroft2006close;@matthes2016benchmarking].
 
@@
 For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded].
-Confounding effects may cause little overall change on species distributions, with land-use pressure occurring across the species' environmental gradient.
-Predictive models for Garry oak (*Quercus garryana*) using only contemporary data [@pellatt2012using] may be confounded by land-use conversion that excludes Garry oak from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.
+Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient.
+For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.
+These confounding effects may hinder detection of climatic forcing of historical species distributions.
 Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
@@
 
-`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
+`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
 Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.
@@
 Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].
-In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].
-Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.).
+In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].
+Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change and land use, and we diagnose the relative influence of historical climate and land use changes on these shifts.
-We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.
+We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.
 We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.
 We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change.
-We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts.
+We assess whether climate and vegetation changes are compounding, counteracting or confounding.
 We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.
@@
 Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset].
-We briefly review the methods here.
-PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history].
-At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree.
+We briefly review these methods.
+PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history].
+At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees.
 Goring et al.
-[-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell.
+[-@goring2016preset] cleaned and aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell.
 This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset].
@@
 Nonetheless, differences persist.
-The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets.
-Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA.
+The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA plots are sited within forested or partially forested landscapes [@woudenberg2010forest].
+while the PLS sampling is uniform.
+Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the assignments of 'absence' within the FIA.
 However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general.
@@
 [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.
-FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset].
-PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
 
-Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.
-In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.
-For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.
+Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.
+Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.
+The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.
+Both the FIA and PLS datasets both have good coverage in the upper Midwest, with   8km grid cells with at least one FIA plot and XXX  grid cells with at least one PLS point.
+In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.
+We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA data (Supplementary Table  2).
+For all species, the range within the upper Midwest incompletly samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.
 
@@
 
-We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] to generate historical and modern climate data.
-The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.
+We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] for the historical and modern climate data.
+The PRISM climate variables are available at 800m resolution and are interpolated from station locations using elevation, aspect and other data.
 We resample estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) to the 8 x 8km grid used for PLS and FIA data.
-These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
-Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.
+These climate variables represent three major climatic controls on plant distributions: summer warmth, winter minimum temperatures, and moisture availability; these or similar bioclimatic variables are widely used in plant distributional modeling [@woodward1987climate;@iverson1998predicting;@maiorano2013building].
 
-Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.
+Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 and 1985 - 2014.
 There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.
 The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
-To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).
-Baker et al.
-[-@baker1985minnesota] show long term increases in mean annual temperature at Ft.
-Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas [@burnette2010daily] show increasing temperatures in all seasons, with faster increases during winter (0.10^o^C decade^-1^) than during summer (0.07^o^C decade^-1^).
-This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.
-The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study.
+To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].
+
+The use of alternative modern climate normals reflects two reasonable choices for whether to pick climate normals corresponding to time of sampling versus time of establishment.
+The 1985-2014 normal encompasses the period of sampling (1998 - 2015) for the FIA data.
+The 1950  - 1979 normal encompases the time of establishment of FIA trees, which have an average age of 50 years in this region based on the raw stand data.
+We report results for the 1985 - 2014 normals, with the 1950-1980 results show in in Supplmentary information.
 
@@
 
-Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) of tree species presence ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.
-Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions).
-Hellinger distance for two discrete probability distributions is defined as:
+Distributions of tree genera within climate space are shown using box plots and described as univariate probability density functions (PDFs) of tree species presence in vegetation ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.
+Differences among univariate PDFs are quantified for each climate variable using Hellinger distances [@matthes2016benchmarking] among the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions).
+The Hellinger distance for two discrete probability distributions is:
 
@@
 
-where $P$ and $Q$ are the distributions, with a common discrete index $i$.
+where $P$ and $Q$ are the distributions, with a common index $i$.
 
-The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-These conditions are observed or recorded in data.
-The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.
-The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.
-We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.
+The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.
+The effect of climate change on shifting PDFs is $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The effect of vegetation change on shifting PDFs is similarly described as $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
+Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with EuroAmerican settlement.
 
-We develop an attribution index, ($\delta_{d} = d_{c} - d_{v}$), such that negative values of $\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting tree distributions.
+We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.
 
-To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.
-For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to land-use change) distributions are significantly different.
+To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).
+For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.
 If both changes are significant and of the same sign, then the shift is compounding.
 If both are significant and in opposite directions then the shift is counteracting.
-If either or both changes are not significant, then the change is defined as confounding.
+If climate change is significant, but land use is not significant, then the change is defined as confounding.
+<!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed.
+-->
 
@@
 
-All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`.
+All analyses use R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`.
 All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate).
@@
 
-Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C since the 1895-1924 normal (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C.
-Regionally, $P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.
-The Palmer Drought Severity Index is closely coherent with $P_{ann}$ trends from 1895 onward and indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid-19^th^ century climates.
-When long term records [@burnette2010daily; @baker1985minnesota] are considered, $\Delta t_{min}$ increases to approximately `r round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `r round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `r round(clim_change[2], 1) - 0.15`^o^C.
+Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals).
+$P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.
+The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century precipitation is a reasonable proxy for mid-19^th^ century precipitation.
+Long term observer and military fort observations suggest warming since the early 19^th^ century, with increases in mean annual temperature at Fort Snelling, Minnesota [@baker1985minnesota] and Fort Leavenworth, Kansas [@burnette2010daily].
+The latter shows faster increases during winter (`r round(clim_change[4], 1) + 0.5`^o^C) than summer(`r round(clim_change[2], 1) + 0.35`^o^C).
+Hence, our estimates of historical climate change may be too conservative for $T_{min}$ and $T_{diff}$, and too large for $T_{max}$.
 
@@
 #  This takes a bit of time.
+if ('loss_melt.rds' %in% list.files('data/output/')) {
+  loss_melt <- readRDS('data/output/loss_melt.rds')
+} else {
   loss_melt <- loss_plot()
+  saveRDS(loss_melt, file = 'data/output/loss_melt.rds')
+}
 
@@
 
-For most tree genera, historical range losses dominate over range gains and are spatially structured (Table 1, Figure 3).
+For most tree genera, historical range losses dominate over gains and are spatially structured (Supplemental Table 1, Figure 3).
 Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
 For *Ulmus* and *Larix*, however, losses occur throughout the range.
-Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.
-Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.
+Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the upper Midwest.
+Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and average `r round(mean(loss_table$Gain),1)`% across all taxa.
 Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
-Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.
+*Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.
 Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
@@
 
-boxplot_clim <- clim_boxplots(clim_table = allveg,
+boxplot_allveg <- clim_boxplots(clim_table = allveg,
                               filename = "boxplot_corrected_all")
 
-boxplot_clim <- clim_boxplots(clim_table = newveg,
+boxplot_plss <- clim_boxplots(clim_table = newveg,
                               filename = "boxplot_corrected_pls")
@@
 
-The joint effects of historical climate change and land use on tree-climate distributions are illustrated for *Larix* (Figure 2).
-For *Larix* we see clear shifts in climate space when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change plays a strong role in shifting the distribution of *Larix* in climate space across all climate variables.
-Historical land use contributes to shifting distributions, particularly for $T_{max}$, where losses at the highest $T_{max}$ values are visible from pre-settlement to modern (Fig.
-2d).
-However, for *Larix*, land use overall appears to exert a weaker influence than climate change.
+The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (Figure 2).
+For *Quercus* we see tree-climate PDFs when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change is primarily responsible for shifting the distribution of *Quercus* in climate space.
+**DOUBLE CHECK THIS**
 
-Shifts in tree-climate distributions are significant for all genera and all climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons (all p < 0.0001).
-The directionality of shifts between historical and modern climate distributions is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ is higher (`r fig_nums("boxplot", display = "cite")`).
-Across all genera, modern $P_{ann}$ is higher than in the past.
-*Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while the smallest increase occurs for *Larix*.
-Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and declines least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
-*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), while the smallest change is for *Larix* ($\delta_{Pann}$ = 42mm).
+Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all p < 0.0001).
+The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).
+*Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while *Larix* has the smallest.
+Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
+*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm).
+These patterns are unchanged by sensitivity analysis in hich the historical temperature changes are augmentes by 1950 - 1990 temperature trends reported from the historical records (Supplemental Information).
 
@@
 hell_plot     <- hellinger_plot()
+# For suplemental data:
+hell_plot_pls <- hellinger_plot(newveg, 'hellinger_pls')
+hell_plot_fia <- hellinger_plot(topveg, 'hellinger_fia')
 
@@
 
-Both the total shift in realized tree-climate distributions and the attribution of these changes to historical land use or climate change varies strongly among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
+Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary  among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical land use on shifting climatic niches is important for many taxa, this shift is not uniform across taxa.
-The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than climate shifts.
+Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa.
+The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts **that are more due** to land-use than climate shifts.
 The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
 The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change.
-For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use, and the niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributable to changing climate.
+For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use.
+The niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.
 
@@
 
-The majority of climate - land-use interactions are compounding (n = 30) or confounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).
-Compounding interactions, where both climate and land use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space.
+**The majority of climate - land-use interactions are confounding (n = 30) or compounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).**  Compounding interactions, where both climate and land use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space.
 Compounding interactions are also strong for $P_{ann}$ and are caused by both land use and climate change shifting species to a wetter portion of climate space than they occupied historically.
@@
 Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
-Analysis included in the supplemental material indicates that a shift to an earlier baseline (using the estimated change rate from [@burnette2010daily] in Manhattan, Kansas) results in no change in this pattern (see "R/confounding_table.R" in S1).
+The patterns of confounding and compounding interactions are insenitive to the incorporation of estimates of 19^th^ century temperature changes (Supplementary Information Table 3).
+The apparent $T_{max}$ cooling and numer of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Suplementary Table 4).
 
@@
 
-*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions with $P_{ann}$, $T_{diff}$ and $T_{min}$, with compounding interactions with $T_{max}$ (`r tab_nums("compTable", display = "cite")`).
+*Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions for $P_{ann}$, $T_{diff}$ and $T_{min}$, and compounding interactions for $T_{max}$ (`r tab_nums("compTable", display = "cite")`).
 Each of these taxa experiences range losses, but patterns of gain and loss differ among taxa.
@@
 
+## Key Findings and Implications
+
 This work reinforces the importance of historical datasets for examining relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
@@
 Our analyses demonstrate that species-climate relationships are changing as a result of land use and climate change over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
-Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global].
+Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over this century and the next [@sala2000global].
 While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse;@rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
 
-Previous studies have shown that the realized climate niches of tree species shifted during the climate changes accompanying the last deglaciation.
-To a first order, many species accommodated past climate changes by shifting their ranges; *i.e.* they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
+Previous studies have shown that the realized climate niches of tree species shifted during the climate changes accompanying the last deglaciation [@pearman2008niche;@maiorano2013building;@veloz2012no].
+Many species accommodated past climate changes by shifting their ranges; *i.e.* they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.
 Range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
-However, species tracking of past climatic change was imperfect and other studies indicates shifts in species realized niches over time [@maiorano2013building;@pearman2008niche].
-Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], occupying non-analogue climates [@veloz2012no], or shifts in competition resulting from changing patterns of co-occurrence through time [@maiorano2013building].
+However, species tracking of past climatic change was imperfect, leading to shifts in species realized niches over time [@maiorano2013building;@pearman2008niche].
+Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], non-analogue climates [@veloz2012no], or shifts in competition resulting from changing species associations [@maiorano2013building].
 This study adds to prior work by showing that the land use changes of the last several centuries can also significantly modify species-climate relationships.
 
-In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.
-First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, declining summer temperatures, and higher precipitation.
-These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@estrada2013statistically].
-Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress].
+In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three processes.
+First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, higher precipitation and, perhaps, declining summer temperatures.
+These trends are produced by a combination of natural variability and anthropogenic change; the latter signal strengthens beginning in the second half of the 20^th^ century [@estrada2013statistically].
+Second, tree species distributions presumably have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress].
 Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges (Figure 1f).
-Agricultural conversion has largely eliminated open forests in warm and dry regions [@rhemtulla2007], while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times.
-Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient.
-While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of *Tsuga*.
+Agricultural conversion has greatly reduced the extent of savanna and open forests southern Wisconsin and Minnesota, causing tree species to occupy cooler habitats with higher precipitation than historically.
+While fire and logging altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for *Tsuga* (`r fig_nums("gain_loss", display="cite")`), a shade tolerant and late successional taxon.
 
-In addition to land use changes, the near total regional collapse of *Ulmus* distributions provides another case in which human agency (in this case, accidentally introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem], although reintroduction efforts are underway.
+In addition to land use changes, the near total regional collapse of *Ulmus* distributions provides another mechanism by which human agency (introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem].
 *Castanea dentata* was also effectively wiped out by chestnut blight, with reintroduction efforts underway, and *Tsuga canadensis* and *Fraxinus* spp.
-are now experiencing rapid mortality due to exotic pests.
-These pathogen-induced changes in tree distributions do not align neatly with the land use/climate attribution index described here, but they do fall under a broader set of vegetation changes caused by anthropogenic changesfactors, both inadvertent and intentional.
+are now experiencing rapid mortality due to exotic pests [@castello1995pathogens;@schlarbaum1998three].
+These pathogen-induced changes in tree distributions point to a broader set of vegetation changes caused by anthropogenic changes, both inadvertent and intentional.
 
-Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.
-In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
-If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
-Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
+Anthropogenic agency may also be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale;@groffman2014ecological].
+Human agency may also extend forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
+Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].
+
+Compounding and confounding interactions between historical climate change and land use have important implications for the management and genetics of forest communities.
+The observation that historical realized niches differ from modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
+Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 with $P_{ann}$ projections ranging from -17 to 110mm from a 1980s baseline [@wicci2011climate].
+Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates.
+The higher temperatures and lower precipitation values associated with taxa distributions for the pre-settlement period may suggest greater resilience to climate change than might be assessed using modern data only.
+This evidence for historical resilience to warmer and dryer climates is particularly critical for species that reach their range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).
+
+However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.
+In addition, adaptation to future climate change may be impeded by the historical reduction of locally adapted populations due to past land use change.
+Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation], so the historical loss of genotypes adapted to warmer and drier conditions may mean more rapid losses at the trailing edges of the species range.
+Similarly, long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] may be lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
+Net gene flow in populations along the remaining trailing edge may be weighted towards individuals maladapted to warmer conditions which could amplify the effects of climate change among natural populations [@kremer2012long].
+
+Correlational species-distribution models are likely to be the most affected by the use of contemporary observational datasets that are biased by historical climate and land-use change and climatic disequilibrium.
+However, even mechanistic models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.* @wullschleger2014plant].
+Hence, the predictive ability of both correlative and process based models might be sensitive to the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.
+One solution is to combine historical and contemporary data when building species distribution models [@maiorano2013building].
+
+## Potential Limitations
+
+Several potential limitations are associated with this study: 1) differences in sampling design between FIA and PLS datasets and effects on inferred tree distributions, 2) choice of time period for the historical climate normal, and 3) choice of time period  for the modern climate normal.
+We address each in turn and argue that the main finding of this paper – that tree-climate relationships have shifted over the last 100 years in response to historic land use and climate change – is robust.
+
+The first point is the most critical, because it implies that some of the reported range losses may be because the FIA dataset is underrepresenting tree distributions in unforested or semi-forested regions.
+If so, estimates of forest loss (Figure 3) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the upper Midwest.
+Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States (Bell et al.
+2014, Nieto-Lugilde et al.
+2015, Wang et al.
+2016).
 Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
 As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.
+These results are generally insensitive to choice of climatic normal for the historic and modern time periods, except for analyses related to Tmax.
+PDSI trends suggest small or now changes in precipitation over the 19th century (Fig.
+1), while the reported 19th-century increases in temperature mean that the estimates of changes in Tmin and Tdif reported here are conservative.
+Similarly, the patterns of change for t_min, t_diff and p_ann remain robust to choice of modern climate normal (1950 – 1980 vs.
+1995 - 2014) (Supplementary Information).
+The reported change in t_max, however, is reduced in both cases, suggesting that reported climate changes related to this variable are sensitive to choice of climatic normal.
+Note that the effects of historic land use on species-climate relationships are insensitive to choice of climate normal (because climate is held constant when calculating dv), so the effects of land use on species shifting to cooler portions of tmax (Figure 5, Table 1),  Independent analyses using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships following Euro-American settlement in the region (St.
+Jacques et al.
+2008), which further indicates that vegetation-climate relationships have changed since settlement.
 
-Artifacts of the choice of climate normal (1995 - 2014) for the modern data mean an exaggerated effect for $t_{max}$, however, supplemental analysis shows that the patterns of change for the other variables, $t_{min}$, $t_{diff}$ and $p_{ann}$ remain robust.
-Analysis using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships [@jacques2008pre] following Euro-American settlement in the region providing further support that modern vegetaton-climate relationships have changed since settlement.
+## Summary
 
-Anthropogenic agency may also be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale;@groffman2014ecological].
-Human agency may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
-Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].
-However, usage of species distributional data from highly managed ecosystems must account for the intensive management that may be required to maintain species beyond their realized niche.
-
-Compounding and confounding interactions between historical climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.
-On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
-Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate] and $P_{ann}$ projections range from -17 to 110mm by 2055 relative to 1980s baselines, based on fourteen global climate models from the Climate Model Intercomparison Project [@wicci2011climate].
-Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates.
-These higher temperatures and lower precipitation across taxa distributions during the pre-settlement period indicate the potential for greater resilience to climate change than might be assessed using modern data only.
-This evidence for historical resilience to warmer and dryer climates may be particularly critical for species that reach their northern or western range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).
-
-However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.
-In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of adapted populations due to land use change and loss of genetic diversity.
-Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation].
-Hence, the loss of genotypes adapted to warmer and drier conditions (the trailing edge of distributions for many taxa in the Upper Midwest) caused by land use may mean more rapid losses at the trailing edges of the species range.
-Long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
-Thus net gene flow in populations along the remaining trailing edge will be from the central and leading populations, individuals maladapted to warmer conditions.
-This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].
-
-Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by the use of contemporary observational datasets that are biased by historical climate and land-use change and climatic disequilibrium.
-However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.* @wullschleger2014plant].
-In this case, both correlative and process based models might be affected by the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.
-
-In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
-The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
+Historical datasets and multitemporal studies such as this oneadd to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
+Land use over the past two centuries and on-going legacies, combined with the likelihood that tree distributions are lagging recent climate changes, create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
+The main conclusions of this study are robust to changes in vegetation sampling design over time and choice of historic and modern climatic normals, although both concerns are important and deserve further study.
 Given the importance of modern distributional datasets for predicting species range shifts, compounding interactions between historical climate and land use change may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.

### 2016-12-02 SupplementaryInformation.Rmd
I had to soft-reset a bunch of changes to unde the accidental inclusion of two large RDS files.
Changes include all revisions up to the last pass-off to @iceageecologist & include code changes to improve the modularity of the code.

---
title: "Untitled"
author: "Simon Goring"
date: "November 29, 2016"
output: word_document
---

# Spatial Autocorrelation

We include the effects of autocorrelation in the t-tests using the `SpatialPack` function `modified.ttest`. Unfortunately, the `modified.ttest` only performs a pairwise t-test.  Pairwise t-tests are inappropriate for these analyses, because, for the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison.  In this case, only the overlapping points would be included, and it would show zero change since climate is unchanged.

To overcome this challenge, we obtained a modified estimate of degrees of freedom (df) that accounts for spatial autocorrelation based on Moran's I from `modified.ttest` for each of the two spatial processes. To then produce a difference of the means that is not pairwise, we obtain the corrected df from `modified.ttest`, and recalculate the p-value for the t-test using the t estimate from the `t.test`. The corresponding code is available in [`confounding_table.R`](https://github.com/PalEON-Project/Composition_Climate/blob/master/R/confounding_table.R).

# Climate changes:  1895-1924 vs. 1950-1980

If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased XXoC between 1895-1924 and 1950-1980, T_max changed by XXoC, while T_diff changed by XXoC

```{r}

```

**Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era. Taxa are sorted by the percent area lost (Column 3).  In this analysis, all PLS and FIA grid cells were used and a taxon was assumed have been gained/lost for a given grid cell if that grid cell had data in one time period but not the other.

```{r, results = 'as-is', warning=FALSE, message=FALSE, echo=FALSE}

loss_melt <- readRDS('data/output/loss_melt.rds')
loss_table <- as.data.frame.matrix(table(loss_melt[[1]]$variable,
                                         loss_melt[[1]]$value))
loss_table <- data.frame(loss_table, 
                         Modern = round((loss_table$Gain + loss_table$Presence) * .064, 0),
                         Historical = round((loss_table$Loss + loss_table$Presence ) * .064, 0))

loss_table[,1:3] <- round(analogue::tran(loss_table[,1:3], method = "proportion") * 100,1)

sci_names <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")

rownames(loss_table) <- paste0("*", sci_names, "*")

knitr::kable(loss_table)

```

**Supplemental Table 2**:  As Supplemental Table 1, except that in this analysis, all estimates are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Figure 1.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results='as-is'}

if ('s2loss_table.rds' %in% list.files('data/output')) {
  s2loss_melt <- readRDS('data/output/s2loss_table.rds')
} else {

  s2loss_melt <- loss_plot(overlap = TRUE)
  saveRDS(s2loss_melt, file = 'data/output/s2loss_table.rds')
}

s2loss_table <- as.data.frame.matrix(table(s2loss_melt[[1]]$variable,
                                         s2loss_melt[[1]]$value))
s2loss_table <- data.frame(s2loss_table, 
                         Modern = round((s2loss_table$Gain + s2loss_table$Presence) * .064, 0),
                         Historical = round((s2loss_table$Loss + s2loss_table$Presence ) * .064, 0))

s2loss_table[,1:3] <- round(analogue::tran(s2loss_table[,1:3], method = "proportion") * 100,1)

sci_names <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")

rownames(s2loss_table) <- paste0("*", sci_names, "*")

knitr::kable(s2loss_table, justify = "left", style = "rmarkdown",
                     emphasize.rownames = FALSE)

```

**Supplemental Table 3**:  As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al. 2010, Baker et al. 1985).  See Supplemental Figure 2 for more information.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results = 'as-is'}

conf_pls <- readRDS('data/output/conf_pls.RDS')

knitr::kable(conf_pls)
```

**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1980 instead of 1990-2014.  See also Supplemental Figures 3, 4.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results = 'as-is'}

conf_fia <- readRDS('data/output/conf_fia.RDS')

knitr::kable(conf_fia)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

s2loss_melt[[2]]

```

**Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Table 2.

![](figures/boxplot_corrected_pls.png)

**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of using 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e. the time of the PLS surveys).  No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate warming, with estimates of 19th century warming at Fort Leavenworth, Kansas of 0.10oC decade-1 in the winter and 0.07oC decade-1 in the summer (Burnette et al. 2010).  Given this and the ~50 year offset between PLS and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35oC for t_max, +0.5oC for t_min, and +0.15oC for t_diff. This would imply that the reported regional increase in T_min between historical to modern periods (3.3^o^C, Fig. 1) should be further increased to 3.8oC, the reported regional decrease in T_max  (-0.8^o^C) should be reduced to -0.45oC), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C). Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for Tmin and Tdiff and too large for Tmax.  The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig. 4.

![](figures/boxplot_corrected_fia.png)

**Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.

![](figures/hellinger_pls.png)

**Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.

![](figures/hellinger_fia.png)

### 2016-12-02 R/confounding_table.R

@@ -150,6 +150,6 @@ get_conf <- function(x) {
 }
 
 conf_table <- get_conf(tests)
-conf_pls   <- get_conf(tests_pls)
-conf_fia   <- get_conf(tests_fia)
-conf_all   <- get_conf(tests_all)
+conf_pls   <- get_conf(tests_pls); saveRDS(conf_pls, file = 'data/output/conf_pls.RDS')
+conf_fia   <- get_conf(tests_fia); saveRDS(conf_fia, file = 'data/output/conf_fia.RDS')
+conf_all   <- get_conf(tests_all); saveRDS(conf_all, file = 'data/output/conf_all.RDS')

### 2016-12-02 R/figures/gain_loss_plot.R

@@ -1,11 +1,15 @@
 #  Plotting for taxa that shows the change between historical and modern distributions.
 
-get_loss_gain <- function(unit_raster, best.taxa, pls_data, agg_dens){
+get_loss_gain <- function(unit_raster, best.taxa, pls_data, agg_dens, overlap = FALSE){
   loss.gain <- data.frame(cell = 1:ncell(unit_raster))
   
   agg_dens[is.na(agg_dens)] <- 0
   
-  for(i in 1:length(best.taxa)) {
+  if (overlap == TRUE) {
+    pls_data <- pls_data %>% filter(cell %in% agg_dens$cell)
+  }
+  
+  for (i in 1:length(best.taxa)) {
     loss.gain[,(i + 1)] <- NA
     colnames(loss.gain)[i + 1] <- best.taxa[i]
     
@@ -28,7 +32,7 @@ get_loss_gain <- function(unit_raster, best.taxa, pls_data, agg_dens){
   loss.melt
 }
 
-loss_plot <- function(){
+loss_plot <- function(overlap = FALSE){
   
   if(model_proj == '+init=epsg:4326'){
     ext <- c(-98, -83, 42, 50)
@@ -38,7 +42,7 @@ loss_plot <- function(){
     ext <- c(-100000, 1050000, 600000, 1600000)
   }
   
-  loss_melt <- get_loss_gain(unit_raster, best.taxa, pls_data, agg_dens)
+  loss_melt <- get_loss_gain(unit_raster, best.taxa, pls_data, agg_dens, overlap = overlap)
 
   loss_melt$rgb <- brewer.pal(n = 3, 'Dark2')[as.numeric(factor(loss_melt$value, 
                                                                 levels = c('Presence', 'Loss', 'Gain')))]

### 2016-12-02 R/hellinger_dists.R

@@ -22,14 +22,14 @@ hellinger <- function(clim1, clim2){
   
 }
 
-hellinger_plot <- function(){
+hellinger_plot <- function(climtable = vegclim_table, filename = 'hellinger'){
   
-  vegclim_table$taxon <- mapvalues(vegclim_table$taxon, from = levels(vegclim_table$taxon),
+  climtable$taxon <- mapvalues(climtable$taxon, from = levels(climtable$taxon),
                                 to = c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
                                        "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
                                        "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus"))
   
-  best.taxa <- as.character(unique(vegclim_table$taxon))
+  best.taxa <- as.character(unique(climtable$taxon))
   
   all.hellinger <- ldply(1:length(best.taxa), function(x){
         ldply(1:4, function(y){
@@ -37,7 +37,7 @@ hellinger_plot <- function(){
           tree     <- best.taxa[x]
           clim_var <- climate$name[y]
           
-          data.subset <- vegclim_table %>% filter(taxon == tree & climate %in% clim_var & data > 0)
+          data.subset <- climtable %>% filter(taxon == tree & climate %in% clim_var & data > 0)
           
           VHCH <- data.subset %>% filter(base == 'PLSS' & c.ref == 'PLSS') %>% select(clim) %>% unlist
           VHCM <- data.subset %>% filter(base == 'PLSS' & c.ref == 'FIA') %>% select(clim) %>% unlist
@@ -85,9 +85,9 @@ hellinger_plot <- function(){
              label = 'Land Use Dominates', family = 'serif',
              fontface = 'bold', size = 3)
   
-  ggsave(plot = hell_plot, filename = 'figures/hellingerplot.png', 
+  ggsave(plot = hell_plot, filename = paste0('figures/',filename,'.png'), 
          width = 6.7, height = 6, dpi = 300)
-  ggsave(plot = hell_plot, filename = 'Final_Figures/hellingerplot.pdf', 
+  ggsave(plot = hell_plot, filename = paste0('Final_Figures/',filename,'.pdf'), 
          width = 6.7, height = 6, dpi = 300)
   
   list(hell_plot, all.hellinger)

### 2016-12-02 Composition_Ecology.Rmd
Inclusion of @iceageecologist's changes, and my first big pass through with edits.

@@
 
-(143 words)
-
 Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood.
-Here we show that tree-climate relationships for 15 common tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change.
+We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
@@
 These distributions of species in geographic and environmental space, combined with future climate scenarios, are used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for climate-driven shifts in habitat suitability [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting].
-Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
+Forest inventories offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
 
 Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling].
-Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses of species to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
-Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
+Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
+Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
 In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.
@@
 ```
-![](figures/Fig_1_Composite.png)
+![](Final_Figures/figure_one.tiff)
 
-`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.
-Modern climate shows lower $T_{max}$, but higher $T_{min}$ relative to early-19^th^ Century normals, along with increasing $P_{ann}$.
+`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions with a LOESS smooth superimposed for illustration.
+Modern climate (normals indicated by rectangles).
 PDSI shows strong coherence with $P_{ann}$ in this region.
@@
 
-Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
-Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
+Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
+Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
 Rhemtulla et al.
@@
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern US over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.
 We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.
@@
 The effect of vegetation change on shifting PDFs is similarly described as $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
-Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with EuroAmerican settlement.
+Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with Euro-American settlement.
 
@@
 
-```{r echo = FALSE}
-tab_nums("tab_gain_loss", "Gain and loss (%) of the analyzed tree genera since the pre-settlement era, based on presence or absence within each 8x8km grid cell.
-Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.
-The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.
-Modern and historical occurence are reported, followed by percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
-Modern and Historical indicate represent range area (100s of km^2^) occupied by the tree taxon for the pre-settlement or modern era.
-Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
-
-```
-
 For most tree genera, historical range losses dominate over gains and are spatially structured (Supplemental Table 1, Figure 3).
@@
 
-`r tab_nums("tab_gain_loss", display = "full")`
-
 ```{r table1, results="asis", echo=FALSE, eval=TRUE}
@@
 rownames(loss_table) <- paste0("*", sci_names, "*")
-
-pander::pandoc.table(loss_table, justify = "left", style = "rmarkdown",
-                     emphasize.rownames = FALSE)
 ```

### 2016-12-02 Composition_Ecology.Rmd
Revisions ready to pass back to @iceageecologist.
Generally minor text changes, some changes to the `bib` file.  Filled in values,

Cleaned up the first `if` statement in `checker` in `confounding_table`.

Removed sets of files as well to clean up the repo.

@@
 
-* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land use, niche shift, Public Land Survey System, realized niche
+* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land-use, niche shift, Public Land Survey System, realized niche
 
@@
 Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood.
-We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change.
+We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land-use and climate change.
 Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
-A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables.
+A new attribution method implicates both historical climate change and land-use in these shifts, with the relative importance varying among genera and climate variables.
 Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions.
@@
 Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling].
-Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence].
+Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land-use, and other historical legacies [@svenning2015influence].
 Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
@@
 
-Historical land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
-Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
+Historical land-use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche].
+Anthropogenic land-use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].
 Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].
@@
 Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient.
-For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.
+For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land-use has not systematically shifted Garry oak distributions in climate space.
 These confounding effects may hinder detection of climatic forcing of historical species distributions.
-Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species.
+Counteracting effects occur when climate and land-use change act in opposite directions, narrowing the available geographic and climatic space available for species.
 For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.
@@
 (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed.
-Land use data from the NLCD [@jin2013comprehensive].*")`
+land-use data from the NLCD [@jin2013comprehensive].*")`
 
-Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
+Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
 Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization].
@@
 [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin.
-Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].
+Wildland ecosystems have been converted to agricultural and urban land-use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].
 In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].
@@
 
-Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern US over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.
+Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern US over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land-use changes and pathogen outbreaks (for *Ulmus*) on these shifts.
 We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.
@@
 Goring et al.
-[-@goring2016preset] cleaned and aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell.
+[-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell.
 This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset].
@@
 A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*.
-We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.
+We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.
 
-Modern forest data are from the US FIA program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest].
-The FIA uses a nationally standardized sampling procedure that has evolved over time.
-In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
+Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest].
+The FIA uses a nationally standardized sampling procedure; in the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].
 
@@
 Nonetheless, differences persist.
-The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA plots are sited within forested or partially forested landscapes [@woudenberg2010forest].
+The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data, and FIA plots are sited within forested or partially forested landscapes.
 while the PLS sampling is uniform.
@@
 The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.
-Both the FIA and PLS datasets both have good coverage in the upper Midwest, with   8km grid cells with at least one FIA plot and XXX  grid cells with at least one PLS point.
+Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5172 8km grid cells with at least one FIA plot and 7376 grid cells with at least one PLS point.
 In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.
@@
 
-Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 and 1985 - 2014.
+Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data.
+We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplmentary Information.
 There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.
@@
 
-The use of alternative modern climate normals reflects two reasonable choices for whether to pick climate normals corresponding to time of sampling versus time of establishment.
-The 1985-2014 normal encompasses the period of sampling (1998 - 2015) for the FIA data.
-The 1950  - 1979 normal encompases the time of establishment of FIA trees, which have an average age of 50 years in this region based on the raw stand data.
-We report results for the 1985 - 2014 normals, with the 1950-1980 results show in in Supplmentary information.
-
 ## Shifts in Species Distributions in Climate Space
 
-![](figures/figureFour_panels.tiff)
+![](Final_Figures/figure_two.tiff)
 
@@
 The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.
-The effect of climate change on shifting PDFs is $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-The effect of vegetation change on shifting PDFs is similarly described as $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
-Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with Euro-American settlement.
+The effect of climate change on shifting PDFs ($d_{c}$) is the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
+Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.
 
-We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.
+We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land-use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.
 
-To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).
+To assess whether the effects of historical land-use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).
 For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.
@@
 If both are significant and in opposite directions then the shift is counteracting.
-If climate change is significant, but land use is not significant, then the change is defined as confounding.
+If climate change is significant, but land-use is not significant, then the change is defined as confounding.
 <!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed.
@@
 
-For most tree genera, historical range losses dominate over gains and are spatially structured (Supplemental Table 1, Figure 3).
+All tree genera show higher losses than gains (Supplemental Table 1, Figure 3).
 Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3).
@@
 Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and average `r round(mean(loss_table$Gain),1)`% across all taxa.
-Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
-*Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.
+Spatial patterns of gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).
+*Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plot and stand heterogeneity, rather than systematic range gains.
 Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.
@@
 
-```{r table1, results="asis", echo=FALSE, eval=TRUE}
+```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
 
@@
                "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")
+rownames(loss_table) <- paste0("*", sci_names, "*")
 
-rownames(loss_table) <- paste0("*", sci_names, "*")
-```
-
-```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
 source("R/resampling_climate.R")
@@
 
-`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States.
-In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile.
-These shifts in realized climate distributions are produced by the joint effects of historical climate change and land-use change.")`
-
-The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (Figure 2).
-For *Quercus* we see tree-climate PDFs when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change is primarily responsible for shifting the distribution of *Quercus* in climate space.
-**DOUBLE CHECK THIS**
+`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States produced by the joint effects of historical climate change and land-use change.
+The central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile.")`
 
@@
 Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
+The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (`r fig_nums("fourPanel", display = "cite")`).
+For *Quercus* we see changes due to both both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the atrtibution index (`r fig_nums("hellPlot", display = "cite")`a).
+
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
 Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa.
-The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts **that are more due** to land-use than climate shifts.
-The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
+The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts that are due more to land-use than climate shifts (`r fig_nums("hellPlot", display = "cite")`a).
+The greatest land-use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
 The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change.
-For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use.
+For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land-use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land-use.
 The niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.
@@
 ```{r tests, warning=FALSE, message=FALSE, echo=FALSE, results="asis", eval=TRUE}
+
 source("R/confounding_table.R")
@@
 
-pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
+colnames(conf_table) <- c("Taxon", "$P_{ann}$", "$T_{max}$", "$T_{diff}$", "$T_{min}$")
+
+conf_table[,1] <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
+               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
+               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")
+
+knitr::kable(conf_table)
 
@@
 
-**The majority of climate - land-use interactions are confounding (n = 30) or compounding (n = 28), while few (n = 2) show counteracting effects (`r tab_nums("compTable", display = "cite")`).**  Compounding interactions, where both climate and land use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space.
-Compounding interactions are also strong for $P_{ann}$ and are caused by both land use and climate change shifting species to a wetter portion of climate space than they occupied historically.
-These analyses are insensitive to the type of correction applied to the *p*-value, although running the analysis without a correction factor reduces the counts of insignificant (confounding) relationships and increases the number of  counteracting and compounding interactions.
-Patterns of interaction, by species, for $P_{ann}$ and $T_{diff}$ are significantly correlated (${\chi^{2}}_{4}$ = 15.22), but are not significantly correlated for other climate variables.
+The majority of climate-vegetation interactions are confounding ($-$; n = 30) or compounding ($\\Box$; n = 23).
+None show counteracting effects (`r tab_nums("compTable", display = "cite")`).
+Compounding interactions, where both climate and land-use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space ($-_c$).
+Compounding interactions are also strong for $P_{ann}$ and are caused by both land-use and climate change shifting species to a wetter portion of climate space ($+_c$) than they occupied historically.
 The patterns of confounding and compounding interactions are insenitive to the incorporation of estimates of 19^th^ century temperature changes (Supplementary Information Table 3).
 The apparent $T_{max}$ cooling and numer of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Suplementary Table 4).
-
-Counteracting interactions, where climate shifts and land use shifts show opposite signs relative to baseline climate, effectively narrowing the realized climatic niche from both directions, exist only for *Ulmus* ($T_{max}$) and *Larix* ($T_{min}$).
 
@@
 This work reinforces the importance of historical datasets for examining relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
-Modern surveys sample only a subset of historical distributions, both spatially and within multidimensional climate space.
-Our analyses demonstrate that species-climate relationships are changing as a result of land use and climate change over the past few centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
-Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over this century and the next [@sala2000global].
+Modern surveys sample a subset of historical distributions, both spatially and within multidimensional climate space.
+Our analyses demonstrate that species-climate relationships are changing as a result of land-use and climate change over the past two centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+land-use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over this century and the next [@sala2000global].
 While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse;@rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
@@
 Range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].
-However, species tracking of past climatic change was imperfect, leading to shifts in species realized niches over time [@maiorano2013building;@pearman2008niche].
-Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], non-analogue climates [@veloz2012no], or shifts in competition resulting from changing species associations [@maiorano2013building].
-This study adds to prior work by showing that the land use changes of the last several centuries can also significantly modify species-climate relationships.
+Mechanisms for realized niche shifts include species lagging rapid climate change [@ordonez2013realized], non-analogue climates [@veloz2012no], or shifts in competition resulting from changing species associations [@maiorano2013building].
+This study adds to prior work by showing that the land-use changes of the last several centuries can also significantly modify species-climate relationships.
 
-In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three processes.
+In this study, shifts in realized climate niches and the underlying interactions between climate change and land-use emerge from three processes.
 First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, higher precipitation and, perhaps, declining summer temperatures.
@@
 Second, tree species distributions presumably have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress].
-Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges (Figure 1f).
-Agricultural conversion has greatly reduced the extent of savanna and open forests southern Wisconsin and Minnesota, causing tree species to occupy cooler habitats with higher precipitation than historically.
+Third, anthropogenic land-use has selectively eliminated portions of geographic and environmental space from species' potential ranges (Figure 1f).
+Agricultural conversion has greatly reduced the extent of savanna and open forests in southern Wisconsin and Minnesota, causing tree species to occupy cooler habitats with higher precipitation than historically.
 While fire and logging altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for *Tsuga* (`r fig_nums("gain_loss", display="cite")`), a shade tolerant and late successional taxon.
 
-In addition to land use changes, the near total regional collapse of *Ulmus* distributions provides another mechanism by which human agency (introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem].
+In addition to land-use changes, the near total regional collapse of *Ulmus* distributions provides another mechanism by which human agency (introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem].
 *Castanea dentata* was also effectively wiped out by chestnut blight, with reintroduction efforts underway, and *Tsuga canadensis* and *Fraxinus* spp.
 are now experiencing rapid mortality due to exotic pests [@castello1995pathogens;@schlarbaum1998three].
-These pathogen-induced changes in tree distributions point to a broader set of vegetation changes caused by anthropogenic changes, both inadvertent and intentional.
+Pathogen-induced changes in tree distributions point to a broader set of vegetation changes caused by anthropogenic changes, both inadvertent and intentional.
 
 Anthropogenic agency may also be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale;@groffman2014ecological].
-Human agency may also extend forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.
+Human agency may also extend forest ecosystems at historical range margins, or preserve historical range margins from land-use conversion to agriculture as windrows, or isolated woodlots.
 Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].
 
-Compounding and confounding interactions between historical climate change and land use have important implications for the management and genetics of forest communities.
+Compounding and confounding interactions between historical climate change and land-use have important implications for the management and genetics of forest communities.
 The observation that historical realized niches differ from modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
@@
 
-However, projected temperature rises by 2100 far exceed the shifts due to historical land use and climate change.
-In addition, adaptation to future climate change may be impeded by the historical reduction of locally adapted populations due to past land use change.
+Projected temperatures for 2100 far exceed the shifts due to historical land-use and climate change.
+In addition, adaptation to future climate change may be impeded by the historical reduction of locally adapted populations due to past land-use change.
 Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation], so the historical loss of genotypes adapted to warmer and drier conditions may mean more rapid losses at the trailing edges of the species range.
-Similarly, long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] may be lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.
+Similarly, long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] may be lost when land-use conversion to agricultural production results in the extirpation of locally adapted populations.
 Net gene flow in populations along the remaining trailing edge may be weighted towards individuals maladapted to warmer conditions which could amplify the effects of climate change among natural populations [@kremer2012long].
@@
 Several potential limitations are associated with this study: 1) differences in sampling design between FIA and PLS datasets and effects on inferred tree distributions, 2) choice of time period for the historical climate normal, and 3) choice of time period  for the modern climate normal.
-We address each in turn and argue that the main finding of this paper – that tree-climate relationships have shifted over the last 100 years in response to historic land use and climate change – is robust.
+We address each in turn and argue that the main finding of this paper – that tree-climate relationships have shifted over the last 100 years in response to historic land-use and climate change – is robust.
 
-The first point is the most critical, because it implies that some of the reported range losses may be because the FIA dataset is underrepresenting tree distributions in unforested or semi-forested regions.
+The first point is the most critical.
+Some of the reported range losses may be the result of incomplete sampling of tree distributions in unforested or semi-forested regions.
 If so, estimates of forest loss (Figure 3) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the upper Midwest.
-Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States (Bell et al.
-2014, Nieto-Lugilde et al.
-2015, Wang et al.
-2016).
-Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
+Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@bell2014early;@nieto2015close;@wang2016landscape].
+Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments based on FIA data.
 As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.
-These results are generally insensitive to choice of climatic normal for the historic and modern time periods, except for analyses related to Tmax.
-PDSI trends suggest small or now changes in precipitation over the 19th century (Fig.
-1), while the reported 19th-century increases in temperature mean that the estimates of changes in Tmin and Tdif reported here are conservative.
-Similarly, the patterns of change for t_min, t_diff and p_ann remain robust to choice of modern climate normal (1950 – 1980 vs.
-1995 - 2014) (Supplementary Information).
-The reported change in t_max, however, is reduced in both cases, suggesting that reported climate changes related to this variable are sensitive to choice of climatic normal.
-Note that the effects of historic land use on species-climate relationships are insensitive to choice of climate normal (because climate is held constant when calculating dv), so the effects of land use on species shifting to cooler portions of tmax (Figure 5, Table 1),  Independent analyses using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships following Euro-American settlement in the region (St.
-Jacques et al.
-2008), which further indicates that vegetation-climate relationships have changed since settlement.
+
+Our results are generally insensitive to the choice of climatic normal, except for analyses related to $T_{max}$.
+Reported 19th-century increases in temperature mean that the estimates of changes in $T_{min}$ and $T_{diff}$ are likely conservative.
+Similarly, patterns of change for $T_{min}$, $T_{diff}$ and $P_{ann}$ remain robust to choice of modern climate normal (1950 -– 1980 vs.
+1995 -- 2014) (Supplementary Information).
+The reported change in $T_{max}$, however, is reduced in both cases, suggesting that reported climate changes related to this variable are sensitive to choice of climatic normal.
+Note that the effects of historic land-use on species-climate relationships are insensitive to choice of climate normal (because climate is held constant when calculating $d_v$), so the effects of land-use on species shifting to cooler portions of $T_{max}$ (Figure 5, Table 1),  Independent analyses using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships following Euro-American settlement in the region [@jacques2008pre], which further indicates that vegetation-climate relationships have changed since settlement.
 
@@
 
-Historical datasets and multitemporal studies such as this oneadd to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
-Land use over the past two centuries and on-going legacies, combined with the likelihood that tree distributions are lagging recent climate changes, create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
+Historical datasets and multitemporal studies such as this one add to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
+land-use over the past two centuries and on-going legacies, combined with the likelihood that tree distributions are lagging recent climate changes, create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
 The main conclusions of this study are robust to changes in vegetation sampling design over time and choice of historic and modern climatic normals, although both concerns are important and deserve further study.
-Given the importance of modern distributional datasets for predicting species range shifts, compounding interactions between historical climate and land use change may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.
+Given the importance of modern distributional datasets for predicting species range shifts, compounding interactions between historical climate and land-use change may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.
 Both empirical and process-based vegetation models need to combine, whenever possible, both contemporary and historical information on tree distributions, tree-climate relationships, and their respective shifts through time.

### 2016-12-02 SupplementaryInformation.Rmd
Revisions ready to pass back to @iceageecologist.
Generally minor text changes, some changes to the `bib` file.  Filled in values,

Cleaned up the first `if` statement in `checker` in `confounding_table`.

Removed sets of files as well to clean up the repo.

@@
 
-If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased XXoC between 1895-1924 and 1950-1980, T_max changed by XXoC, while T_diff changed by XXoC
-
-```{r}
-
-```
+If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased 0.5^o^C between 1895-1924 and 1950-1980, T_max changed by 0.35^o^C, while T_diff changed by -0.15^o^C
 

### 2016-12-02 R/confounding_table.R

@@ -106,47 +106,41 @@ tests_all <- do.call(rbind.data.frame,
 checker <- function(x) {
   # x will be the row:
   
+  csig <- p.adjust(as.numeric(x[4]), method = "bonferroni", n = 120) < 0.05
   
-  if (p.adjust(as.numeric(x[4]), method = "bonferroni", n = 120) < 0.05 & 
-      p.adjust(as.numeric(x[6]), method = "bonferroni", n = 120) < 0.05) {
-    
-    # both p values are significant:
-    if ((as.numeric(x[3]) * as.numeric(x[5])) > 0) {
-      # A simple test to see if they're both positive
-      return(ifelse(as.numeric(x[3]) > 0, '$\\Box$ ($+_c$, $+_v$)', '$\\Box$ ($-_c$,$-_v$)'))
-      
-    } else {
-      # They are in different directions:
-      return(ifelse(as.numeric(x[3]) > 0, '$\\bigcirc$ ($+_c$,$-_v$)', '$\\bigcirc$ ($-_c$,$+_v$)'))
-    }
-    
+  if (csig == TRUE & as.numeric(x[3]) > 0) { clim <- "$+_c$" }
+  if (csig == TRUE & as.numeric(x[3]) < 0) { clim <- "$-_c$" }
+  if (csig == FALSE) { "$._c$" }
+  
+  veg  <- ifelse(p.adjust(as.numeric(x[6]), method = "bonferroni", n = 120) < 0.05, 
+                 ifelse(as.numeric(x[3]) > 0, "$+_v$", "$-_v$"),
+                 "$._v$")
+  
+  if (clim == "$._c$") {
+    # There's not change in climate.
+    symb <- "*ns*"
   } else {
-    # If one or the either is not significant:
-    if (as.numeric(x[4]) > (0.05 / 120) & as.numeric(x[6]) > (0.05 / 120)) {
-      # If neither element shows significant change
-      return('n/a ($._c$, $._v$)')
+    if (substr(clim, 1, 3) == substr(veg, 1, 3)) {
+      # They're equivalent and significant.
+      symb <- "$\\Box$"
     } else {
-      if (as.numeric(x[4]) < (0.05 / 120)) {
-        # If climate change is significant, but the land use change is not significant.
-        if (as.numeric(x[3]) > 0) {
-          return('n/a ($+_c$, $._v$)')
-        } else {
-          return('n/a ($-_c$, $._v$)')
-        }
+      if (veg == "$._v$") {
[…]

### 2016-12-02 Composition_Ecology.Rmd
Edited thru SuppInfo, light edits to main text

@@
 * Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
-* John W Williams: Co-conceived paper, and co-wrote the paper
+* John W Williams: Co-conceived paper and co-wrote the paper
 
@@
 ---------------------
-The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
+The authors would like to thank the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
 In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group.

### 2016-12-02 SupplementaryInformation.Rmd
Edited thru SuppInfo, light edits to main text

--- 2016-12-02 SupplementaryInformation.Rmd
@@
 ---
-title: "Untitled"
+title: "Supplementary Information"
 author: "Simon Goring"
@@
 Unfortunately, the `modified.ttest` only performs a pairwise t-test.
-Pairwise t-tests are inappropriate for these analyses, because, for the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison.
-In this case, only the overlapping points would be included, and it would show zero change since climate is unchanged.
+Pairwise t-tests are inappropriate for these analyses, because, for the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison, only the overlapping points would be included, and it would show zero change since climate is unchanged.
 
@@
 
-# Climate changes:  1895-1924 vs. 1950-1980
+# Climate changes between 1895-1924 and 1950-1980  (vs. main paper analysis of 1895-1924 and 1990-2014)
 
-If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased XXoC between 1895-1924 and 1950-1980, T_max changed by XXoC, while T_diff changed by XXoC
+If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: ($T_{min}$) increased 0.5^o^C between 1895-1924 and 1950-1980, ($T_{max}$) increased by 0.35^o^C, while T_diff changed by -0.15^o^C
 
-```{r}
-
-```
-
+# Supplemental Tables
 **Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era.
-The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
-Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.
+The first three columns (Gain, Loss, Presence) are expressed as percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras.
+Modern and Historical indicate represent range area (units: 10^2^ km^2^) occupied by the tree taxon for the pre-settlement or modern era.
 Taxa are sorted by the percent area lost (Column 3).
@@
 
-**Supplemental Table 2**:  As Supplemental Table 1, except that in this analysis, all estimates are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.
-See also Supplemental Figure 1.
+**Supplemental Table 2**:  As Supplemental Table 1, except that in this analysis, all estimates are restricted to the subset of 8 km gridcells that had at least one observation in both the PLS and FIA datasets.
+For most taxa, the percentages are similar (within a few percentage points) between Supplemental Tables 1 and 2, except for Quercus, which has a substantially lower estimate of percent area lost (ST1: 46% vs.
+ST2: 25.7%)  See also Supplemental Figure 1.
 
@@
 
-**Supplemental Table 3**:  As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al.
+<!-- COM: scientific names needed in ST3 -->
+
+**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al.
 2010, Baker et al.
 1985).
-See Supplemental Figure 2 for more information.
+All entries in this table are identical to those in Table 1, indicating that the assessments of interaction strength are insensitive to this adjustment to mid-19th-century climates.
+See Supplemental Figure 2 for further details about this analysis.
 
@@
 
+<!-- COM: ST4: scientific names needed -->
+<!-- COM: ST4: Correct column names and formatting for climate variables.  e.g. ppt should be PAnn -->
+<!-- COM: Correct formatting for variable names in text.  I have tried to do this where I have seen them, but may have missed some variables.   -->
 **Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1980 instead of 1990-2014.
+The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged.
+For ($T_{max}$), the sign of the climate change flips from negative (cooling) to positive (warming), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.
+(Note that directionality of ($T_{min}$) is unchanged, consistent with 20^th^- and 21^st^-century warming).
 See also Supplemental Figures 3, 4.
@@
 
+# Supplemental Figures
 **Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.
@@
 
-**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of using 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e.
-the time of the PLS surveys).
-No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate warming, with estimates of 19th century warming at Fort Leavenworth, Kansas of 0.10oC decade-1 in the winter and 0.07oC decade-1 in the summer (Burnette et al.
+**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e.
+a climate normal closer to the time of the PLS surveys).
+No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate 19th-century warming, with estimates of  warming at Fort Leavenworth, Kansas of 0.10oC decade^-1^ in the winter and 0.07oC decade^-1^ in the summer (Burnette et al.
 2010).
-Given this and the ~50 year offset between PLS and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35oC for t_max, +0.5oC for t_min, and +0.15oC for t_diff.
-This would imply that the reported regional increase in T_min between historical to modern periods (3.3^o^C, Fig.
-1) should be further increased to 3.8oC, the reported regional decrease in T_max  (-0.8^o^C) should be reduced to -0.45oC), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C).
-Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for Tmin and Tdiff and too large for Tmax.
+Given this and the ~50 year offset between PLS observations and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35^o^C for ($T_{max}$), +0.5^o^C for ($T_{min}$), and +0.15^o^C for t_diff.
+This would imply that the reported regional increase in ($T_{min}$) between historical to modern periods (3.3^o^C, Fig.
+1) should be further increased to 3.8^o^C, the reported regional decrease in ($T_{max}$)  (-0.8^o^C) should be reduced to -0.45^o^C), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C).
+Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for ($T_{min}$) and ($T_{diff}$) and too large for ($T_{max}$).
 The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig.
-4.
+4, suggesting that the findings of this paper are largely insensitive to this adjustment to account for 19th-century rises in temperature.
 
@@
 **Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.
+The results for ($T_{min}$) and ($P_{ann}$) are largely unchanged between this analysis and that shown in Fig.
+4, indicating that the finding of tree-climate distribution shifts is robust to choice of modern cliamte normal.
+However, ($T_{max}$) shows much closer overlap between the PLS and FIA periods, suggesting that the reported difference in tree-climate distributions for ($T_{max}$) is sensitive to choice of modern climate normal.
 
@@
 **Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.
+Few differences are apparent between this figure and Figure 5, suggesting that the attribution of tree-climate distribution shifts to changes in climate versus changes in vegetation is insensitive to choice of modern climate normal.
 

### 2016-12-02 Composition_Ecology.Rmd
Finished read-through of main text, pushing edits

@@
 
-* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land-use, niche shift, Public Land Survey System, realized niche
+* Anthropocene, climate change, climate disequilibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land-use, niche shift, Public Land Survey System, realized niche
 
@@
 # Abstract
-
-Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood.
+<!-- removed reference to 'higher Tmax' given results of sensitivity analyses.  Also took away the statement about 'reinforces shift to cooler conditions, again given ' 146 words after rewording-->
+Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate relationships are poorly understood.
 We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land-use and climate change.
-Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall.
+Realized niches have shifted towards higher minimum temperatures and higher rainfall.
 A new attribution method implicates both historical climate change and land-use in these shifts, with the relative importance varying among genera and climate variables.
-Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions.
-The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.
+Most climate/land-use interactions are compounding, in which historical land-use reinforces  shifts in species-climate relationships toward wetter distributions, or confounding, in which land use complicates shifts towards warmer distributions.
+Compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.
 
@@
 Forest inventories offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
-
+<!-- deleted Ordonez 2013 reference to shorten references -->
 Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling].
 Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land-use, and other historical legacies [@svenning2015influence].
-Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
+Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes.
 In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.
-In practice, however, the parameterization of these models is poorly constrained for processes operating at long time scales [@moorcroft2006close;@matthes2016benchmarking].
+In practice, however, the parameterization of these models is poorly constrained for slow ecological processes [@moorcroft2006close;@matthes2016benchmarking].
 
@@
 Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.
-For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded].
+For example, warming has been compounded by land-use conversion at the lowest elevations in the Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded].
 Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient.
@@
 land-use data from the NLCD [@jin2013comprehensive].*")`
-
+<!-- WICCI reference below:  Fix author in bibtex to WICCI, i.e. (WICCI, 2011) -->
 Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
@@
 We assess whether climate and vegetation changes are compounding, counteracting or confounding.
-We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.
+We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based solely on modern distributional data.
 
@@
 PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history].
-At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees.
+At each survey point, surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees.
 Goring et al.
-[-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell.
+[-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (122 trees) per cell.
 This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset].
 Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera.
-A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*.
-We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.
+A few common names, such as "Ironwood" can not be clearly resolved to a single tree genus, and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*.
+We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in geographic or climate space of all cells with recorded presence for a specific taxon.
+<!-- inserted 'geographic or' because we often refer to range gains and losses in the paper -->
 
@@
 Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence].
-Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera.
+Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically, by aggregating taxa to the level of genera.
 Nonetheless, differences persist.
@@
 The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.
-Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5172 8km grid cells with at least one FIA plot and 7376 grid cells with at least one PLS point.
-In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.
-We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA data (Supplementary Table  2).
-For all species, the range within the upper Midwest incompletly samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.
+Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5,172 8km grid cells with at least one FIA plot and 7,376 grid cells with at least one PLS point.
+
+In mapping tree distributions in the upper Midwest we consider three classes: "gain," where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss," where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence," where a tree was present within a grid cell in both FIA and PLS eras.
+We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA observations (Supplementary Table  2).
+For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.
 
@@
 
-Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data.
-We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplmentary Information.
-There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.
+Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternatives for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data.
+We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplementary Information.
+There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatiotemporal resolution for this time period for all climate variables.
 The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].
@@
 ![](Final_Figures/figure_two.tiff)
-
+<!-- Is Figure 2 for Larix or Quercus?  Legend here says Larix, later in text says Quercus -->
 `r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships.
@@
 To assess whether the effects of historical land-use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).
-For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.
+For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different (Figure 2).
 If both changes are significant and of the same sign, then the shift is compounding.
@@
 Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals).
-$P_{ann}$ has increased by `r round(clim_change[1], 1)`mm.
-The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century precipitation is a reasonable proxy for mid-19^th^ century precipitation.
-Long term observer and military fort observations suggest warming since the early 19^th^ century, with increases in mean annual temperature at Fort Snelling, Minnesota [@baker1985minnesota] and Fort Leavenworth, Kansas [@burnette2010daily].
+$P_{ann}$ increased by `r round(clim_change[1], 1)`mm.
+The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.
+Long term observer and military fort observations suggest warming during the 19^th^ century, with increases in mean annual temperature at Fort Snelling, Minnesota [@baker1985minnesota] and Fort Leavenworth, Kansas [@burnette2010daily].
 The latter shows faster increases during winter (`r round(clim_change[4], 1) + 0.5`^o^C) than summer(`r round(clim_change[2], 1) + 0.35`^o^C).
-Hence, our estimates of historical climate change may be too conservative for $T_{min}$ and $T_{diff}$, and too large for $T_{max}$.
+Hence, the PRISM-based estimates of historical climate change may be too conservative for $T_{min}$ and $T_{diff}$, and too large for $T_{max}$.
 
@@
 *Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historical range, or beyond it.
+These patterns of range gains and losses are generally insensitive to choice of all gridcells versus the subset of gridcells with data in both time periods, except for *Quercus*, which has a much lower estimate of range loss when only the subset is used (46% vs.
+25.7%, Figure 3, Table 1, Supplementary Figure 1, Supplementary Table 2).
 
@@
 *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm).
-These patterns are unchanged by sensitivity analysis in hich the historical temperature changes are augmentes by 1950 - 1990 temperature trends reported from the historical records (Supplemental Information).
+These patterns are unchanged by sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).
 
@@
 For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
-
+<!-- Does Fig. 2 show Larix or Quercus?  This paragraph states Quercus; legend states Larix -->
 Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
 The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (`r fig_nums("fourPanel", display = "cite")`).
-For *Quercus* we see changes due to both both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the atrtibution index (`r fig_nums("hellPlot", display = "cite")`a).
+For *Quercus* we see changes due to both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the attribution index $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`a).
 
+<!-- Note that figure numbering is still off - Fig 6 should be Fig 5.  Needs fixing prior to submission-->
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
-Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa.
-The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts that are due more to land-use than climate shifts (`r fig_nums("hellPlot", display = "cite")`a).
+Attribution analyses based on $\Delta_{d} (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa.
+The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen precipitation niche shifts that are slightly more attributable to land-use than climate shifts (`r fig_nums("hellPlot", display = "cite")`a).
 The greatest land-use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$).
@@
 For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land-use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land-use.
-The niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.
+The $T_{min}$ shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.
 
@@
 ```
-
-The majority of climate-vegetation interactions are confounding ($-$; n = 30) or compounding ($\\Box$; n = 23).
-None show counteracting effects (`r tab_nums("compTable", display = "cite")`).
-Compounding interactions, where both climate and land-use shifts push species distributions in the same direction along the climate gradient, are particularly strong for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space ($-_c$).
+<!-- Given new Table 1, have rewritten to state that all climate-veg interaction are compounding or confounding -->
+<!-- Last sentence in para may need tweaking, if Supp Table 4 is changed - see comments there. -->
+All climate-vegetation interactions are compounding ($\\Box$; n = 30) or confounding ($-$; n = 30), with no counteracting effects found (`r tab_nums("compTable", display = "cite")`).
+Compounding interactions, where both climate and land-use shifts push species distributions in the same direction along the climate gradient, are strongest for $T_{max}$ (`r tab_nums("compTable", display = "cite")`), and are consistent with the observed trend of cooling $T_{max}$ (`r fig_nums("land_use_clim", display = "cite")`, `r fig_nums("boxplots", display = "cite")`) and losses concentrated in the southern portions of most species ranges (`r fig_nums("gain_loss", display = "cite")`); both tend to shift tree-climate distributions to cooler portions of climate space ($-_c$).
 Compounding interactions are also strong for $P_{ann}$ and are caused by both land-use and climate change shifting species to a wetter portion of climate space ($+_c$) than they occupied historically.
-The patterns of confounding and compounding interactions are insenitive to the incorporation of estimates of 19^th^ century temperature changes (Supplementary Information Table 3).
-The apparent $T_{max}$ cooling and numer of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Suplementary Table 4).
+The patterns of confounding and compounding interactions are insensitive to alternative estimates of 19^th^ century temperature changes (Supplementary Information Table 3).
+However, the apparent $T_{max}$ cooling and number of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Supplementary Table 4).
 
@@
 These taxa all show significant losses in range along the southern and western margins of their distribution (`r fig_nums("gain_loss", display="cite")`), with few gains in those regions.
-
 *Fraxinus*, *Tsuga* and *Tilia* all show confounding interactions for $P_{ann}$, $T_{diff}$ and $T_{min}$, and compounding interactions for $T_{max}$ (`r tab_nums("compTable", display = "cite")`).
@@
 
-This work reinforces the importance of historical datasets for examining relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
+This work reinforces the utility of historical datasets for examining relationships between species and climate, and their importance for forecasting future species distributions [@moritz2013future;@maguire2015modeling].
 Modern surveys sample a subset of historical distributions, both spatially and within multidimensional climate space.
-Our analyses demonstrate that species-climate relationships are changing as a result of land-use and climate change over the past two centuries, and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
-land-use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over this century and the next [@sala2000global].
-While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse;@rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
+Our analyses demonstrate that species-climate relationships are changing as a result of land-use and climate change over the past two centuries, and add to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat].
+Land-use change reduces the correlational structure between species and climate [@devineau2011arid] and has been one of the strongest drivers of changes in global biodiversity over the 20th and 21st centuries [@sala2000global].
+While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse;@rhemtulla2009legacies], the legacies of land-use are likely to persist, both in terms of forest composition [@goring2016preset;@kujawa2016theeffect], and the correlational structure between species and climate.
 
@@
 In this study, shifts in realized climate niches and the underlying interactions between climate change and land-use emerge from three processes.
-First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, higher precipitation and, perhaps, declining summer temperatures.
+First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, higher precipitation, and variable summer temperatures.
 These trends are produced by a combination of natural variability and anthropogenic change; the latter signal strengthens beginning in the second half of the 20^th^ century [@estrada2013statistically].
@@
 Agricultural conversion has greatly reduced the extent of savanna and open forests in southern Wisconsin and Minnesota, causing tree species to occupy cooler habitats with higher precipitation than historically.
-While fire and logging altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for *Tsuga* (`r fig_nums("gain_loss", display="cite")`), a shade tolerant and late successional taxon.
+While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for *Tsuga* (`r fig_nums("gain_loss", display="cite")`), a shade-tolerant and late-successional taxon.
 
-In addition to land-use changes, the near total regional collapse of *Ulmus* distributions provides another mechanism by which human agency (introduced Dutch elm disease) has affected distributions through near extirpation [@barnes1979stem].
+In addition to land use, the regional collapse of *Ulmus* distributions points to the importance of forest pathogens as mechanism by which human agency (introduced Dutch elm disease) has affected tree-climate relationships [@barnes1979stem].
 *Castanea dentata* was also effectively wiped out by chestnut blight, with reintroduction efforts underway, and *Tsuga canadensis* and *Fraxinus* spp.
 are now experiencing rapid mortality due to exotic pests [@castello1995pathogens;@schlarbaum1998three].
-Pathogen-induced changes in tree distributions point to a broader set of vegetation changes caused by anthropogenic changes, both inadvertent and intentional.
+Hence, land use is not the sole driver of vegetation changes reported here, although it the dominant one.
 
-Anthropogenic agency may also be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale;@groffman2014ecological].
-Human agency may also extend forest ecosystems at historical range margins, or preserve historical range margins from land-use conversion to agriculture as windrows, or isolated woodlots.
-Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].
-
+<!-- Fix the WICCI ref, as above. -->
 Compounding and confounding interactions between historical climate change and land-use have important implications for the management and genetics of forest communities.
-The observation that historical realized niches differ from modern niches suggests that some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
+Some tree genera may have more resilience to climate change than expected based on contemporary observations alone.
 Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 with $P_{ann}$ projections ranging from -17 to 110mm from a 1980s baseline [@wicci2011climate].
-Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates, and $P_{ann}$ from 42 to 86mm lower than modern estimates.
-The higher temperatures and lower precipitation values associated with taxa distributions for the pre-settlement period may suggest greater resilience to climate change than might be assessed using modern data only.
+Every tree genera examined in this study had historical $T_{max}$ values from 0.4 to 1.3^o^C higher than modern estimates (primarily attributable to land use change, Figure 5), and $P_{ann}$ from 42 to 86mm lower than modern estimates (primarily attributable to climate change, Figure 5).
+The higher temperatures and lower precipitation associated with pre-settlement tree distributions may suggest greater resilience to climate change than might be assessed using modern data only.
 This evidence for historical resilience to warmer and dryer climates is particularly critical for species that reach their range limits in the upper Midwest (e.g., *Tsuga*, *Fagus*).
 
-Projected temperatures for 2100 far exceed the shifts due to historical land-use and climate change.
-In addition, adaptation to future climate change may be impeded by the historical reduction of locally adapted populations due to past land-use change.
+Conversely, adaptation to future climate change may be impeded by the historical reduction of locally adapted populations due to past land-use change.
 Local adaptation is an important component of tree responses to climate change [@aitken2008adaptation], so the historical loss of genotypes adapted to warmer and drier conditions may mean more rapid losses at the trailing edges of the species range.
-Similarly, long distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] may be lost when land-use conversion to agricultural production results in the extirpation of locally adapted populations.
-Net gene flow in populations along the remaining trailing edge may be weighted towards individuals maladapted to warmer conditions which could amplify the effects of climate change among natural populations [@kremer2012long].
+Similarly, long-distance gene flow of adaptive traits to central and leading populations from the trailing edge [@hu2006seed] may be lost when land-use conversion to agricultural production results in the extirpation of locally adapted populations.
+Net gene flow in populations along the remaining populations may be weighted towards individuals maladapted to warmer conditions, which could amplify the effects of climate change at the trailing edge [@kremer2012long].
 
@@
 However, even mechanistic models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.* @wullschleger2014plant].
-Hence, the predictive ability of both correlative and process based models might be sensitive to the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.
+Hence, the predictive ability of both correlative and process based models should be sensitive to the shifting climatic niches demonstrated here, and the underlying processes of spatially biased land-use conversion, changing climates, and climatic disequilibrium.
 One solution is to combine historical and contemporary data when building species distribution models [@maiorano2013building].
@@
 Several potential limitations are associated with this study: 1) differences in sampling design between FIA and PLS datasets and effects on inferred tree distributions, 2) choice of time period for the historical climate normal, and 3) choice of time period  for the modern climate normal.
-We address each in turn and argue that the main finding of this paper – that tree-climate relationships have shifted over the last 100 years in response to historic land-use and climate change – is robust.
+We address each in turn and explain why the main finding of this paper – that tree-climate relationships have shifted over the last few centuries in response to historic land-use and climate change – is robust.
 
-The first point is the most critical.
-Some of the reported range losses may be the result of incomplete sampling of tree distributions in unforested or semi-forested regions.
+The first point is the most critical, because some of the reported range losses may be an artifact of incomplete FIA sampling of tree distributions in unforested or semi-forested regions.
 If so, estimates of forest loss (Figure 3) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the upper Midwest.
+However, the range gain/loss/presence estimates are not qualitatively affected by analysis of all gridcells (Figure 3, Table 1) versus the subset of gridcells with both PLS and FIA data (Supplementary Information).
+However, it remains possible that FIA plots have a lower detectability than PLS data because of the fewer plots per 8km gridcell.
 Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@bell2014early;@nieto2015close;@wang2016landscape].
 Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments based on FIA data.
-As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.
 
-Our results are generally insensitive to the choice of climatic normal, except for analyses related to $T_{max}$.
-Reported 19th-century increases in temperature mean that the estimates of changes in $T_{min}$ and $T_{diff}$ are likely conservative.
+Our results are generally insensitive to the choice of climatic normal (potential limtitations 2 & 3), except for analyses related to $T_{max}$.
+The reported 19th-century warming from US forts means that the PRISM-based estimates of changes in $T_{min}$ and $T_{diff}$ are conservative.
 Similarly, patterns of change for $T_{min}$, $T_{diff}$ and $P_{ann}$ remain robust to choice of modern climate normal (1950 -– 1980 vs.
 1995 -- 2014) (Supplementary Information).
-The reported change in $T_{max}$, however, is reduced in both cases, suggesting that reported climate changes related to this variable are sensitive to choice of climatic normal.
-Note that the effects of historic land-use on species-climate relationships are insensitive to choice of climate normal (because climate is held constant when calculating $d_v$), so the effects of land-use on species shifting to cooler portions of $T_{max}$ (Figure 5, Table 1),  Independent analyses using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships following Euro-American settlement in the region [@jacques2008pre], which further indicates that vegetation-climate relationships have changed since settlement.
+The reported change in $T_{max}$, however, is affected in both cases, suggesting that actual $T_{max}$ trends are small and sensitive to choice of climatic normal.
+Note that the effects of historic land-use on species-climate relationships are insensitive to choice of climate normal (because climate is held constant when calculating $d_v$), so the effects of land-use on species shifting to cooler portions of $T_{max}$ (Figure 5, Table 1) is not affected by choice of climate normal for $T_{max}$.
+Independent analyses using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships following Euro-American settlement in the region [@jacques2008pre], which further indicates that vegetation-climate relationships have changed since settlement.
 
@@
 
-Historical datasets and multitemporal studies such as this one add to our ability to understand the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
-land-use over the past two centuries and on-going legacies, combined with the likelihood that tree distributions are lagging recent climate changes, create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
+Historical datasets and multitemporal studies such as this one add to our understanding of the stability and dynamic of species-climate distributions and their adaptive potential in the face of climate change.
+Land use over the past two centuries and its legacies, combined with the likelihood that tree distributions are lagging recent climate changes, create the strong likelihood that tree-climate distributions in the upper Midwest have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.
 The main conclusions of this study are robust to changes in vegetation sampling design over time and choice of historic and modern climatic normals, although both concerns are important and deserve further study.
@@
 ---------------------
-The authors would like to thank the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
+The authors would like to thank the large number of individuals who have worked to undertake the Public Lands System Survey, to compile the original surveyor notes, to digitize and standardize the survey results, and finally, to assist in interpreting and compiling the data in its present form.
 In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group.
-The authors than Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript.
+The authors thank Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript.
 This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.

### 2016-12-02 SupplementaryInformation.Rmd
Finished read-through of main text, pushing edits

@@
 
-# Climate changes between 1895-1924 and 1950-1980  (vs. main paper analysis of 1895-1924 and 1990-2014)
+# Climate changes between 1895-1924 and 1950-1979  (vs. main paper analysis of 1895-1924 and 1990-2014)
 
-If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: ($T_{min}$) increased 0.5^o^C between 1895-1924 and 1950-1980, ($T_{max}$) increased by 0.35^o^C, while T_diff changed by -0.15^o^C
+If a 1950-1979 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1979, the regional climate changes are as follows: ($T_{min}$) increased 0.5^o^C between 1895-1924 and 1950-1979, ($T_{max}$) increased by 0.35^o^C, while T_diff changed by -0.15^o^C
 
@@
 <!-- COM: Correct formatting for variable names in text.  I have tried to do this where I have seen them, but may have missed some variables.   -->
-**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1980 instead of 1990-2014.
+<!-- COM: For TMax, why does the sign of the vegetation change flip from negative to positive between Table 1 and this Supp Table 4?
+I can understand why the sign of the climate change would alter as we play around with different climate normals, but the sign of the vegetation change should be stable.
+The other climate variables seem to be unaffected.
+This may be the bug you mentioned by email.
+-->
+**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1979 instead of 1990-2014.
 The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged.
@@
 
-**Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.
+**Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1979 instead of 1990-2014.
 The results for ($T_{min}$) and ($P_{ann}$) are largely unchanged between this analysis and that shown in Fig.
@@
 
-**Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.
+**Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1979 instead of 1990-2014.
 Few differences are apparent between this figure and Figure 5, suggesting that the attribution of tree-climate distribution shifts to changes in climate versus changes in vegetation is insensitive to choice of modern climate normal.

### 2016-12-03 Response_to_reviewers.Rmd
Revised response to reviews, minor edits to main text

@@
 
-We submit for your consideration the revised manuscript "Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States."
+We submit for your consideration the revised manuscript "*Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States.*"
 
-We have carefully revised the manuscript in response to reviewer comments and have paid particular attention to the potential limitations associated with 1) comparing the PLS and FIA data and 2) choice of climate normal.
+We have carefully revised the manuscript in response to reviewer comments.
+In these revisions, we have paid particular attention to the potential limitations raised by the reviewers regarding the 1) comparing the PLS and FIA data and 2) choice of climate normal.
 We completed several new analyses to test the sensitivity of our results to choice of climate normal and created a separate subsection within the Discussion assessing the robustness of our findings.
-
-We have edited the manuscript for clarity and to respond to the other points raised by the reviewers.
+<!-- be sure to fill in the word count -->
+We have also edited the manuscript for clarity and to respond to the other points raised by the reviewers.
 We include below a point-by-point response to all reviews.
 The revised version has XX  words and 6 figures and tables.
+The main text of revised manuscript is now slightly beyond the 5,000 word limit, so please advise whether we should trim further to meet the target.
 We have expanded the Supplementary Information to show several additional analyses.
@@
 This lets the reader see for themselves what goes into 'confounding'.
-We also added to the Introduction (4th paragraph) the sentence “These confounding effects may make it more difficult to detect climatic forcing of historical species distributions.”  As it turns out, in the primary analyses presented here, all instances of confounding involve a significant change in climate and no significant change in vegetation (Table 1)
+We also added to the Introduction (4th paragraph) the sentence “*These confounding effects may hinder detection of climatic forcing of historical species distributions.*”  As it turns out, in the primary analyses presented here, all instances of confounding involve a significant change in climate and no significant change in vegetation (Table 1)
 
@@
 
-This Goring et al. paper is now accepted.  We have added the following paragraph:
+This Goring et al. paper is now accepted and in press.  We have added the following paragraph:
 
-*FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset].
-PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].
-Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.
-We have selected the 15 most common tree species in the region .
-.
-.
-*
+*Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.
+Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.
+The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.
+Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5,172 8km grid cells with at least one FIA plot and 7,376 grid cells with at least one PLS point.*
+
+And have added a section on 'Potential Limitations' to the Discussion.
 
@@
 
-We have changed this sentence in particular to read: "*The difference between the PLSS and modern tree distributions ($d_{V}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest*"" and changed other references to $d_{l}$ to $d_{v}$.
+We have changed all references from $d_{l}$ to $d_{v}$.
 
@@
 We've changed the table to make things more clear.
+'Confounding' is now restricted to situations when there is significant effects of climate change on the tree distributions but no significant effect of land use.
 
@@
 
-We have added the paragraph:  *For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.* to indicate this limitation.
+We have added the paragraph:  *For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.* to address this point.
 
@@
 
-We have added the sentence: *While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of* Tsuga.
+We have added the sentence: *While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for * Tsuga * (`r fig_nums("gain_loss", display="cite")`), a shade-tolerant and late-successional taxon.*
 
@@
 
-I have corrected the names so that only scientific names are displayed.
+We have corrected the names so that only scientific names are displayed.
 
@@
 
-Referee 2 was more critical of the paper than Reviewer 1.
+> This paper compares tree/climate relationships between two time periods and concludes that modern forest data sets may not be adequate for quantifying the realized niches of tree species because tree populations may not be in equilibrium with environment under rapid land use and climate change and the data sets have limitations in estimating tree presence and absence.
+This is a very important point and the paper will likely be of very high interest.
+This comparison of pre-settlement times with modern times is unique in the species distribution modeling literature and nicely demonstrates the key points.
+
+> The paper is well written and the graphics are generally very good.
+
+We thank Reviewer 2 for the positive feedback and constructive criticism.  We have performed several new analyses and revised existing analyses in response.  We describe these further below.
 
@@
 We thank the reviewer for these comments.
-We believe that we now more fully address the limitations -- We have added references to long term weather records and used their early 19th century estimates as the basis for some sensitivity analysis (see below), we have used a moving window to examine the effect of 21st century climate change on the choice of the modern climate normal.
-We have also strengthed the section that indicates prior support for widespread land use change, and its impacts on tree species distributions in the region.
-Lastly, in the **Discussion** we have reitterated what we believe are some of the limitations of the current study.
+We now more fully address the limitations through multiple solutions.
+1) We have added references to long term weather records and used their early 19th century estimates as the basis for sensitivity analysis (see below).
+2) We have experimented with two modern climate normals, representing time of sampling and time of establishment.
+3) We strengthened the literature reiew section that indicates prior support for widespread land use change, and its impacts on tree species distributions in the region.
+4) Lastly, in the Discussion we have reitterated what we believe are some of the limitations of the current study.
 
@@
 
-We have replaced instances of "northern" with north-central.
+Here and elsewhere, we have adopted ‘upper Midwestern US’ throughout ms. as the most spatially precise geographic place name.
 
@@
 
+Please see our response above to Reviewer 1 and the expanded treatment of this question in the Methods and Discussion.
+
 > 181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.
@@
 
-We have added the sentences: *These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].
-Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.*
-
-**JWW to add Woodward - moisture availability -**
+We have added the sentences: *These climate variables represent three major climatic controls on plant distributions: summer warmth, winter minimum temperatures, and moisture availability; these or similar bioclimatic variables are widely used in plant distributional modeling [@woodward1987climate;@iverson1998predicting;@maiorano2013building].*
 
@@
 
-We have added context to this choice and the implications.
-Using long term records reported in Baker et al (1985) and Burnett et al.
-(2010) we indicate that the long term trend is overall warming in the region.
-We had previously used PDSI since this was the only available long term record we had access to.
-This change resulted in the inclusion of the folowing senteces in the Methods:
+We respectfully disagree that it is a serious stretch to use 1895-1915 climates to represent 1830-1910 climates.
+The PDSI data shows no evidence of long-term trend between 1800 and 1910, based on a LOESS line fitted to the time series shown in Fig.
+1 (see other responses below).
 
-*Baker et al.
-(1985) show long term increases in mean annual temperature at Ft.
-Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas Burnette et al.
-(2010) show increasing temperatures in all seasons, with faster increases during winter (0.10oC decade-1) than during summer (0.07oC decade-1).
-This implies that historical estimates may be biased by +0.35oC for $t_{max}$, by +0.5oC for $t_{min}$ and +0.15oC for $t_{diff}$.*
+To further address this point, we have added a review of early meteorological records from US military forts.
+These records do indicate some warming in the region (Baker et al, 1985, Burnett et al.
+2010).
+We have added the following text to the first paragraph of the results.
 
-and in the Results:
+*Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al.
+1985) and Fort Leavenworth, Kansas (Burnette et al.
+2010).
+The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1).
+Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).*
 
-*When long term records (Burnette et al., 2010; Baker et al., 1985) are considered, $\Delta t_{min}$ increases to approximately `round(clim_change[4], 1) + 0.5`^o^C, $\Delta t_{max}$ to `round(clim_change[2], 1) + 0.35`^o^C and $\Delta t_{diff}$ to `round(clim_change[2], 1) - 0.15`^o^C.*
+We have used these findings as the basis of a simple sensitivity analysis to the Supplementary Information (Supplementary Figure 3) in which adjust the historic climate data and historic-modern differences based on 19th-century trends observed in the US fort data.
+This analysis is summarized by the following new sentence in the results:
 
-Interestingly, these shifts in PLS baseline climate do not result in significant changes to the overall climate envelope for the tree species (compare the paper figure with the figure below), indicating the influence of land use is much larger than for climate for most taxa, and supercedes the shifts from the early 19th to the early 20th century.
-
-![](figures/boxplot_corrected_pls.png)
-
-In addition, the change, while large, remains small enough that the shifts related to land use remain larger than changes due to climate, and the directionality of all changes remains constant.
-We have added the code into the supplemental material.
-
-In the results we note:
-
-*Analysis included in the supplemental material indicates that a shift to an earlier baseline (using the estimated change rate from [@burnette2010daily] in Manhattan, Kansas) results in no change in this pattern (see "R/confounding_table.R" in S1).*
+*These patterns are unchanged by sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).*
 
@@
 
-We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks, we believe that our contribution further underscores this point.
+We have addressed this point by adding an analysis in which we use climate normals for 1950-1980 for the modern climate instead of 1990-2014.
+We report the results from the latter in the main paper, and have added several new figures to the supplementary information.
+We’ve also analyzed tree age in the FIA data, which shows a mean age of 52, 56, and 49 years in WI, MN, and MI.
+Based on this, the 1950-1980 is a good alternative time window for the time of tree establishment.
+We also note that this paper is not intended as an exercise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks.
+We’ve added the following text to the manuscript:
 
-The normals for the region, beginning with a normal ending 1980 and continuing to a normal ending 2014, give a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.
-These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively, but, as pointed out in Baker *et al*.
-(2016; doi:10.1111/gcb.13273) uncertainty in the underlying estimates are often overlooked.
-Given the variability within windows, and the relatively young age of most stands in WI, MN and MI (FIA data shows mean ages of 52, 56 and 49 years respectively), we believe that the choice of normal is acceptable.
-Having said that, we examined the importance using both shifts in PLS era, and shifts in the modern climate normal.
-The largest effect is in $t_{max}$, since the most recent normals all show a significant increase in $t_{max}$.
-
-We have added the sentence: *"The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study."*
-
-![](figures/boxplot_corrected_all.png)
+*The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.
+The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here.
+The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region.
+We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.*
 
@@
 
-I believe that this is addressed in the response to the prior comment.
-We have added a LOESS curve to Figure 1, we have selected a normal that reflects the period of sampling, and have provided additional support to the choice of the modern normal.
+We believe that this point is now addressed in responses to other comments.
+We have added a LOESS curve to Figure 1 (in response to a later  comment), we have selected an alternate modern climate normal that represents the period of establishment, and have provided additional support to the choice of the modern normal.
+Note too that we specifically chose 30-year climatologies in order to smooth out the effects of interannual and interdecadal-scale variability on our results.
 
@@
 We apreciate this point and thank the reviewer for bringing it up.
-The challenge in this case is that a multivariate SDM blends the climate variables.
-Further, a multi-model SDM approach might weight climate variables differently across models, through time, making sensitivity analysis for the impact of each climate variable challenging.
+Please note that this analysis is not intended to focus on the construction of SDMs, but rather to show the joint effects of historical land use and climate change on tree-climate relationships.
+We respectfully believe that a multivariate SDM would obscure these shifting relationships by blending together the climate variables.
+Further, a multi-model SDM approach might weight climate variables differently across models, through time.
+We believe that there is value and insight in focusing on individual climate variables and the distribution of trees along individual axes.
 
-We point out that this analysis is not intended to focus on the construction of SDMs, but rather to show the relationship between land use and the individual climate variables, through the distribution of tree species.
-Thus, disentagling land use and climate shifts on a set of multivariate combination of variables from each time slice becomes extremely challenging, and, is effectively a separate paper.
-
-What this paper intends to point to is that land use is impacting climate variables differentially across species as a result of land use, and that this may have implications for SDM construction.
-Shifts in the strength of SDMs between time scales and potential utility of climate variables in SDM construction is a separate analysis and paper.
+Shifts in the strength of SDMs between time scales and potential utility of climate variables in SDM construction is worthwhile, but a separate analysis and paper.
 
@@
 
-We have included spatial autocorrelation in the t-tests using the `SpatialPack` function`modified.ttest`.
-Unfortunately, the `modified.ttest` only performs a pairwise t-test, which is inappropriate in this situation since we would not be able to assess the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison since only the overlapping points would be included and we would assess zero change.
-
-The `modified.ttest` calculates a t value, but also calculates a modified $df$ that takes into account spatial autocorrelation based on Moran's I for each of the two spatial processes.
-To then produce a difference of the means that is not pairwise, we obtain the corrected $df$ from `modified.ttest`, and recalculate the $p$-value for the t-test using the t estimate from the t.test.
-This is all coded in the `confounding_table.R` file.
+We have now included spatial autocorrelation in this analysis, and describe the methods in the Supplementary Information.
 
@@
 We have added a LOESS curve onto Figure 1, representing a span of 20 years, along with the SE fit.
-We have added light shading over the normal boxes to highlight the normal period.
-As indicated above, we have provided a more detailled analysis of the choice of normal periods and described the effect of the choice on outcomes.
+We have added light shading over boxes corresponding to the chosen climate normals.
+As indicated above, we have provided a more detailed analysis of the choice of normal periods and described the effect of the choice on outcomes.
+During the figure revisions, we removed the red dots.
 
@@
 
-I have added an explicit statement to quantify losses in the paragraph beginning at **l**104.
-There is strong support for extensive loss of coverage since the PLSS for certain species, and expansion for others, all referenced in this paragraph.
-These patterns are well supported in the literature.
+These losses are well supported in the literature.
 Rhemtulla et al (2007) cite estimates of land use conversion up to 79% for southern deciduous savanna, and Rhemtulla et al (2009) show that losses in oak-hickory forest complexes surpass 5,000,000 ha.
 As noted in the new sentences: "*Rhemtulla et al.
-[-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin.
-Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (Figure Xf), where 79% of deciduous savanna was converted to cropland [@rhemtulla2007regional].*"
-
-Given the extensive support for broad scale land use change, and the associated impacts of fire & logging, we believe that these patterns are real.
+[-@rhemtulla2009historical] estimate that Pinus strobus and Tsuga canadensis now occupy only 4% of their original coverage in Wisconsin.
+Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (Figure 1f), where 79% of deciduous savanna was converted to cropland [@rhemtulla2007regional].*"
 
@@
 
-As above, we believe that these patterns are real.  We extend this point however beginning at **l**371, where we discuss the potential use of isolated trees or trees unsampled in the FIA data.
+As above, we believe that these patterns are real.  We have added text to the Methods and Discussion about  differences in the FIA and PLS data and potential effects on detecting low-density trees.
 
@@
 
-We have updated this figure to make the comparison clearer.
-
-![](figures/figure_two.png)
+We have updated this figure to clarify the comparison.
 
@@
 
-See the following note.
+Please see the response to the next comment.
 
@@
 
-I believe the reviewer possibly overstates the case for limitations.
-There is strong support for changes since EuroAmerican settlement, both in vegetation-climate relationships, in the spatial extent of vegetation cover, and in pollen-vegetation relationships in the region, indicating a long-term shift in vegetation representation in climate space.
-However, we have qualified the statements made in this section, and added a short paragraph pointing to the potential effects of the choice of climate normal as well.
-
-*Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.
-In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.
-If so, estimates of forest loss (`fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest.
-Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].
-Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.
-As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.*
-
-*Artifacts of the choice of climate normal (1995 - 2014) for the modern data mean an exaggerated effect for $t_{max}$, however, supplemental analysis shows that the patterns of change for the other variables, $t_{min}$, $t_{diff}$ and $p_{ann}$ remain robust.
-Analysis using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships [@jacques2008pre] following Euro-American settlement in the region providing further support that modern vegetaton-climate relationships have changed since settlement.*
+We have created a standalone section in the Discussion (and the various sensitivity analyses described above) that more fully works through these points.
+Our takeaway message here is that the limitations are real and need considering, but do not detract from the main finding that historic land use and climate changes in this region have been large and have interacted to alter tree-climate relationships.
 
@@
 
-Referee 3 was most critical of the paper.
+> Goring and Williams investigated the overall distribution changes and relative influence of historic land use and climate change on these changes in the upper Midwestern United States.
+This is an interesting topic because climate change has been changing tree species distribution, meanwhile land use change may be interacting with climate change to compound, confound, or contract these changes.
+However, the manuscript is not very compelling for several reasons:
+
+We thank the reviewer for these comments and hope that the revised manuscript addresses these concerns.
 
@@
 
-We have changed this throughout the paper.
+See response to Reviewer 2.  We have changed the title to ‘upper Midwestern United States’.
 
@@
 
-We believe the title should stand as is.
-We are not interested in the changes in spatial distributions, except inasmuch as they affect the climate-space occupied by the tree species, thus the "tree-climate relationship".
+We respectfully believe the title should stand as is.
+This paper does indeed present evidence for both changes in geographic distributions and tree-climate relationships.
+However, we believe the latter is of more interest to the Ecology Letters readership and have focused the paper’s attention on it.
 
@@
 
-We appreciate the opinion of the reviewer in this context. We address the first point by changing the introductory sentence to:
+We appreciate the clarification of the reviewer regarding the first point and have addressed it by changing the introductory sentence to:
 
@@
 
-We believe that the issue of stability or extending the dist-clim relationship is a matter of opinion.
+We respectfully disagree with the reviewer regarding the second point.
+We believe that the central aspect of this paper is its analysis of the stability of tree-climate relationships over the past two centuries.
 The paper shows that, for many taxa, land use following EuroAmerican settlement in the north-central US limits the representitiveness of modern inventory-based distribution models.
-We do believe that these historical datases can be used to extend distribution-climate relationships.
-The exlicit tests in this paper focus on shifts in climate-distribution relationships and the attribution for these shifts, thus, the stabilty of the relationships.
+We do not attempt here to use historical and contemporary data to build better predictions for the future, but have inserted a pointer to other papers where that has been done:
+
+*One solution is to use both historical and contemporary data when building species distribution models (Nogues-Bravo et al. 2016 Maiorano et al. 2013).*
 
@@
 We agree.
-The impact of pathogen outbreaks are significant for *Ulmus*, and (outside the current range) for *Castanea*, however, given the current data sources, they are difficult to properly assess.
+Pathogen outbreaks cannot be ignored, because their impacts are significant for Ulmus, and (outside the current range) for Castanea, but they are not central to this paper.
 We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions in the Discussion and in the Introduction.
+We’ve also added several references about the past effects of pathogens on tree mortality and forest composition.
 
@@
 
+The reviewer is pointing out an important distinction:  what we can observe is historical climate change and vegetation change.
+Yet we often make inferences about land use because, in this region, forest composition has been so heavily altered by past and current land use.
 Previous authors have clearly indicated that EuroAmerican settlement in the north-central United States can be almost entirely attributed to human agency.
-In responding to Reviewer 2, we make the same point, and make this point again below.
-EuroAmerican Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture.
-Thus, VH, from the PLSS forest survey (ignoring climate data for now), represents forest structure and distribution in the absence of settlement impacts.
+Euro-American Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture.
+Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement.
+Initial clearance for agricultural and urban development was contemporaneous with massive logging operations, and, in the late 1800s, widespread fire originiating from human sources.
+Given this, and the fact that very little "old growth" forest remains, modern forests are indelibly imprinted with the signal of human land use, whether agricultural conversion or timbering.
+The IPCC considers forestry to be a component of land use.
+The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging.
+While sucession processes occurred since clearance and settlement, these were directly affected by the initial land clearance.
+For example, Rhemtulla et al.
+(2009) indicate that the loss of hemlock was caused both by logging, and recruitment failure as the result of the loss of seed source because of the extensive logging, thus both cause & effect are rooted in EuroAmerican settlement.
+
+Thus, VH, from the PLSS forest survey (ignoring climate data for now), mostly represents forest structure and distribution in the absence of settlement impacts.
 VM represents a modern landscape, sampled over the last decade as part of the FIA program.
-Distribution loss is almost entirely attributable to anthropogenic agency, either loss of forested land (oak-hickory forest lost 5x10^6^ ha of coverage since settlement due to agricultural conversion, as pointed out in the response to Reviewer 2), thus we feel confident in ascribing change from VC -> VM to land use.
-We believe that changes in the introduction and discussion, strengthening the argument for widespread vegetation change help resolve this issue.
-
-To further clarify, we have modified the explanation of Figure 2 to read, in part:
-
-*The difference between the PLSS and modern tree distributions ($d_{v}$) is attributed largely to land use change, and we will refer to it as such throughout.
-As described above, widespread post-settlement disturbance in the upper Midwest included conversion to agriculture, logging and widespread fire in the late 19th century, largely attributed to slash burning and logging.
-When the vegetation change is overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) we show the change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.*
+Differences between VH and VM is almost entirely attributable to anthropogenic agency, either loss of forested land (oak-hickory forest lost 5x106 ha of coverage since settlement due to agricultural conversion, as pointed out in the response to Reviewer 2) or logging, tree plantations, or other forms of forestry use.
+Hence we are confident in ascribing most changes from VC to VM to land use.
+To further clarify, we have modified the explanation of Figure 2 in the Methods as follows:
+*The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
+Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.*
 
@@
 
-We believe that our argument is valid, but may have been unclear in the framing.
-First, we believe that attributing the distributional changes from VH to VM to land use is justified.
-Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement.
-Initial clearance for agricultural and urban development was contemporaneous with massive logging operations, and, in the late 1800s, widespread fire originiating from human sources.
-Given this, and the fact that very little "old growth" forest remains, modern forests are indellibly imprinted with the signal of human land use, whether agricultural conversion or timbering.
-The IPCC considers forestry to be a component of land use.
-The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging.
-While sucession processes occurred since clearance and settlement, these were directly affected by the initial land clearance.
-For example, Rhemtulla et al.
-(2009) indicate that the loss of hemlock was caused both by logging, and recruitment failure as the result of the loss of seed source because of the extensive logging, thus both cause & effect are rooted in EuroAmerican settlement.
+See response to prior comment
 
@@
 
-We have changed this throughout.
-We except descriptions of methods applied in earlier papers, or in the preparation of data prior to the current study (for example, when summarizing the methods of Goring et al.
-2016).
+In our experience, either present or past tense is acceptable, as long as verb tense is used consistently.  We have not changed tense, but are happy to comply with any request by Ecology Letters.
 
@@
 The four paragraphs of interaction subsection in the results section were not concise and like bullets.
+
+We have ensured that each paragraph in the results has a clear and informative topic sentence.
 
@@
 
-We have clarified these variables: "*Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) .
+We have clarified the description of these variables in the Methods: "*Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) .
 .
@@
 
-*In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.*
+*In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "presence", where a tree was present within a grid cell in both FIA and PLS eras.*
 
@@
 
+We’ve amended a sentence in Methods to read “*Distributions of tree genera within climate space are shown using box plots and described as univariate probability density functions (PDFs) of tree species presence*”
+
 > The approach used in Figure 5 was not described in the materials and method section but was only in the title of Figure 5.
 More detailed information should be given to better analyze the relative importance of land use change and climate change effects.
+
+The approach shown in Fig. 5 is described in the section of Methods ‘Shifts of Species Distributions in Climate Space.’   We have edited it for clarity.
 
@@
 
-*where $P$ and $Q$ are the distributions, with a common discrete index $i$.*
+*where $P$ and $Q$ are the distributions, with a common index $i$.*
 
 > 3) Again, how the effects land use and exotic pathogens were included in the analyses?
+
+See responses to prior comments.
 
@@
 
-We have attempted to clarify this description by re-writing this paragraph:
+We revised this section of the methods for clarity, in part by re-writing this paragraph:
 
-*The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-These conditions are observed or recorded in data.
-The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.
-The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `fig_nums("fourPanel", display = "cite")`a; left column).
-We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.
-We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.*
+*The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
+The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.
+The effect of climate change on shifting PDFs ($d_{c}$) is the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
+The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
+Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.*
 
@@
 
-<!-- catch them up? but also make sure it's clarified better. -->
+We have modified the t-tests to include spatial autocorrelation and expanded their description in the Supplementary Information.
 
@@
 I suggest that the authors should set the results in a broader context and logically discuss the implications for distribution-climate relationships, distribution modeling, historical database, and management.
+
+We have heavily revised the discussion in response to all reviewers’ comments.
+Note that we were unsure about how to best address this comment, because it seems to ask that the discussion be both broadened and narrowed.
+We hope that the revised discussion addresses these concerns.
+
 > 1) L316-L322 were not the findings of this study. How were these points related to the findings? They read like introduction and should not be in discussion section.
 
-<!-- This is just reframing & revisiting the argument. -->
+This paragraph is intended to review mechanisms by which species-climate relationships may change over time. We have not changed this text.
 
@@
 
-This paragraph is intended to provide stronger links to other mechanisms by which species may not be in equilibrium with occupied climate space.
-We believe that the final sentence of the paragraph is sufficient as is.
+We have added two pointers back to figure 1, to better link this discussion to that figure.
 
@@
 
+See prior response to prior comments.
+
 ## Specific comments:
@@
 
-The sentence refers to the "stability and representativeness", which is singular.
+Fixed.
 
@@
 
-The new attribution index is within the current paper, we believe that the present tense is valid.
+This edit goes back to prior stylistic disagreements about choice of present vs. past tense.  We have kept the abstract in the present tense.
 
@@
 
-Again, this is presnt tense as it refers to the current study.
+See response to prior comment.
 
@@
 
+We have edited the figures to address these comments.
+
 > 5) All figure titles were too long. Many of these contents were approach and should be put in methods section.
 
-> Overall, this is an interesting topic.
-Especially, the interaction between land use change and climate change on distribution changes is the most interesting part, because the overall distribution changes have been extensively studied.
-However, its rigor and clarity (especially methods and discussion sections) do not meet the standard for Ecology Letters.
-I do not think it warrants publication in its present form.
+5)	We have added more information to the Methods section, while also retaining much of this information in the Figure legends.
+We prefer a style that uses longer legends, to facilitate figure interpretation.
 
 ******************************************
-Editor
-Editors Comments for the Author(s):
-The reviews highlight the importance and novelty of the main idea - that past land-use complicates our estimates of forest community responses to climate change.
-However, they express concern about the way that the results are presented.
-There are many suggestions for clarification that should be addressed in revision.
-In particular, it's important to address the limitations of comparing the two data types, as this was a concern expressed by each reviewer.

### 2016-12-03 Composition_Ecology.Rmd
Revised response to reviews, minor edits to main text

@@
 
-In mapping tree distributions in the upper Midwest we consider three classes: "gain," where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss," where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence," where a tree was present within a grid cell in both FIA and PLS eras.
+In mapping tree distributions in the upper Midwest we consider three classes: "gain," where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss," where a tree was present in a grid cell during the PLS but absent in the FIA; and "presence," where a tree was present within a grid cell in both FIA and PLS eras.
 We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA observations (Supplementary Table  2).
@@
 To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].
+<!-- Re-inserted paragraph below (I think it's important to explain why we picked the two modern normals that we did.  Note that there is a missing ref below, highlighted in bold.) -->
+The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.
+The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here.
+The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (**REF**).
+We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.
 
@@
 The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.
-Long term observer and military fort observations suggest warming during the 19^th^ century, with increases in mean annual temperature at Fort Snelling, Minnesota [@baker1985minnesota] and Fort Leavenworth, Kansas [@burnette2010daily].
-The latter shows faster increases during winter (`r round(clim_change[4], 1) + 0.5`^o^C) than summer(`r round(clim_change[2], 1) + 0.35`^o^C).
-Hence, the PRISM-based estimates of historical climate change may be too conservative for $T_{min}$ and $T_{diff}$, and too large for $T_{max}$.
+Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al.
+1985) and Fort Leavenworth, Kansas (Burnette et al.
+2010).
+The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1).
+Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).
 
@@
 *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm).
-These patterns are unchanged by sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).
+These patterns are unchanged by a sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).
 

### 2016-12-04 Composition_Ecology.Rmd
Removed the header block of `Supplemental Material`.
Changed heading types for Acknowledgements & references in main paper.

--- 2016-12-02 Composition_Ecology.Rmd
@@
 * Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
-* John W Williams: Co-conceived paper and co-wrote the paper
+* John W Williams: Co-conceived paper, and co-wrote the paper
 
@@
 
-Acknowledgements
----------------------
-The authors would like to thank the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
+# Acknowledgements
+
+The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form.
 In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group.
@@
 
-References
----------------
+# References

### 2016-12-04 Composition_Ecology.Rmd

--- 2016-12-03 Composition_Ecology.Rmd
@@
 # Abstract
-<!-- removed reference to 'higher Tmax' given results of sensitivity analyses.  Also took away the statement about 'reinforces shift to cooler conditions, again given ' 146 words after rewording-->
+
 Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate relationships are poorly understood.
@@
 Forest inventories offer rich observational datasets that have been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].
-<!-- deleted Ordonez 2013 reference to shorten references -->
+
 Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling].
@@
 land-use data from the NLCD [@jin2013comprehensive].*")`
-<!-- WICCI reference below:  Fix author in bibtex to WICCI, i.e. (WICCI, 2011) -->
+
 Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate].
@@
 We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in geographic or climate space of all cells with recorded presence for a specific taxon.
-<!-- inserted 'geographic or' because we often refer to range gains and losses in the paper -->
 
@@
 To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].
-<!-- Re-inserted paragraph below (I think it's important to explain why we picked the two modern normals that we did.  Note that there is a missing ref below, highlighted in bold.) -->
+
 The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.
 The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here.
-The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (**REF**).
+The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (FIA DataMart: http://apps.fs.fed.us/fiadb-downloads/datamart.html, accessed 1/12/2016).
 We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.
@@
 ![](Final_Figures/figure_two.tiff)
-<!-- Is Figure 2 for Larix or Quercus?  Legend here says Larix, later in text says Quercus -->
+
 `r fig_nums("fourPanel", "(a) The 2x2 factorial design used to diagnose the effects of historical climate and vegetation change on observed shifts in tree-climate relationships.
 Observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row).
-(b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`
+(b-e) Representative normalized probability densities for *Quercus* for each of the four combinations of climate and vegetation states, plotted relative to $T_{max}$ (b), $T_{diff}$ (c), $T_{min}$ (d), and $P_{ann}$ (e).")`
 
@@
 If both are significant and in opposite directions then the shift is counteracting.
-If climate change is significant, but land-use is not significant, then the change is defined as confounding.
-<!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed.
--->
+If climate change is significant, but land-use is not significant, then the change is defined as confounding, as is the converse, since in both cases, the climate signal is weakened because of vegetation change.
 
@@
 $P_{ann}$ increased by `r round(clim_change[1], 1)`mm.
-The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.
+The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward ($r$ = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.
 Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al.
@@
 2010).
-The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1).
-Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).
+The latter shows faster increases during winter (0.10^o^C decade^-1^) than summer (0.07^o^C decade^-1^).
+Hence, our estimates of historic climate change may be too conservative for $T_{min}$ and $T_{diff}$ and too large for $T_{max}$ (Supplementary Information).
 
@@
 
-Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all p < 0.0001).
+Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all $p$ < 0.0001).
 The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).
-*Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while *Larix* has the smallest.
-Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).
-*Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm).
+*Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{T_{min}}$ = 4.2 ^o^C), while *Larix* has the smallest.
+Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{T_{max}}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{T_{max}}$ = -0.4 ^o^C).
+*Ulmus* shows the largest increase in mean precipitation ($\delta_{P_{ann}}$ = 86mm), and *Larix* the smallest ($\delta_{P_{ann}}$ = 42mm).
 These patterns are unchanged by a sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).
@@
 For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`
-<!-- Does Fig. 2 show Larix or Quercus?  This paragraph states Quercus; legend states Larix -->
+
 Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).
@@
 
-<!-- Note that figure numbering is still off - Fig 6 should be Fig 5.  Needs fixing prior to submission-->
 Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).
@@
 ```
-<!-- Given new Table 1, have rewritten to state that all climate-veg interaction are compounding or confounding -->
-<!-- Last sentence in para may need tweaking, if Supp Table 4 is changed - see comments there. -->
+
 All climate-vegetation interactions are compounding ($\\Box$; n = 30) or confounding ($-$; n = 30), with no counteracting effects found (`r tab_nums("compTable", display = "cite")`).
@@
 The patterns of confounding and compounding interactions are insensitive to alternative estimates of 19^th^ century temperature changes (Supplementary Information Table 3).
-However, the apparent $T_{max}$ cooling and number of compounding interactions are reduced when 1950 - 1980 climate normals are used instead of 1985-2014 normals (Supplementary Table 4).
+However, the 1950 - 1980 climate normal results in non-significant $T_{max}$ changes, resulting in a loss of all compounding interactions for $T_{max}$ (Supplementary Table 4).
+All $T_{max}$ interactions are then confounding, with significant shifts for vegetation changes (all positive), except for *Pinus* and *Thuja/Juniperus*, where changes of vegetation in climate space are non-significant.
 
@@
 
-<!-- Fix the WICCI ref, as above. -->
 Compounding and confounding interactions between historical climate change and land-use have important implications for the management and genetics of forest communities.
@@
 
-Acknowledgements
----------------------
+# Acknowledgements
+
 The authors would like to thank the large number of individuals who have worked to undertake the Public Lands System Survey, to compile the original surveyor notes, to digitize and standardize the survey results, and finally, to assist in interpreting and compiling the data in its present form.

### 2016-12-04 SupplementaryInformation.Rmd
Removed the header block of `Supplemental Material`.
Changed heading types for Acknowledgements & references in main paper.

@@
 ---
-title: "Supplementary Information"
-author: "Simon Goring"
-date: "November 29, 2016"
-output: word_document
+output:
+  word_document:
+    reference_docx: word_template.docx
 ---
@@
 
-To overcome this challenge, we obtained a modified estimate of degrees of freedom (df) that accounts for spatial autocorrelation based on Moran's I from `modified.ttest` for each of the two spatial processes.
-To then produce a difference of the means that is not pairwise, we obtain the corrected df from `modified.ttest`, and recalculate the p-value for the t-test using the t estimate from the `t.test`.
+To overcome this challenge, we obtained a modified estimate of degrees of freedom ($df$) that accounts for spatial autocorrelation based on Moran's I from `modified.ttest` for each of the two spatial processes.
+To then produce a difference of the means that is not pairwise, we obtain the corrected df from `modified.ttest`, and recalculate the p-value for the t-test using the $t$ estimate from the `t.test`.
 The corresponding code is available in [`confounding_table.R`](https://github.com/PalEON-Project/Composition_Climate/blob/master/R/confounding_table.R).
@@
 # Supplemental Tables
+
 **Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era.
@@
 
-<!-- COM: scientific names needed in ST3 -->
-
-**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al.
+**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported from early meteorological time series at observing stations and US military forts (Burnette et al.
 2010, Baker et al.
@@
 
+conf_pls$taxa <- paste0("*", sci_names, "*")
+colnames(conf_pls) <- c("Taxon", "$P_{ann}$", "$T_{max}$", "$T_{diff}$", "$T_{min}$")
+
 knitr::kable(conf_pls)
@@
 
-<!-- COM: ST4: scientific names needed -->
-<!-- COM: ST4: Correct column names and formatting for climate variables.  e.g. ppt should be PAnn -->
-<!-- COM: Correct formatting for variable names in text.  I have tried to do this where I have seen them, but may have missed some variables.   -->
-<!-- COM: For TMax, why does the sign of the vegetation change flip from negative to positive between Table 1 and this Supp Table 4?
-I can understand why the sign of the climate change would alter as we play around with different climate normals, but the sign of the vegetation change should be stable.
-The other climate variables seem to be unaffected.
-This may be the bug you mentioned by email.
--->
 **Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1979 instead of 1990-2014.
 The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged.
-For ($T_{max}$), the sign of the climate change flips from negative (cooling) to positive (warming), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.
+For ($T_{max}$), the sign of the climate change flips from negative (cooling, $-_c$) to non-significant ($._c$), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.
 (Note that directionality of ($T_{min}$) is unchanged, consistent with 20^th^- and 21^st^-century warming).
@@
 conf_fia <- readRDS('data/output/conf_fia.RDS')
+
+conf_fia$taxa <- paste0("*", sci_names, "*")
+colnames(conf_fia) <- c("Taxon", "$P_{ann}$", "$T_{max}$", "$T_{diff}$", "$T_{min}$")
 
@@
 # Supplemental Figures
+
 **Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.
@@
 
-**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e.
+**Supplemental Figure 2**:  As Figure 4 in the main text, but with an adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e.
 a climate normal closer to the time of the PLS surveys).

### 2016-12-05 Response_to_reviewers.Rmd
Minor text modifications to the Response.  Nothing major.

@@
 
-We have added the sentence: *While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for * Tsuga * (`r fig_nums("gain_loss", display="cite")`), a shade-tolerant and late-successional taxon.*
+We have added the sentence: *While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for * Tsuga * (Figure 3), a shade-tolerant and late-successional taxon.*
 
@@
 
-We have changed the x axis to a log10 scale and added more tick marks to the x-axis to make it clear that it is non-linear.
+We have changed the $x$-axis to a $log_{10}$ scale and added more tick marks to the $x$-axis to make it clear that it is non-linear.
 
@@
 2010).
-The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1).
-Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).*
+The latter shows faster increases during winter (0.10^o^C decade-1) than summer (0.07^o^C decade^-1^).
+Hence, our estimates of historic climate change may be too conservative for $T_{min}$ and $T_{diff}$ and too large for $T_{max}$ (Supplementary Information).*
 
@@
 
-*The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right).
-The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.
-The effect of climate change on shifting PDFs ($d_{c}$) is the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column).
-The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a).
-Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.*
+*The total observed shift (d_tot) between a historical and modern PDF is the Hellinger distance between V_M C_M and V_H C_H (Figure 2a: top left to bottom right).
+The conditions V_M C_H and V_H C_M are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.
+The effect of climate change on shifting PDFs (d_c) is the difference between PDFs using the early climate normals (C_H) and modern climate normals (C_M) overlaid on PLS vegetation (e.g., V_H C_H versus V_H C_M; Figure 2a; left column).
+The effect of vegetation change on shifting PDFs is d_v, the Hellinger distance between V_H C_H and V_M C_H (top row of Figure 2a).
+Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by d_v to the land-use change associated with Euro-American settlement.*
 
@@
 We have modified the t-tests to include spatial autocorrelation and expanded their description in the Supplementary Information.
+The responsible variables are explained more clearly in the methods now.
+We compare VHCH and VHCM to estimate the significance and direction of the climate shift (vegetation is fixed, climate layer changes), and compare VHCH to VMCH to estimate the change in vegetation (keeping vegetation constant).
 
@@
 
-5)	We have added more information to the Methods section, while also retaining much of this information in the Figure legends.
+We have added more information to the Methods section, while also retaining much of this information in the Figure legends.
 We prefer a style that uses longer legends, to facilitate figure interpretation.
